---
title: "00 — Load data (per-site processed datasets, 2026-08-31 draft snapshot)"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    fig-width: 10
    fig-height: 6
execute:
  warning: false
  message: false
---

{{< include _disclaimer.qmd >}}

# Purpose

This notebook is the entry point for the LEVANTE longitudinal analysis
project. It downloads the scores data from Redivis, caches it locally,
attaches LEVANTE-standard labels and palettes, and then summarizes:

- which sites have data (and how much),
- which tasks are covered at each site/dataset,
- age distributions,
- longitudinal coverage (how many children have repeat assessments).

**Data source (2026-08-31 update, presentation snapshot).** The unified
`levante_data_latest` dataset is still at v1_2 (June); the loaders bind the
**per-site processed datasets** directly, and this snapshot pulls each
dataset's **unreleased "next" draft** — the absolute newest processed data
(`common.R::levante_site_specs`; the cached rds is the reproducibility anchor
since drafts are mutable; provenance in
`data/scores_sites_2026-08-31_meta.csv`). Relative to the June release this
adds, cumulatively: **rural Colombia's second wave** (repeat child×task pairs
32 → 1,239; ~9.6-month gap, led by ROAR + Math);
**Leipzig's third wave, which began arriving 2026-08-29** — 39 children
re-tested on the full battery through Aug 30, already scored in the draft
(**167 three-wave child×task cells**, the project's first); **Argentina (UTDT
main) coming online at scale** — 788 children, 12 tasks, ages 4.8–8.4
(es-AR); **Western's June–August backlog entering processing** (+863 scored rows vs
the 08-24 snapshot — doubling Western's longitudinal sample to 263 repeat
pairs / 46 children); Sheffield more than tripling (755 rows); and
Boston's downward extension growing to 42 children (chapter 10). The per-site
scores are all under the updated (registry v2_3) scoring models. Cape Coast
and Sheffield intl-YS still have no scored rows. Note: trial-level pulls (and
therefore the ROAR-Word engagement filter) remain at the 08-24 released
versions — the filter does not yet cover the newest runs.

```{r setup}
source(here::here("common.R"))
scores_raw <- load_levante_scores()
trials_swr <- load_levante_trials(task_ids = "swr")
```

Sanity check: row counts per site / dataset.

```{r counts}
scores_raw |>
  count(site_label, dataset_label, name = "n_rows") |>
  arrange(desc(n_rows)) |>
  knitr::kable()
```

# Cleaning

Two cleaning steps are applied here so that they propagate consistently
through 02 / 03 / `tasks/` notebooks. See the "Known data issues" section
at the bottom of this notebook for the full justification.

```{r cleaning}
# New in the 2026-08 per-site processing: an `exclusion` column marks runs the
# pipeline could not score (no_data / incomplete / few_valid_trials /
# duplicate). These rows carry no score; dropping them is what un-inflates the
# handful of rural kids that otherwise show 8-10 "runs" of one task.
basic_cleaned <- scores_raw |>
  filter(is.na(exclusion)) |>
  filter(!is.na(age), age >= 2, age <= 14) |>
  filter(is.na(score) | (score > -8 & score < 8))

cleaned <- clean_levante_scores(basic_cleaned, trials_swr)
scores  <- cleaned$scores

cleaned$report |> knitr::kable(caption = "Cleaning report")

glue("Kept {nrow(scores)} of {nrow(scores_raw)} score rows (raw → cleaned).")
```

# Inventory by site / dataset

```{r site-summary}
site_summary <- scores |>
  group_by(site_label, dataset_label) |>
  summarise(
    n_children = n_distinct(user_id),
    n_runs     = n_distinct(run_id),
    n_tasks    = n_distinct(task_id),
    n_rows     = n(),
    age_min    = min(age, na.rm = TRUE),
    age_max    = max(age, na.rm = TRUE),
    .groups    = "drop"
  ) |>
  arrange(desc(n_children))

knitr::kable(site_summary, digits = 1)
```

```{r site-bar, fig.height=4}
site_summary |>
  mutate(dataset_label = fct_reorder(dataset_label, n_children)) |>
  ggplot(aes(x = dataset_label, y = n_children, fill = site_label)) +
  geom_col() +
  geom_text(aes(label = n_children), hjust = -0.1, size = 3) +
  coord_flip() +
  scale_fill_site(name = NULL) +
  labs(x = NULL, y = "Children with at least one scored task",
       title = "Sample size by dataset") +
  expand_limits(y = max(site_summary$n_children) * 1.1) +
  theme(legend.position = "bottom")
```

## Age distribution per dataset

```{r age-dist, fig.height=6}
scores |>
  distinct(site_label, dataset_label, user_id, age) |>
  ggplot(aes(x = age, fill = site_label)) +
  geom_histogram(binwidth = 0.5, color = "white") +
  facet_wrap(~dataset_label, scales = "free_y") +
  scale_fill_site(guide = "none") +
  scale_y_continuous(expand = expansion(0, 0)) +
  labs(x = "Age (years)", y = "Number of children",
       title = "Age distribution per dataset")
```

## Task × dataset coverage

Which tasks are present at which dataset? Cell value = number of scored runs;
empty cells mean no data.

```{r coverage, fig.height=5}
scores |>
  count(dataset_label, task_label, task_category, name = "n_runs") |>
  ggplot(aes(x = task_label, y = dataset_label, fill = task_category)) +
  geom_tile(color = "white") +
  geom_text(aes(label = n_runs), size = 3) +
  scale_fill_task(guide = "none") +
  facet_grid(. ~ task_category, scales = "free_x", space = "free_x") +
  labs(x = NULL, y = NULL,
       title = "Task × dataset coverage (cell = number of scored runs)") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.spacing = unit(0.1, "lines"))
```

# Longitudinal coverage

LEVANTE's design targets ≥ 3 timepoints per child. Here a "timepoint" =
distinct `run_id` within a `user_id` × `task_id`. The question: how much
repeat data do we actually have?

```{r repeats-per-task}
repeats_task <- scores |>
  group_by(dataset_label, user_id, task_id, task_label, task_category) |>
  summarise(n_runs = n_distinct(run_id),
            age_span = diff(range(age, na.rm = TRUE)),
            .groups = "drop")

repeats_task |>
  count(dataset_label, n_runs, name = "n_child_task") |>
  ggplot(aes(x = factor(n_runs), y = n_child_task, fill = dataset_label)) +
  geom_col() +
  facet_wrap(~dataset_label, scales = "free_y") +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  labs(x = "Times a (child, task) pair was administered",
       y = "Count of (child, task) pairs",
       title = "Per-(child, task) repeat-administration counts")
```

```{r repeats-per-dataset}
scores |>
  group_by(dataset_label, user_id, task_id) |>
  summarise(n_runs = n_distinct(run_id),
            age_gap_m = diff(range(age, na.rm = TRUE)) * 12,
            .groups = "drop") |>
  filter(n_runs > 1) |>
  group_by(dataset_label) |>
  summarise(
    n_repeat_pairs        = n(),
    n_kids_with_any_repeat = n_distinct(user_id),
    median_gap_m          = median(age_gap_m),
    max_gap_m             = max(age_gap_m),
    .groups = "drop"
  ) |>
  arrange(desc(n_repeat_pairs)) |>
  knitr::kable(digits = 1,
               col.names = c("Dataset", "Repeat (child, task) pairs",
                             "Kids with any repeat",
                             "Median gap (mo)", "Max gap (mo)"))
```

# Score-by-age at a glance

Cross-sectional age gradient by task and site. This is intentionally a
quick-look plot — 01 fits proper longitudinal models.

```{r score-age, fig.height=10, fig.width=12}
scores |>
  filter(!is.na(score), task_id %in% core_task_ids) |>
  levante_age_score_plot()
```

# Save

We persist two artifacts for downstream notebooks:

- `data/scores_raw.rds` — the raw `levante_data_latest` pull, used by
  `01_data_integrity.qmd` so that the audit reflects what the data look
  like *before* we apply project-level fixes.
- `data/scores_all_sites.rds` — the cleaned tibble. **This is the canonical
  scores table used by 02, 03, and the `tasks/` notebooks.** Cleaning
  decisions are described in *Known data issues* below.

```{r save}
write_rds(scores_raw, here("data", "scores_raw.rds"))
write_rds(scores,     here("data", "scores_all_sites.rds"))
write_rds(site_summary, here("data", "site_summary.rds"))
write_rds(cleaned$roar_drops, here("data", "roar_word_dropped_runs.rds"))

glue("Saved {nrow(scores)} cleaned score rows ({nrow(scores_raw)} raw) from {nrow(site_summary)} datasets to data/.")
```

# Known data issues

## 2026-08 data-update notes (per-site processed datasets)

New issues observed in the per-site processing that the loaders / cleaning
handle, plus one substantive change to be aware of:

- **Re-scored with updated models (registry v2_3 vs June's v1_14) — an
  intentional, across-the-board scoring-model update (confirmed by the DCC).**
  On overlapping runs, new-vs-old scores correlate r ≥ 0.99 for most tasks but
  **r = 0.88 for Theory of Mind** and **r = 0.94 for Hearts & Flowers** (max
  per-run |Δθ| up to ~3 logits) — the largest shifts, plausibly including the
  ToM item-identity repairs documented in `tasks/tom_reality_check_bug`.
  Task-level ToM conclusions from June (the `tasks/` ToM chapters) predate the
  re-scoring and should be re-checked before being quoted.
- **ToM placeholder rows:** every theory-of-mind run carries an all-NA
  duplicate row; `load_levante_scores_sites()` drops them (912 rows).
- **`site` is NA** in the newest processing for rural, Sheffield, and Boston;
  backfilled from `levante_site_specs`.
- **`exclusion` column** replaces silent row absence for unscoreable runs
  (dropped in cleaning above; 1,167 rows).
- The per-site scores tables now carry `time_started`/`time_finished` for
  recently-processed datasets (rural, Sheffield, Boston) but not yet for the
  older ones.

**Now on `levante_data_latest` v1.2 (corrected).** The major IRT
column-order scoring bug we found (see `reports/rlevante_handoff.md`) has been
fixed in levantemodels and the data re-released; we verified v1.2 published scores
match our independent corrected scoring (r = 1.000). This section logs the
remaining issues and how each is handled: (a) **fixed here** (cleaning applied
in this notebook), (b) **resolved upstream** (fixed in the v1.2 release), and
(c) **flagged only** (the right fix lives elsewhere — scoring pipeline or
validator).

## Resolved upstream (v1.2)

**0. IRT column-order scoring bug (FIXED).** `levantemodels::score_irt` did not
reorder response columns to the model's item order before `mirt::fscores`,
which matches by position — so IRT-scored tasks were mis-scored, worst for
CAT and guessing-floor items. This drove most of what we had attributed to
"CAT-vs-non-CAT step shifts": after the fix, the Leipzig non-adaptive-minus-
adaptive mean shift dropped from **+2.4 → +0.30** (Math) and **+1.1 → +0.10**
(Vocab), and apparent longitudinal *declines* became sensible positive growth.
Full writeup and validation: `reports/rlevante_handoff.md`,
`tasks/math_eap_vs_ml.html`, `tasks/vocab.html`.

For deep-dive details see:

- `01_data_integrity.html` — per-(site, task) audit.
- `reports/adaptive_missingness.html` — short DCC-team report on the
  `adaptive` flag.
- `tasks/roar_word.html` — trial-level investigation of ROAR-Word floor
  scores.
- `tasks/math.html` — trial-level investigation of the Bogotá math "RTM."

## Fixed here

**1. `adaptive` flag missing (235 runs in v1.2; was 196 in v1.0).** Still
concentrated in early-beta task_versions of Bogotá Memory and Leipzig/Western
Math. The trial table shows zero `theta_estimate` writes for any of these
runs, consistent with non-adaptive administration. Confirmed with the LEVANTE
DCC: all were non-adaptive. (Not yet fixed upstream in v1.2.)

**Fix applied:** `common.R::backfill_adaptive_na()` sets `adaptive = FALSE`
on these rows during cleaning.

**2. ROAR-Word non-engagement.** A small subset of ROAR-Word runs show
trial-level accuracy below chance (40%) and/or median RT under 500 ms
(faster than plausible word reading). These produce IRT score floors at
θ = −6 with SE = 37 and are not informative. They are distinct from genuine
non-readers (who attempt items at chance with normal RTs and also floor —
those are kept).

**Fix applied:** `common.R::roar_word_engagement_drops()` drops runs with
`pct_correct < 0.4 OR median_rt_ms < 500`. The dropped run-ids and the
reason for each are saved to `data/roar_word_dropped_runs.rds` for
downstream traceability.

## Flagged only — not corrected in this repo

These deserve attention at the data-coordinating-center / scoring-pipeline
level rather than in analysis code.

**3. Residual CAT-vs-non-CAT differences (much reduced in v1.2).** With the
scoring bug fixed (#0), the dramatic mode shifts are gone — residual
non-adaptive-minus-adaptive mean shifts are now all < 0.75 logits (Leipzig
Math +0.30, Vocab +0.10; the largest is Leipzig Pattern Matching +0.73). What
remains is plausibly genuine: mild ceiling on the easy non-adaptive forms plus
**EAP shrinkage** — short CAT sessions carry less information, so EAP pulls
them slightly harder toward the group mean, which can look like mild
regression to the mean across a mode boundary. This is a small effect and a
known property of EAP (see the EAP-vs-WLE discussion); not a bug. Every
two-wave Bogotá math kid is still non-adaptive at T1 → CAT at T2, so interpret
those θ deltas with the residual shrinkage in mind.

**4. Memory task scoring — RESOLVED (2026-08).** The release-notes concern
(grid size 2×2 vs 3×3) is handled: grid is a calibrated item dimension, and
the Leipzig wave-1→2 grid switch shows continuous within-child θ while raw
sumscores break across grids (`tasks/memory` §4). 01 now marks Memory OK;
the operational rule is *use θ, never cross-grid sumscores*.

**5. Same & Different scoring — RESOLVED (2026-08).** The corrected scoring
landed in the registry v2_3 re-scoring and is final per the DCC. 01 upgrades
SDS to OK (OK-mode at Bogotá, which still has no T2).

**6. Pattern Matching / Sentence Understanding training effects.** Both
show systematic T2 > T1 within mode-stable subsets per the release notes.
We can't distinguish training from developmental change with two waves.

**7. Bogotá T2 ROAR assignment is mutually exclusive.** At T2, individual
kids in Bogotá were assigned **either** swr **or** sre **or** pa — almost
never two of them. Pairwise (swr, sre) overlap at T2 = 1 child (vs 584 at
T1). This breaks any multi-indicator Language LGC that uses more than one
ROAR task at Bogotá T2. 03 falls back to a swr + trog two-indicator
LGC with metric invariance, which is identified.

# Notes

- The bogota and rural Colombia samples share `site = pilot_uniandes_co` but
  stay separable via `dataset`. We carry both labels (`site_label`,
  `dataset_label`) downstream.
- Data come from the per-site processed datasets pinned in
  `common.R::levante_site_specs` (snapshot 2026-08-24) — update the specs and
  the snapshot constant when new versions land, or pass `version = "v1_2"` to
  the loaders to reproduce the June unified pull.
- Compare what we see here to <https://levante-framework.github.io/levante-datapage/>
  as a sanity check on the inventory.
