---
title: "01 — Data integrity audit (Leipzig & Bogotá)"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    fig-width: 10
    fig-height: 6
execute:
  warning: false
  message: false
---

{{< include _disclaimer.qmd >}}

::: {.callout-note}
## Now on corrected data (v1.2)
Re-run on `levante_data_latest` **v1.2** (IRT column-order scoring bug fixed).
The large adaptive/non-adaptive **mode shifts** this audit originally flagged
were mostly that bug: they have collapsed from ~+2.4 (Leipzig Math) and +1.1
(Vocab) to < 0.75 logits across the board. The "OK-mode" cautions below are
therefore much less pressing. As of 2026-08 the Memory and Same-&-Different
scoring concerns are also resolved (Memory: grid handled, switch-test passes —
tasks/memory §4; SDS: corrected scoring landed in registry v2_3, final). The
genuinely remaining issues are ROAR-Word non-engagement and a small residual
EAP-shrinkage difference across the mode boundary.
:::

# Purpose

Before fitting growth models (02) or SEM (03), audit data integrity on the two
longitudinal sites and produce a **per-(site, task) verdict**: usable as-is,
usable with a filter, usable within one administration mode only, or drop.

Known issues from the LEVANTE data release notes:

> - Adaptive vs non-adaptive ability scores aren't fully comparable; CAT
>   scores tend to be systematically lower because high-ability children ran
>   out of time on non-adaptive variants and never saw harder items
>   (ceiling-effect upward bias on non-adaptive).
> - Non-adaptive variants were more common at earlier waves → confounds
>   longitudinal change with administration-mode shift.
> - Pattern Matching and Sentence Understanding show systematic T2 > T1
>   within both modes — could be training or developmental.
> - Memory: known scoring bug, do not trust until next release.
> - Same & Different: scoring will be updated next release.
> - Surveys: still under development, use with caution.

Concerns originally flagged from the v1.0 spaghetti plots — math/vocab
"depressed in adaptive", shape-rotation decreases, Bogotá math
"regression-to-mean compression", pattern-matching downward trends — turned
out to be **overwhelmingly the column-order scoring bug** (see
`reports/rlevante_handoff.html`). On v1.2 they have collapsed (mode shifts now
< 0.75 logits). What this audit now isolates are the *genuinely* remaining
issues: ROAR-Word non-engagement and the unflagged `adaptive` rows (the
Memory and Same-&-Different scoring flags were resolved in 2026-08 — see the
verdict table).

```{r setup}
source(here::here("common.R"))

# This notebook intentionally reads the RAW scores tibble (pre-cleaning)
# so the audit reflects the data as released. The cleaning decisions
# documented here are applied in 00_load_data, which produces the cleaned
# scores_all_sites.rds that 02 / 03 / tasks/ consume.
scores_all <- read_rds(here("data", "scores_raw.rds"))

audit_datasets <- c("pilot_mpieva_de_main", "pilot_uniandes_co_bogota")
scores <- scores_all |>
  filter(dataset %in% audit_datasets, !is.na(score), !is.na(age))
```

# Score scales by task

ROAR-Sentence and ROAR-Word are scored differently from the LEVANTE core
tasks. ROAR-Word is a CAT theta (`ability_cat`), ROAR-Sentence is a
guessing-adjusted raw count, ROAR-Phoneme is on the standard IRT theta scale.
This affects how we should think about "extreme" scores.

```{r score-types}
scores |>
  count(dataset_label, task_label, score_type) |>
  pivot_wider(names_from = score_type, values_from = n, values_fill = 0L) |>
  knitr::kable()
```

# 1. Coverage: adaptive × wave per (dataset, task)

For each (dataset, task), count runs by `(wave, adaptive)`. This shows which
cells are vulnerable to the CAT-vs-non-CAT confound.

```{r waved-data}
waved <- scores |>
  group_by(dataset, user_id, task_id) |>
  arrange(age) |>
  mutate(wave = row_number()) |>
  ungroup() |>
  filter(wave <= 2)
```

```{r coverage-table, fig.width=12, fig.height=7}
cov_tbl <- waved |>
  mutate(adaptive_lbl = case_when(
    adaptive == TRUE  ~ "adaptive",
    adaptive == FALSE ~ "non-adapt",
    TRUE              ~ "unknown"
  )) |>
  count(dataset_label, task_label, task_category, wave, adaptive_lbl,
        name = "n_runs")

ggplot(cov_tbl,
       aes(x = factor(wave), y = n_runs, fill = adaptive_lbl)) +
  geom_col(position = "stack") +
  facet_grid(dataset_label ~ task_category + task_label,
             scales = "free_x", space = "free_x") +
  scale_fill_manual(values = c(adaptive = "#d33682",
                               `non-adapt` = "#268bd2",
                               unknown = "grey70")) +
  labs(x = "Wave", y = "Runs", fill = NULL,
       title = "Run counts by wave × adaptive (longitudinal datasets)") +
  theme(strip.text.x = element_text(angle = 0, size = 8),
        legend.position = "bottom",
        panel.spacing.x = unit(0.05, "lines"))
```

```{r coverage-tab-pretty}
cov_tbl |>
  unite("cell", wave, adaptive_lbl, sep = "_") |>
  pivot_wider(names_from = cell, values_from = n_runs, values_fill = 0L) |>
  arrange(dataset_label, task_category, task_label) |>
  select(dataset_label, task_category, task_label,
         any_of(c("1_non-adapt","1_adaptive","1_unknown",
                  "2_non-adapt","2_adaptive","2_unknown"))) |>
  knitr::kable()
```

# 2. Score distributions by adaptive

Histograms of all scored runs (T1+T2 pooled) split by adaptive flag. If CAT
and non-CAT are on the same scale, the distributions should overlap. Where
they don't, that's the artifact.

```{r score-dist, fig.height=10, fig.width=12}
waved |>
  filter(!is.na(adaptive)) |>
  ggplot(aes(x = score, fill = adaptive)) +
  geom_histogram(bins = 40, alpha = 0.7, position = "identity") +
  facet_grid(dataset_label ~ task_category + task_label,
             scales = "free", space = "free_x") +
  scale_fill_manual(values = c(`FALSE` = "#268bd2", `TRUE` = "#d33682"),
                    labels = c("non-adaptive", "adaptive"),
                    name = NULL) +
  labs(x = "Score", y = "Count",
       title = "Score distributions by adaptive flag") +
  theme(strip.text.x = element_text(angle = 0, size = 7),
        legend.position = "bottom",
        panel.spacing.x = unit(0.05, "lines"))
```

```{r score-dist-stats}
waved |>
  filter(!is.na(adaptive)) |>
  group_by(dataset_label, task_label, adaptive) |>
  summarise(n = n(),
            mean = mean(score, na.rm = TRUE),
            sd   = sd(score, na.rm = TRUE),
            p10  = quantile(score, 0.1, na.rm = TRUE),
            p90  = quantile(score, 0.9, na.rm = TRUE),
            .groups = "drop") |>
  pivot_wider(names_from = adaptive,
              values_from = c(n, mean, sd, p10, p90)) |>
  mutate(mean_shift_NA_minus_A = mean_FALSE - mean_TRUE) |>
  arrange(dataset_label, task_label) |>
  knitr::kable(digits = 2)
```

# 3. Within-child change by adaptive transition

For each child with T1 and T2, classify the (T1 mode → T2 mode) transition,
then compute the delta. A clean "developmental change" estimate requires the
mode to stay the same across waves.

```{r transitions}
pairs <- waved |>
  group_by(dataset, dataset_label, user_id, task_id, task_label, task_category) |>
  filter(n_distinct(wave) == 2) |>
  summarise(
    score_t1 = score[wave == 1],
    score_t2 = score[wave == 2],
    age_t1   = age[wave == 1],
    age_t2   = age[wave == 2],
    adapt_t1 = adaptive[wave == 1],
    adapt_t2 = adaptive[wave == 2],
    .groups  = "drop"
  ) |>
  mutate(
    delta   = score_t2 - score_t1,
    age_gap = (age_t2 - age_t1) * 12,
    transition = case_when(
      is.na(adapt_t1) | is.na(adapt_t2) ~ "unknown",
      adapt_t1 == FALSE & adapt_t2 == FALSE ~ "NA→NA",
      adapt_t1 == FALSE & adapt_t2 == TRUE  ~ "NA→A",
      adapt_t1 == TRUE  & adapt_t2 == TRUE  ~ "A→A",
      adapt_t1 == TRUE  & adapt_t2 == FALSE ~ "A→NA"
    )
  )

pairs |>
  count(dataset_label, task_label, transition) |>
  pivot_wider(names_from = transition, values_from = n, values_fill = 0L) |>
  arrange(dataset_label, task_label) |>
  knitr::kable()
```

```{r delta-by-transition}
pairs |>
  filter(transition %in% c("NA→NA", "NA→A", "A→A")) |>
  group_by(dataset_label, task_label, transition) |>
  filter(n() >= 10) |>
  summarise(n = n(),
            mean_delta = mean(delta, na.rm = TRUE),
            sd_delta   = sd(delta, na.rm = TRUE),
            median_age_gap_m = median(age_gap, na.rm = TRUE),
            .groups = "drop") |>
  arrange(dataset_label, task_label, transition) |>
  knitr::kable(digits = 2)
```

```{r delta-hist, fig.height=10, fig.width=12}
pairs |>
  filter(transition %in% c("NA→NA", "NA→A", "A→A")) |>
  group_by(dataset_label, task_label, transition) |>
  filter(n() >= 10) |>
  ungroup() |>
  ggplot(aes(x = delta, fill = transition)) +
  geom_histogram(bins = 30, alpha = 0.7, position = "identity") +
  geom_vline(xintercept = 0, lty = 3) +
  facet_grid(dataset_label ~ task_category + task_label,
             scales = "free", space = "free_x") +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "Score T2 − T1", y = "Children",
       title = "Within-child change by adaptive transition",
       fill = NULL) +
  theme(strip.text.x = element_text(angle = 0, size = 7),
        legend.position = "bottom",
        panel.spacing.x = unit(0.05, "lines"))
```

# 4. Regression-to-mean (Galton's regression)

For each (dataset, task) with ≥ 30 two-wave kids, fit `score_t2 ~ score_t1`
restricted to mode-stable transitions. A slope of 1 means no regression to
mean (perfect persistence; differences reflect actual growth). A slope < 1
means high T1 scorers regress down and low T1 scorers regress up. The shift
from slope-1 is approximately `1 - reliability + change_dependence`.

```{r rtm}
rtm <- pairs |>
  filter(transition %in% c("NA→NA", "A→A")) |>
  group_by(dataset_label, task_label, transition) |>
  filter(n() >= 30) |>
  summarise(
    n = n(),
    galton_slope     = coef(lm(score_t2 ~ score_t1))[2],
    galton_se        = summary(lm(score_t2 ~ score_t1))$coefficients[2, 2],
    intercept_shift  = coef(lm(score_t2 ~ score_t1))[1],
    cor_t1_t2        = cor(score_t1, score_t2),
    .groups = "drop"
  )

knitr::kable(rtm, digits = 3)
```

```{r rtm-plot, fig.height=8, fig.width=12}
pairs |>
  filter(transition %in% c("NA→NA", "A→A")) |>
  group_by(dataset_label, task_label, transition) |>
  filter(n() >= 30) |>
  ungroup() |>
  ggplot(aes(x = score_t1, y = score_t2, color = transition)) +
  geom_abline(slope = 1, intercept = 0, lty = 3, color = "grey50") +
  geom_point(alpha = 0.35, size = 0.8) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_grid(dataset_label ~ task_category + task_label, scales = "free") +
  scale_color_brewer(palette = "Set2") +
  labs(x = "Score T1", y = "Score T2",
       title = "Galton's regression: T2 on T1 (mode-stable transitions)",
       subtitle = "Dotted = identity; coloured line = OLS fit",
       color = NULL) +
  theme(strip.text.x = element_text(angle = 0, size = 7),
        legend.position = "bottom",
        panel.spacing.x = unit(0.05, "lines"))
```

# 5. Extreme low / high scores

Flag runs with |score| > 4 on the IRT theta scale. For ROAR-Word
(`ability_cat`) and ROAR-Phoneme this is the same θ scale; for ROAR-Sentence
the scale is "guessing-adjusted raw" so a separate threshold (|x| > 4 SDs of
that task) applies.

```{r extreme-scores}
extreme <- waved |>
  group_by(dataset, task_id, task_label, score_type) |>
  mutate(z = if_else(score_type == "guessing_adjusted_number_correct_scaled",
                     (score - mean(score, na.rm = TRUE)) /
                       sd(score, na.rm = TRUE),
                     score)) |>
  ungroup() |>
  filter(abs(z) > 4)

extreme |>
  count(dataset_label, task_label, name = "n_extreme") |>
  arrange(desc(n_extreme)) |>
  knitr::kable()
```

```{r extreme-details}
extreme |>
  group_by(dataset_label, task_label) |>
  summarise(
    n         = n(),
    median_age = median(age),
    median_num_attempted = median(num_attempted, na.rm = TRUE),
    median_score_se = median(score_se, na.rm = TRUE),
    pct_adaptive = mean(adaptive, na.rm = TRUE) * 100,
    score_range = paste(round(min(score), 1), round(max(score), 1), sep = " to "),
    .groups = "drop"
  ) |>
  arrange(dataset_label, task_label) |>
  knitr::kable(digits = 2)
```

```{r extreme-vs-num}
extreme |>
  ggplot(aes(x = num_attempted, y = score, color = adaptive)) +
  geom_point(alpha = 0.6) +
  facet_wrap(vars(dataset_label, task_label), scales = "free") +
  scale_color_manual(values = c(`FALSE` = "#268bd2", `TRUE` = "#d33682"),
                     labels = c("non-adapt", "adaptive"),
                     na.value = "grey60", name = NULL) +
  labs(x = "Number of items attempted", y = "Score",
       title = "Extreme scores vs. number of items attempted",
       subtitle = "If extreme scores cluster at low num_attempted, they're likely scoring noise") +
  theme(legend.position = "bottom")
```

# 6. ROAR-Word super-low scores in Bogotá: zoom in

The ROAR-Word `ability_cat` distribution in Bogotá showed a clump of very low
scores. Let's look more closely.

```{r roar-word-zoom}
rw_bog <- scores |>
  filter(dataset == "pilot_uniandes_co_bogota",
         task_id == "swr")

rw_bog |>
  ggplot(aes(x = age, y = score, color = adaptive)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "loess", se = FALSE) +
  geom_hline(yintercept = -4, lty = 3, color = "red") +
  scale_color_manual(values = c(`FALSE` = "#268bd2", `TRUE` = "#d33682"),
                     na.value = "grey60", name = NULL) +
  labs(x = "Age", y = "ROAR-Word ability_cat θ",
       title = "ROAR-Word scores by age (Bogotá)",
       subtitle = "Red dotted = floor candidate (-4)")
```

```{r roar-word-very-low}
rw_bog |>
  filter(score < -3) |>
  arrange(score) |>
  select(user_id, age, run_id, score, score_se, num_attempted, adaptive,
         scoring_model, max_incorrect, max_time) |>
  head(15) |>
  knitr::kable(digits = 2)
```

```{r roar-word-attempted}
rw_bog |>
  ggplot(aes(x = num_attempted, fill = score < -3)) +
  geom_histogram(binwidth = 5, alpha = 0.7, position = "identity") +
  scale_fill_manual(values = c(`FALSE` = "steelblue", `TRUE` = "firebrick"),
                    labels = c("score ≥ -3", "score < -3"),
                    name = NULL) +
  labs(x = "Number of items attempted", y = "Runs",
       title = "ROAR-Word num_attempted, split by extreme-low score (Bogotá)")
```

# 7. Score SE distribution — flag unreliable estimates

Per-run IRT SE: huge values mean the ability estimate is unreliable
(insufficient information). Compare across (site, task).

```{r score-se}
waved |>
  ggplot(aes(x = score_se)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  facet_grid(dataset_label ~ task_category + task_label, scales = "free") +
  geom_vline(xintercept = 1, lty = 3, color = "red") +
  labs(x = "score_se", y = "Runs",
       title = "Score SE distribution (red = 1.0 threshold)") +
  theme(strip.text.x = element_text(angle = 0, size = 7),
        panel.spacing.x = unit(0.05, "lines"))
```

```{r score-se-counts}
waved |>
  group_by(dataset_label, task_label) |>
  summarise(n = n(),
            n_se_gt_1 = sum(score_se > 1, na.rm = TRUE),
            pct_se_gt_1 = mean(score_se > 1, na.rm = TRUE) * 100,
            n_se_na = sum(is.na(score_se)),
            .groups = "drop") |>
  arrange(desc(pct_se_gt_1)) |>
  knitr::kable(digits = 2)
```

# 8. Pattern Matching & Sentence Understanding: T2 > T1 in both modes

Release notes call these out as systematically higher at T2 within both
adaptive and non-adaptive. Plot T1 vs T2 mean by age band to see if it
looks more like training or real development.

```{r pmsu-zoom, fig.height=6, fig.width=10}
pairs |>
  filter(task_id %in% c("matrix-reasoning", "trog"),
         transition %in% c("NA→NA", "A→A")) |>
  ggplot(aes(x = score_t1, y = score_t2, color = transition)) +
  geom_abline(slope = 1, intercept = 0, lty = 3, color = "grey50") +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", se = TRUE) +
  facet_grid(dataset_label ~ task_label, scales = "free") +
  scale_color_brewer(palette = "Set2") +
  labs(x = "T1 score", y = "T2 score",
       title = "Pattern Matching & Sentence Understanding: T2 vs T1",
       color = NULL) +
  theme(legend.position = "bottom")
```

# 9. Memory (Bogotá): super-low scores

```{r memory-zoom}
mem_bog <- scores |>
  filter(dataset == "pilot_uniandes_co_bogota", task_id == "memory-game")

mem_bog |>
  ggplot(aes(x = age, y = score)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", se = FALSE) +
  geom_hline(yintercept = -3, lty = 3, color = "red") +
  labs(x = "Age", y = "Memory IRT θ",
       title = "Memory scores by age (Bogotá)")
```

```{r memory-low}
mem_bog |>
  filter(score < -3) |>
  arrange(score) |>
  select(user_id, age, run_id, score, score_se, num_attempted, adaptive,
         scoring_model) |>
  head(15) |>
  knitr::kable(digits = 2)
```

# 10. Verdict table

Status codes:

- **OK**: usable as-is; growth signal trustworthy
- **OK-mode**: usable if you restrict to one administration mode (or split
  by it). NA→NA or A→A subsamples only.
- **CAVEAT**: usable but caveat-it heavily; effect size on growth is partly
  measurement artifact.
- **DROP**: known broken per release notes or local diagnostics — do not use
  for growth/SEM until next data release.

```{r verdict}
verdict <- tribble(
  ~dataset_label,            ~task_label,             ~status,    ~notes,
  # Leipzig
  "Germany — Leipzig",       "Hearts and Flowers",    "OK",       "Always non-adaptive at both waves; clean.",
  "Germany — Leipzig",       "Memory",                "OK",       "2026-08: DROP lifted. Grid is a calibrated item dimension, and the wave-1→2 grid switch shows continuous within-child θ (tasks/memory §4). Do not use raw sumscores across grids.",
  "Germany — Leipzig",       "Same and Different",    "OK",       "2026-08: corrected scoring landed (registry v2_3, DCC-confirmed final).",
  "Germany — Leipzig",       "Pattern Matching",      "OK",       "v1.2: bug fixed. Minor residual mode shift (+0.73); note T2>T1 training concern.",
  "Germany — Leipzig",       "Shape Rotation",        "OK",       "v1.2: bug fixed. Residual mode shift small.",
  "Germany — Leipzig",       "Math",                  "OK",       "v1.2: bug fixed; non-adapt−adapt shift now +0.30 (was +2.4). Mild EAP shrinkage across mode boundary.",
  "Germany — Leipzig",       "Vocabulary",            "OK",       "v1.2: bug fixed; shift now +0.10 (was +1.1). Clean.",
  "Germany — Leipzig",       "Sentence Understanding","OK",       "v1.2: bug fixed. Note T2>T1 training concern (release notes).",
  "Germany — Leipzig",       "Stories (ToM)",         "OK",       "No CAT at this site; modest signal.",
  "Germany — Leipzig",       "ROAR-Sentence",         "CAVEAT",   "Different score type (guessing-adjusted raw); thin T2.",
  "Germany — Leipzig",       "ROAR-Word",             "OK",       "ability_cat scale; engagement filter applied in cleaning.",

  # Bogotá
  "Colombia — Bogotá",       "Hearts and Flowers",    "OK",       "Always non-adaptive.",
  "Colombia — Bogotá",       "Memory",                "OK",       "2026-08: DROP lifted (see tasks/memory §4); super-low scores were the v1.0 bug. Sumscores not comparable across grids — use θ.",
  "Colombia — Bogotá",       "Same and Different",    "OK-mode",  "2026-08: corrected scoring landed (final). Still zero T2 at Bogotá — cross-sectional use only.",
  "Colombia — Bogotá",       "Pattern Matching",      "OK",       "v1.2: bug fixed. Residual mode shift −0.40.",
  "Colombia — Bogotá",       "Shape Rotation",        "OK",       "v1.2: bug fixed.",
  "Colombia — Bogotá",       "Math",                  "OK",       "v1.2: bug fixed; the apparent 'RTM' was the bug. Two-wave kids are NA→CAT, so mind mild EAP shrinkage.",
  "Colombia — Bogotá",       "Vocabulary",            "OK",       "v1.2: bug fixed. Thin T2 (n≈30).",
  "Colombia — Bogotá",       "Sentence Understanding","OK",       "Mostly mode-stable; positive growth.",
  "Colombia — Bogotá",       "Stories (ToM)",         "DROP",     "Zero T2 data (cross-sectional only).",
  "Colombia — Bogotá",       "ROAR-Sentence",         "OK",       "n≈165 with repeats; raw scale stable.",
  "Colombia — Bogotá",       "ROAR-Word",             "OK",       "Engagement filter applied (accuracy<.4 or median RT<500ms); floor cases are genuine non-readers.",
  "Colombia — Bogotá",       "ROAR-Phoneme",          "OK",       "Clean.",

  # Caquetá/Boyacá (rural Colombia) — wave 2 landed in the 2026-08 update.
  # Repeats concentrate in ROAR + Math; other tasks are cross-sectional so far.
  "Colombia — Caquetá/Boyacá", "Math",                 "OK",       "2026-08: 204 repeat pairs, cleanly CAT→CAT (mode-stable); mind EAP shrinkage.",
  "Colombia — Caquetá/Boyacá", "ROAR-Word",            "OK",       "2026-08: 201 repeat pairs, non-adaptive both waves; engagement filter applied.",
  "Colombia — Caquetá/Boyacá", "ROAR-Sentence",        "OK",       "2026-08: 174 repeat pairs; raw (guessing-adjusted) scale, non-IRT.",
  "Colombia — Caquetá/Boyacá", "ROAR-Phoneme",         "OK",       "2026-08: 210 repeat pairs, the largest ROAR-Phoneme longitudinal cell anywhere.",
  "Colombia — Caquetá/Boyacá", "Sentence Understanding","CAVEAT",  "Only 6 repeat pairs so far — cross-sectional for now.",
  "Colombia — Caquetá/Boyacá", "Vocabulary",           "CAVEAT",   "Only 4 repeat pairs so far — cross-sectional for now.",
  "Colombia — Caquetá/Boyacá", "Stories (ToM)",        "CAVEAT",   "Cross-sectional; also re-scored under registry v2_3 (see 00).",
  "Colombia — Caquetá/Boyacá", "Memory",               "CAVEAT",   "2026-08: scoring concern resolved (tasks/memory §4) but only 2 repeat pairs — cross-sectional for now.",
  "Colombia — Caquetá/Boyacá", "Same and Different",   "CAVEAT",   "2026-08: corrected scoring landed (final); cross-sectional so far.",
  "Colombia — Caquetá/Boyacá", "Hearts and Flowers",   "CAVEAT",   "4 repeat pairs; cross-sectional for now.",
  "Colombia — Caquetá/Boyacá", "Pattern Matching",     "CAVEAT",   "5 repeat pairs; cross-sectional for now.",
  "Colombia — Caquetá/Boyacá", "Shape Rotation",       "CAVEAT",   "1 repeat pair; cross-sectional for now."
)

verdict |>
  arrange(dataset_label, task_label) |>
  knitr::kable()
```

# 11. Recommended filter helpers

These functions encode the verdict above. They are exported to be used by
02 and 03. Anything not flagged is left alone.

```{r filter-helpers}
#' Drop runs from tasks marked DROP and (optionally) restrict mode-sensitive
#' tasks to within-mode pairs.
clean_for_growth <- function(scores, mode_strict = TRUE) {
  drops <- tribble(
    ~dataset,                       ~task_id,
    "pilot_mpieva_de_main",         "memory-game",
    "pilot_uniandes_co_bogota",     "memory-game",
    "pilot_uniandes_co_bogota",     "same-different-selection",
    "pilot_uniandes_co_bogota",     "theory-of-mind"
  )
  out <- scores |> anti_join(drops, by = c("dataset", "task_id"))
  if (mode_strict) {
    # Within each (dataset, child, task) keep only runs whose mode matches
    # that child's first run on that task (so we never mix modes within a
    # child's longitudinal trajectory).
    out <- out |>
      group_by(dataset, user_id, task_id) |>
      arrange(age) |>
      filter(is.na(adaptive) | adaptive == first(na.omit(adaptive))) |>
      ungroup()
  }
  out
}

#' Drop runs that look like truncated administrations (very few items
#' attempted) where this is most likely to confound the IRT score.
drop_short_runs <- function(scores, min_attempted = 5) {
  scores |>
    filter(is.na(num_attempted) | num_attempted >= min_attempted)
}

write_rds(verdict, here("data", "task_site_verdict.rds"))
```

# Summary (v1.2, updated for the 2026-08 per-site snapshot)

**2026-08 update.** The audit now runs on the per-site processed datasets
(see 00). What's new for integrity purposes: (a) **rural Colombia's second
wave** — repeats concentrate in ROAR (pa/swr/sre) + Math, with Math cleanly
CAT→CAT (no mode transition to worry about), while the other core tasks remain
cross-sectional there; (b) the new sites (Sheffield, MPIB-YS, UTDT-YS, Boston)
are single-wave, so no growth verdicts yet — their rows will be added when
repeats land; (c) scores were **re-scored under registry v2_3** — ToM changed
most (r = 0.88 vs June scores), so ToM-specific conclusions should be
re-validated; (d) unscoreable runs now arrive as explicit `exclusion` rows
(dropped in 00's cleaning) rather than being silently absent.

With the column-order scoring bug fixed upstream, most tasks are now **OK** —
the large mode-confound caveats from the v1.0 audit were the bug, not the
data. Three genuinely-remaining issues survive:

1. **Memory (`memory-game`) — DROP lifted (2026-08).** Grid is a calibrated
   item dimension, and the Leipzig wave-1→2 grid switch shows continuous
   within-child θ while raw sumscores visibly break (tasks/memory §4). Use θ,
   never cross-grid sumscores. The old super-low Bogotá scores were the v1.0
   bug.
2. **Same & Different — resolved (2026-08).** Corrected scoring landed in
   registry v2_3 (DCC-confirmed final). Still zero T2 at Bogotá →
   cross-sectional there.
3. **ROAR-Word non-engagement — handled in cleaning.** A subset of runs floor
   at θ = −6: ~17 are rushers (median RT < 500 ms), the rest genuine
   near-chance non-readers. `00` applies the filter (accuracy < 0.4 OR median
   RT < 500 ms). Floor rates: rural Colombia 16% > Western 7.3% > Bogotá 5.5%
   > Leipzig 1.6%.

Plus two lower-grade caveats: the **unflagged `adaptive`** rows (235, backfilled
to non-adaptive in cleaning — confirmed with DCC) and a small **EAP-shrinkage**
difference for the few tasks whose two-wave data crosses a CAT/non-CAT boundary
(notably Bogotá math). Training-vs-development for Pattern Matching & Sentence
Understanding (T2 > T1, release notes) remains an open question, not a defect.

The verdict table is saved to `data/task_site_verdict.rds`. Cleaning helpers
(`backfill_adaptive_na`, `roar_word_engagement_drops`, `clean_levante_scores`)
now live in `common.R` and are applied in `00_load_data`.
