---
title: "02 — Longitudinal traces and simple growth models"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    fig-width: 10
    fig-height: 6
execute:
  warning: false
  message: false
---

{{< include _disclaimer.qmd >}}

::: {.callout-note}
## Now on corrected data (v1.2)
Re-run on `levante_data_latest` **v1.2**, which fixes the IRT column-order
scoring bug. The spurious within-child *declines* seen in earlier (v1.0)
renders are gone: all within-child slopes are now positive (e.g. Math ≈ 0.6
θ/yr, Vocab/EF/reasoning ≈ 0.1–0.4). Any text below that frames a task as
"declining" reflects the old buggy data and is superseded by the corrected
figures.
:::

::: {.callout-note}
## 2026-08 data update: rural Colombia's second wave
The 2026-08 snapshot adds **Caquetá/Boyacá** as a fourth longitudinal dataset
(~213 repeat kids at a ~9.6-month gap, concentrated in ROAR + Math). Its
within-child growth is clean and — notably — **almost exactly matches the
cross-sectional age gradient**: Math 0.45 θ/yr within vs 0.45 across, ROAR-Word
0.47 vs 0.49, ROAR-Phoneme 0.28 vs 0.28, ROAR-Sentence 0.22 vs 0.23. Math
repeats are CAT→CAT (no mode transition), making this the cleanest
within-child growth replication in the project so far.
:::

# Purpose

00 inventoried what data we have; 01 audited per-(site, task) data integrity
and produced a verdict table at `data/task_site_verdict.rds`. This notebook
focuses on the datasets that actually have repeat assessments and asks:

1.  Do individual children show consistent growth in IRT ability scores
    across waves?
2.  Are within-individual growth rates broadly consistent with the
    cross-sectional age gradient in that dataset?

We restrict attention to the four datasets with meaningful longitudinal
signal as of the 2026-08 per-site snapshot:

- **Germany — Leipzig** (mpieva-de): ~3- and 7-month retest, ~226 kids.
- **Colombia — Bogotá** (uniandes-co-bogota): ~15-month gap, ~420 kids.
- **Colombia — Caquetá/Boyacá** (uniandes-co-rural): **new in the 2026-08
  update** — a ~9.6-month-gap second wave, ~213 repeat kids. Repeats are
  concentrated in **ROAR (pa/swr/sre) and Math** (~170–210 pairs each; Math is
  cleanly CAT→CAT); the other core tasks are still cross-sectional here.
- **Canada — Ontario** (western-ca): ~12-month gap, still only ~10–21 repeat
  kids per task, so it falls below the ≥ 30-per-task threshold for the mixed
  models and mainly appears in the spaghetti/cross-sectional views.

```{r setup}
source(here::here("common.R"))
library(lme4)
library(broom.mixed)
library(ggrepel)

scores_all <- read_rds(here("data", "scores_all_sites.rds"))
```

# Restrict to longitudinal datasets

```{r long-datasets}
long_datasets <- c("pilot_mpieva_de_main",
                   "pilot_uniandes_co_bogota",
                   "pilot_uniandes_co_rural",
                   "pilot_western_ca_main")

scores <- scores_all |>
  filter(dataset %in% long_datasets,
         !is.na(score), !is.na(age))

scores |>
  count(dataset_label, task_label) |>
  pivot_wider(names_from = dataset_label, values_from = n,
              values_fill = 0L) |>
  knitr::kable()
```

# Adaptive vs non-adaptive scores

Pilot data mixes non-adaptive and computer-adaptive (CAT) administrations of
the same task. Under v1.0's scoring bug this produced large spurious step
jumps across the mode boundary; on v1.2 (fixed) the residual mode shift is
small (< 0.75 logits — see 01), attributable to mild ceiling on the easy
non-adaptive forms plus EAP shrinkage. We still flag the mode split below so
within-mode vs across-mode comparisons stay visible.

```{r adaptive-mix}
scores |>
  group_by(dataset_label, task_label, adaptive) |>
  summarise(n_runs = n_distinct(run_id), .groups = "drop") |>
  pivot_wider(names_from = adaptive, values_from = n_runs, values_fill = 0L,
              names_prefix = "adaptive_") |>
  knitr::kable()
```

# Spaghetti plots

Every child with ≥ 2 runs on a given task, one line per child, raw IRT score.
Color marks adaptive vs non-adaptive runs.

```{r repeat-data}
repeats <- scores |>
  group_by(dataset, user_id, task_id) |>
  filter(n_distinct(run_id) > 1) |>
  ungroup() |>
  arrange(dataset, user_id, task_id, age)

glue("Repeat (dataset, child, task) cells: {n_distinct(paste(repeats$dataset, repeats$user_id, repeats$task_id))}")
```

```{r spaghetti-de, fig.height=7}
repeats |>
  filter(dataset == "pilot_mpieva_de_main") |>
  ggplot(aes(x = age, y = score, group = user_id, color = adaptive)) +
  geom_line(alpha = 0.4) +
  geom_point(alpha = 0.4, size = 0.6) +
  facet_wrap(vars(task_category, task_label), scales = "free_y") +
  scale_color_manual(values = c(`TRUE` = "firebrick", `FALSE` = "steelblue"),
                     na.value = "grey60") +
  labs(title = "Germany — Leipzig spaghetti (kids with ≥ 2 runs)",
       x = "Age (years)", y = "IRT ability")
```

```{r spaghetti-co, fig.height=7}
repeats |>
  filter(dataset == "pilot_uniandes_co_bogota") |>
  ggplot(aes(x = age, y = score, group = user_id, color = adaptive)) +
  geom_line(alpha = 0.3) +
  geom_point(alpha = 0.3, size = 0.6) +
  facet_wrap(vars(task_category, task_label), scales = "free_y") +
  scale_color_manual(values = c(`TRUE` = "firebrick", `FALSE` = "steelblue"),
                     na.value = "grey60") +
  labs(title = "Colombia — Bogotá spaghetti (kids with ≥ 2 runs)",
       x = "Age (years)", y = "IRT ability")
```

```{r spaghetti-co-rural, fig.height=7}
repeats |>
  filter(dataset == "pilot_uniandes_co_rural") |>
  ggplot(aes(x = age, y = score, group = user_id, color = adaptive)) +
  geom_line(alpha = 0.3) +
  geom_point(alpha = 0.3, size = 0.6) +
  facet_wrap(vars(task_category, task_label), scales = "free_y") +
  scale_color_manual(values = c(`TRUE` = "firebrick", `FALSE` = "steelblue"),
                     na.value = "grey60") +
  labs(title = "Colombia — Caquetá/Boyacá spaghetti (kids with ≥ 2 runs; new in 2026-08)",
       x = "Age (years)", y = "IRT ability")
```

```{r spaghetti-ca, fig.height=7}
repeats |>
  filter(dataset == "pilot_western_ca_main") |>
  ggplot(aes(x = age, y = score, group = user_id, color = adaptive)) +
  geom_line(alpha = 0.5) +
  geom_point(alpha = 0.5, size = 0.8) +
  facet_wrap(vars(task_category, task_label), scales = "free_y") +
  scale_color_manual(values = c(`TRUE` = "firebrick", `FALSE` = "steelblue"),
                     na.value = "grey60") +
  labs(title = "Canada — Ontario spaghetti (kids with ≥ 2 runs)",
       x = "Age (years)", y = "IRT ability")
```

# Simple growth models per (dataset, task)

For each (dataset, task), fit

`score ~ age_c + (1 | user_id)`

with `age_c` = age − mean(age). With mostly 2 timepoints the random-slope
variance is unidentifiable, so we read the fixed effect of `age_c` as the
average within-individual slope. Per-child heterogeneity is examined
separately below.

Where both adaptive and non-adaptive versions are present we include
`adaptive` as a fixed effect.

```{r fit-models}
fit_growth <- function(df) {
  df <- df |> mutate(age_c = age - mean(age, na.rm = TRUE))
  has_both_adaptive <- length(unique(na.omit(df$adaptive))) > 1
  fm <- if (has_both_adaptive) {
    score ~ age_c + adaptive + (1 | user_id)
  } else {
    score ~ age_c + (1 | user_id)
  }
  tryCatch(
    lmer(fm, data = df, REML = TRUE,
         control = lmerControl(optimizer = "bobyqa",
                               optCtrl = list(maxfun = 2e5))),
    error = function(e) {
      message(glue("  fit failed: {conditionMessage(e)}"))
      NULL
    }
  )
}

growth_data <- scores |>
  group_by(dataset, task_id) |>
  filter(n_distinct(user_id[duplicated(user_id) | duplicated(user_id, fromLast = TRUE)]) >= 30) |>
  ungroup()

growth_fits <- growth_data |>
  group_by(dataset, dataset_label, task_id, task_label, task_category) |>
  nest() |>
  mutate(fit = map(data, fit_growth)) |>
  filter(!map_lgl(fit, is.null))

glue("Fitted {nrow(growth_fits)} (dataset, task) growth models.")
```

```{r extract-coefs}
within_slopes <- growth_fits |>
  mutate(coef_tbl = map(fit, \(m) broom.mixed::tidy(m, effects = "fixed",
                                                    conf.int = TRUE))) |>
  select(dataset_label, task_label, task_category, coef_tbl) |>
  unnest(coef_tbl) |>
  filter(term == "age_c") |>
  select(dataset_label, task_label, task_category,
         within_slope = estimate, within_slope_se = std.error,
         within_ci_lo = conf.low, within_ci_hi = conf.high)

knitr::kable(within_slopes, digits = 3)
```

# Cross-sectional age gradients for comparison

OLS slope of score on age within each (dataset, task), using each child's
first observation.

```{r cross-section-slopes}
cross_slopes <- scores |>
  group_by(dataset, user_id, task_id) |>
  arrange(age) |>
  slice(1) |>
  ungroup() |>
  group_by(dataset, dataset_label, task_id, task_label, task_category) |>
  filter(n() >= 30) |>
  summarise(
    cross_slope = coef(lm(score ~ age))[2],
    cross_n     = n(),
    .groups = "drop"
  )

slope_compare <- within_slopes |>
  inner_join(cross_slopes, by = c("dataset_label", "task_label",
                                   "task_category"))

knitr::kable(slope_compare, digits = 3)
```

```{r slope-compare-plot, fig.height=5}
slope_compare |>
  ggplot(aes(x = cross_slope, y = within_slope, color = task_category)) +
  geom_abline(slope = 1, intercept = 0, lty = 3, color = "grey50") +
  geom_hline(yintercept = 0, lty = 3, color = "grey80") +
  geom_vline(xintercept = 0, lty = 3, color = "grey80") +
  geom_errorbar(aes(ymin = within_ci_lo, ymax = within_ci_hi),
                alpha = 0.5, width = 0) +
  geom_point(size = 2.5) +
  geom_text_repel(aes(label = task_label), size = 3,
                  show.legend = FALSE, max.overlaps = 30) +
  facet_wrap(~dataset_label) +
  scale_colour_task() +
  labs(x = "Cross-sectional slope (per year)",
       y = "Within-child slope (per year)",
       title = "Within-child growth vs cross-sectional age gradient",
       subtitle = "Dotted line = identity",
       color = NULL) +
  theme(legend.position = "bottom")
```

# Fitted trajectories

Population-mean model curves overlaid on raw spaghetti, faceted by task.

```{r fitted-curves, fig.height=10, fig.width=12}
predict_band <- function(fit, df) {
  if (is.null(fit)) return(NULL)
  age_grid <- seq(min(df$age, na.rm = TRUE),
                  max(df$age, na.rm = TRUE), length.out = 40)
  age_mean <- mean(df$age, na.rm = TRUE)
  newd <- tibble(age = age_grid, age_c = age_grid - age_mean)
  if ("adaptive" %in% all.vars(formula(fit))) newd$adaptive <- FALSE
  newd$score <- predict(fit, newdata = newd, re.form = NA,
                        allow.new.levels = TRUE)
  newd |> select(age, score)
}

fitted_lines <- growth_fits |>
  mutate(pred = map2(fit, data, predict_band)) |>
  select(dataset_label, task_label, task_category, pred) |>
  unnest(pred)

ggplot() +
  geom_line(data = repeats,
            aes(x = age, y = score, group = user_id),
            alpha = 0.15, color = "grey40") +
  geom_line(data = fitted_lines,
            aes(x = age, y = score, color = dataset_label),
            linewidth = 1) +
  facet_wrap(vars(task_category, task_label), scales = "free_y") +
  scale_color_brewer(palette = "Set1", name = NULL) +
  labs(title = "Fitted growth trajectories on top of raw spaghetti",
       x = "Age (years)", y = "IRT ability") +
  theme(legend.position = "bottom")
```

# Child-level slopes

Per-child OLS slopes for kids with ≥ 2 runs (rise / run when N = 2).

```{r per-child-slopes, fig.height=8}
child_slopes <- repeats |>
  group_by(dataset, dataset_label, task_id, task_label, task_category, user_id) |>
  filter(n_distinct(run_id) >= 2, diff(range(age)) > 0.1) |>
  summarise(
    n_runs    = n_distinct(run_id),
    age_mean  = mean(age),
    age_span  = diff(range(age)),
    slope     = coef(lm(score ~ age))[2],
    .groups = "drop"
  ) |>
  filter(is.finite(slope))

child_slopes |>
  group_by(dataset_label, task_label) |>
  filter(n() >= 30) |>
  ungroup() |>
  ggplot(aes(x = slope, fill = dataset_label)) +
  geom_histogram(bins = 30, alpha = 0.7, position = "identity") +
  geom_vline(xintercept = 0, lty = 3) +
  facet_wrap(vars(task_category, task_label), scales = "free") +
  scale_fill_brewer(palette = "Set1", name = NULL) +
  labs(title = "Per-child OLS slopes (per year)",
       x = "Slope of score on age",
       y = "Children") +
  theme(legend.position = "bottom")
```

```{r slope-summary}
child_slopes |>
  group_by(dataset_label, task_label) |>
  summarise(n = n(), mean_slope = mean(slope), sd_slope = sd(slope),
            .groups = "drop") |>
  filter(n >= 30) |>
  arrange(dataset_label, task_label) |>
  knitr::kable(digits = 3)
```

# Save outputs

```{r save}
write_rds(within_slopes,  here("data", "growth_within_slopes.rds"))
write_rds(slope_compare,  here("data", "growth_slope_comparison.rds"))
write_rds(child_slopes,   here("data", "growth_child_slopes.rds"))

glue("Saved within-slope and per-child slope tables to data/.")
```

# Open questions for 02

- The CO-bog 1-year gap means most kids only have 2 timepoints. That's enough
  to estimate a slope per child, but the per-individual SE is large. The
  latent-growth approach in 02 pools across tasks/constructs.
- mpieva-de's 3- and 7-month gaps give a within-year change estimate that
  should be small in absolute terms; per-child OLS slopes there are noisy.
- Tasks with adaptive/non-adaptive mixing (vocab in mpieva-de especially)
  might bias estimates; flagged again in 02.
