---
title: "04 — Construct structure: emergent dimensions, differentiation, and the role of g"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 9
    fig-height: 5.5
execute:
  warning: false
  message: false
---

{{< include _disclaimer.qmd >}}

# The question

03 found the core tasks resolve into a **strong general factor** with only
weakly-separable domains. Here we interrogate the construct structure with the
**full measure set** — treating ROAR (reading: word/sentence/phoneme) and MEFS
(executive function) as *real measures of the underlying structure*, not just
external validators. They are not LEVANTE innovations, but they are valid
measures that help reveal the true dimensional structure.

Plan:

1. **Emergent structure (ESEM)** — what dimensions do the data prefer, with all
   13 measures and cross-loadings allowed?
2. **Developmental differentiation** — how do the factor correlations change
   with age, under **both** the theory-driven structure **and** the emergent
   structure (local SEM)?
3. **A shared-speed (method) artifact?** — does general response speed account
   for the cross-task covariation?

```{r setup}
source(here::here("common.R"))
library(lavaan); library(psych)
sc <- read_rds(here("data/scores_all_sites.rds"))
sites4 <- c("pilot_mpieva_de_main","pilot_uniandes_co_bogota",
            "pilot_uniandes_co_rural","pilot_western_ca_main")

# MEFS (raw total_score, age-increasing) — from the levante-pilots scoring outputs;
# not in levante_data_latest. Merge by user_id.
mefs <- read_rds(here("../levante-pilots/02_scoring_outputs/scores/scores_mefs.rds")) |>
  filter(metric_type == "total_score") |>
  group_by(user_id) |> arrange(age) |> slice(1) |> ungroup() |>
  select(user_id, mefs = metric_value)

fo <- sc |> filter(dataset %in% sites4, task_id %in% c(core_task_ids,"swr","sre","pa"),
                   !is.na(score)) |>
  group_by(user_id, short) |> arrange(age) |> slice(1) |> ungroup()
age1 <- fo |> group_by(user_id) |> arrange(age) |> slice(1) |> ungroup() |> select(user_id, age)
xw <- fo |> select(user_id, short, score) |> distinct(user_id, short, .keep_all = TRUE) |>
  pivot_wider(names_from = short, values_from = score) |>
  left_join(mefs, by = "user_id") |> left_join(age1, by = "user_id")

vars13 <- c("matrix","mrot","math","hf","mg","sds","mefs","trog","vocab","tom","swr","sre","pa")
```

Pooled cross-sectional sample (first observation per child, four data-rich
sites): **N = `r nrow(xw)`** total; **`r sum(!is.na(xw$mefs))`** with MEFS,
**`r sum(!is.na(xw$swr))`** with ROAR-Word. Reading and MEFS enter as
first-class measures alongside the nine core tasks.

# 1. Emergent structure (ESEM, 13 measures)

```{r esem, cache=TRUE}
R <- lavCor(xw |> select(all_of(vars13)), missing = "fiml"); N <- nrow(xw)
map_dfr(1:5, function(nf){
  f <- suppressWarnings(fa(R, nfactors = nf, n.obs = N, fm = "ml", rotate = "oblimin"))
  tibble(factors = nf, RMSEA = f$RMSEA[1], TLI = f$TLI, BIC = f$BIC)
}) |> knitr::kable(digits = 3, caption = "EFA fit by number of factors (FIML correlation, 13 measures)")
```

```{r esem-load}
f2 <- suppressWarnings(fa(R, nfactors = 2, n.obs = N, fm = "ml", rotate = "oblimin"))
L <- unclass(f2$loadings); colnames(L) <- c("F1","F2")
as.data.frame(round(L, 2)) |> rownames_to_column("measure") |>
  mutate(kind = case_when(measure %in% c("swr","sre","pa") ~ "reading (ROAR)",
                          measure == "mefs" ~ "EF (MEFS)", TRUE ~ "core")) |>
  knitr::kable(caption = glue::glue("2-factor ESEM loadings (oblimin; factor r = {round(f2$Phi[1,2],2)})"))
```

Two correlated dimensions (r ≈ 0.67) emerge. They map onto a broad
**fluid / nonverbal** factor (matrix, mrot, the memory tasks sds & mg, **MEFS**)
and a **verbal / literacy** factor (vocab, ToM, **the ROAR reading measures**),
with math split between them. The placement of the *new* measures is exactly as
theory would predict — **MEFS lands on the fluid side** (.59) and **all three
ROAR reading measures on the verbal/literacy side** (.49–.84) — so they sharpen
rather than disrupt the two-factor picture.

The rotation is *not* perfectly clean, though, and the messiness is informative:
TROG (grammar) loads with the fluid cluster (.69) and Hearts & Flowers with the
literacy cluster (.54). There is **no separate EF factor** — the four EF-ish
tasks (hf, sds, mg, MEFS) scatter across both dimensions rather than cohering.
This cross-domain bleed is the signature of a strong positive manifold: with so
much shared general variance, a two-factor rotation recovers a verbal-vs-nonverbal
split rather than the reasoning/EF/language theory. (BIC keeps improving past two
factors, but the extra factors are small and measure-specific.)

# 2. Developmental differentiation

03 selected a **two-factor** structure (fluid / verbal) as the parsimony-preferred,
proper model, so we carry just that forward — here on the full measure set, with
MEFS in *fluid* and the ROAR reading tasks in *verbal*. We track the
**fluid–verbal correlation** as a smooth function of age (kernel-weighted local
SEM, bandwidth 1.5 yr): if the two abilities differentiate, it should fall with age.

```{r losem, cache=TRUE}
emergent2 <- "fluid =~ matrix + mrot + math + hf + mg + sds + mefs
              verbal =~ trog + vocab + tom + swr + sre + pa"

losem_cor <- function(model, pairs, ages = seq(6.5, 11.5, by = 0.5), h = 1.5) {
  map_dfr(ages, function(a0) {
    w <- dnorm((xw$age - a0) / h)
    fit <- tryCatch(cfa(model, cbind(xw, .w = w), std.lv = TRUE, missing = "fiml",
                        sampling.weights = ".w"), error = function(e) NULL)
    if (is.null(fit) || !lavInspect(fit, "converged")) return(NULL)
    cc <- lavInspect(fit, "cor.lv")
    bind_cols(focal_age = a0,
              map_dfr(pairs, ~tibble(pair = paste(.x, collapse="–"), r = cc[.x[1], .x[2]])))
  })
}
ld_emerg <- losem_cor(emergent2, list(c("fluid","verbal")))
```

```{r losem-plot, fig.height=4.6}
ld_emerg |>
  ggplot(aes(focal_age, r)) +
  geom_line(linewidth = 1, color = "#268bd2") + geom_point(size = 2, color = "#268bd2") +
  scale_x_continuous(breaks = 7:11) + coord_cartesian(ylim = c(0.75, 1)) +
  labs(x = "Focal age (years)", y = "Fluid–verbal correlation",
       title = "The two abilities differentiate with age",
       subtitle = "Fluid–verbal correlation falls from ≈ 0.96 (age 7) to ≈ 0.87 (age 11)")
```

```{r losem-tab}
ld_emerg |> filter(focal_age %in% c(7, 9, 11)) |>
  transmute(focal_age, `fluid–verbal r` = round(r, 2)) |>
  knitr::kable(caption = "Fluid–verbal correlation by focal age")
```

The fluid and verbal abilities **differentiate with age**: their correlation
falls from ≈ 0.96 at age 7 to ≈ 0.87 at age 11 (June render: 0.93 → 0.83 —
the 2026-08 rural addition raises the overall level of integration but leaves
the decline intact). So even the two broad,
heavily-correlated dimensions pull apart as children get older — the
developmental de-correlation (Spearman's "law of diminishing returns") is visible
at the coarsest level of structure. And per `03` this is **structural, not
measurement drift**: loadings stay ~age-invariant (metric ΔCFI ≈ −0.01) while the
factor correlation changes.

## Does it replicate across sites?

For cross-site comparability we use a core **8-task** 2-factor (fluid =
matrix / mrot / math / H&F / memory; verbal = TROG / vocab / ToM) — dropping
Same-&-Different (sparse adaptive coverage in age-split subsamples) and the
MEFS / ROAR measures (coverage varies by site). We compare the fluid–verbal
correlation in younger vs older children at each site, and trace it continuously
by site (local SEM).

```{r site-prep}
xws <- xw |> left_join(distinct(fo, user_id, dataset), by = "user_id") |>
  mutate(site = recode(dataset, pilot_mpieva_de_main = "Germany", pilot_uniandes_co_bogota = "Bogotá-urban",
                       pilot_uniandes_co_rural = "Bogotá-rural", pilot_western_ca_main = "Western"))
m8 <- "fluid =~ matrix+mrot+math+hf+mg\n verbal =~ trog+vocab+tom"
fv <- function(d){ f <- suppressWarnings(tryCatch(cfa(m8, d, std.lv = TRUE, missing = "fiml"), error = function(e) NULL))
  if (is.null(f) || !lavInspect(f, "converged")) return(NA_real_); lavInspect(f, "cor.lv")["fluid","verbal"] }
```

```{r site-youngold}
xws |> group_by(site) |> group_modify(function(d, ...){ md <- median(d$age)
  tibble(n = nrow(d), younger = fv(filter(d, age < md)), older = fv(filter(d, age >= md))) }) |> ungroup() |>
  mutate(differentiation = younger - older, across(c(younger, older, differentiation), \(x) round(x, 3))) |>
  knitr::kable(caption = "Fluid–verbal correlation younger vs older, by site (split at site median age)")
```

```{r site-losem, fig.height=4.6}
losem_site <- function(d, ages = seq(6.5, 11.5, .5), h = 1.5) map_dfr(ages, function(a0){ w <- dnorm((d$age - a0)/h)
  f <- suppressWarnings(tryCatch(cfa(m8, cbind(d, .w = w), std.lv = TRUE, missing = "fiml", sampling.weights = ".w"), error = function(e) NULL))
  if (is.null(f) || !lavInspect(f, "converged")) return(NULL)
  tibble(focal_age = a0, r = lavInspect(f, "cor.lv")["fluid","verbal"]) })
xws |> group_by(site) |> group_modify(~losem_site(.x)) |> ungroup() |>
  ggplot(aes(focal_age, r, color = site)) +
  geom_line(linewidth = 1) + geom_point(size = 1.6) +
  scale_color_brewer(palette = "Set1", name = NULL) +
  scale_x_continuous(breaks = 7:11) + coord_cartesian(ylim = c(0.6, 1)) +
  labs(x = "Focal age (years)", y = "Fluid–verbal correlation",
       title = "Differentiation with age, by site",
       subtitle = "Declines in Germany & Western; flatter in Bogotá-urban") +
  theme(legend.position = "bottom")
```

The differentiation **replicates across sites**: in every site the younger
children show a higher fluid–verbal correlation than the older (differentiation
Δ ≈ 0.06–0.16). It is clearest in **Bogotá-rural (Δ 0.16)** and **Western
(Δ 0.12)**, and more modest in **Germany (Δ 0.06)** and **Bogotá-urban
(Δ 0.07)**. The *overall* correlation sits in a narrow band across sites
(≈ 0.83–0.86), so the **level** of integration is similar everywhere even though
the developmental **slope** varies. (Bogotá-rural's continuous local SEM didn't
converge across the full age grid — sparse coverage — so it appears in the table
but not the curve; its younger/older split confirms the same pattern.)

# 3. Is the g-loading a shared-speed (method) artifact?

A general response-speed factor (same 4-AFC tablet format) could inflate
cross-task covariation. We build a general-speed factor from per-person median
log RTs across the 8 timed core tasks (Memory has no RT) and relate it to
cognitive g.

```{r rt, cache=TRUE}
rttasks <- setdiff(core_task_ids, "memory-game")
fr_runs <- sc |> filter(dataset %in% sites4, task_id %in% rttasks) |>
  group_by(user_id, task_id) |> arrange(age) |> slice(1) |> ungroup() |> pull(run_id)
trrt <- load_levante_trials(task_ids = rttasks) |>
  filter(run_id %in% fr_runs, !is.na(rt_numeric), rt_numeric > 200, rt_numeric < 30000) |>
  group_by(run_id, user_id, short) |> summarise(lrt = median(log(rt_numeric)), .groups = "drop")
rtwide <- trrt |> group_by(short) |> mutate(z = as.numeric(scale(lrt))) |> ungroup() |>
  select(user_id, short, z) |> distinct(user_id, short, .keep_all = TRUE) |>
  pivot_wider(names_from = short, values_from = z)
cm <- cor(rtwide |> select(-user_id) |> as.matrix(), use = "pairwise")
gfit <- cfa("g =~ matrix+mrot+math+hf+mg+sds+trog+vocab+tom", xw, std.lv=TRUE, missing="fiml")
gsc <- tibble(user_id = xw$user_id, g = as.numeric(lavPredict(gfit)[, "g"]))
spd <- rtwide |> mutate(speed = rowMeans(across(-user_id), na.rm = TRUE)) |> select(user_id, speed)
sg <- gsc |> inner_join(spd, by = "user_id")
tibble(metric = c("mean cross-task RT correlation", "cor(cognitive g, slowness)"),
       value = c(round(mean(cm[upper.tri(cm)], na.rm=TRUE), 3),
                 round(cor(sg$g, sg$speed, use="complete.obs"), 3))) |>
  knitr::kable(caption = "A general speed factor exists but is ~orthogonal to ability")
```

A general speed factor exists (RTs correlate ≈ 0.34 across tasks) but is
**nearly orthogonal to cognitive g** (r ≈ 0.15, positive — slower, more
deliberate children if anything score slightly higher). The g-loading is
**not** a shared-speed/method artifact.

# Synthesis

With ROAR and MEFS included as real measures, the structure sharpens but the
big picture is unchanged and now better-anchored:

| Finding | Evidence |
|---|---|
| ≈ **two** correlated dimensions, not three | ESEM: fluid (incl. MEFS) + verbal/literacy (incl. ROAR), r ≈ 0.67 |
| Reasoning & EF do **not** separate | fused at r ≈ 0.92, flat across all ages; no distinct EF factor in ESEM |
| External measures behave as theory predicts | MEFS → fluid side (.59); ROAR reading → verbal/literacy side (.49–.84) |
| Dimensions **differentiate with age** | fluid–verbal correlation falls 0.96→0.87 (age 7→11, 2026-08 data); structural, not loading drift (03); replicates across sites (Δ 0.06–0.15) |
| g-loading is substantive | not a site-mean effect (03), not a speed artifact (§3) |

The battery, at these ages, measures a **strong fluid/executive ability and a
separable verbal/literacy ability**, with the verbal dimension differentiating
from the fluid core as children get older. Adding MEFS and the ROAR measures
*confirms* this two-dimensional picture rather than revealing more domains — and
notably, reading relates to the whole structure (g), not just to language
(language–reading ≈ reasoning–reading), so even with reading as a first-class
measure the differential structure is fluid-vs-verbal, not finer.

**Next:** a longitudinal second-order model (does *g* grow; does the
fluid/verbal split widen within child over time?) as waves accrue; and bringing
MEFS into the cross-language DIF / measurement-invariance work now that it's a
modeled measure.

```{r save}
write_rds(list(esem2 = f2, losem_emergent = ld_emerg,
               rt = list(mean_r = mean(cm[upper.tri(cm)], na.rm=TRUE),
                         g_speed = cor(sg$g, sg$speed, use="complete.obs"))),
          here::here("data/construct_structure_04.rds"))
```
