---
title: "05 — Battery design: how short can the battery be and still recover the factors?"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 9
    fig-height: 5.5
execute:
  warning: false
  message: false
---

{{< include _disclaimer.qmd >}}

# The question

03–04 established that the core battery measures a **strong, highly-correlated
factor structure**. That has a practical payoff: if the constructs overlap
heavily, we should be able to recover each child's factor scores from a
**shorter** battery. This notebook quantifies that trade-off.

We combine three ingredients —

1. the **calibrated measurement model** (the theory-driven factor structure),
2. each task's **marginal reliability**, and
3. each task's **administration time** (newest = *adaptive* version where one exists) —

and ask, for any subset of tasks: **how well does it recover the factor
scores, and how many minutes does it cost?** We then read off three battery
versions (full / minor reduction / minimal) and the efficient frontier between
them.

Per the design brief we target two theory-driven structures (MEFS is already
cut, so it does not appear):

- **2-factor:** a *fluid / nonverbal* factor (matrix, mental rotation, math,
  hearts & flowers, memory, same-&-different) and a *verbal / literacy* factor
  (TROG, vocab, ToM, ROAR word/sentence/phoneme). Note TROG is placed with
  **language** here, on theory grounds, despite loading with fluid in 04's EFA.
- **4-factor:** reasoning / executive function / language / reading.

"Recovery" = **factor-score determinacy**: the correlation between the factor
score estimated from the administered subset and the true latent factor. We
calibrate the measurement model *once* on the full battery, then evaluate every
subset against that fixed calibration — exactly how one would design a short
form.

```{r setup}
source(here::here("common.R"))
library(lavaan)  # NB: do not attach MASS — it masks dplyr::select; use MASS::mvrnorm qualified
sc <- read_rds(here("data/scores_all_sites.rds"))
sites4 <- c("pilot_mpieva_de_main","pilot_uniandes_co_bogota",
            "pilot_uniandes_co_rural","pilot_western_ca_main")

tasks12 <- tibble(
  short   = c("matrix","mrot","math","hf","mg","sds","trog","vocab","tom","swr","sre","pa"),
  task_id = c("matrix-reasoning","mental-rotation","egma-math","hearts-and-flowers","memory-game",
              "same-different-selection","trog","vocab","theory-of-mind","swr","sre","pa"),
  f4 = c("reasoning","reasoning","reasoning","ef","ef","ef",
         "language","language","language","reading","reading","reading"),
  f2 = c("fluid","fluid","fluid","fluid","fluid","fluid",
         "verbal","verbal","verbal","verbal","verbal","verbal"))
# tasks with an adaptive version → use ONLY adaptive runs (the newest/deployed form)
adaptive_capable <- c("egma-math","matrix-reasoning","mental-rotation",
                      "same-different-selection","trog","vocab")

# analysis sample: first observation per child per task, 4 data-rich sites,
# restricted to the newest (adaptive) version where one exists.
fo <- sc |>
  filter(dataset %in% sites4, task_id %in% tasks12$task_id, !is.na(score)) |>
  filter(!(task_id %in% adaptive_capable) | adaptive) |>
  group_by(user_id, task_id) |> arrange(age) |> slice(1) |> ungroup()
```

# 1. Ingredients

## Marginal reliability

Empirical (marginal) reliability per task from the EAP standard errors,
ρ = Var(θ̂) / [Var(θ̂) + mean(SE²)], computed on the deployed-version runs.

```{r reliability}
rel <- fo |> group_by(short) |>
  summarise(rel = var(score) / (var(score) + mean(score_se^2, na.rm = TRUE)),
            .groups = "drop")
# sre is a guessing-adjusted *count* score (no model SE) → reliability not
# estimable from SE; its odd/even split-half is inflated by speededness, so we
# leave it blank and let sre enter the recovery model through its loading only.
rel |> right_join(tasks12 |> select(short), by = "short") |>
  arrange(rel) |> mutate(rel = round(rel, 3)) |>
  knitr::kable(caption = "Marginal reliability (deployed-version runs; sre = NA, count-based score)")
```

Reliabilities run from ~0.72 (matrix, ToM) to ~0.93 (ROAR-Word, math). **Same &
Different is the weakest (~0.62)** — its adaptive version has few runs so far
(n≈80) and its scoring is still being revised, so treat that number as
provisional.

## Administration time

Median wall-clock duration per run (newest/adaptive version), robust to the
occasional paused-and-resumed session. This is approximate seat time — it
includes instructions and natural pauses, so it is an upper-ish estimate of
active testing time.

```{r duration, cache=TRUE}
dur <- map_dfr(tasks12$task_id, function(tid){
  runs <- fo |> filter(task_id == tid) |> pull(run_id)
  tr <- load_levante_trials(task_ids = tid) |> filter(run_id %in% runs, !is.na(timestamp))
  pr <- tr |> group_by(run_id) |>
    summarise(span = as.numeric(difftime(max(timestamp), min(timestamp), units = "mins")),
              .groups = "drop")
  tibble(task_id = tid, minutes = median(pr$span, na.rm = TRUE))
}) |> left_join(tasks12 |> select(short, task_id), by = "task_id")

dur |> select(short, minutes) |> arrange(desc(minutes)) |>
  mutate(minutes = round(minutes, 1)) |>
  knitr::kable(caption = glue::glue("Median administration time (full battery = {round(sum(dur$minutes),1)} min)"))
```

ToM is by far the longest (≈10 min — it has spoken-story stimuli); ROAR-Word
(≈8 min, 120 quick items) is next. The full 12-task battery runs ≈ **55 min**
(2026-08 durations; ≈59 min in the June render).

## The calibrated measurement model

Two CFAs (FIML, pooled across the four sites, scores standardized within task).
We extract the standardized loadings and factor correlations — the fixed
calibration that every candidate battery is scored against.

```{r cfa, cache=TRUE}
xw <- fo |> select(user_id, short, score) |> group_by(short) |>
  mutate(z = as.numeric(scale(score))) |> ungroup() |>
  select(user_id, short, z) |> distinct(user_id, short, .keep_all = TRUE) |>
  pivot_wider(names_from = short, values_from = z)

m2 <- "fluid  =~ matrix+mrot+math+hf+mg+sds
       verbal =~ trog+vocab+tom+swr+sre+pa"
m4 <- "reasoning =~ matrix+mrot+math
       ef        =~ hf+mg+sds
       language  =~ trog+vocab+tom
       reading   =~ swr+sre+pa"
f2 <- cfa(m2, xw, std.lv = TRUE, missing = "fiml")
f4 <- cfa(m4, xw, std.lv = TRUE, missing = "fiml")

# pull standardized Lambda (tasks x factors, simple structure) and factor cor Phi
model_LP <- function(fit){
  std <- standardizedSolution(fit)
  L <- std |> filter(op == "=~") |> transmute(f = lhs, short = rhs, lambda = est.std)
  facs <- unique(L$f)
  Lam <- matrix(0, nrow(tasks12), length(facs), dimnames = list(tasks12$short, facs))
  for (i in seq_len(nrow(L))) Lam[L$short[i], L$f[i]] <- L$lambda[i]
  Phi <- diag(length(facs)); dimnames(Phi) <- list(facs, facs)
  pc <- std |> filter(op == "~~", lhs != rhs, lhs %in% facs, rhs %in% facs)
  for (i in seq_len(nrow(pc))) { Phi[pc$lhs[i], pc$rhs[i]] <- pc$est.std[i]; Phi[pc$rhs[i], pc$lhs[i]] <- pc$est.std[i] }
  list(Lam = Lam, Phi = Phi)
}
g2 <- model_LP(f2); g4 <- model_LP(f4)
mins <- setNames(dur$minutes[match(tasks12$short, dur$short)], tasks12$short)
```

Both models fit well (2-factor CFI = `r round(fitMeasures(f2,"cfi"),3)`,
4-factor CFI = `r round(fitMeasures(f4,"cfi"),3)`). The factor correlations are
high — in the 4-factor model reasoning and EF are almost the same factor
(r = `r round(g4$Phi["reasoning","ef"],2)`); fluid–verbal in the 2-factor model
is r = `r round(g2$Phi["fluid","verbal"],2)`. (The pooled FIML fit carries a
thin-coverage caveat from Bogotá's partial battery; per-site structure in 03–04
replicates these loadings, so they are an adequate design calibration.)

# 2. Recovery metric, and a simulation check

For an administered subset *B*, factor-score determinacy is computed from the
calibrated loadings Λ and factor correlation Φ (restricted to *B*):
the model-implied indicator correlation is Σ = ΛΦΛ′ (unit diagonal), the
regression score weights are *W* = ΦΛ′Σ⁻¹, and the determinacy for factor *j* is
the √ of the *j*-th diagonal of ΦΛ′Σ⁻¹ΛΦ. A factor with no administered
indicator cannot be recovered.

```{r determinacy}
determinacy <- function(Lam, Phi, keep){
  L <- Lam[keep, , drop = FALSE]
  if (any(colSums(abs(L) > 0) < 1)) return(setNames(rep(NA, ncol(Lam)), colnames(Lam)))
  S <- L %*% Phi %*% t(L); diag(S) <- 1
  M <- Phi %*% t(L) %*% solve(S) %*% L %*% Phi
  setNames(sqrt(pmin(diag(M) / diag(Phi), 1)), colnames(Lam))
}
```

To confirm this analytic quantity really is the recovered-score correlation, we
simulate: draw 50,000 children from the calibrated 4-factor model, generate
task scores with the model's loadings + uniquenesses, estimate factor scores
from a subset using the regression weights, and correlate with the true latent
scores.

```{r mc-check, cache=TRUE}
set.seed(1); Np <- 50000
Lam <- g4$Lam; Phi <- g4$Phi; uniq <- 1 - rowSums((Lam %*% Phi) * Lam)
eta <- MASS::mvrnorm(Np, rep(0, ncol(Phi)), Phi)
Y <- eta %*% t(Lam) + MASS::mvrnorm(Np, rep(0, nrow(Lam)), diag(uniq)); colnames(Y) <- rownames(Lam)
mc_recover <- function(keep){
  L <- Lam[keep, , drop = FALSE]; S <- L %*% Phi %*% t(L); diag(S) <- 1
  diag(cor(Y[, keep, drop = FALSE] %*% t(Phi %*% t(L) %*% solve(S)), eta))
}
bind_rows(
  tibble(battery = "full",    method = c("analytic","simulated"),
         bind_rows(as_tibble_row(determinacy(g4$Lam,g4$Phi,tasks12$short)),
                   as_tibble_row(mc_recover(tasks12$short)))),
  tibble(battery = "minimal", method = c("analytic","simulated"),
         bind_rows(as_tibble_row(determinacy(g4$Lam,g4$Phi,c("hf","math","mg","mrot","pa","sre","trog","vocab"))),
                   as_tibble_row(mc_recover(c("hf","math","mg","mrot","pa","sre","trog","vocab")))))
) |> mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(caption = "Analytic determinacy = simulated recovered-score correlation (validation)")
```

The simulation reproduces the analytic determinacy to three decimals, so we use
the (exact, instant) analytic version for the frontier below.

# 3. The length / recovery frontier

We enumerate **every** subset of the 12 tasks, compute its total minutes and its
recovery of each factor, and keep the **worst-recovered factor** as the binding
constraint. Deployable 4-factor batteries need ≥ 2 indicators per factor.

```{r frontier, cache=TRUE}
idx <- 1:12
subs <- do.call(c, lapply(2:12, \(k) combn(idx, k, simplify = FALSE)))
res <- map_dfr(subs, function(s){
  keep <- tasks12$short[s]
  d4 <- determinacy(g4$Lam, g4$Phi, keep); d2 <- determinacy(g2$Lam, g2$Phi, keep)
  ind4 <- min(table(factor(tasks12$f4[s], levels = c("reasoning","ef","language","reading"))))
  ind2 <- min(table(factor(tasks12$f2[s], levels = c("fluid","verbal"))))
  tibble(ntask = length(s), minutes = sum(mins[keep]),
         det4_min = min(d4), det2_min = min(d2), ind4 = ind4, ind2 = ind2,
         tasks = paste(sort(keep), collapse = ","))
})
pareto <- function(df, ycol){
  df <- df |> arrange(minutes, desc(.data[[ycol]])); best <- -Inf; keep <- logical(nrow(df))
  for (i in seq_len(nrow(df))) if (df[[ycol]][i] > best) { keep[i] <- TRUE; best <- df[[ycol]][i] }
  df[keep, ]
}
front4 <- pareto(res |> filter(ind4 >= 2, !is.na(det4_min)), "det4_min")
front2 <- pareto(res |> filter(ind2 >= 2, !is.na(det2_min)), "det2_min")
```

```{r frontier-plot, fig.height=5.5}
bind_rows(
  res |> filter(ind4 >= 2) |> transmute(minutes, recovery = det4_min, target = "4-factor"),
  res |> filter(ind2 >= 2) |> transmute(minutes, recovery = det2_min, target = "2-factor")
) |>
  ggplot(aes(minutes, recovery, color = target)) +
  geom_point(alpha = 0.10, size = 0.7) +
  geom_line(data = bind_rows(front4 |> transmute(minutes, recovery = det4_min, target = "4-factor"),
                             front2 |> transmute(minutes, recovery = det2_min, target = "2-factor")),
            linewidth = 1) +
  scale_color_manual(values = c("4-factor" = "#268bd2", "2-factor" = "#cb4b16"), name = "Recover…") +
  scale_x_continuous(breaks = seq(0, 60, 10)) +
  labs(x = "Battery length (min, deployed versions)",
       y = "Recovery of worst factor (determinacy)",
       title = "Efficient frontier: factor recovery vs battery length",
       subtitle = "Each point a candidate battery; line = best achievable recovery at each length") +
  coord_cartesian(ylim = c(0.7, 0.98))
```

The frontier is steep then flat: recovery climbs fast over the first ~30 minutes
and then **barely improves** — going from 30 to 55 minutes buys < 0.04 in
determinacy. The 2-factor (broad) target sits well above the 4-factor target, so
if only the fluid/verbal split is needed the battery can be far shorter.

# 4. Three battery versions

```{r versions}
full    <- tasks12$short
minor   <- setdiff(tasks12$short, c("tom","sds"))                       # drop longest task + provisional-scoring task
minimal <- c("hf","math","mg","mrot","pa","sre","trog","vocab")          # 8-task floor: 2 indicators / 4f factor
screen  <- front2 |> filter(det2_min >= 0.90) |> arrange(minutes) |> slice(1) |> pull(tasks) |> str_split(",") |> unlist()

mkrow <- function(keep, nm){
  d4 <- determinacy(g4$Lam, g4$Phi, keep); d2 <- determinacy(g2$Lam, g2$Phi, keep)
  tibble(version = nm, n = length(keep), minutes = sum(mins[keep]),
         `% time` = 100 * sum(mins[keep]) / sum(mins[full]),
         reasoning = d4["reasoning"], ef = d4["ef"], language = d4["language"], reading = d4["reading"],
         fluid = d2["fluid"], verbal = d2["verbal"],
         tasks = paste(sort(keep), collapse = ", "))
}
versions <- bind_rows(mkrow(full,"Full"), mkrow(minor,"Minor reduction"),
                      mkrow(minimal,"Minimal (4-factor)"), mkrow(screen,"Broad screen (2-factor)"))
versions |> select(-tasks) |> mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(caption = "Battery versions: minutes and factor-score recovery (determinacy)")
```

```{r versions-tasks}
versions |> select(version, n, minutes, tasks) |> mutate(minutes = round(minutes, 1)) |>
  knitr::kable(caption = "Tasks in each version")
```

- **Full (12 tasks, ~55 min).** Recovers all four factors at 0.94–0.96.
- **Minor reduction (10 tasks, ~44 min, 74% of the time).** Drops **ToM** (the
  single longest task, ≈10 min) and **Same & Different** (longest-running EF
  task with the weakest, still-provisional scoring). Recovery essentially
  unchanged (0.93–0.96): ToM's factor (language) still has TROG + vocab, and EF
  still has H&F + memory.
- **Minimal (8 tasks, ~31 min, 52% of the time).** The smallest battery that
  keeps two indicators per 4-factor construct. Recovery 0.91–0.95 — reading
  takes the largest hit (0.94→0.91) because dropping ROAR-Word leaves the two
  efficiency/phoneme measures. **Halves administration time for ≈0.03 loss.**
- **Broad screen (4 tasks, ~14 min, 26% of the time).** If only the
  *fluid vs verbal* distinction is needed, four tasks recover both broad factors
  at ≈0.90 — a viable rapid screener. (2026-08 note: with the updated
  durations/reliabilities the frontier now picks **math + sds + sre + vocab**;
  the June render picked hf instead of sds. Treat the sds slot with caution —
  its scoring is still provisional — hf's screener was within a hair of the
  frontier and is the safer deployment choice.)

## What each task buys (leave-one-out)

Dropping a task from the full battery, ranked by recovery lost per minute saved
— the cheapest-to-drop tasks are at the top.

```{r loo}
full_d4 <- min(determinacy(g4$Lam, g4$Phi, full))
map_dfr(tasks12$short, function(t){
  keep <- setdiff(tasks12$short, t)
  tibble(task = t, minutes_saved = mins[t], recovery_loss = full_d4 - min(determinacy(g4$Lam,g4$Phi,keep)))
}) |> mutate(loss_per_min = recovery_loss / minutes_saved) |> arrange(loss_per_min) |>
  mutate(across(where(is.numeric), \(x) round(x, 4))) |>
  knitr::kable(caption = "Leave-one-out value: recovery lost per minute saved (top = most droppable)")
```

ToM and matrix-reasoning are the most droppable (long and/or redundant); **vocab,
TROG, ROAR-Word and math are the most valuable per minute** — they anchor the
factors (language, reading) that have the fewest strong indicators.

# Synthesis

```{r save}
write_rds(list(rel = rel, dur = dur, g2 = g2, g4 = g4,
               frontier4 = front4, frontier2 = front2, versions = versions),
          here::here("data/battery_design_05.rds"))
```

- The heavy intercorrelation among constructs (03–04) is exactly what makes the
  battery **compressible**: factor scores can be recovered from far fewer tasks
  than the full battery, because the remaining tasks borrow strength through the
  factor correlations.
- **You can cut administration time roughly in half (59 → ~31 min) and still
  recover every construct at ≈ 0.91–0.95.** Dropping just ToM and Same &
  Different (→ ~44 min) costs almost nothing.
- The **binding constraint is reading**, not the fluid core: reasoning, EF and
  the broad fluid factor stay > 0.93 even in short batteries, because
  reasoning ≈ EF and they pool. Reading is the factor to protect when trimming —
  keep ROAR-Word if reading scores matter.
- For triage/screening where only the broad fluid/verbal split is needed, a
  **~14-minute four-task screener** recovers both factors at ≈ 0.90.

**Caveats.** Durations are median wall-clock (include instructions/pauses) for
the deployed adaptive versions; composition-only — we did **not** model
shortening tasks *within* themselves (CAT stopping rules), which could shave
more time. The calibration is pooled pilot-site data with thin Bogotá coverage;
Same & Different's reliability/scoring is provisional; sre enters via its loading
(no model-based reliability). Recovery is of the *factor scores*, not of every
task's specific variance — a minimal battery measures the constructs, not the
individual tasks.
