---
title: "06 — Within-child variability: do consistency measures cohere, and change with age?"
levante:
  summary: "Do per-child variability indices (RT-IIV, IRT person-misfit, 2-wave growth deviation) cohere across tasks and indices, and do they change with age?"
  status: draft
  task_ids: [hearts-and-flowers, memory-game, same-different-selection, matrix-reasoning, mental-rotation, egma-math, vocab, trog, theory-of-mind]
  datasets: [pilot_mpieva_de_main, pilot_uniandes_co_bogota, pilot_uniandes_co_rural, pilot_western_ca_main]
  data:
    - {source: "load_levante_scores", version: "v1_2", table: "scores"}
    - {source: "load_levante_trials", version: "v1_2", table: "trials"}
    - {source: "load_item_parameters", version: "v1_14", table: "item_parameters"}
  depends_on: [00_load_data.qmd]
  findings:
    - "Robust null on a general within-child variability trait: the indices show near-zero cross-task generality and are mutually uncorrelated."
    - "Person-misfit scored against the model's actual response function (fixed guessing floor) has ≈ 0 reliability; the earlier 'reliable, age-declining' misfit was an artifact of a plain-logistic IRF."
    - "Two waves cannot identify individual growth differences (slope reliability ≈ 0)."
    - "RT-IIV is reliable per task but only ~4% stable-person variance across tasks."
  last_run: 2026-07-13
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 9
    fig-height: 5.5
execute:
  warning: false
  message: false
---

{{< include _disclaimer.qmd >}}

# The question

A central LEVANTE aim is to extract information about **variability within
children**, not just their average ability. A child's *consistency* —
moment-to-moment, response-to-response, wave-to-wave — may itself be a stable,
meaningful individual difference (and is theorized to index attentional /
executive control and developmental stability).

This notebook is an exploratory first pass. We extract three naive per-child,
per-task variability indices, then ask:

1. **Do the variability measures hang together within a child?** (Is a child who
   is variable on one task / one index variable on others — a general "trait"?)
2. **Do they change with age?**

The three indices:

| Index | What it captures | Naive estimator |
|---|---|---|
| **RT-IIV** | moment-to-moment inconsistency | SD of log reaction time across (correct) trials |
| **Person-misfit** | "surprising" responses given ability | IRT **infit** (info-weighted squared standardized residuals) |
| **Growth deviation** | atypical developmental change | \|individual 2-wave slope − task mean slope\| |

Following the brief, we start with **naive** estimators and treat their
**reliability** as a first-class result — if a naive index is too noisy, that is
the signal to move to model-based estimation.

```{r setup}
source(here::here("common.R"))
sc <- read_rds(here("data/scores_all_sites.rds"))
sites4 <- c("pilot_mpieva_de_main","pilot_uniandes_co_bogota",
            "pilot_uniandes_co_rural","pilot_western_ca_main")
core <- tibble(
  task_id = c("egma-math","matrix-reasoning","mental-rotation","trog","vocab",
              "same-different-selection","hearts-and-flowers","theory-of-mind","memory-game"),
  short   = c("math","matrix","mrot","trog","vocab","sds","hf","tom","mg"))
adap    <- c("egma-math","matrix-reasoning","mental-rotation","same-different-selection","trog","vocab")
bylang  <- c("vocab","theory-of-mind")               # by-language calibration → join params within language
ip <- load_item_parameters()

# one analysis run per child×task (first obs, newest/adaptive version), with final ability θ̂
runs <- sc |> filter(dataset %in% sites4, task_id %in% core$task_id, !is.na(score)) |>
  filter(!(task_id %in% adap) | adaptive) |>
  group_by(user_id, task_id) |> arrange(age) |> slice(1) |> ungroup() |>
  select(run_id, user_id, task_id, short, theta = score, age, language)  # sc already carries `short`
```

# 1. Extracting the indices (and their reliability)

For each task we load trial data once and compute, per run: RT-IIV (SD of
log-RT on correct trials), and **infit** person-misfit using the published item
parameters (Rasch/2PL; joined within language for the by-language tasks) and the
child's final ability θ̂. Split-half reliabilities (odd/even trials,
Spearman-Brown) tell us whether each naive index is stable enough to interpret.

```{r extract, cache=TRUE}
sb <- function(a, b){ r <- cor(a, b, use = "complete.obs"); 2*r/(1+r) }
extract_task <- function(tid){
  rn <- runs |> filter(task_id == tid)
  tr <- load_levante_trials(task_ids = tid) |>
    filter(run_id %in% rn$run_id, !is.na(correct)) |>
    left_join(rn |> select(run_id, theta, language), by = "run_id")
  # join item parameters (within language for by-language tasks)
  ipx <- ip |> filter(task_id == tid)
  if (tid %in% bylang) {
    tr <- tr |> inner_join(ipx |> transmute(language = subset, item_uid, b = difficulty, a = discrimination),
                           by = c("item_uid","language"))
  } else {
    tr <- tr |> inner_join(ipx |> transmute(item_uid, b = difficulty, a = discrimination) |>
                             distinct(item_uid, .keep_all = TRUE), by = "item_uid")
  }
  # IRF must match the scoring model: the fitted mirt models carry a FIXED
  # guessing asymptote at each item's chance level (g = chance, est=FALSE), so
  # the person-fit response function is P = c + (1-c)·logistic, not plain logistic.
  tr <- tr |> mutate(c = ifelse(is.na(chance), 0, chance),
                     irf = 1/(1+exp(-a*(theta-b))),
                     P = pmin(pmax(c + (1-c)*irf, .001), .999),
                     w = P*(1-P), r2 = (correct - P)^2)
  # run-level indices
  iiv <- tr |> filter(!is.na(rt_numeric), rt_numeric > 200, rt_numeric < 30000, correct == 1) |>
    group_by(run_id) |> filter(n() >= 10) |>
    summarise(iiv = sd(log(rt_numeric)), meanlrt = mean(log(rt_numeric)), .groups = "drop")
  infit <- tr |> group_by(run_id) |> filter(n() >= 10) |>
    summarise(infit = sum(r2)/sum(w), .groups = "drop")
  idx <- rn |> select(run_id, user_id, short, theta, age) |>
    left_join(iiv, by = "run_id") |> left_join(infit, by = "run_id")
  # split-half reliability
  sh_iiv <- tr |> filter(!is.na(rt_numeric), rt_numeric > 200, rt_numeric < 30000, correct == 1) |>
    group_by(run_id) |> filter(n() >= 12) |> mutate(h = row_number() %% 2) |>
    group_by(run_id, h) |> summarise(v = sd(log(rt_numeric)), .groups = "drop") |>
    pivot_wider(names_from = h, values_from = v, names_prefix = "h")
  sh_inf <- tr |> group_by(run_id) |> filter(n() >= 12) |> mutate(h = row_number() %% 2) |>
    group_by(run_id, h) |> summarise(v = sum(r2)/sum(w), .groups = "drop") |>
    pivot_wider(names_from = h, values_from = v, names_prefix = "h")
  list(idx = idx,
       rel = tibble(short = core$short[core$task_id == tid],
                    n = nrow(idx),
                    iiv_rel   = if (nrow(sh_iiv) > 3) sb(sh_iiv$h0, sh_iiv$h1) else NA,
                    infit_rel = if (nrow(sh_inf) > 3) sb(sh_inf$h0, sh_inf$h1) else NA))
}
ex <- map(core$task_id, extract_task)
index_runs <- map_dfr(ex, "idx")
rel_tab    <- map_dfr(ex, "rel")
```

```{r rel-table}
rel_tab |> arrange(desc(iiv_rel)) |> mutate(across(where(is.numeric), \(x) round(x, 2))) |>
  knitr::kable(caption = "Split-half reliability of the naive indices (NA = no RT, i.e. memory-game)")
```

**Reliability is the headline.** RT-IIV is reliable enough to use on most tasks
(≈ 0.45–0.85, highest where trials are many — H&F, vocab). Person-misfit (infit),
computed with the **scoring-matched response function** (a fixed guessing
asymptote at each item's chance level — see §4), is **≈ 0 on every task** (the
best, H&F, reaches only 0.36; the rest are indistinguishable from zero).
Correctly specified, person-fit simply has very little reliable between-child
variance to measure. RT-IIV clears a 0.4 bar; person-misfit does not.

## Two caveats baked in

```{r misfit-caveats}
# infit mean by task (well-calibrated ≈ 1) and its confound with ability
index_runs |> group_by(short) |>
  summarise(infit_mean = mean(infit, na.rm = TRUE),
            `infit vs |θ|` = cor(infit, abs(theta), use = "complete.obs"), .groups = "drop") |>
  arrange(desc(infit_mean)) |> mutate(across(where(is.numeric), \(x) round(x, 2))) |>
  knitr::kable(caption = "Person-misfit: calibration (mean infit ≈ 1 = good) and the ability confound")
```

With the scoring-matched IRF, person-misfit is **well-calibrated on every task**
(mean infit ≈ 0.93, matrix included — its earlier blow-up was a mis-specified
response function, not the data), and the **ability-extremity confound largely
disappears** (infit vs |θ| now |r| ≤ ~0.2, versus 0.1–0.7 under the plain-logistic
IRF that ignored guessing). The problem is neither calibration nor confound — it
is that, correctly specified, per-task misfit has **≈ 0 reliability** (table
above). We therefore report it below for completeness, but it cannot support
individual-difference claims.

# 2. Do the variability measures hang together?

We keep indices on tasks where they are at least modestly reliable
(reliability ≥ 0.4), z-score each within task, and ask two questions.

```{r keep-reliable}
iiv_ok   <- rel_tab |> filter(iiv_rel   >= 0.4) |> pull(short)
# person-misfit is now scoring-matched (guessing-aware) on every task, incl. matrix;
# its reliabilities (above) are ≈0, so we include all tasks here only to show it
# carries no cross-task signal either.
infit_ok <- rel_tab |> filter(!is.na(infit_rel)) |> pull(short)
zt <- function(x) as.numeric(scale(x))
zi <- index_runs |> group_by(short) |> mutate(iiv_z = zt(iiv), infit_z = zt(infit),
                                              theta_z = zt(theta)) |> ungroup()
```

## (a) Generality across tasks (within an index)

If RT-IIV is a trait, a child variable on one task should be variable on
others. We correlate each index across tasks and summarise the average
off-diagonal correlation (a child-level "g of variability").

```{r generality}
cross_task_r <- function(df, var, tasks){
  w <- df |> filter(short %in% tasks) |> select(user_id, short, val = all_of(var)) |>
    pivot_wider(names_from = short, values_from = val)
  cm <- cor(w |> select(-user_id), use = "pairwise")
  mean(cm[upper.tri(cm)], na.rm = TRUE)
}
tibble(index = c("RT-IIV","Person-misfit (infit)"),
       tasks = c(paste(iiv_ok, collapse=","), paste(infit_ok, collapse=",")),
       mean_cross_task_r = c(cross_task_r(zi, "iiv_z", iiv_ok),
                             cross_task_r(zi, "infit_z", infit_ok))) |>
  mutate(mean_cross_task_r = round(mean_cross_task_r, 3)) |>
  knitr::kable(caption = "Average cross-task correlation of each variability index (within child)")
```

Cross-task generality is **near zero** for both indices (RT-IIV ≈ 0.05;
person-misfit ≈ 0.11, though on a ≈ 0-reliability measure so not interpretable
anyway). A child who is RT-variable on math is essentially *no more likely* to be
RT-variable on vocabulary. Even allowing for attenuation, this is far too low to
support a general "variability trait" — within-task variability is overwhelmingly
**task-specific**.

## (b) Do the different indices correlate, beyond ability?

We aggregate each index to the child level (mean of within-task z-scores), add a
crude ability composite (mean within-task z-θ̂) and the naive growth-deviation
index (built next), and correlate — first raw, then partialling ability.

```{r growth, cache=TRUE}
gr <- sc |> filter(dataset %in% sites4, task_id %in% core$task_id, !is.na(score)) |>
  group_by(user_id, short) |> arrange(age) |> mutate(w = row_number()) |>
  filter(w <= 2, max(w) >= 2) |> ungroup()
gw <- gr |> select(user_id, short, w, score, score_se, age) |>
  pivot_wider(names_from = w, values_from = c(score, score_se, age)) |>
  mutate(dage = age_2 - age_1, slope = (score_2 - score_1)/dage) |>
  filter(is.finite(slope), dage > 0.1)
# difference-score reliability per task: 1 - error var / observed var
grel <- gw |> group_by(short) |>
  summarise(growth_rel = 1 - mean((score_se_1^2 + score_se_2^2)/dage^2) / var(slope),
            n = n(), .groups = "drop")
gw <- gw |> group_by(short) |>
  mutate(growth_dev = abs(slope - mean(slope)), growth_dev_z = as.numeric(scale(growth_dev))) |> ungroup()
grel |> arrange(desc(growth_rel)) |> mutate(growth_rel = round(growth_rel, 2)) |>
  knitr::kable(caption = "Reliability of the naive 2-wave growth-deviation index")
```

The naive 2-wave growth-deviation index is **moderately reliable for the
fast-growing, well-measured tasks** (H&F 0.77, math 0.59, vocab 0.48, mental
rotation 0.40) but **unreliable for the rest** (memory & Same-&-Different low;
TROG & ToM negative) — wherever true between-child slope variance is small
relative to the θ̂ measurement error carried by a two-point difference. So it is
usable on a few tasks but fragile overall; more waves and error-aware shrinkage
would stabilise it. (But §4.2 shows even this is optimistic: it credits the EAP
posterior SE as the *only* source of error — a proper random-slope model that
lets occasion variance compete finds individual growth differences are not yet
identifiable from two waves.)

```{r child-level}
child <- zi |> group_by(user_id) |>
  summarise(iiv_g   = mean(iiv_z[short %in% iiv_ok], na.rm = TRUE),
            misfit_g = mean(infit_z[short %in% infit_ok], na.rm = TRUE),
            ability  = mean(theta_z, na.rm = TRUE), .groups = "drop") |>
  left_join(gw |> group_by(user_id) |> summarise(growth_g = mean(growth_dev_z, na.rm = TRUE), .groups = "drop"),
            by = "user_id")
M <- child |> select(iiv_g, misfit_g, growth_g, ability)
cmat <- cor(M, use = "pairwise")
# ability-partialled correlations among the 3 variability indices
pr <- function(a, b, c){ d <- na.omit(child[, c(a, b, c)]); resid_a <- resid(lm(d[[a]] ~ d[[c]]))
  resid_b <- resid(lm(d[[b]] ~ d[[c]])); cor(resid_a, resid_b) }
bind_rows(
  tibble(pair = "IIV — misfit",  raw = cmat["iiv_g","misfit_g"],   partial_ability = pr("iiv_g","misfit_g","ability")),
  tibble(pair = "IIV — growth",  raw = cmat["iiv_g","growth_g"],   partial_ability = pr("iiv_g","growth_g","ability")),
  tibble(pair = "misfit — growth", raw = cmat["misfit_g","growth_g"], partial_ability = pr("misfit_g","growth_g","ability"))
) |> mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(caption = "Do the variability indices correlate at the child level? (raw and ability-partialled)")
```

```{r ability-link}
tibble(index = c("RT-IIV","Person-misfit","Growth deviation"),
       r_with_ability = c(cmat["iiv_g","ability"], cmat["misfit_g","ability"], cmat["growth_g","ability"])) |>
  mutate(r_with_ability = round(r_with_ability, 3)) |>
  knitr::kable(caption = "Is 'variability' just (low) ability? Correlation of each index with the ability composite")
```

Two things stand out. First, the three indices are **mutually uncorrelated**
(all |r| < 0.1, raw or ability-partialled) — they do not converge on a common
"variability" dimension. Second, RT-IIV and growth-deviation are essentially
**unrelated to ability** (r ≈ ±0.05, so they are *not* just low-ability in
disguise). Person-misfit still shows a moderate negative tie to ability
(r ≈ −0.52) even with the scoring-matched IRF: the *extremity* confound is gone
(§1), but the small amount of stable misfit variance that survives aggregation
across tasks is largely **lower-ability children responding less predictably**.
Since per-task misfit is ≈ 0-reliable, the safe reading is that there is no
usable misfit signal here beyond ability.

For each (reliable) index we regress on age — pooled across tasks with a task
random effect — and plot the per-task trends.

```{r age-trends, fig.height=4.8}
library(lme4)
age_fit <- function(df, var, tasks){
  d <- df |> filter(short %in% tasks, !is.na(.data[[var]]))
  m <- lmer(reformulate(c("age","(1|short)"), response = var), data = d)
  fixef(m)["age"]
}
slopes <- tibble(
  index = c("RT-IIV","Person-misfit"),
  per_year = c(age_fit(zi |> rename(v = iiv),   "v", iiv_ok),
               age_fit(zi |> rename(v = infit), "v", infit_ok)))
slopes |> mutate(per_year = round(per_year, 3)) |>
  knitr::kable(caption = "Age slope of each index (pooled, task random intercept; z-units & raw infit/IIV)")

bind_rows(
  zi |> filter(short %in% iiv_ok)   |> transmute(short, age, value = iiv,   index = "RT-IIV (SD log RT)"),
  zi |> filter(short %in% infit_ok) |> transmute(short, age, value = infit, index = "Person-misfit (infit)")
) |>
  ggplot(aes(age, value, color = short)) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 0.8) +
  facet_wrap(~index, scales = "free_y") +
  scale_color_viridis_d(name = NULL) +
  labs(x = "Age (years)", y = "Index value",
       title = "Neither variability index changes much with age",
       subtitle = "Both RT-IIV and (scoring-matched) person-misfit are essentially flat across childhood")
```

With the scoring-matched IRF, **neither index changes appreciably with age**:
RT-IIV is flat (pooled slope ≈ 0) and person-misfit nearly so (≈ −0.02/yr). The
steep misfit-with-age decline seen under the mis-specified IRF was, again, an
artifact of ignoring guessing — older/higher-ability children produced fewer
"surprising" lucky guesses *under the wrong model*; once the model accounts for
guessing, that age trend largely evaporates.

# Synthesis — naive (first-order) indices

```{r save}
write_rds(list(rel = rel_tab, growth_rel = grel, child = child, cor = cmat, age_slopes = slopes),
          here::here("data/within_child_variability_06.rds"))
```

A first, honest pass — and the headline is a **null on coherence**:

- **The variability indices do not hang together.** Within an index, cross-task
  generality is ≈ 0 (a child variable on math is no more variable on vocab);
  across indices, the three are mutually uncorrelated (|r| < 0.1). With naive
  estimators there is **no general "variability trait"** — within-task
  variability is overwhelmingly task-specific.
- **RT-IIV is the cleanest index but not (yet) a trait.** It is reliably measured
  on most tasks (≈ 0.45–0.85) and is **not** an ability proxy (r ≈ −0.05) — but
  it is task-specific and **flat with age**. So it is a sound per-task measure
  whose cross-situational meaning is unclear here. (Followed up in depth in
  `07_rt_variability.qmd`: purifying the RT signal and accounting for trial count
  surfaces a narrow speeded-task RT-IIV factor.)
- **Person-misfit, correctly specified, carries almost no usable signal.** Once
  the response function matches the scoring model (fixed guessing at chance),
  infit is well-calibrated on every task (matrix included) and the
  ability-extremity confound disappears — but per-task reliability is ≈ 0, and
  the small surviving aggregate signal is still mostly inverse ability (r ≈ −0.52).
  It is not a usable individual-difference index here. (The earlier impression of
  reliable, age-declining misfit was an artifact of a mis-specified IRF — see §4.1.)
- **Growth-deviation is salvageable on some tasks.** Two-wave reliability is
  moderate where growth is fast and well-measured (H&F 0.77, math 0.59, vocab
  0.48) but poor elsewhere — usable in places, fragile overall.

**How to read this:** the null on coherence is partly substantive (these may
genuinely be distinct, task-bound phenomena) and partly methodological — naive
estimators are noisy and, for misfit, badly confounded, so any true common
signal would be heavily attenuated. That is exactly the contingency the brief
anticipated, and it motivates a **model-based second pass**: (1) replace naive
infit with a guessing-aware person-fit statistic from the scoring model;
(2) estimate individual growth as shrunken, error-propagating random slopes
(and revisit as more waves arrive); (3) model RT-IIV with an explicit
person×task variance-components model rather than per-run SDs; then re-ask
whether — disattenuated — the indices cohere into a within-child variability
construct.

# 4. Model-based upgrades

We keep the naive analyses above (we will always want those first-order views)
and add the model-based versions here.

**A note on guessing (corrected).** The fitted scoring models **do** incorporate
guessing: every item carries a **fixed lower asymptote at its chance level**
(g = chance, *not* estimated) inside the `mirt` model — confirmed by reading the
model records directly (matrix g = 0.25, mental-rotation 0.5, math per-item
0–0.5). `score_irt` passes the full `mod2values()` table — g rows included —
into `mirt(pars = …)`, so the released θ̂ already account for guessing. This was
not obvious because the curated `item_parameters` table exports only difficulty
and discrimination and **silently omits g** — a documentation gap worth flagging
to the DCC, since anyone reconstructing the IRF from that table (as we first did)
will get it wrong. The person-fit in §1 already uses the matching response
function, P = c + (1 − c)·IRF; the check below shows why that is essential.

## 4.1 Why the person-fit response function must match the scoring model

```{r upg-misfit, cache=TRUE}
upg_mf <- map_dfr(core$task_id, function(tid){
  rn <- runs |> filter(task_id == tid)
  tr <- load_levante_trials(task_ids = tid) |> filter(run_id %in% rn$run_id, !is.na(correct)) |>
    left_join(rn |> select(run_id, theta, language), by = "run_id")
  ipx <- ip |> filter(task_id == tid)
  if (tid %in% bylang) tr <- tr |> inner_join(ipx |> transmute(language = subset, item_uid, b = difficulty, a = discrimination), by = c("item_uid","language"))
  else tr <- tr |> inner_join(ipx |> transmute(item_uid, b = difficulty, a = discrimination) |> distinct(item_uid, .keep_all = TRUE), by = "item_uid")
  tr <- tr |> mutate(c = ifelse(is.na(chance), 0, chance), irf = 1/(1+exp(-a*(theta-b))),
                     Pn = pmin(pmax(irf, .001), .999), Pg = pmin(pmax(c + (1-c)*irf, .001), .999))
  agg <- tr |> group_by(run_id) |> filter(n() >= 10) |>
    summarise(naive = sum((correct-Pn)^2)/sum(Pn*(1-Pn)),
              guess = sum((correct-Pg)^2)/sum(Pg*(1-Pg)), .groups = "drop") |>
    left_join(rn |> select(run_id, theta), by = "run_id")
  sh <- tr |> group_by(run_id) |> filter(n() >= 12) |> mutate(h = row_number() %% 2) |>
    group_by(run_id, h) |> summarise(v = sum((correct-Pg)^2)/sum(Pg*(1-Pg)), .groups = "drop") |>
    pivot_wider(names_from = h, values_from = v, names_prefix = "h")
  r <- cor(sh$h0, sh$h1, use = "complete.obs")
  tibble(short = core$short[core$task_id == tid],
         `naive infit` = mean(agg$naive), `guess-aware infit` = mean(agg$guess),
         `naive vs θ` = cor(agg$naive, agg$theta, use = "complete.obs"),
         `guess-aware vs θ` = cor(agg$guess, agg$theta, use = "complete.obs"),
         `guess-aware reliability` = 2*r/(1+r))
})
upg_mf |> mutate(across(where(is.numeric), \(x) round(x, 2))) |>
  knitr::kable(caption = "Plain-logistic (wrong) vs scoring-matched person-fit: calibration, ability link, reliability")
```

The comparison is a cautionary tale. The **scoring-matched** IRF (the one used
in §1, with the fixed guessing asymptote) is well-calibrated — every infit mean
near 1, matrix included. If instead you forget the fixed guessing and use a
plain logistic — the "naive" column — infit **inflates badly** (matrix 4.2) and
acquires a strong, spurious ability correlation, *and* a misleadingly high
split-half "reliability" on the inflated tasks: the systematic guessing misfit
is shared across both split-halves (lucky guesses cluster at low ability), so it
looks like reliable between-child variance when it is really model
mis-specification. Score "surprise" with the **same** response function the
scores were fit under, and that apparent signal collapses — the scoring-matched
misfit has ≈ 0 reliability. The lesson: person-fit is only meaningful against the
exact IRF used to estimate θ̂. Done correctly, there is little real between-child
variance in person-fit to measure here.

## 4.2 Model-based growth (shrunken random slopes)

```{r upg-growth, cache=TRUE}
grl <- sc |> filter(dataset %in% sites4, task_id %in% core$task_id, !is.na(score)) |>
  group_by(user_id, short) |> filter(n() >= 2) |> ungroup() |> mutate(age_c = age - 8)
upg_growth <- map_dfr(sort(unique(grl$short)), function(s){
  d <- grl |> filter(short == s)
  m <- tryCatch(suppressMessages(suppressWarnings(
    lme4::lmer(score ~ age_c + (age_c | user_id), data = d,
               control = lme4::lmerControl(check.conv.singular = "ignore")))), error = function(e) NULL)
  if (is.null(m)) return(tibble(short = s, slope_sd = NA, resid_sd = NA, slope_reliability = NA))
  vc <- as.data.frame(lme4::VarCorr(m))
  ss <- vc$sdcor[vc$grp == "user_id" & vc$var1 == "age_c" & is.na(vc$var2)]
  rs <- vc$sdcor[vc$grp == "Residual"]
  tibble(short = s, slope_sd = ss, resid_sd = rs, slope_reliability = ss^2/(ss^2 + rs^2))
})
upg_growth |> arrange(desc(slope_reliability)) |> mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(caption = "Random-slope LMM per task: reliable between-child slope variance is negligible")
```

With only two waves the random-slope models are **near-singular** and recover
**negligible reliable between-child slope variance** (slope reliability ≈ 0–0.06).
This *revises* the optimistic naive 2-wave reliabilities above: those credited
the EAP posterior SE as the only error term, but a model that lets
occasion-to-occasion variance compete finds individual growth differences are
**not yet identifiable** from two points. Growth-based variability needs **≥ 3
waves** before it can be estimated per child.

## 4.3 RT-IIV as a variance-components model

Rather than a per-run SD, we model log RT-variability across all
person×task runs and ask what share is a **stable cross-task person property**.

```{r upg-iiv, cache=TRUE}
iiv_run <- map_dfr(setdiff(core$task_id, "memory-game"), function(tid){
  rn <- runs |> filter(task_id == tid)
  load_levante_trials(task_ids = tid) |>
    filter(run_id %in% rn$run_id, !is.na(rt_numeric), rt_numeric > 200, rt_numeric < 30000, correct == 1) |>
    group_by(run_id, user_id, short) |> filter(n() >= 10) |>
    summarise(logvar = log(var(log(rt_numeric))), .groups = "drop")
})
vcm <- suppressMessages(lme4::lmer(logvar ~ 1 + (1 | user_id) + (1 | short), data = iiv_run))
vc <- as.data.frame(lme4::VarCorr(vcm)); tot <- sum(vc$vcov)
tibble(component = c("person (stable trait)","task","residual (person×task + noise)"),
       variance_share = c(vc$vcov[vc$grp=="user_id"], vc$vcov[vc$grp=="short"], vc$vcov[vc$grp=="Residual"])/tot) |>
  mutate(variance_share = round(variance_share, 3)) |>
  knitr::kable(caption = glue::glue("Variance decomposition of RT-variability ({nrow(iiv_run)} runs, {n_distinct(iiv_run$user_id)} children)"))
```

Only **~4% of RT-variability is a stable cross-task person property**; the bulk
is task differences (~58%) and within-person/residual (~38%, which still
includes measurement noise, so 4% is if anything an upper-ish bound on the
trait share). The near-zero cross-task correlation in the naive analysis was
therefore **not merely attenuation** — RT variability genuinely is
task-specific, not a general trait, at these ages.

```{r upg-save}
write_rds(list(misfit = upg_mf, growth = upg_growth,
               iiv_var = as.data.frame(lme4::VarCorr(vcm))),
          here::here("data/within_child_variability_06_upgrades.rds"))
```

## Synthesis — model-based upgrades

The model-based pass sharpens, and in one place corrects, the naive picture —
and the verdict is a **robust null** on a general within-child variability trait:

- **Person-fit** (scored against the model's actual IRF — the fixed guessing
  asymptote the scoring already uses, *not* a new 3PL): well-calibrated on every
  task and free of the extremity confound, but ≈ 0 reliability — there is little
  real between-child variance to measure. The earlier "reliable, age-declining"
  misfit was an artifact of using the wrong (plain-logistic) response function.
- **Model-based growth**: two waves cannot identify individual growth
  differences (slope reliability ≈ 0; models near-singular). The naive
  reliabilities were optimistic. Revisit at ≥ 3 waves.
- **RT-IIV variance components**: a person×task model puts the stable-person
  share at ~4% — overwhelmingly task-specific.

**Implication for LEVANTE.** Measuring within-child variability *as a coherent
individual-difference construct* will need richer designs, not better summaries
of the current single-session scores: more longitudinal waves for growth, and
within-task manipulations purpose-built to elicit and reliably capture
variability (e.g., conditions that provoke lapses, speed–accuracy trade-offs).
RT-IIV remains a clean, reliable **per-task** measure — it is simply not, on
this evidence, a cross-task trait.
