---
title: "07 — RT intra-individual variability: purification and cross-task structure"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 9
    fig-height: 5.5
execute:
  warning: false
  message: false
---

{{< include _disclaimer.qmd >}}

# The question

In `06` the most promising within-child variability index was **RT
intra-individual variability (RT-IIV)** — but the *naive* per-run SD of log-RT
showed almost no cross-task coherence (mean r ≈ 0.05) and only ~4% stable-person
variance. Two things could be hiding a real signal: (a) the raw SD is
**contaminated** by systematic structure (within-task time-course,
autocorrelation, administration differences), and (b) RT-IIV is a second-moment
estimate, so it is **noisy when trial counts are low**.

This notebook does the cleanup the feedback asked for: purify the RT signal, then
re-examine whether **cross-task RT-IIV cohere** — both across all tasks and
restricted to tasks/runs with enough trials.

```{r setup}
source(here::here("common.R"))
library(splines); library(psych)
sc <- read_rds(here("data/scores_all_sites.rds"))
sites4 <- c("pilot_mpieva_de_main","pilot_uniandes_co_bogota",
            "pilot_uniandes_co_rural","pilot_western_ca_main")
rt_tasks <- tibble(
  task_id = c("egma-math","matrix-reasoning","mental-rotation","trog","vocab",
              "same-different-selection","hearts-and-flowers","theory-of-mind","swr","sre","pa"),
  short   = c("math","matrix","mrot","trog","vocab","sds","hf","tom","swr","sre","pa"))
adap <- c("egma-math","matrix-reasoning","mental-rotation","same-different-selection","trog","vocab")

# ---- analysis parameter: which trials feed RT-IIV (carried throughout) ----
USE_TRIALS <- "all"      # "all" or "correct"

runs <- sc |> filter(dataset %in% sites4, task_id %in% rt_tasks$task_id, !is.na(score)) |>
  filter(!(task_id %in% adap) | adaptive) |>
  group_by(user_id, task_id) |> arrange(age) |> slice(1) |> ungroup() |>
  select(run_id, user_id, task_id, age, language)

# ---- purification helpers (the pipeline §1.4 selects) -----------------------
# Detrend log-RT for the within-task time-course AND item difficulty (the IRT b
# of each administered item), since harder items take longer; difficulty is
# joined per item (within language for by-language tasks). Tasks without item
# parameters (sre) fall back to trial-only detrend.
ip <- load_item_parameters(); bylang <- c("vocab","theory-of-mind","swr","pa")
join_difficulty <- function(tr, tid){
  ipx <- ip |> filter(task_id == tid); tr <- tr |> select(-any_of("difficulty"))
  if (nrow(ipx) == 0) return(tr |> mutate(difficulty = NA_real_))
  if (tid %in% bylang) tr |> left_join(ipx |> transmute(language = subset, item_uid, difficulty), by = c("item_uid","language"))
  else tr |> left_join(ipx |> transmute(item_uid, difficulty) |> distinct(item_uid, .keep_all = TRUE), by = "item_uid")
}
detrend_resid <- function(tr){
  has_d <- "difficulty" %in% names(tr) && sum(is.finite(tr$difficulty)) > 0.5 * nrow(tr)
  if (has_d) tr <- tr |> mutate(difficulty = ifelse(is.finite(difficulty), difficulty, median(difficulty, na.rm = TRUE)))
  rhs <- "ns(trial_number, 4)"
  if (has_d) rhs <- paste(rhs, "+ difficulty + difficulty:trial_number")
  if (n_distinct(tr$dataset) > 1) rhs <- paste(rhs, "+ dataset")
  tr$resid <- resid(lm(as.formula(paste("lrt ~", rhs)), data = tr)); tr
}
```

The trial-set choice (**`r USE_TRIALS`** trials) is a parameter, applied
everywhere below; a correct-only sensitivity check appears in §1.4.

# 1. Purifying the RT signal

**What actually needs removing.** IIV here is the SD of log-RT *within a run*.
Because the SD is location-invariant, any nuisance that is **constant within a
run** — the child's mean speed, the run's mean, a **site/administration
intercept** — drops out automatically (and in log-space a multiplicative
site/device slowdown is just such an additive shift). What inflates the
within-run SD is structure that **varies across trials within a run**: the
**trial-order time-course** (warm-up, fatigue), the **item-difficulty sequence**
(harder items run longer — especially on non-speeded tasks), and
**autocorrelation**. So those are what we detrend and filter — the specific
combination is chosen systematically in §1.4. Site/age, which can shift the
*level* of a child's IIV (administration medium, development), are removed at the
**IIV level** instead (§3).

## 1.1 Detrending (trial-order and item difficulty)

Log-RT drifts over the course of a task (warm-up/fatigue) and tracks item
difficulty (harder items take longer). We fit a task-level model
`log RT ~ ns(trial_number, 4) + difficulty + difficulty:trial + site` — using the
fitted IRT difficulty (`b`) of each administered item — and work with the
residuals. (§1.4 shows the difficulty term is what most improves the measure,
especially on non-speeded tasks; it falls back to trial-only where item
parameters are unavailable, i.e. ROAR-Sentence.)

```{r extract, cache=TRUE}
purify_task <- function(tid, use_trials){
  rn <- runs |> filter(task_id == tid)
  tr <- load_levante_trials(task_ids = tid) |>
    filter(run_id %in% rn$run_id, !is.na(rt_numeric), rt_numeric > 200, rt_numeric < 30000,
           !is.na(trial_number)) |>
    left_join(rn |> select(run_id, age, language), by = "run_id")
  if (use_trials == "correct") tr <- tr |> filter(correct == 1)
  tr <- tr |> mutate(lrt = log(rt_numeric)) |> join_difficulty(tid) |> detrend_resid()
  # AR(1) innovations within run (pooled rho per task)
  tr <- tr |> group_by(run_id) |> arrange(trial_number, .by_group = TRUE) |>
    mutate(lag_resid = lag(resid)) |> ungroup()
  rho <- cor(tr$resid, tr$lag_resid, use = "complete.obs")
  tr <- tr |> mutate(innov = resid - rho * lag_resid)
  trend <- tr |> mutate(tn = pmin(trial_number, quantile(trial_number, .95))) |>
    group_by(short = rt_tasks$short[rt_tasks$task_id == tid], tn) |>
    summarise(m = mean(lrt - mean(lrt)), .groups = "drop")
  per_run <- tr |> group_by(run_id, user_id) |>
    summarise(ntrial = n(),
              iiv_raw   = sd(lrt),
              iiv_detr  = sd(resid),
              iiv_innov = sd(innov, na.rm = TRUE), .groups = "drop") |>
    filter(ntrial >= 8) |>
    mutate(short = rt_tasks$short[rt_tasks$task_id == tid])
  list(short = rt_tasks$short[rt_tasks$task_id == tid], rho = rho,
       per_run = per_run, trend = trend, med_n = median(per_run$ntrial))
}
ex <- map(rt_tasks$task_id, purify_task, use_trials = USE_TRIALS)
iiv <- map_dfr(ex, "per_run")
trend <- map_dfr(ex, "trend")
```

```{r trend-plot, fig.height=4.3}
trend |> ggplot(aes(tn, m, color = short)) +
  geom_line(linewidth = 0.8) + scale_color_viridis_d(name = NULL) +
  geom_hline(yintercept = 0, linetype = 2, color = "grey60") +
  labs(x = "Trial number", y = "Mean log-RT (centered)",
       title = "Within-task time-course of log-RT (the trend we detrend out)",
       subtitle = "Most tasks slow slightly then drift; speeded tasks (swr/hf) speed up with warm-up")
```

## 1.2 Autocorrelation

Consecutive RTs are positively autocorrelated, so the raw SD mixes
trial-to-trial noise with slow drift. We estimate a per-task AR(1) coefficient
on the detrended residuals (within run) and also form the **innovations**
(uncorrelated component) as a more aggressive purification.

## 1.3 How much each step changes the IIV

```{r purify-summary}
sb <- function(a, b){ r <- cor(a, b, use = "complete.obs"); 2*r/(1+r) }
rel_one <- function(tid, use_trials){
  rn <- runs |> filter(task_id == tid)
  tr <- load_levante_trials(task_ids = tid) |>
    filter(run_id %in% rn$run_id, !is.na(rt_numeric), rt_numeric > 200, rt_numeric < 30000, !is.na(trial_number)) |>
    left_join(rn |> select(run_id, language), by = "run_id") |> mutate(lrt = log(rt_numeric))
  if (use_trials == "correct") tr <- tr |> filter(correct == 1)
  tr <- tr |> join_difficulty(tid) |> detrend_resid()
  sh <- tr |> group_by(run_id) |> filter(n() >= 12) |> mutate(h = row_number() %% 2) |>
    group_by(run_id, h) |> summarise(v = sd(resid), .groups = "drop") |>
    pivot_wider(names_from = h, values_from = v, names_prefix = "h")
  sb(sh$h0, sh$h1)
}
summ <- map_dfr(seq_along(ex), function(i){
  e <- ex[[i]]
  tibble(short = e$short, median_trials = e$med_n, n_runs = nrow(e$per_run), ar1_rho = e$rho,
         `cor(raw,detrended)` = cor(e$per_run$iiv_raw, e$per_run$iiv_detr),
         `cor(detrended,AR-innov)` = cor(e$per_run$iiv_detr, e$per_run$iiv_innov),
         iiv_reliability = rel_one(rt_tasks$task_id[i], USE_TRIALS))
})
summ |> arrange(desc(iiv_reliability)) |> mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(caption = "Purification diagnostics per task (AR(1) ρ, step effects, split-half reliability)")
```

Autocorrelation is substantial (ρ ≈ 0.26–0.62). The trial+difficulty detrending
now moves the IIV substantially on difficulty-sensitive tasks (mental rotation
`cor(raw, detrended)` ≈ 0.76, matrix/math ≈ 0.85) and little on the speeded tasks
(≈ 0.96–1.0); the AR-innovation IIV stays highly correlated with the detrended
one (≥ 0.87). We carry the **detrended** IIV forward as primary (autocorrelation
retained, per §1.4) and use the AR-innovation only as a robustness variant.

## 1.4 Choosing the pipeline: a systematic 2⁴ comparison

The purification involves four binary choices: **detrend by trial number?**,
**all vs correct trials?**, **detrend by item difficulty?** (the IRT `b`), and
**remove autocorrelation?** (AR(1) innovations). We run all **2⁴ = 16**
combinations and score each on **split-half reliability** (first vs second half)
and **longitudinal test–retest** (wave-1 vs wave-2, Leipzig + Bogotá), then pick
the setting that jointly maximizes the two. (ROAR-Sentence lacks item parameters,
so it is omitted from this comparison.)

```{r factorial, cache=TRUE}
long_sites <- c("pilot_mpieva_de_main","pilot_uniandes_co_bogota","pilot_uniandes_co_rural")
fac_tasks <- rt_tasks |> filter(short != "sre") |> mutate(speeded = short %in% c("hf","swr","pa"))
scor <- function(x, y){ ok <- is.finite(x) & is.finite(y); if (sum(ok) < 3 || sd(x[ok]) == 0 || sd(y[ok]) == 0) return(NA_real_); cor(x[ok], y[ok]) }
grid <- expand_grid(trial = c(FALSE,TRUE), trials = c("all","correct"), diff = c(FALSE,TRUE), ar = c(FALSE,TRUE))
load_long <- function(tid){
  wv <- sc |> filter(dataset %in% long_sites, task_id == tid, !is.na(score)) |>
    group_by(user_id) |> arrange(age) |> mutate(wave = row_number(), nwave = max(wave)) |>
    filter(nwave >= 2, wave <= 2) |> ungroup() |> select(run_id, wave, language)
  load_levante_trials(task_ids = tid) |>
    filter(run_id %in% wv$run_id, !is.na(rt_numeric), rt_numeric > 200, rt_numeric < 30000, !is.na(trial_number), !is.na(correct)) |>
    left_join(wv, by = "run_id") |> mutate(lrt = log(rt_numeric)) |> join_difficulty(tid)
}
eval_setting <- function(tr0, s){
  tr <- if (s$trials == "correct") filter(tr0, correct == 1) else tr0
  if (nrow(tr) < 150) return(c(rel = NA, retest = NA))
  has_d <- s$diff && sum(is.finite(tr$difficulty)) > 0.5*nrow(tr)
  if (has_d) tr <- tr |> mutate(difficulty = ifelse(is.finite(difficulty), difficulty, median(difficulty, na.rm = TRUE)))
  rhs <- if (s$trial) "ns(trial_number,3)" else NULL
  if (has_d) rhs <- c(rhs, "difficulty", if (s$trial) "difficulty:trial_number")
  if (n_distinct(tr$dataset) > 1) rhs <- c(rhs, "dataset")
  form <- if (is.null(rhs)) lrt ~ 1 else as.formula(paste("lrt ~", paste(rhs, collapse = "+")))
  tr$resid <- resid(lm(form, tr))
  tr <- tr |> group_by(run_id) |> arrange(trial_number, .by_group = TRUE) |> mutate(half = if_else(trial_number <= median(trial_number), 1L, 2L))
  if (s$ar) { tr <- tr |> mutate(lr = lag(resid)); rho <- scor(tr$resid, tr$lr); tr <- tr |> mutate(val = resid - rho*lr) } else tr <- tr |> mutate(val = resid)
  tr <- ungroup(tr)
  per <- function(by) tr |> group_by(across(all_of(by))) |> summarise(iiv = sd(val, na.rm = TRUE), n = n(), .groups = "drop") |> filter(n >= 8)
  rr <- per(c("run_id","half")) |> select(run_id, half, iiv) |> pivot_wider(names_from = half, values_from = iiv, names_prefix = "h")
  rel <- scor(rr$h1, rr$h2); rel <- if (is.na(rel)) NA else 2*rel/(1+rel)
  pr <- per("run_id") |> left_join(distinct(tr, run_id, user_id, wave), by = "run_id")
  rw <- pr |> select(user_id, wave, iiv) |> pivot_wider(names_from = wave, values_from = iiv, names_prefix = "w") |> filter(!is.na(w1), !is.na(w2))
  c(rel = rel, retest = scor(rw$w1, rw$w2))
}
facres <- map_dfr(seq_len(nrow(fac_tasks)), function(i){
  tr0 <- load_long(fac_tasks$task_id[i])
  map_dfr(seq_len(nrow(grid)), function(j){ m <- eval_setting(tr0, grid[j,])
    bind_cols(grid[j,], tibble(short = fac_tasks$short[i], speeded = fac_tasks$speeded[i], rel = m["rel"], retest = m["retest"])) })
})
```

```{r factorial-table}
facres |> group_by(trial, trials, diff, ar) |>
  summarise(reliability = mean(rel, na.rm = TRUE), test_retest = mean(retest, na.rm = TRUE), .groups = "drop") |>
  mutate(joint = (reliability + test_retest)/2) |> arrange(desc(joint)) |> head(6) |>
  mutate(across(c(reliability, test_retest, joint), \(x) round(x, 3))) |>
  knitr::kable(caption = "Top 6 of 16 purification settings (mean across tasks, by joint = (reliability + test–retest)/2)")
```

```{r factorial-marginals}
map_dfr(c("trial","diff","ar","trials"), function(f){
  facres |> group_by(level = as.character(.data[[f]])) |>
    summarise(reliability = mean(rel, na.rm = TRUE), test_retest = mean(retest, na.rm = TRUE), .groups = "drop") |>
    mutate(choice = f, .before = 1) }) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(caption = "Marginal effect of each choice (averaging over the other three)")
```

```{r factorial-diff-speeded}
facres |> group_by(tasks = if_else(speeded, "speeded", "non-speeded"), `difficulty detrend` = if_else(diff, "yes", "no")) |>
  summarise(reliability = mean(rel, na.rm = TRUE), test_retest = mean(retest, na.rm = TRUE), .groups = "drop") |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(caption = "Difficulty-detrending gains concentrate in the non-speeded tasks")
```

The factorial settles all four choices:

- **All trials** is the biggest single win, on both metrics — error trials
  (impulsive errors, lapses) carry attentional-variability signal, and including
  them steadies the second-moment estimate.
- **Difficulty-detrending helps**, validating the idea that part of RT
  variability is the item-difficulty *sequence*, not the child (reliability
  0.40→0.45, test–retest 0.18→0.19). The gain is concentrated in the
  **non-speeded** tasks (reliability 0.37→0.43, where RT is difficulty-driven) and
  adds nothing on the fast, difficulty-insensitive speeded tasks.
- **Trial-order detrending** helps modestly.
- **AR removal is a trade-off**: it raises split-half reliability (0.39→0.46) but
  *lowers* test–retest (0.19→0.18), because the autocorrelation it strips is partly
  a **person-stable** property. The single highest *joint* score removes it (its
  split-half gain outweighs its small test–retest cost on average), but for a
  **trait** measure — where longitudinal stability is the point — we **keep** it.

**Chosen pipeline (used throughout):** `log RT ~ ns(trial) + difficulty +
difficulty:trial + site`, on **all trials**, with the autocorrelation retained —
prioritizing longitudinal stability while remaining near the top on reliability.

# 2. Reliability depends on trial count

```{r rel-ntrials, fig.height=4.3}
summ |> ggplot(aes(median_trials, iiv_reliability)) +
  geom_smooth(method = "lm", se = FALSE, color = "grey70", linewidth = 0.7) +
  geom_point(aes(size = n_runs), alpha = 0.7, color = "#268bd2") +
  ggrepel::geom_text_repel(aes(label = short), size = 3.2) +
  scale_x_continuous(trans = "log10") +
  labs(x = "Median trials per run (log scale)", y = "Split-half reliability of RT-IIV",
       size = "n runs", title = "RT-IIV reliability rises with trial count",
       subtitle = "swr (120 trials) ≈ 0.94; difficulty-detrending also lifts several moderate-trial tasks")
```

Reliability still rises with trial count (ROAR-Word, 120 trials, ≈ 0.94; the
shortest tasks pa and mental rotation, ~ 19–26 trials, ≈ 0.5–0.57), but
difficulty-detrending has lifted several **moderate-trial non-speeded tasks** —
matrix (31 trials) and ToM (33) now reach ≈ 0.77–0.79 — loosening the
relationship. The practical lesson stands (enough trials still matters), and it
motivates the trial-count filter in §3.

# 3. Do cross-task RT-IIV cohere?

We build the child × task matrix of (detrended) RT-IIV — keeping site and age
variance in (we examine those as effects of interest in §4, not removing them) —
and examine the correlation matrix and factor structure, first across **all
tasks**, then restricted to tasks that clear a reliability/trial-count bar.

```{r crosstask}
iiv_adj <- iiv |> group_by(short) |> mutate(iiv_z = as.numeric(scale(iiv_detr))) |> ungroup()
wide <- iiv_adj |> select(user_id, short, iiv_z) |> distinct(user_id, short, .keep_all = TRUE) |>
  pivot_wider(names_from = short, values_from = iiv_z)
cmat <- cor(wide |> select(-user_id), use = "pairwise")
cor_heatmap <- function(cm, title){
  m <- cm; diag(m) <- NA
  df <- as.data.frame(m) |> rownames_to_column("t1") |> pivot_longer(-t1, names_to = "t2", values_to = "r")
  mid <- mean(range(df$r, na.rm = TRUE))
  df |> mutate(t1 = factor(t1, levels = rev(rownames(cm))), t2 = factor(t2, levels = rownames(cm)), dark = r > mid) |>
    ggplot(aes(t2, t1, fill = r)) + geom_tile(color = "white") +
    geom_text(aes(label = ifelse(is.na(r), "", sprintf("%.2f", r)), color = dark), size = 3, show.legend = FALSE) +
    scale_fill_viridis_c(na.value = "grey92", name = "r") +
    scale_color_manual(values = c(`TRUE` = "grey10", `FALSE` = "white"), na.value = "grey40") +
    coord_fixed() + labs(x = NULL, y = NULL, title = title) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), panel.grid = element_blank())
}
cor_heatmap(cmat, "Cross-task correlations of purified RT-IIV (all tasks)")
```

```{r fa-all}
ord <- rt_tasks$short
fa_report <- function(cm, N, label){
  fits <- map_dfr(1:3, function(nf){ f <- suppressWarnings(fa(cm, nfactors = nf, n.obs = N, fm = "ml", rotate = "oblimin"))
    tibble(factors = nf, TLI = f$TLI, RMSEA = f$RMSEA[1]) })
  list(fits = fits, load = suppressWarnings(fa(cm, nfactors = 2, n.obs = N, fm = "ml", rotate = "oblimin")))
}
mean_off <- function(cm) mean(cm[upper.tri(cm)], na.rm = TRUE)
allfa <- fa_report(cmat, nrow(wide), "all")
allfa$fits |> mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(caption = glue::glue("EFA fit, all tasks (mean off-diagonal r = {round(mean_off(cmat),3)})"))
L <- unclass(allfa$load$loadings); colnames(L) <- paste0("F", 1:ncol(L))
as.data.frame(round(L, 2)) |> rownames_to_column("task") |>
  knitr::kable(caption = "2-factor loadings, all tasks")
```

```{r fa-filtered}
keep <- summ |> filter(iiv_reliability >= 0.6, median_trials >= 25, n_runs >= 150) |> pull(short)
widef <- iiv_adj |> filter(short %in% keep) |> select(user_id, short, iiv_z) |>
  distinct(user_id, short, .keep_all = TRUE) |> pivot_wider(names_from = short, values_from = iiv_z)
cmatf <- cor(widef |> select(-user_id), use = "pairwise")
cor_heatmap(cmatf, glue::glue("Reliable subset: {paste(keep, collapse=', ')} (mean off-diagonal r = {round(mean_off(cmatf),3)})"))
fff <- fa_report(cmatf, nrow(widef), "filtered")
fff$fits |> mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(caption = "EFA fit, reliable subset")
Lf <- unclass(suppressWarnings(fa(cmatf, nfactors = 1, n.obs = nrow(widef), fm = "ml"))$loadings)
as.data.frame(round(Lf, 2)) |> rownames_to_column("task") |> arrange(desc(ML1)) |>
  knitr::kable(caption = "1-factor loadings, reliable subset")
```

The picture is **modest coherence, led by the speeded tasks but now broader**:

- Difficulty-purification lifts the cross-task correlation to ≈ 0.13 (from naive
  0.05), and to ≈ 0.20 in the reliable subset — higher than before difficulty was
  removed. A single general RT-IIV factor still does **not** fit the full set
  (1-factor TLI ≈ 0.34).
- The 2-factor solution shows a **broad RT-IIV factor** led by the speeded tasks
  (ROAR-Word .88, ROAR-Sentence .54, Hearts & Flowers .54) but now with the
  **non-speeded tasks loading on it too** (matrix .47, mental rotation .42,
  math .41) — purifying their difficulty-driven RT brought them partly into the
  common factor, so the earlier sharp speeded-vs-cognitive split has softened. The
  second factor is essentially the small-sample Same-&-Different artifact (n ≈ 78,
  loading .94), which drops out under the reliability filter.
- Restricting to the **reliable, high-trial tasks** (matrix, vocab, H&F, ToM,
  ROAR-Word, ROAR-Sentence) raises the mean correlation to ≈ 0.20 and yields a
  passable single factor still **dominated by the speeded tasks** (ROAR-Word .95,
  H&F .54, ROAR-Sentence .44, ToM .37) with the cognitive tasks weaker (matrix .29,
  vocab .15).

# 4. Site and age differences in RT-IIV

Rather than scrubbing site and age out of the IIV, we examine them directly: do
children respond more variably at some **sites** (administration medium, setting,
population) or at younger **ages**?

```{r site-age-prep}
library(lme4)
iiv_sa <- iiv |> left_join(runs |> select(run_id, age), by = "run_id") |>
  left_join(sc |> distinct(run_id, dataset), by = "run_id") |>
  mutate(site = recode(dataset,
           pilot_mpieva_de_main = "DE (Leipzig)", pilot_uniandes_co_bogota = "CO (Bogotá-urban)",
           pilot_uniandes_co_rural = "CO (Bogotá-rural)", pilot_western_ca_main = "CA (Western)"),
         iiv_z = ave(iiv_detr, short, FUN = \(x) as.numeric(scale(x))), age_c = age - 9)
```

## 4.1 Age

```{r age-plot, fig.height=4.3}
slope <- fixef(lmer(iiv_detr ~ age + (1|short), data = iiv_sa))["age"]
iiv_sa |> ggplot(aes(age, iiv_detr, color = short)) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 0.7) +
  scale_color_viridis_d(name = NULL) +
  labs(x = "Age (years)", y = "Purified RT-IIV (SD detrended log-RT)",
       title = glue::glue("Purified RT-IIV vs age (pooled slope {round(slope,3)} log-RT-SD/yr)"),
       subtitle = "Modest overall decline — concentrated in the speeded tasks")
```

```{r age-slopes}
iiv_sa |> group_by(short) |> summarise(age_slope = coef(lm(iiv_detr ~ age))[2], .groups = "drop") |>
  arrange(age_slope) |> mutate(age_slope = round(age_slope, 4)) |>
  knitr::kable(caption = "Per-task age slope of RT-IIV (negative = more consistent with age)")
```

RT-IIV declines with age overall, but the decline is **concentrated in the
speeded tasks** — ROAR-Word (−0.051/yr), ROAR-Sentence (−0.029), Hearts & Flowers
(−0.027) — exactly the tasks that carry the cross-task RT-IIV signal (§3), while
the untimed cognitive tasks are essentially flat. Developmental gains in
response consistency show up where RT-IIV is a clean measure.

This is far sharper at the **factor** level. Scoring the §3 two-factor structure
cross-sectionally (mean within-task z, ≥ 2 tasks) and regressing on age:

```{r factor-age, fig.height=4.3}
speeded <- c("swr","sre","hf"); cognitive <- c("matrix","mrot","trog","math")
fac_age <- iiv_sa |>
  mutate(fac = case_when(short %in% speeded ~ "Speeded (swr/sre/hf)",
                         short %in% cognitive ~ "Untimed-cognitive (matrix/mrot/trog/math)", TRUE ~ NA_character_)) |>
  filter(!is.na(fac)) |> group_by(user_id, age, fac) |>
  summarise(score = mean(iiv_z), ntask = n(), .groups = "drop") |> filter(ntask >= 2)
fac_age |> ggplot(aes(age, score, color = fac)) +
  geom_smooth(method = "loess", se = TRUE, linewidth = 0.9) +
  scale_color_manual(values = c("Speeded (swr/sre/hf)" = "#268bd2",
                                "Untimed-cognitive (matrix/mrot/trog/math)" = "#cb4b16"), name = NULL) +
  labs(x = "Age (years)", y = "RT-IIV factor score (mean within-task z)",
       title = "Speeded RT-IIV falls steeply with age; untimed-cognitive only shallowly",
       subtitle = "Cross-sectional, the §3 two-factor scores") +
  theme(legend.position = "bottom")
```

```{r factor-age-slopes}
fac_age |> group_by(fac) |>
  summarise(n = n(), age_slope = coef(lm(score ~ age))[2], r_with_age = cor(score, age), .groups = "drop") |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(caption = "Cross-sectional age slope of each RT-IIV factor score")
```

The two factors **differ sharply on age**. The speeded factor declines strongly
(slope ≈ −0.19 z/yr, r ≈ −0.43 — about 1.3 SD lower from age 5–7 to 11–13), the
classic developmental rise in processing/attentional consistency. The
untimed-cognitive factor declines only **weakly** (slope ≈ −0.06, r ≈ −0.16) —
shallower by a factor of three; once difficulty is purified out it shows a faint
maturation signal but nothing like the speeded factor. So the speeded RT-IIV
factor remains the one that is most coherent (§3), most stable (§5), and most
developmentally graded.

## 4.2 Site

```{r site-means, fig.height=4.3}
iiv_sa |> group_by(short, site) |> summarise(m = mean(iiv_z), se = sd(iiv_z)/sqrt(n()), .groups = "drop") |>
  ggplot(aes(site, m, color = site)) +
  geom_hline(yintercept = 0, linetype = 2, color = "grey70") +
  geom_pointrange(aes(ymin = m - se, ymax = m + se), size = 0.3) +
  facet_wrap(~short, nrow = 2) + scale_color_brewer(palette = "Set1", guide = "none") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 6)) +
  labs(x = NULL, y = "RT-IIV (within-task z)", title = "RT-IIV by site, per task",
       subtitle = "Bogotá (urban & rural) consistently above Leipzig / Western")
```

```{r site-model}
m_site <- lmer(iiv_z ~ dataset + age_c + (1|short), data = iiv_sa)
vc <- as.data.frame(VarCorr(lmer(iiv_z ~ (1|short) + (1|dataset), data = iiv_sa)))
co <- summary(m_site)$coefficients
tibble(term = c("CO Bogotá-urban vs DE","CO Bogotá-rural vs DE","CA Western vs DE","age (per year)"),
       estimate = co[c("datasetpilot_uniandes_co_bogota","datasetpilot_uniandes_co_rural",
                       "datasetpilot_western_ca_main","age_c"), "Estimate"],
       t = co[c("datasetpilot_uniandes_co_bogota","datasetpilot_uniandes_co_rural",
                "datasetpilot_western_ca_main","age_c"), "t value"]) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(caption = glue::glue(
    "Site & age effects on RT-IIV (z), age-adjusted; site explains {round(vc$vcov[vc$grp=='dataset']/sum(vc$vcov)*100,1)}% of total IIV variance"))
```

There are clear **group-level site differences**: children in Bogotá — both
urban (+0.29 SD) and rural (+0.38 SD) — are substantially **more RT-variable**
than Leipzig even after age adjustment, while Western ≈ Leipzig. These are large,
consistent mean shifts, though site accounts for only ~4.6% of *total* RT-IIV
variance (individual differences dominate). The cause is not identifiable here —
it could be administration medium (devices, supervision, setting), population, or
language — but it is a clear caution against treating RT-IIV as directly
comparable across sites without accounting for it, and a flag for the
measurement-invariance of variability indices.

# 5. Longitudinal stability

Split-half reliability (§2) is a *within-session* property — it says nothing
about whether a child's RT-IIV is stable over time. The real trait test is
**test–retest stability** across waves. We use the longitudinal samples
(**Leipzig** + **Bogotá** urban & rural), take each child's first two
administrations per task, purify as before, and correlate wave-1 with wave-2
RT-IIV.

```{r stability, cache=TRUE}
long_sites <- c("pilot_mpieva_de_main","pilot_uniandes_co_bogota","pilot_uniandes_co_rural")
wv <- sc |> filter(dataset %in% long_sites, task_id %in% rt_tasks$task_id, !is.na(score)) |>
  group_by(user_id, task_id) |> arrange(age) |> mutate(wave = row_number(), nwave = max(wave)) |>
  filter(nwave >= 2, wave <= 2) |> ungroup() |> select(run_id, user_id, task_id, age, dataset, wave, language)

stab <- map_dfr(rt_tasks$task_id, function(tid){
  rw <- wv |> filter(task_id == tid)
  tr <- load_levante_trials(task_ids = tid) |>
    filter(run_id %in% rw$run_id, !is.na(rt_numeric), rt_numeric > 200, rt_numeric < 30000, !is.na(trial_number)) |>
    left_join(rw |> select(run_id, language), by = "run_id") |> mutate(lrt = log(rt_numeric))
  if (nrow(tr) < 50) return(NULL)
  tr <- tr |> join_difficulty(tid) |> detrend_resid()
  pr <- tr |> group_by(run_id) |> filter(n() >= 8) |> summarise(iiv = sd(resid), .groups = "drop") |>
    left_join(rw, by = "run_id")
  w <- pr |> select(user_id, wave, iiv, age) |>
    pivot_wider(names_from = wave, values_from = c(iiv, age), names_sep = "") |>
    filter(!is.na(iiv1), !is.na(iiv2))
  sh <- tr |> filter(run_id %in% (rw |> filter(wave == 1) |> pull(run_id))) |>
    group_by(run_id) |> filter(n() >= 12) |> mutate(h = row_number() %% 2) |>
    group_by(run_id, h) |> summarise(v = sd(resid), .groups = "drop") |>
    pivot_wider(names_from = h, values_from = v, names_prefix = "h")
  rsh <- cor(sh$h0, sh$h1, use = "complete.obs")
  tibble(short = rt_tasks$short[rt_tasks$task_id == tid], n = nrow(w),
         gap_yr = median(w$age2 - w$age1), within_session = 2*rsh/(1+rsh),
         test_retest = cor(w$iiv1, w$iiv2))
}) |> filter(n >= 30)
```

```{r stability-table}
pooled <- map_dfr(rt_tasks$task_id, function(tid){
  rw <- wv |> filter(task_id == tid)
  tr <- load_levante_trials(task_ids = tid) |>
    filter(run_id %in% rw$run_id, !is.na(rt_numeric), rt_numeric > 200, rt_numeric < 30000, !is.na(trial_number)) |>
    left_join(rw |> select(run_id, language), by = "run_id") |> mutate(lrt = log(rt_numeric))
  if (nrow(tr) < 50) return(NULL)
  tr <- tr |> join_difficulty(tid) |> detrend_resid()
  tr |> group_by(run_id) |> filter(n() >= 8) |> summarise(iiv = sd(resid), .groups = "drop") |>
    left_join(rw, by = "run_id") |> mutate(short = rt_tasks$short[rt_tasks$task_id == tid])
}) |> group_by(short) |> mutate(iivz = as.numeric(scale(iiv))) |> ungroup()
pw <- pooled |> select(user_id, short, wave, iivz) |> pivot_wider(names_from = wave, values_from = iivz, names_prefix = "w") |>
  filter(!is.na(w1), !is.na(w2))
pooled_r <- cor(pw$w1, pw$w2)
stab |> mutate(disattenuated = test_retest/within_session) |> arrange(desc(test_retest)) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(caption = glue::glue("Longitudinal test–retest stability of RT-IIV (Leipzig + Colombia urban/rural; pooled across tasks r = {round(pooled_r,3)}, n = {nrow(pw)})"))
```

```{r stability-plot, fig.height=4.3}
stab |> select(short, `within-session (split-half)` = within_session, `across-wave (test–retest)` = test_retest) |>
  pivot_longer(-short) |>
  ggplot(aes(value, reorder(short, value), color = name)) +
  geom_line(aes(group = short), color = "grey75") + geom_point(size = 2.6) +
  scale_color_manual(values = c("within-session (split-half)" = "#93a1a1", "across-wave (test–retest)" = "#268bd2"), name = NULL) +
  labs(x = "Reliability / stability", y = NULL,
       title = "RT-IIV is internally reliable but only moderately stable over time",
       subtitle = "Gap between within-session and across-wave = occasion-specific (state) variance") +
  theme(legend.position = "bottom")
```

## Test–retest by site

The same wave-1 vs wave-2 pairs, split by site. Two things are visible at
once: **the speeded tasks (blue) are the stable ones at every site** (r ≈
0.3–0.5, vs ≈ 0.1–0.2 for the untimed tasks), and **rural Colombia's second
wave adds a large new mass of speeded pairs at the top of that range**
(speeded r ≈ 0.51) — which is what moves the aggregate stability numbers, and
what makes the speeded factor-score composite (next section) finally
computable on a real sample.

```{r stability-site-scatter, fig.height=4.6, fig.width=10.5}
site_lbl <- c(pilot_mpieva_de_main = "Germany — Leipzig",
              pilot_uniandes_co_bogota = "Colombia — Bogotá (urban)",
              pilot_uniandes_co_rural = "Colombia — Caquetá/Boyacá (rural)")
speeded_set <- c("swr", "sre", "hf")
pw_site <- pooled |>
  select(user_id, dataset, short, wave, iivz) |>
  pivot_wider(names_from = wave, values_from = iivz, names_prefix = "w") |>
  filter(!is.na(w1), !is.na(w2)) |>
  mutate(site = factor(site_lbl[dataset], levels = site_lbl),
         cluster = if_else(short %in% speeded_set, "speeded (swr/sre/hf)", "untimed-cognitive"))
site_r <- pw_site |> group_by(site) |>
  summarise(n = n(), r = cor(w1, w2),
            n_speeded = sum(cluster != "untimed-cognitive"),
            r_speeded = cor(w1[cluster != "untimed-cognitive"], w2[cluster != "untimed-cognitive"]),
            .groups = "drop")
ggplot(pw_site, aes(w1, w2)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey75", lty = 3) +
  geom_point(aes(colour = cluster), alpha = 0.45, size = 1.1) +
  geom_smooth(method = "lm", se = FALSE, colour = "grey25", linewidth = 0.7) +
  geom_text(data = site_r, aes(x = -2.1, y = 3.4,
            label = glue::glue("r = {round(r, 2)} (n = {n})\nspeeded r = {round(r_speeded, 2)} (n = {n_speeded})")),
            hjust = 0, size = 3.1) +
  facet_wrap(~site) +
  scale_colour_manual(values = c(`speeded (swr/sre/hf)` = "#268bd2",
                                 `untimed-cognitive` = "#93a1a1"), name = NULL) +
  coord_equal(xlim = c(-2.3, 3.7), ylim = c(-2.3, 3.7)) +
  labs(x = "RT-IIV, wave 1 (z within task)", y = "RT-IIV, wave 2 (z within task)",
       title = "RT-IIV test–retest by site: the speeded tasks are the stable ones everywhere",
       subtitle = "Each point = one child × task; purified RT-IIV; ~1-year gap (rural ≈ 9.6 mo)") +
  theme(legend.position = "bottom")
```

RT-IIV is **partly trait-like, partly state**. Pooled across tasks the
test–retest correlation is ≈ 0.24, and even the best tasks — ROAR-Word (0.55,
now on n = 389 with rural Colombia's second wave), Hearts & Flowers (0.46),
Same-&-Different (0.40), ROAR-Sentence (0.39) — sit well below their
within-session reliabilities (0.6–0.93). The drop is the gap between the two
series in the plot: a large share of a child's RT-IIV on any given day is
**occasion-specific**, not a stable trait. Two things sharpen the read:

- The **speeded tasks** (ROAR-Word, H&F, ROAR-Sentence) that anchored the
  cross-task factor (§3) are again the most stable — and with the 2026-08 rural
  retest data ROAR-Word overtakes H&F as the single most stable task.
- Even **disattenuated** for within-session unreliability, stability tops out
  around 0.5–0.6 — so the *true* RT-IIV trait still carries substantial
  year-to-year change (partly real developmental change: the speeded-task gaps
  span ~0.8–1.3 yr vs ~0.3–0.5 yr for the core tasks).

## Do the two-factor scores hold up better?

Pooling tasks into the §3 two-factor scores (speeded = swr/sre/hf;
untimed-cognitive = matrix/mrot/trog/math) *should* cancel task-specific noise
and stabilise the estimate — if the factors carry real shared variance. We score
each child's two RT-IIV factors per wave (mean of available within-task z-IIVs,
≥ 2 tasks) and test their wave-to-wave stability against the single tasks.

```{r factor-stability}
speeded <- c("swr","sre","hf"); cognitive <- c("matrix","mrot","trog","math")
fc <- pooled |>
  mutate(fac = case_when(short %in% speeded ~ "Speeded (swr/sre/hf)",
                         short %in% cognitive ~ "Cognitive (matrix/mrot/trog/math)", TRUE ~ NA_character_)) |>
  filter(!is.na(fac)) |> group_by(user_id, wave, fac) |>
  summarise(score = mean(iivz), ntask = n(), .groups = "drop") |> filter(ntask >= 2)
fac_stab <- fc |> select(user_id, wave, fac, score) |>
  pivot_wider(names_from = wave, values_from = score, names_prefix = "w") |>
  filter(!is.na(w1), !is.na(w2)) |> group_by(fac) |> summarise(n = n(), factor_score_r = cor(w1, w2), .groups = "drop")
contrast <- tibble(fac = c("Speeded (swr/sre/hf)","Cognitive (matrix/mrot/trog/math)"),
  mean_single_task_r = c(mean(stab$test_retest[stab$short %in% speeded]),
                         mean(stab$test_retest[stab$short %in% cognitive])),
  best_single_task_r = c(max(stab$test_retest[stab$short %in% speeded]),
                         max(stab$test_retest[stab$short %in% cognitive])))
fac_stab |> left_join(contrast, by = "fac") |> mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(caption = "Two-factor RT-IIV factor-score stability vs single-task stability")
```

**2026-08 update: this table reverses the June conclusion for the speeded
factor.** In the June render only 57 children had ≥ 2 speeded tasks repeated —
a *coverage* artifact (H&F repeats lived mainly in Leipzig, ROAR repeats in
Bogotá-urban, and almost no child had both) — and the speeded factor score
(≈ 0.20) sat *below* H&F alone, which we then read as "the durable RT-IIV
signal is task-specific." Rural Colombia's second wave repairs exactly that
coverage gap: its children repeat ROAR-Word *and* ROAR-Sentence together,
giving **n = 224**, and the speeded factor score's stability jumps to
**≈ 0.63 — above the best single task (ROAR-Word, 0.55)**. Aggregation *does*
help once the tasks actually co-occur: the speeded RT-IIV factor is now the
most stable within-child variability index in the project. The cognitive
factor is unchanged in its verdict (factor 0.24, n = 312, vs best single task
0.26 — no gain), consistent with §3: the durable shared variance is specific
to the speeded cluster, not a general RT-IIV trait.

# Synthesis

```{r save}
write_rds(list(summary = summ, cor_all = cmat, cor_reliable = cmatf,
               keep = keep, use_trials = USE_TRIALS,
               stability = stab, stability_pooled = pooled_r, factor_stability = fac_stab),
          here::here("data/rt_variability_07.rds"))
```

- **Purification helps, modestly.** The pipeline chosen in §1.4 — detrending the
  within-task time-course **and item difficulty**, on all trials, autocorrelation
  retained — roughly triples cross-task RT-IIV coherence versus the naive per-run
  SD (≈ 0.05 → ≈ 0.13 across all tasks, ≈ 0.20 in the reliable subset). The
  difficulty term is the key addition, and it pays off most on the non-speeded
  tasks whose RT is difficulty-driven.
- **Trial count still matters.** RT-IIV reliability runs from ≈ 0.5 (≤ ~25 trials)
  to ≈ 0.94 (120 trials), though difficulty-detrending now lifts several
  moderate-trial non-speeded tasks (matrix, ToM ≈ 0.77–0.79). Designs that want
  this measure should still ensure enough trials.
- **Coherence is modest, led by the speeded tasks but now broader.** RT-IIV forms
  a **broad factor** dominated by the speeded tasks (ROAR-Word / Hearts & Flowers /
  ROAR-Sentence) onto which the difficulty-purified non-speeded tasks (matrix,
  mental rotation, math) now also load — not a strong single general trait, but no
  longer two cleanly separate strands.
- **Site and age differences are real.** RT-IIV declines with age (speeded factor
  ≈ −0.19 z/yr; untimed-cognitive only ≈ −0.06). And children in **Bogotá are
  markedly more RT-variable** than Leipzig (urban +0.29, rural +0.38 SD,
  age-adjusted; Western ≈ Leipzig) — a large, consistent group-level shift (though
  site is only ~4.6% of total IIV variance) that may reflect administration medium,
  setting, or population and cautions against treating RT-IIV as directly
  cross-site comparable.
- **Moderately stable over time; the speeded composite is the keeper.** Across
  ~1-year waves (Leipzig + Colombia urban/rural), RT-IIV test–retest is ≈ 0.24
  pooled — well below within-session reliability, so much of a child's RT-IIV
  on a given day is occasion-specific. But (2026-08 revision, with rural's
  second wave repairing the old coverage gap) the **speeded factor score
  (swr/sre/hf) reaches test–retest ≈ 0.63 (n = 224), above any single task** —
  aggregation does help once the speeded tasks co-occur within child. The
  cognitive factor still shows no aggregation gain (≈ 0.24): the durable signal
  is the speeded cluster, not a general RT-IIV trait.
- **Implication.** If LEVANTE wants RT-IIV as an individual-difference measure,
  the leverage is in **speeded tasks with many trials**; reading it off untimed or
  short tasks will mostly capture task-specific strategy noise. This is a sharper,
  more optimistic read than 06's flat null — purification + trial-count awareness
  surfaces a real (if narrow) speeded-task RT-IIV factor, and (as of the 2026-08
  data) its **composite is decently stable (≈ 0.63)** — a usable partly-stable
  index, though still not a fixed trait.
