---
title: "08 — Accuracy variability: MSSD, the accuracy-level confound, and whether anything survives it"
levante:
  summary: "Does accuracy variability (MSSD, the trial-to-trial flip rate) carry reliable individual differences once its mechanical dependence on accuracy level is removed?"
  status: draft
  task_ids: [egma-math, matrix-reasoning, mental-rotation, trog, vocab, same-different-selection, hearts-and-flowers, theory-of-mind, memory-game, swr, sre, pa]
  datasets: [pilot_mpieva_de_main, pilot_uniandes_co_bogota, pilot_uniandes_co_rural, pilot_western_ca_main]
  data:
    - {source: "load_levante_scores", version: "v1_2", table: "scores"}
    - {source: "load_levante_trials", version: "v1_2", table: "trials"}
    - {source: "load_item_parameters", version: "v1_14", table: "item_parameters"}
  depends_on: [00_load_data.qmd, 07_rt_variability.qmd]
  findings:
    - "Raw and detrended MSSD are reliable and stable only because they re-encode accuracy level (confound ≈ −0.7)."
    - "The de-confounded index (excess over independence ≡ lag-1 autocorrelation) has reliability ≈ 0.03 and stability ≈ 0.02."
    - "Do not use accuracy MSSD as a variability measure; the leverage is RT on speeded, high-trial tasks (07)."
  last_run: 2026-07-13
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 9
    fig-height: 5.5
execute:
  warning: false
  message: false
---

{{< include _disclaimer.qmd >}}

# The question

`07` showed RT intra-individual variability has a real (if narrow) signal once
purified. This notebook is the **accuracy** companion. Accuracy is binary, so the
RT tools (iSD, CV, ex-Gaussian) do not apply — the variability-indices tutorial
(van Langen, Aristodemou & Kievit) flags **MSSD** (mean square of successive
differences) as the index built for 0/1 data. For a binary sequence MSSD is just
the **trial-to-trial flip rate**: how often the child switches right↔wrong on
successive trials.

The central worry is a confound: for an independent Bernoulli sequence the flip
rate is mechanically **2·p·(1−p)**, peaking at p = 0.5 and falling to 0 at
floor/ceiling. So raw MSSD largely re-encodes **accuracy level** (≈ ability). We
therefore compute three variants and — per the brief — carry the **reliability
and longitudinal-stability** checks through to see which, if any, survives
de-confounding:

1. **raw MSSD** — the flip rate (no correction)
2. **detrended MSSD** — flip-variability of residuals after removing the
   within-task trial-order trend (handles warm-up/fatigue, not accuracy level)
3. **excess over independence** — observed flip rate minus the child's *own*
   independence expectation 2·p·(1−p) (handles accuracy level; equivalent to the
   scale-free **lag-1 autocorrelation** of the accuracy sequence)

```{r setup}
source(here::here("common.R"))
library(splines)
sc <- read_rds(here("data/scores_all_sites.rds"))
sites4 <- c("pilot_mpieva_de_main","pilot_uniandes_co_bogota","pilot_uniandes_co_rural","pilot_western_ca_main")
long_sites <- c("pilot_mpieva_de_main","pilot_uniandes_co_bogota","pilot_uniandes_co_rural")
acc_tasks <- tibble(
  task_id = c("egma-math","matrix-reasoning","mental-rotation","trog","vocab","same-different-selection",
              "hearts-and-flowers","theory-of-mind","memory-game","swr","sre","pa"),
  short   = c("math","matrix","mrot","trog","vocab","sds","hf","tom","mg","swr","sre","pa"))
adap <- c("egma-math","matrix-reasoning","mental-rotation","same-different-selection","trog","vocab")
scor <- function(x, y){ ok <- is.finite(x) & is.finite(y)
  if (sum(ok) < 3 || sd(x[ok]) == 0 || sd(y[ok]) == 0) return(NA_real_); cor(x[ok], y[ok]) }

# per-run accuracy-variability indices for a task, given a set of runs (+ optional grouping col)
acc_index <- function(tid, runs_tbl, extra = character()){
  tr <- load_levante_trials(task_ids = tid) |>
    filter(run_id %in% runs_tbl$run_id, !is.na(correct), !is.na(trial_number)) |>
    left_join(runs_tbl |> select(run_id, all_of(extra)), by = "run_id") |>
    arrange(run_id, trial_number)
  if (nrow(tr) < 150 || n_distinct(tr$correct) < 2) return(NULL)
  ns_ <- n_distinct(tr$dataset)
  form <- if (ns_ > 1) correct ~ ns(trial_number, 3) + dataset else correct ~ ns(trial_number, 3)
  tr$phat <- predict(suppressWarnings(glm(form, binomial, tr)), type = "response")
  tr |> group_by(run_id) |>
    mutate(lc = lag(correct), resid = correct - phat, lresid = lag(resid),
           flip = as.numeric(correct != lc), dresid = resid - lresid,
           half = if_else(trial_number <= median(trial_number), 1L, 2L)) |> ungroup()
}
agg_run <- function(d, by = c("run_id")){
  d |> group_by(across(all_of(by))) |>
    summarise(acc = mean(correct), ntrial = n(),
              raw = mean(flip, na.rm = TRUE),
              detrended = mean(dresid^2, na.rm = TRUE),
              excess = mean(flip, na.rm = TRUE) - 2*mean(correct)*(1-mean(correct)),
              .groups = "drop") |>
    mutate(short = unique(d$short))
}
```

# 1. Computing the index and the accuracy-level confound

Cross-sectional sample: first observation per child per task, four data-rich
sites, newest (adaptive) version where one exists — the same frame as 07.

```{r extract-xs, cache=TRUE}
runs_xs <- sc |> filter(dataset %in% sites4, task_id %in% acc_tasks$task_id, !is.na(score)) |>
  filter(!(task_id %in% adap) | adaptive) |>
  group_by(user_id, task_id) |> arrange(age) |> slice(1) |> ungroup() |>
  select(run_id, user_id, task_id, short, age)   # sc already carries `short`
ex_xs <- map(acc_tasks$task_id, ~acc_index(.x, runs_xs |> filter(task_id == .x)))
trial_xs <- set_names(ex_xs, acc_tasks$short)
idx_xs <- map_dfr(ex_xs, ~if (is.null(.x)) NULL else agg_run(.x) |> filter(ntrial >= 10) |>
                    left_join(runs_xs |> select(run_id, user_id, age), by = "run_id"))
```

```{r confound, fig.height=4.3}
conf <- idx_xs |> group_by(short) |>
  summarise(raw = scor(raw, acc), detrended = scor(detrended, acc), excess = scor(excess, acc), .groups = "drop") |>
  pivot_longer(-short, names_to = "variant", values_to = "r")
conf |> mutate(variant = factor(variant, c("raw","detrended","excess"))) |>
  ggplot(aes(r, reorder(short, r), color = variant)) +
  geom_vline(xintercept = 0, linetype = 2, color = "grey60") +
  geom_point(size = 2.4) +
  scale_color_manual(values = c(raw = "#cb4b16", detrended = "#b58900", excess = "#268bd2"), name = NULL) +
  labs(x = "Correlation of variability index with accuracy level", y = NULL,
       title = "Raw and detrended MSSD are ≈ accuracy; excess-over-independence is not",
       subtitle = "Detrending the trial-order trend does nothing to the accuracy confound") +
  theme(legend.position = "bottom")
```

Raw MSSD correlates ≈ **−0.7** with accuracy (more accurate children flip
less), and **detrending the trial-order trend changes this essentially nothing**
(raw ≈ detrended on every task) — the confound is the accuracy *level*, not a
time trend. The **excess-over-independence** index removes it (confound ≈ 0):
once we subtract each child's own 2·p·(1−p) expectation, the flip rate is no
longer a proxy for ability.

# 2. Does anything survive de-confounding? (reliability & stability)

The key test the brief asked for: do the de-confounded indices keep good
**internal reliability** and **longitudinal stability**, or does removing the
accuracy confound also remove the signal? Internal reliability uses a
**first-half/second-half** split (MSSD is order-dependent, so odd/even would
break the successive structure); stability is wave-1 vs wave-2 in the
longitudinal samples (Leipzig + Bogotá).

```{r perf, cache=TRUE}
vars <- c("raw","detrended","excess")
# internal reliability (first vs second half) on the cross-sectional sample
rel_tab <- map_dfr(acc_tasks$short, function(s){
  d <- trial_xs[[s]]; if (is.null(d)) return(NULL)
  hh <- agg_run(d, by = c("run_id","half")) |> filter(ntrial >= 5)
  tibble(short = s, variant = vars, reliability = map_dbl(vars, function(v){
    w <- hh |> select(run_id, half, val = all_of(v)) |> pivot_wider(names_from = half, values_from = val, names_prefix = "h")
    r <- scor(w$h1, w$h2); if (is.na(r)) NA else 2*r/(1+r) }))
})
# longitudinal test-retest (DE + Bogotá)
wv <- sc |> filter(dataset %in% long_sites, task_id %in% acc_tasks$task_id, !is.na(score)) |>
  group_by(user_id, task_id) |> arrange(age) |> mutate(wave = row_number(), nwave = max(wave)) |>
  filter(nwave >= 2, wave <= 2) |> ungroup() |>
  select(run_id, user_id, task_id, short, wave)   # sc already carries `short`
ret_tab <- map_dfr(acc_tasks$task_id, function(tid){
  s <- acc_tasks$short[acc_tasks$task_id == tid]
  d <- acc_index(tid, wv |> filter(task_id == tid), extra = "wave"); if (is.null(d)) return(NULL)
  pr <- agg_run(d, by = c("run_id","wave")) |> filter(ntrial >= 10) |>
    left_join(wv |> select(run_id, user_id), by = "run_id")
  tibble(short = s, variant = vars, test_retest = map_dbl(vars, function(v){
    w <- pr |> select(user_id, wave, val = all_of(v)) |> pivot_wider(names_from = wave, values_from = val, names_prefix = "w") |>
      filter(!is.na(w1), !is.na(w2)); scor(w$w1, w$w2) }))
})
perf <- rel_tab |> left_join(ret_tab, by = c("short","variant")) |>
  left_join(conf |> rename(acc_confound = r), by = c("short","variant"))
```

```{r perf-summary}
perf |> group_by(variant) |>
  summarise(`mean |accuracy confound|` = mean(abs(acc_confound), na.rm = TRUE),
            `mean reliability` = mean(reliability, na.rm = TRUE),
            `mean test–retest` = mean(test_retest, na.rm = TRUE), .groups = "drop") |>
  mutate(variant = factor(variant, vars)) |> arrange(variant) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(caption = "Across tasks: de-confounding (excess) trades away the reliable, stable signal")
```

```{r perf-plot, fig.height=4.3}
perf |> select(short, variant, reliability, test_retest) |>
  pivot_longer(c(reliability, test_retest), names_to = "metric", values_to = "value") |>
  mutate(variant = factor(variant, vars),
         metric = recode(metric, reliability = "internal reliability", test_retest = "longitudinal test–retest")) |>
  ggplot(aes(value, short, color = variant)) +
  geom_vline(xintercept = 0, linetype = 2, color = "grey70") +
  geom_point(size = 2.2) + facet_wrap(~metric) +
  scale_color_manual(values = c(raw = "#cb4b16", detrended = "#b58900", excess = "#268bd2"), name = NULL) +
  labs(x = NULL, y = NULL, title = "Reliable and stable only while confounded with accuracy",
       subtitle = "raw/detrended (orange/gold) perform well but are ≈ ability; excess (blue) de-confounds but collapses to ~0") +
  theme(legend.position = "bottom")
```

This is the headline. **Raw and detrended MSSD are reliably (≈ 0.44) and stably
(test–retest ≈ 0.38) measured** — on par with RT-IIV — **but only because they
re-encode accuracy** (confound ≈ −0.7). The moment we remove the accuracy-level
dependency, the **excess** index de-confounds (≈ 0) but its reliability falls to
≈ 0.03 and its stability to ≈ 0.02. There is essentially
**no stable, individual-differences signal in accuracy variability beyond
accuracy level itself.** (The scale-free lag-1 autocorrelation gives the same
verdict.) This mirrors `06`'s person-misfit result — once correctly de-confounded,
"surprise"/instability in accuracy carries little reliable trait variance.

## Does a better ability anchor (theta) rescue it?

The observed within-run proportion correct is a noisy ability anchor — and for
adaptive tasks it is pinned near 0.5 regardless of ability. A principled
alternative is to baseline against the **IRT model's predicted P per trial**
(from the child's θ̂, item difficulty, and the fixed guessing asymptote the
scoring uses), so the independence expectation reflects the actual
item-difficulty sequence. We compare this theta-based excess to the observed-p
excess on the longitudinal sample.

```{r theta-baseline, cache=TRUE}
ip <- load_item_parameters(); bylang_t <- c("vocab","theory-of-mind","swr","pa")
irt_tasks <- setdiff(acc_tasks$task_id, "sre")   # sre has no IRT parameters
theta_cmp <- map_dfr(irt_tasks, function(tid){
  w <- wv |> filter(task_id == tid) |> left_join(sc |> distinct(run_id, theta = score, language), by = "run_id")
  tr <- load_levante_trials(task_ids = tid) |> filter(run_id %in% w$run_id, !is.na(correct), !is.na(trial_number)) |>
    left_join(w |> select(run_id, wave, theta, language), by = "run_id") |> arrange(run_id, trial_number)
  ipx <- ip |> filter(task_id == tid)
  if (tid %in% bylang_t) tr <- tr |> inner_join(ipx |> transmute(language = subset, item_uid, b = difficulty, a = discrimination), by = c("item_uid","language"))
  else tr <- tr |> inner_join(ipx |> transmute(item_uid, b = difficulty, a = discrimination) |> distinct(item_uid, .keep_all = TRUE), by = "item_uid")
  if (nrow(tr) < 150) return(NULL)
  tr <- tr |> mutate(cg = ifelse(is.na(chance), 0, chance), P = pmin(pmax(cg + (1-cg)/(1+exp(-a*(theta-b))), .001), .999)) |>
    group_by(run_id) |> mutate(lc = lag(correct), lP = lag(P), flip = as.numeric(correct != lc),
      flip_exp_t = P*(1-lP) + lP*(1-P), half = if_else(trial_number <= median(trial_number), 1L, 2L)) |> ungroup()
  agg <- function(d) d |> summarise(ntrial = n(), th = first(theta),
    obs_p = mean(flip, na.rm = TRUE) - 2*mean(correct)*(1-mean(correct)),
    theta_irf = mean(flip - flip_exp_t, na.rm = TRUE), .groups = "drop")
  pr <- tr |> group_by(run_id, wave) |> agg() |> filter(ntrial >= 10) |> left_join(sc |> distinct(run_id, user_id), by = "run_id")
  hh <- tr |> group_by(run_id, half) |> agg() |> filter(ntrial >= 5)
  one <- function(v){
    rh <- hh |> select(run_id, half, val = all_of(v)) |> pivot_wider(names_from = half, values_from = val, names_prefix = "h")
    rr <- scor(rh$h1, rh$h2); rw <- pr |> select(user_id, wave, val = all_of(v)) |> pivot_wider(names_from = wave, values_from = val, names_prefix = "w") |> filter(!is.na(w1), !is.na(w2))
    tibble(baseline = v, conf_theta = scor(pr[[v]], pr$th), reliability = if (is.na(rr)) NA else 2*rr/(1+rr), test_retest = scor(rw$w1, rw$w2)) }
  bind_rows(one("obs_p"), one("theta_irf")) |> mutate(short = acc_tasks$short[acc_tasks$task_id == tid])
})
theta_cmp |> group_by(baseline) |>
  summarise(`mean |confound w/ θ|` = mean(abs(conf_theta), na.rm = TRUE),
            `mean reliability` = mean(reliability, na.rm = TRUE),
            `mean test–retest` = mean(test_retest, na.rm = TRUE), .groups = "drop") |>
  mutate(baseline = recode(baseline, obs_p = "observed p", theta_irf = "theta-IRF predicted P")) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(caption = "De-confounding baseline: observed p vs theta-IRF — neither leaves reliable signal")
```

It does **not** rescue the index. The theta-IRF baseline is, if anything, *more*
confounded (≈ −0.2 vs +0.1) and has **negative** internal reliability (noise),
with only a marginal, untrustworthy uptick in test–retest on a near-zero-reliability
base — most likely because IRF miscalibration on some tasks (e.g. Same-&-Different,
ROAR-Word) leaks ability back into the residual. The conclusion is therefore
**robust to the de-confounding method**: by either a person's own accuracy or a
model-based θ̂ anchor, there is no stable accuracy-instability signal beyond ability.

# 3. Reliability vs trial count

```{r rel-ntrials, fig.height=4.3}
medn <- idx_xs |> group_by(short) |> summarise(median_trials = median(ntrial), .groups = "drop")
rel_tab |> filter(variant %in% c("raw","excess")) |> left_join(medn, by = "short") |>
  ggplot(aes(median_trials, reliability, color = variant)) +
  geom_hline(yintercept = 0, linetype = 2, color = "grey70") +
  geom_point(size = 2.3) + scale_x_continuous(trans = "log10") +
  scale_color_manual(values = c(raw = "#cb4b16", excess = "#268bd2"), name = NULL) +
  labs(x = "Median trials per run (log)", y = "internal reliability",
       title = "More trials buys reliability for raw MSSD (= accuracy), not for the de-confounded index",
       subtitle = "excess-over-independence stays near zero regardless of trial count") +
  theme(legend.position = "bottom")
```

For raw MSSD, reliability climbs with trial count (it inherits the reliability of
the accuracy estimate). For the de-confounded **excess** index, more trials do
**not** help — there is little true signal to estimate.

# 4. Cross-task structure

```{r crosstask}
mean_off <- function(cm) mean(cm[upper.tri(cm)], na.rm = TRUE)
wide_v <- function(v) idx_xs |> select(user_id, short, val = all_of(v)) |>
  distinct(user_id, short, .keep_all = TRUE) |> pivot_wider(names_from = short, values_from = val)
cmr <- cor(wide_v("raw") |> select(-user_id), use = "pairwise")
cme <- cor(wide_v("excess") |> select(-user_id), use = "pairwise")
tibble(structure = c("raw MSSD (≈ accuracy)","excess-over-independence"),
       `mean cross-task r` = round(c(mean_off(cmr), mean_off(cme)), 3)) |>
  knitr::kable(caption = "Raw MSSD coheres across tasks (it is ability/g); de-confounded does not")
```

Raw accuracy-MSSD shows substantial cross-task coherence — but that is just the
**positive manifold of ability** showing through (each task's flip rate tracks
its accuracy, and accuracies are intercorrelated). The de-confounded index shows
near-zero cross-task structure, so there is no "accuracy-instability factor"
analogous to the speeded RT-IIV factor.

# 5. Site and age

```{r site-age, fig.height=4.3}
sa <- idx_xs |> left_join(sc |> distinct(run_id, dataset), by = "run_id") |>
  group_by(short) |> mutate(raw_z = as.numeric(scale(raw)), excess_z = as.numeric(scale(excess))) |> ungroup()
bind_rows(sa |> transmute(age, value = raw_z, index = "raw MSSD (≈ accuracy)"),
          sa |> transmute(age, value = excess_z, index = "excess-over-independence")) |>
  ggplot(aes(age, value, color = index)) +
  geom_smooth(method = "loess", se = TRUE, linewidth = 0.9) +
  scale_color_manual(values = c("raw MSSD (≈ accuracy)" = "#cb4b16", "excess-over-independence" = "#268bd2"), name = NULL) +
  labs(x = "Age (years)", y = "Accuracy-variability (within-task z)",
       title = "Raw MSSD falls with age (mirroring rising accuracy); excess is flat",
       subtitle = "The only developmental signal in accuracy variability is the accuracy increase itself") +
  theme(legend.position = "bottom")
```

```{r save}
write_rds(list(confound = conf, perf = perf, cor_raw = cmr, cor_excess = cme),
          here::here("data/accuracy_variability_08.rds"))
```

Raw MSSD declines with age — but only because accuracy rises with age (so flips
fall toward ceiling); the de-confounded index is flat. There is no developmental
signal in accuracy *instability* once accuracy level is removed.

# Synthesis

- **Accuracy variability is essentially ability in disguise.** Raw and detrended
  MSSD are reliable, stable, age-graded, and cross-task coherent — but every one
  of those properties is inherited from **accuracy level** (confound ≈ −0.7; the
  flip rate is mechanically 2·p·(1−p)). Detrending the trial-order trend does not
  touch this.
- **Nothing reliable survives de-confounding.** The excess-over-independence
  index (≡ lag-1 autocorrelation) cleanly removes the accuracy confound, but its
  reliability (≈ 0.03) and longitudinal stability (≈ 0.02) collapse to zero, and
  it shows no cross-task or age structure. There is no stable individual-difference
  trait of "accuracy instability beyond accuracy level" at these ages.
- **This parallels `06` and contrasts `07`.** It is the model-free confirmation
  of `06`'s person-misfit result (de-confounded accuracy "surprise" ≈ 0 reliable
  signal). And it dissociates from RT: in `07`, RT variability retained a real
  (if narrow, speeded-task) signal *beyond* mean RT, whereas accuracy variability
  retains nothing beyond accuracy level.
- **Practical recommendation.** Do **not** treat accuracy MSSD as a variability
  measure — used raw it is an (inverse) ability score, and de-confounded it is
  noise. If within-child variability is the target, the leverage is in **RT** on
  speeded, high-trial tasks (`07`), not accuracy. (Faint exception: detrended
  accuracy autocorrelation on Hearts & Flowers shows a sliver of reliable, stable
  signal — consistent with H&F being the cleanest attentional-lapse task — but it
  is too weak to build on.)
