---
title: "09 — LEVANTE network: enrollment ages & longitudinal coverage"
levante:
  summary: "Planning view of the whole LEVANTE network (not analyzed data): enrollment ages and projected longitudinal age coverage for all 23 funded grants, transcribed from the Q1 2026 Advisory Committee deck."
  status: draft
  task_ids: []
  datasets: []
  data: []
  findings:
    - "Projected coverage is densest at ages 6–11 (peak ~9) and thin at 2–5 and 13+."
    - "Upper tails are upper bounds: accelerated-longitudinal sites are projected per year-cohort, which overstates their upper extent."
  last_run: 2026-07-13
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 10
    fig-height: 7
execute:
  warning: false
  message: false
---

{{< include _disclaimer.qmd >}}

# Purpose

A planning view of the **whole LEVANTE network** — pilots, RfP1, and RfP2 — not
the analyzed data. We transcribe each funded site's main PI, data-collection
country, recruitment setting, target age range, and planned wave schedule from
the Q1 2026 Advisory Committee deck (`papers/Q1 LEVANTE Advisory committee
meeting.pdf`), write it to a CSV, and then visualize (1) the **enrollment ages**
each study recruits and (2) the **eventual age coverage** the accumulated waves
will produce.

```{r setup}
source(here::here("common.R"))
library(tidyverse)
```

# The site table

Transcribed from the deck's per-cohort tables and per-site annex cards.
`age_min`/`age_max` are the recruitment age band; `n_waves`/`cadence` are the
planned wave schedule; `school_based` marks school/classroom/daycare recruitment
(vs. home, lab, online, or community-cohort). See `notes` for deck
inconsistencies and the few cases where the cohort table and annex card disagree.

```{r site-data}
sites <- tribble(
  ~cohort,       ~pi,                ~institution,                          ~country,                       ~age_min, ~age_max, ~school_based, ~longitudinal, ~n_waves, ~cadence,    ~setting,            ~notes,
  "Pilots '23",  "Manuel Bohn",      "Leuphana University Lüneburg",        "Germany",                      5, 12, FALSE, TRUE,  3,  "annual",    "online / family",   "Psychometric pilot; >=3 testing sessions.",
  "Pilots '23",  "Julián Mariño",    "Universidad de los Andes",            "Colombia",                     5, 12, TRUE,  TRUE,  3,  "annual",    "school",            "Psychometric pilot; >=3 testing sessions.",
  "Pilots '23",  "Daniel Ansari",    "Western University",                  "Canada",                       5, 12, TRUE,  TRUE,  3,  "annual",    "school",            "Psychometric pilot; >=3 testing sessions.",
  "Pilots '23",  "Ben Domingue",     "Stanford University",                 "(secondary analysis)",         NA, NA, NA,    NA,    NA, "none",      "secondary analysis","Analyzes other sites' pilot data; no recruitment.",
  "Pilots '25",  "Manuel Bohn",      "Leuphana University Lüneburg",        "Germany",                      2, 5,  FALSE, FALSE, 1,  "none",      "home / everyday",   "Vest-camera recordings; per MZF not longitudinal (deck cited 2 waves).",
  "Pilots '25",  "Charles Nelson",   "Boston Children's Hospital",          "United States",                2, 5,  FALSE, FALSE, 1,  "none",      "lab",               "Per MZF not longitudinal (deck cited lab visits + 1-yr follow-up).",
  "Pilots '25",  "Tami Katzir",      "University of Haifa",                 "Israel",                       5, 10, TRUE,  TRUE,  3,  "annual",    "school",            "K/G1/G2 cohorts; three annual waves.",
  "Pilots '25",  "Julián Mariño",    "Universidad de los Andes",            "Colombia",                     2, 5,  TRUE,  FALSE, 1,  "none",      "daycare",           "Per MZF not longitudinal (deck cited 3 annual waves 2026-28).",
  "Pilots '25",  "Ivy Nkrumah",      "University of Cape Coast",            "Ghana",                        5, 12, TRUE,  TRUE,  3,  "annual",    "school",            "With C. Y. Kwaah; three grade-cohorts, annual.",
  "RfP1",        "Alicia Forsberg",  "University of Sheffield",             "UK",                           6, 9,  TRUE,  TRUE,  3,  "annual",    "school / online",   "Annex card lists 5-13; LEVANTE study 6-9 (used). Three annual waves.",
  "RfP1",        "Catherine Lebel",  "University of Calgary",               "Canada",                       5, 6,  FALSE, TRUE,  4,  "annual",    "home",              "PdP cohort; added at 5-6, annual to age 9. Annex lists U Alberta.",
  "RfP1",        "Rogier Kievit",    "Donders Institute (Radboud)",         "Netherlands",                  5, 9,  FALSE, TRUE,  3,  "annual",    "lab / center",      "Two discrete cohorts (SMILEY ~5, Skippy ~9); 3 annual occasions.",
  "RfP1",        "Zoe Ngo",          "Max Planck Inst. Human Development",  "Germany",                      4, 10, FALSE, TRUE,  4,  "15-month",  "lab",               "Two discrete cohorts (4 & 6); 4 time points 15 months apart.",
  "RfP1",        "Iliana Karipidis", "University of Zurich",                "Switzerland",                  5, 12, TRUE,  TRUE,  3,  "annual",    "school",            "Mono/multilingual; annual over ~3 years.",
  "RfP1",        "Cecilia Calero",   "Universidad Torcuato Di Tella",       "Argentina",                    6, 9,  TRUE,  TRUE,  3,  "annual",    "school",            "Cohorts 6-7 & 8-9, three waves each.",
  "RfP2",        "Diego Lucumí",     "Universidad de los Andes",            "Colombia",                     6, 11, TRUE,  TRUE,  3,  "annual",    "school / community","Pacific-coast / PDET municipalities; three waves.",
  "RfP2",        "Dragan Gasevic",   "Monash University",                   "Colombia",                     9, 12, TRUE,  TRUE,  5,  "biannual",  "school",            "Start age 9-10; data every 6 months (deck cites up to 4 yrs).",
  "RfP2",        "Santiago Cueto",   "GRADE",                              "Peru",                         8, 11, TRUE,  TRUE,  4,  "annual",    "school",            "1500 students / 50 schools; 4 measurement points 2026-2029.",
  "RfP2",        "Kwaku Poku Asante","Kintampo Health Research Center",     "Ghana",                        10, 12, FALSE, TRUE,  3,  "annual",    "community cohort",  "GRAPHS rural cohort; annual yrs 1-3.",
  "RfP2",        "Luciano Gasser",   "University of Teacher Education Bern", "Switzerland",                  9, 12, TRUE,  TRUE,  4,  "annual",    "school",            "BELIMA/SWING; four annual waves (T1-T4). Table says 'first graders'; card age 9-12 used.",
  "RfP2",        "Dana McCoy",       "Harvard University",                  "Peru",                         6, 8,  TRUE,  TRUE,  3,  "annual",    "school / home",     "Rural Andean RCT follow-up; LEVANTE waves at ages 6,7,8.",
  "RfP2",        "Debarati Mukherjee","Public Health Foundation of India",  "India",                        6, 10, TRUE,  TRUE,  3,  "annual",    "school / community","MAASTHI cohort; waves 6-10, 7-11, 8-12.",
  "RfP2",        "Laura Di Giunta",  "Sapienza University of Rome",         "Colombia/Ecuador/Mexico/CR",   6, 12, TRUE,  TRUE,  3,  "annual",    "school",            "Arukay, 4 countries; baseline + annual follow-ups. Deck inconsistent re Costa Rica vs Dominican Republic."
) |>
  mutate(cohort = factor(cohort, c("Pilots '23","Pilots '25","RfP1","RfP2")),
         label  = paste0(pi, " · ", cohort),
         step   = recode(cadence, annual = 1, biannual = 0.5, `15-month` = 1.25, none = NA_real_))

write_csv(sites |> select(-step, -label), here::here("09_site_coverage.csv"))
sites |> select(cohort, pi, institution, country, age_min, age_max, school_based, n_waves, cadence) |>
  knitr::kable(caption = "LEVANTE funded sites (23 grants across pilots, RfP1, RfP2)")
```

```{r quick-counts}
sites |> summarise(grants = n(),
                   recruiting = sum(!is.na(age_min)),
                   longitudinal = sum(longitudinal, na.rm = TRUE),
                   school_based = sum(school_based, na.rm = TRUE),
                   countries = n_distinct(country[!is.na(age_min)])) |>
  knitr::kable(caption = "Network at a glance")
```

The network spans **`r nrow(sites)` grants**; `r sum(!is.na(sites$age_min))`
recruit new participants (Stanford does secondary analysis only), of which
**`r sum(sites$longitudinal, na.rm=TRUE)`** are longitudinal and
**`r sum(sites$school_based, na.rm=TRUE)`** use school / classroom / daycare
recruitment. The figures below drop the secondary-analysis site.

```{r enroll-ages}
# Enrollment ages recruited at baseline. Two sites recruit discrete (gapped)
# cohorts rather than a continuous band; everything else recruits across its band.
enroll_ages <- function(pi, a, b) {
  if (pi == "Rogier Kievit") c(5, 9)
  else if (pi == "Zoe Ngo")  c(4, 6)
  else a:b
}

coh <- sites |> filter(!is.na(age_min)) |>
  mutate(enroll = pmap(list(pi, age_min, age_max), enroll_ages)) |>
  select(cohort, pi, label, n_waves, step, enroll) |>
  unnest(enroll) |>
  rename(enroll_age = enroll)
```

# 1. Enrollment ages, by study

A filled square means the study recruits children of that age at baseline.
Studies are grouped by cohort and ordered by youngest age. (Kievit and Ngo
recruit two discrete cohorts, so their intermediate ages are blank.)

```{r fig-enroll, fig.height=7}
cohort_pal <- c("Pilots '23"="#b58900","Pilots '25"="#cb4b16","RfP1"="#268bd2","RfP2"="#2aa198")
ord <- sites |> filter(!is.na(age_min)) |> arrange(cohort, age_min, age_max) |> pull(label)

coh |> mutate(label = factor(label, levels = rev(ord))) |>
  ggplot(aes(enroll_age, label, fill = cohort)) +
  geom_tile(color = "white", width = 0.92, height = 0.82) +
  scale_fill_manual(values = cohort_pal, name = NULL) +
  scale_x_continuous(breaks = 0:13, expand = expansion(add = 0.5)) +
  labs(x = "Age (years)", y = NULL,
       title = "Enrollment ages recruited by each LEVANTE study",
       subtitle = "Filled = age recruited at baseline; colour = cohort") +
  theme(legend.position = "top", panel.grid.major.y = element_blank(),
        axis.text.y = element_text(size = 8))
```

The pilots, RfP1, and RfP2 concentrate in the **school-age 5–12** band, with the
2025 pilots (Bohn, Nelson, Mariño) and RfP1 Lebel anchoring **early childhood
(2–6)** — matching the deck's stated RfP3 priorities (transition to school;
downward extension to 2–5).

# 2. Longitudinal coverage from the planned waves

Each recruited year-cohort is followed forward by its site's planned waves
(dots = waves; cadence: annual, biannual, or 15-month). Within each site the
year-cohorts are stacked, so the **vertical thickness shows how the overlapping
waves accumulate data at each age** — and the histogram below is the marginal of
exactly this density.

```{r fig-coverage, fig.height=10}
site_order <- sites |> filter(!is.na(age_min)) |>
  arrange(cohort, age_min, age_max) |>
  mutate(base_y = n() - row_number() + 1) |>
  select(label, base_y)

coh2 <- coh |>
  arrange(label, enroll_age) |>
  group_by(label) |>
  mutate(k = n(), j = row_number(),
         offset = if_else(k == 1, 0, (j - (k + 1) / 2) / (k - 1) * 0.84)) |>
  ungroup() |>
  left_join(site_order, by = "label") |>
  mutate(y = base_y + offset,
         x_start = enroll_age,
         x_end = enroll_age + if_else(n_waves <= 1 | is.na(step), 0, (n_waves - 1) * step))

dots <- coh2 |>
  mutate(wv = map2(n_waves, step, ~ if (is.na(.x) || .x <= 1 || is.na(.y)) 0L else 0:(.x - 1))) |>
  unnest(wv) |>
  mutate(wave_age = enroll_age + if_else(is.na(step), 0, wv * step))

ggplot() +
  geom_segment(data = coh2, aes(x_start, y, xend = x_end, yend = y, color = cohort),
               linewidth = 0.5, alpha = 0.9) +
  geom_point(data = dots, aes(wave_age, y, color = cohort), size = 0.7) +
  scale_color_manual(values = cohort_pal, name = NULL) +
  scale_y_continuous(breaks = site_order$base_y, labels = site_order$label,
                     expand = expansion(add = 0.7)) +
  scale_x_continuous(breaks = 0:16, expand = expansion(add = 0.3)) +
  labs(x = "Age (years)", y = NULL,
       title = "Longitudinal coverage: each year-cohort followed by its planned waves",
       subtitle = "One line per recruited year-cohort; dots = waves; stacked within site") +
  theme(legend.position = "top", panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank(), axis.text.y = element_text(size = 7))
```

```{r coverage-density, fig.height=3.4}
coh2 |>
  reframe(age = seq(floor(x_start), floor(x_end)), .by = c(label, enroll_age, x_start, x_end)) |>
  filter(age >= x_start, age <= x_end) |>
  count(age, name = "n_cohorts") |>
  ggplot(aes(age, n_cohorts)) +
  geom_col(fill = "#268bd2") +
  scale_x_continuous(breaks = 0:16) +
  labs(x = "Age (years)", y = "# year-cohorts covering age",
       title = "Marginal coverage density (sum of the stacked lines above)")
```

The accumulated waves cover roughly **ages 2–15**, with the **densest coverage in
the 6–11 school-age range**, where the pilots, RfP1, and RfP2 longitudinal sites
pile up. Coverage **thins at the early-childhood end (2–5)** — carried by only the
2025 pilots and Lebel — and at the **older end (13+)**, which rests on a handful
of cohorts aging out of the school-age sites. This is the empirical case for
RfP3's downward-extension priority.

# Notes & caveats

- **Source:** the Q1 2026 Advisory Committee deck (gitignored under `papers/`).
  Ages are the annex cards' "Age range"; wave schedules are the cohort tables'
  "Timeline" column. Where they disagreed we used the more specific value and
  flagged it in `notes` (Forsberg, Gasser, Lebel, Calero).
- **Coverage model.** Each recruited year-cohort is projected forward by the
  site's planned `n_waves` at its `cadence`. For accelerated-longitudinal sites
  whose *documented* age range already spans some follow-up years, this projects
  each year in the band forward independently and so **over-states the upper
  extent** — read the right-hand tails as an upper bound, not a commitment.
- **Not longitudinal** (per MZF): the three young 2025 pilots (Nelson,
  Mariño-2025, Bohn-2025) are shown as single baseline dots, not lines, even
  though the deck text lists follow-up waves for them.
- **Discrete cohorts:** Kievit (~5 & ~9) and Ngo (4 & 6) recruit two separated
  cohorts; their in-between ages are intentionally blank.
- The CSV is written to `09_site_coverage.csv` for reuse.
