---
title: "10 — Downward extension (Boston Children's, ages 2–5): first look at the raw data"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 10
    fig-height: 6
execute:
  warning: false
  message: false
---

{{< include _disclaimer.qmd >}}

# Purpose

The **very first tranche** of downward-extension data — Boston Children's
Hospital (Nelson site), children aged **2–5**, piloting the LEVANTE core tasks
below the 5–12 range they were built for. There are **no validated scoring
models** for this age band yet, so this is the **raw** per-site dataset
(`pilot_bostonchildrens_us_main_raw:7e6c`, pinned to **v4_26**), not
`levante_data_latest`.

Two aims: **(A) hand-score** the tasks from trial-level correctness with simple,
transparent rules; **(B)** describe **accuracy, time in task, and trials
completed** across tasks and ages. Along the way we log data-quality flags —
this is a pilot, so what breaks is as informative as what works.

```{r setup}
source(here::here("common.R"))
library(tidyverse)
library(patchwork)

DS_REF <- "pilot_bostonchildrens_us_main_raw:7e6c"
DS_VER <- "v4_26"

# Raw tables are cached to data/ (git-ignored); pull via levante::get_raw_table()
# if a cache is missing. Pinned to DS_VER for reproducibility.
load_raw <- function(table) {
  path <- here("data", glue("downex_boston_{table}_{DS_VER}.rds"))
  if (!file_exists(path)) {
    dir_create(here("data"))
    write_rds(levante::get_raw_table(table, data_source = DS_REF,
                                     data_source_version = DS_VER), path)
  }
  read_rds(path)
}
users    <- load_raw("users")
runs     <- load_raw("runs")     |> filter(task_id != "schema_row")
trials   <- load_raw("trials")   |> filter(task_id != "schema_row")
variants <- load_raw("variants") |> filter(task_id != "schema_row")

task_labels <- c(`hearts-and-flowers` = "Hearts & Flowers", `memory-game` = "Memory",
                 `same-different-selection` = "Same & Different", `matrix-reasoning` = "Pattern Matching",
                 `mental-rotation` = "Shape Rotation", `egma-math` = "Math", vocab = "Vocabulary",
                 trog = "Sentence Underst.", pa = "ROAR-Phoneme")
```

# 1. Who and what: the sample and the runs

Ages come from `birth_month`/`birth_year` and each run's `time_started`
(month precision). Only children (`user_type == "student"`) are kept.

```{r sample}
kids <- users |> filter(user_type == "student") |>
  select(user_id, birth_month, birth_year, sex)

runs <- runs |>
  inner_join(kids, by = "user_id") |>
  mutate(task = task_labels[task_id],
         age = (year(time_started) - birth_year) + (month(time_started) - birth_month) / 12,
         age_yr = floor(age),
         # time in task: run duration if finished, else first→last trial timestamp
         dur_min = as.numeric(difftime(time_finished, time_started, units = "mins")))

trial_span <- trials |> group_by(run_id) |>
  summarise(span_min = as.numeric(difftime(max(server_timestamp), min(server_timestamp), units = "mins")), .groups = "drop")
runs <- runs |> left_join(trial_span, by = "run_id") |>
  mutate(dur_min = coalesce(dur_min, span_min))

kids_age <- runs |> group_by(user_id, sex) |> summarise(age = min(age), .groups = "drop") |>
  mutate(age_yr = floor(age))
kids_age |> count(age_yr, sex) |> pivot_wider(names_from = sex, values_from = n, values_fill = 0) |>
  mutate(total = rowSums(across(-age_yr))) |>
  knitr::kable(caption = glue("Children by age (years) at first run — N = {nrow(kids_age)}"))
```

```{r runs-by-task}
runs |> group_by(task) |>
  summarise(runs = n(), children = n_distinct(user_id), completed = sum(completed),
            valid = sum(valid_run), adaptive = any(variant_id %in% variants$variant_id[variants$adaptive]),
            .groups = "drop") |>
  arrange(desc(runs)) |>
  knitr::kable(caption = "Runs per task (completed = dashboard flag; valid = passes the pipeline's run validation)")
```

```{r validity}
runs |> count(completed, valid_run, validation_msg_run) |> arrange(desc(n)) |>
  knitr::kable(caption = "Run validity flags. `less_than_10_test_trials` is the pipeline's minimum-length rule.")
```

Most runs are complete and valid, but a sizeable minority end with **fewer than
10 test trials** — the pipeline's validity floor. Where in the age range those
short runs sit is one of the first things to check (§3).

# 2. Trial-level prep and hand-scoring rules

We keep **test trials only** (`is_practice_trial == FALSE`), and apply
minimal, task-specific rules that mirror what `levantemodels::recode_trials()`
does for the released data:

- **Hearts & Flowers**: a response is correct only if **200 ms ≤ RT ≤ 2000 ms**
  (too-fast/too-slow → incorrect); we report raw accuracy too. Blocks are
  `hearts` / `flowers` / `hearts and flowers` (mixed).
- **Vocabulary**: `catch` trials (attention checks) are scored separately from
  `test` trials.
- **Memory (Corsi)**: span score = the **longest sequence length answered
  correctly**, forward and backward.
- **Math**: accuracy overall and by subtask (`corpus_trial_type`).
- **Chance** per trial is derived from the recorded `distractors` field
  (chance = 1 / (n distractors + 1)); Hearts & Flowers is 2-AFC by design; Memory
  is free recall (chance 0); ROAR-Phoneme records no distractors (chance left
  NA).
- Everything else: proportion correct on test trials.

```{r trial-prep}
test <- trials |>
  filter(!is_practice_trial %in% TRUE) |>
  inner_join(runs |> select(run_id, user_id, task, age, age_yr, valid_run), by = c("run_id", "user_id")) |>
  mutate(n_distr = if_else(is.na(distractors) | distractors %in% c("", "{}"), NA_integer_,
                           str_count(distractors, ",") + 1L),
         chance = case_when(task_id == "hearts-and-flowers" ~ 0.5,
                            task_id == "memory-game" ~ 0,
                            !is.na(n_distr) ~ 1 / (n_distr + 1),
                            TRUE ~ NA_real_),
         # H&F RT recode (mirrors levantemodels::recode_hf)
         correct_hand = if_else(task_id == "hearts-and-flowers",
                                correct & !is.na(rt_numeric) & rt_numeric >= 200 & rt_numeric <= 2000,
                                correct))

test |> group_by(task) |>
  summarise(test_trials = n(), runs = n_distinct(run_id),
            chance = paste(sort(unique(round(chance, 2))), collapse = "/"),
            rt_missing = mean(is.na(rt_numeric)), .groups = "drop") |>
  knitr::kable(digits = 2, caption = "Test trials per task, implied chance level(s), and RT missingness (Memory records no RT)")
```

```{r run-summary}
run_sum <- test |> group_by(run_id, user_id, task, task_id, age, age_yr, valid_run) |>
  summarise(n_trials = n(), acc = mean(correct_hand), acc_raw = mean(correct),
            chance = mean(chance), med_rt = median(rt_numeric, na.rm = TRUE) / 1000,
            .groups = "drop") |>
  left_join(runs |> select(run_id, dur_min, completed, num_attempted, num_correct), by = "run_id")
```

# 3. Basic descriptives across tasks and ages

Each point is one run; lines are loess smooths (age is continuous, 2.0–5.8).

## 3.1 Trials completed

```{r fig-trials, fig.height=7}
ggplot(run_sum, aes(age, n_trials)) +
  geom_point(aes(colour = valid_run), alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE, colour = "grey30", span = 1) +
  facet_wrap(~task, scales = "free_y") +
  scale_colour_manual(values = c(`TRUE` = "#268bd2", `FALSE` = "#dc322f"), name = "valid run") +
  labs(x = "Age (years)", y = "Test trials completed",
       title = "Trials completed per run, by age") +
  theme(legend.position = "top")
```

## 3.2 Time in task

```{r fig-time, fig.height=7}
ggplot(run_sum |> filter(!is.na(dur_min)), aes(age, dur_min)) +
  geom_point(aes(colour = valid_run), alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE, colour = "grey30", span = 1) +
  facet_wrap(~task, scales = "free_y") +
  scale_colour_manual(values = c(`TRUE` = "#268bd2", `FALSE` = "#dc322f"), name = "valid run") +
  labs(x = "Age (years)", y = "Minutes in task",
       title = "Time in task per run, by age",
       subtitle = "Run duration (time_finished − time_started); first→last trial span for unfinished runs") +
  theme(legend.position = "top")
```

## 3.3 Accuracy

Dashed line = chance (task-average of the per-trial chance; Math and Same &
Different mix response formats).

```{r fig-acc, fig.height=7}
chance_df <- run_sum |> group_by(task) |> summarise(chance = mean(chance, na.rm = TRUE), .groups = "drop")
ggplot(run_sum, aes(age, acc)) +
  geom_hline(data = chance_df, aes(yintercept = chance), linetype = "dashed", colour = "grey50") +
  geom_point(aes(colour = valid_run, size = n_trials), alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE, colour = "grey30", span = 1) +
  facet_wrap(~task) +
  scale_colour_manual(values = c(`TRUE` = "#268bd2", `FALSE` = "#dc322f"), name = "valid run") +
  scale_size_area(max_size = 4, name = "test trials") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Age (years)", y = "Proportion correct (hand-scored)",
       title = "Accuracy per run, by age") +
  theme(legend.position = "top")
```

## 3.4 Summary table by task × age year

```{r table-by-age}
run_sum |> group_by(task, age_yr) |>
  summarise(runs = n(), pct_valid = mean(valid_run), trials = median(n_trials),
            minutes = median(dur_min, na.rm = TRUE), acc = mean(acc), .groups = "drop") |>
  mutate(across(c(pct_valid, acc), ~ round(.x, 2)), minutes = round(minutes, 1)) |>
  pivot_wider(names_from = age_yr, values_from = c(runs, pct_valid, trials, minutes, acc),
              names_glue = "{.value}_{age_yr}y") |>
  select(task, starts_with("runs"), starts_with("pct_valid"), starts_with("trials"),
         starts_with("minutes"), starts_with("acc")) |>
  knitr::kable(caption = "Per task × age-year: runs, share valid, median trials, median minutes, mean accuracy")
```

# 4. Hand scores per task

## 4.1 Sanity check: hand accuracy vs the dashboard's own count

The `runs` table carries the dashboard's `num_correct`/`num_attempted`. Our
hand-scored accuracy should agree except where our rules deliberately differ
(H&F RT recode; vocab catch trials excluded).

```{r dashboard-check}
run_sum |> filter(!is.na(num_attempted), num_attempted > 0) |>
  mutate(dash_acc = num_correct / num_attempted) |>
  group_by(task) |>
  summarise(runs = n(), r = cor(acc_raw, dash_acc, use = "complete.obs"),
            mean_diff = mean(acc_raw - dash_acc, na.rm = TRUE),
            trials_match = mean(n_trials == num_attempted), .groups = "drop") |>
  knitr::kable(digits = 2, caption = "Raw hand accuracy vs dashboard num_correct/num_attempted (r, mean difference, share of runs with identical trial counts)")
```

Where the two are directly comparable (Shape Rotation, ROAR-Phoneme, Sentence
Understanding, Vocabulary) the hand score reproduces the dashboard count
**exactly** (r = 1, identical trial counts). Memory and Same & Different count
"trials" differently (sequences / multi-response stages), so they don't match
by construction; the remaining tasks have some runs with no dashboard count.

## 4.2 Task-specific scores

```{r hand-scores}
# Vocabulary: test vs catch
vocab_h <- test |> filter(task_id == "vocab") |>
  group_by(run_id, user_id, age, kind = corpus_trial_type) |>
  summarise(n = n(), acc = mean(correct), .groups = "drop")

# Hearts & Flowers: raw vs RT-recoded, by block
hf_h <- test |> filter(task_id == "hearts-and-flowers") |>
  group_by(run_id, user_id, age, block = corpus_trial_type) |>
  summarise(n = n(), acc_raw = mean(correct), acc_rt = mean(correct_hand),
            med_rt = median(rt_numeric, na.rm = TRUE), .groups = "drop") |>
  mutate(block = factor(block, c("hearts", "flowers", "hearts and flowers")))

# Memory: longest correct span, forward / backward
mem_h <- test |> filter(task_id == "memory-game") |>
  separate(corpus_trial_type, into = c("direction", "span"), sep = "-", convert = TRUE) |>
  group_by(run_id, user_id, age, direction) |>
  summarise(n = n(), max_span = if (any(correct)) max(span[correct]) else 0, .groups = "drop")

# Math: by subtask
math_h <- test |> filter(task_id == "egma-math") |>
  group_by(run_id, user_id, age, subtask = corpus_trial_type) |>
  summarise(n = n(), acc = mean(correct), chance = mean(chance), .groups = "drop")

# Same & Different: by trial type
sds_h <- test |> filter(task_id == "same-different-selection") |>
  group_by(run_id, user_id, age, type = corpus_trial_type) |>
  summarise(n = n(), acc = mean(correct), .groups = "drop")
```

```{r fig-hand, fig.height=9}
p_vocab <- ggplot(vocab_h, aes(age, acc, colour = kind)) + geom_point(alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE, span = 1) + geom_hline(yintercept = 0.25, linetype = "dashed", colour = "grey50") +
  ylim(0, 1) + labs(title = "Vocabulary: test vs catch trials", x = "Age", y = "Prop. correct", colour = NULL)
p_hf <- hf_h |> pivot_longer(c(acc_raw, acc_rt), names_to = "scoring", values_to = "acc") |>
  ggplot(aes(age, acc, colour = scoring)) + geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE, span = 1) + geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey50") +
  facet_wrap(~block) + ylim(0, 1) + labs(title = "Hearts & Flowers: raw vs RT-recoded (200–2000 ms), by block", x = "Age", y = "Prop. correct", colour = NULL)
p_mem <- ggplot(mem_h, aes(age, max_span, colour = direction)) + geom_jitter(width = 0.05, height = 0.08, alpha = 0.7) +
  geom_smooth(method = "loess", se = FALSE, span = 1) + labs(title = "Memory: longest correct span", x = "Age", y = "Max span", colour = NULL)
p_math <- ggplot(math_h, aes(age, acc)) + geom_point(alpha = 0.6, aes(size = n)) +
  geom_smooth(method = "loess", se = FALSE, span = 1.2, colour = "grey30") + facet_wrap(~subtask, nrow = 2) +
  ylim(0, 1) + scale_size_area(max_size = 3) + labs(title = "Math: accuracy by subtask", x = "Age", y = "Prop. correct")
(p_vocab + p_mem) / p_hf / p_math + plot_layout(heights = c(1, 1, 1.4))
```

```{r table-hand}
bind_rows(
  vocab_h |> mutate(task = "Vocabulary", part = kind, score = acc),
  hf_h |> mutate(task = "Hearts & Flowers", part = as.character(block), score = acc_rt),
  mem_h |> mutate(task = "Memory (max span)", part = direction, score = max_span),
  math_h |> mutate(task = "Math", part = subtask, score = acc),
  sds_h |> mutate(task = "Same & Different", part = type, score = acc),
  run_sum |> filter(task_id %in% c("matrix-reasoning", "mental-rotation", "trog", "pa")) |>
    mutate(part = "all", score = acc)
) |>
  mutate(age_yr = floor(age)) |>
  group_by(task, part, age_yr) |>
  summarise(n_runs = n(), score = mean(score), .groups = "drop") |>
  mutate(score = round(score, 2)) |>
  pivot_wider(names_from = age_yr, values_from = c(n_runs, score), names_glue = "{.value}_{age_yr}y") |>
  select(task, part, starts_with("n_runs"), starts_with("score")) |>
  knitr::kable(caption = "Hand scores by task component × age year (mean over runs)")
```

## 4.3 The CAT's running θ as an (unvalidated) provisional score

Several tasks ran as **adaptive** variants, so each trial carries the CAT's
running `theta_estimate` on the **5–12 calibrated item bank**. That bank is out
of range for 2–5-year-olds (see the flags below), so this is *not* a score to
use — but its relation to hand accuracy tells us whether the CAT is
discriminating at all down here.

```{r cat-theta}
theta_last <- test |> filter(!is.na(theta_estimate)) |>
  group_by(run_id) |> slice_max(trial_index, n = 1, with_ties = FALSE) |>
  select(run_id, theta_last = theta_estimate) |>
  inner_join(run_sum, by = "run_id") |>
  group_by(task) |> filter(n() >= 5) |> ungroup()   # drop tasks with a handful of CAT runs

ggplot(theta_last, aes(acc, theta_last)) +
  geom_point(aes(colour = age), alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, colour = "grey30") +
  facet_wrap(~task, scales = "free") +
  scale_colour_viridis_c(name = "age") +
  labs(x = "Hand-scored proportion correct", y = "CAT final θ estimate",
       title = "CAT running θ (5–12 item bank) vs hand accuracy, adaptive tasks only")

theta_last |> group_by(task) |>
  summarise(runs = n(), r_theta_acc = cor(theta_last, acc), r_theta_age = cor(theta_last, age),
            mean_theta = mean(theta_last), .groups = "drop") |>
  knitr::kable(digits = 2, caption = "Final CAT θ: agreement with hand accuracy and age, and mean level (on the 5–12 scale)")
```

The CAT does discriminate down here on Math, Shape Rotation, Sentence
Understanding, and Vocabulary (θ–accuracy r ≈ 0.5–0.85), but the θ values are
far below the 5–12 scale (means ≈ −2.7 to −5.2), and a whole row of runs —
mostly the 2–3-year-olds — sits at **exactly θ = −6, the estimator's hard
lower bound**: those are floor codes, not measurements. Same & Different's θ
is unrelated to accuracy (r ≈ 0.1) and many of its runs sit at exactly 0 (the
prior mean, i.e. θ never updated), consistent with its scoring being under
revision.

# 5. Data-quality flags to carry forward

```{r flags}
# How far down the age range does each Math subtask reach, and where is it above chance?
math_h |> group_by(subtask) |>
  summarise(kids = n_distinct(user_id), min_age = min(age), acc_under4 = mean(acc[age < 4]),
            acc_4plus = mean(acc[age >= 4]), chance = mean(chance), .groups = "drop") |>
  arrange(min_age) |>
  knitr::kable(digits = 2, caption = "Math subtasks: youngest child served, and mean accuracy below vs. above age 4 (chance shown). Under 4 nothing reliably clears chance; from 4 on the number subtasks do, while Addition stays at chance throughout")

# Vocabulary catch trials as an engagement gate
vocab_h |> mutate(age_yr = floor(age)) |> group_by(age_yr, kind) |>
  summarise(runs = n(), acc = mean(acc), .groups = "drop") |>
  pivot_wider(names_from = kind, values_from = c(runs, acc)) |>
  knitr::kable(digits = 2, caption = "Vocabulary: catch-trial vs test accuracy by age year")
```

- **Short / invalid runs concentrate at the young end** (see §3.1, §3.4):
  the `<10 test trials` rule removes many 2-year-old runs — for this age band
  the validity floor itself needs re-thinking (a 2-year-old's 8 valid trials
  may be the whole point).
- **2-year-olds are at chance on nearly everything** (Vocabulary, Sentence
  Understanding, Math), and they **fail the Vocabulary catch trials** (≈ 57%
  vs ≥ 95% from age 3) — so much of the 2-year-old data reflects
  non-engagement/incomprehension rather than measured ability. Catch-trial
  accuracy is a usable engagement gate; some 2-year-olds also tap through 60+
  Sentence-Understanding trials at chance (§3.1).
- **CAT item banks are calibrated for 5–12**: the top of the routing behaves
  sensibly (Subtraction / Multiplication reached only 5-year-olds), but the
  bank has **no floor for this band** — Addition items reach 2-year-olds yet
  sit at chance across the whole 2–5 range, and **below age 4 no Math subtask
  reliably clears chance** (the number subtasks do from 4 on). Running θ is
  not interpretable at this age.
- **Memory records no RT** (`rt_numeric` all NA) — known from the main
  pipeline; fine for accuracy/span, no timing possible.
- **Same & Different** staging is irregular in the raw table
  (`first_response`/`second_response`/… plus blank `assessment_stage`), and
  its scoring is under revision anyway (see LEVANTE.md); the SDS numbers here
  are proportion-correct by trial type only.
- **Shape Rotation** trials record **2 distractors (3-AFC)** in this dataset,
  whereas the calibrated models use a 0.5 guessing floor — check which variant
  ran before comparing to 5–12 data.
- `stop_type` is entirely NA in this tranche (no early-stopping info).

# Notes & next steps

- **Data:** `r DS_REF` v`r DS_VER` (raw, per-site). Cached under `data/`
  (git-ignored). Two *other* downward-extension raw datasets exist on Redivis
  — `pilot_langcog_us_downex_raw:a6kb` (Stanford) and
  `partner_sparklab_us_downex_raw:4n9e` — natural comparison sets once the
  Boston picture is clear.
- **This notebook is not wired into the public book** (`_quarto.yml` renders
  `0[0-9]_*.qmd` only): unreleased raw child data with age/sex at N = 31.
  Adding it to the published book is a deliberate decision, not a default.
- **Next:** (i) decide a validity floor appropriate for 2–5; (ii) look at
  practice-trial performance as an engagement gate; (iii) once ≥ 2 sites of
  downward-extension data exist, item-level difficulty for a 2–5 bank; (iv)
  compare with the Stanford / Sparklab downex tranches.
