---
title: "Corrected longitudinal trajectories — math & vocab, CO & DE"
levante:
  summary: "Presentation figures: before/after longitudinal trajectories for Math and Vocabulary at Bogotá and Leipzig, published (v1.0, buggy) vs rescored with the column-order fix."
  status: draft
  task_ids: [egma-math, vocab]
  datasets: [pilot_mpieva_de_main, pilot_uniandes_co_bogota]
  data:
    - {source: "load_levante_scores", version: "v1_0", table: "scores"}
    - {source: "load_levante_trials", version: "v1_0", table: "trials"}
  depends_on: [00_load_data.qmd, tasks/math_eap_vs_ml.qmd, tasks/vocab.qmd]
  findings:
    - "Figures only; written to figures/trajectories_before_after.png and figures/t1_t2_scatter_before_after.png."
  last_run: unknown
format:
  html:
    toc: true
    code-fold: true
    fig-width: 11
    fig-height: 8
execute:
  warning: false
  message: false
---

{{< include ../_disclaimer.qmd >}}

# Purpose

Presentation-ready before/after longitudinal trajectories for the two tasks
and two sites where the column-order scoring bug mattered most. "Published"
= the scores currently in `levante_data_latest` (buggy). "Corrected" = rescored
with the column-order fix, using the same v1_14 IRT models and the same EAP
estimator the pipeline uses (so this previews what the re-run pipeline will
produce; WLE looks essentially identical — see `tasks/math_item_calibration`
and the EAP/WLE checks).

High-res PNGs are written to `figures/` for slides.

```{r setup}
source(here::here("common.R"))
library(levantemodels)
library(mirt)
library(patchwork)

fig_dir <- here::here("figures"); fs::dir_create(fig_dir)

scores <- read_rds(here::here("data", "scores_all_sites.rds"))

cells <- tribble(
  ~task_id,    ~tk,     ~dataset,                   ~site,      ~model_file,
  "egma-math", "math",  "pilot_uniandes_co_bogota", "Colombia", "math_model_record_v1_14.rds",
  "egma-math", "math",  "pilot_mpieva_de_main",     "Germany",  "math_model_record_v1_14.rds",
  "vocab",     "vocab", "pilot_uniandes_co_bogota", "Colombia", "vocab_model_co_v1_14.rds",
  "vocab",     "vocab", "pilot_mpieva_de_main",     "Germany",  "vocab_model_de_v1_14.rds"
)
```

```{r rescore, cache=TRUE}
rescore_cell <- function(task_id, tk, dataset, model_file, ...) {
  mod_rec <- read_rds(here::here("data", model_file))
  tr <- load_levante_trials(task_ids = task_id) |> filter(dataset == !!dataset)
  score_task_irt(tr, tk, dataset, mod_rec, method = "EAP") |>
    transmute(run_id, score_corrected = score)
}

corrected <- cells |>
  mutate(scored = pmap(cells, rescore_cell)) |>
  select(task_id, dataset, scored) |>
  unnest(scored)

# Assemble: published (buggy) + corrected, with age/wave per run
traj <- scores |>
  filter(task_id %in% cells$task_id, dataset %in% cells$dataset) |>
  select(run_id, user_id, age, task_id, task_label, dataset,
         score_published = score) |>
  inner_join(cells |> distinct(task_id, dataset, site),
             by = c("task_id", "dataset")) |>
  left_join(corrected, by = c("run_id", "task_id", "dataset")) |>
  group_by(task_id, dataset, user_id) |>
  arrange(age) |> mutate(wave = row_number()) |> ungroup() |>
  filter(wave <= 2)

# keep only kids with two waves
two_wave <- traj |>
  group_by(task_id, dataset, user_id) |>
  filter(n_distinct(wave) == 2) |> ungroup() |>
  mutate(cell = paste0(task_label, " — ", site))
```

# Before / after trajectories (age on x)

```{r traj-plot, fig.width=11, fig.height=9}
plot_long <- two_wave |>
  pivot_longer(c(score_published, score_corrected),
               names_to = "scoring", values_to = "theta") |>
  mutate(scoring = recode(scoring,
                          score_published = "Published (buggy)",
                          score_corrected = "Corrected") |>
           factor(levels = c("Published (buggy)", "Corrected")),
         theta = pmin(pmax(theta, -6), 6))

p_traj <- ggplot(plot_long, aes(x = age, y = theta)) +
  geom_line(aes(group = user_id), alpha = 0.18, color = "grey30") +
  geom_point(aes(color = factor(wave)), alpha = 0.35, size = 0.8) +
  geom_smooth(method = "lm", se = FALSE, color = "firebrick", linewidth = 1) +
  facet_grid(cell ~ scoring) +
  scale_color_manual(values = c(`1` = "#268bd2", `2` = "#cb4b16"),
                     name = "Wave", labels = c("T1", "T2")) +
  labs(x = "Age (years)", y = "Ability (IRT θ)",
       title = "Longitudinal trajectories: published vs. corrected scoring",
       subtitle = "Each line = one child across two waves; red = population trend") +
  theme(legend.position = "bottom")
p_traj
ggsave(file.path(fig_dir, "trajectories_before_after.png"), p_traj,
       width = 11, height = 9, dpi = 300, bg = "white")
```

# T1 vs T2 scatter (the compression, before/after)

```{r scatter, fig.width=11, fig.height=9}
pairs <- two_wave |>
  select(cell, user_id, wave, score_published, score_corrected) |>
  pivot_longer(c(score_published, score_corrected),
               names_to = "scoring", values_to = "theta") |>
  mutate(scoring = recode(scoring, score_published = "Published (buggy)",
                          score_corrected = "Corrected")) |>
  pivot_wider(names_from = wave, values_from = theta, names_prefix = "t") |>
  mutate(scoring = factor(scoring, levels = c("Published (buggy)", "Corrected")),
         t1 = pmin(pmax(t1, -6), 6), t2 = pmin(pmax(t2, -6), 6))

p_scatter <- ggplot(pairs, aes(x = t1, y = t2)) +
  geom_abline(slope = 1, intercept = 0, lty = 3, color = "grey50") +
  geom_point(alpha = 0.4, color = "#268bd2") +
  geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 0.8) +
  facet_grid(cell ~ scoring) +
  coord_equal() +
  labs(x = "θ at T1", y = "θ at T2",
       title = "T1 → T2 by scoring (dotted = no change; flat slope = spurious compression)",
       subtitle = "Published scoring flattens the slope (apparent regression to mean); corrected restores it") +
  theme(aspect.ratio = 1)
p_scatter
ggsave(file.path(fig_dir, "t1_t2_scatter_before_after.png"), p_scatter,
       width = 11, height = 9, dpi = 300, bg = "white")
```

# Summary table

```{r summary}
pairs |>
  group_by(cell, scoring) |>
  summarise(n = n(),
            mean_change = mean(t2 - t1, na.rm = TRUE),
            galton_slope = coef(lm(t2 ~ t1))[2],
            .groups = "drop") |>
  pivot_wider(names_from = scoring,
              values_from = c(n, mean_change, galton_slope)) |>
  knitr::kable(digits = 2)
```

```{r save-data}
write_rds(two_wave, here::here("data", "corrected_trajectories_long.rds"))
```

Figures saved to `figures/trajectories_before_after.png` and
`figures/t1_t2_scatter_before_after.png`.
