---
title: "Cross-language DIF screen — Math, Pattern Matching, Shape Rotation, H&F, Memory"
levante:
  summary: "QA screen for cross-language DIF on the five multigroup-scalar tasks: per-language Rasch fits (German, Spanish, English), centered and compared item by item."
  status: draft
  task_ids: [egma-math, matrix-reasoning, mental-rotation, hearts-and-flowers, memory-game]
  datasets: all
  data:
    - {source: "load_levante_scores", version: "v1_2", table: "scores"}
    - {source: "load_levante_trials", version: "v1_2", table: "trials"}
  depends_on: [00_load_data.qmd]
  findings:
    - "The five tasks are broadly cross-language invariant; Shape Rotation is near-perfect, Math and Hearts & Flowers lower but item-specific."
    - "No broken items on the scale of the TROG German item; the largest gaps (~1.5–2.2 logits) are worth spot-checking."
    - "Vocabulary and the by-language ROAR tasks still need the anchored multigroup screen."
  last_run: 2026-07-16
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 10
    fig-height: 6
execute:
  warning: false
  message: false
---

{{< include ../_disclaimer.qmd >}}

# Purpose

A QA screen for **differential item functioning across languages** on the five
core tasks calibrated with a **`multigroup_site` scalar** model (so item
difficulties are on a common scale): Math, Pattern Matching (matrix), Shape
Rotation (mental rotation), Hearts & Flowers, Memory. (ToM and TROG were done
separately; the `by_language` tasks — Vocab, ROAR — need the anchored approach
and aren't included here.)

Method (as for TROG): fit a **Rasch model separately per language** on each
child's first session (German = Leipzig, Spanish = Bogotá + rural, English =
Canada), center each language's difficulties, and compare. Large cross-language
gaps flag items that may be mistranslated or mis-keyed.

```{r setup}
source(here::here("common.R"))
library(mirt)
sc <- read_rds(here("data/scores_all_sites.rds"))

lang_of <- function(ds) dplyr::case_when(
  ds == "pilot_mpieva_de_main" ~ "German",
  stringr::str_detect(ds, "uniandes_co") ~ "Spanish",
  ds == "pilot_western_ca_main" ~ "English", TRUE ~ NA_character_)

fit_lang <- function(trd, L) {
  m <- trd |> filter(lang == L) |> distinct(run_id, item_uid, .keep_all = TRUE) |>
    select(run_id, item_uid, correct) |>
    pivot_wider(names_from = item_uid, values_from = correct) |>
    select(-run_id) |> as.matrix()
  m <- m[, colMeans(!is.na(m)) > 0.03, drop = FALSE]
  m <- m[, apply(m, 2, \(x) length(unique(x[!is.na(x)])) > 1), drop = FALSE]
  if (ncol(m) < 5) return(NULL)
  fit <- mirt(m, 1, itemtype = "Rasch", verbose = FALSE, technical = list(NCYCLES = 2000))
  d <- coef(fit, simplify = TRUE)$items[, "d"]
  tibble(item_uid = names(d), b_c = (-d) - mean(-d), lang = L)
}

tasks <- c("egma-math","matrix-reasoning","mental-rotation",
           "hearts-and-flowers","memory-game")
```

```{r run, cache=TRUE}
dif_all <- map_dfr(tasks, function(tk) {
  fr <- sc |> filter(task_id == tk) |> mutate(lang = lang_of(dataset)) |>
    filter(!is.na(lang)) |> group_by(user_id) |> arrange(age) |> slice(1) |>
    ungroup() |> select(run_id, lang)
  trd <- load_levante_trials(task_ids = tk) |> inner_join(fr, by = "run_id") |>
    mutate(correct = as.numeric(correct))
  map_dfr(c("German","Spanish","English"), \(L) fit_lang(trd, L)) |>
    mutate(task = tk)
})
wide <- dif_all |> pivot_wider(names_from = lang, values_from = b_c) |>
  rowwise() |>
  mutate(maxgap = {v <- c_across(any_of(c("German","Spanish","English")));
                   if (sum(!is.na(v)) >= 2) diff(range(v, na.rm = TRUE)) else NA_real_}) |>
  ungroup() |>
  left_join(task_lookup |> select(task_id, task_label), by = c("task" = "task_id"))
```

# Task-level invariance

```{r task-summary}
pair_r <- function(d, a, b) if (all(c(a,b) %in% names(d))) {
  p <- d |> select(all_of(c(a,b))) |> drop_na(); if (nrow(p) > 3) cor(p[[1]], p[[2]]) else NA
} else NA
wide |> group_by(task_label) |>
  group_modify(~ tibble(
    n_items   = sum(!is.na(.x$maxgap)),
    `r DE–ES` = pair_r(.x, "German","Spanish"),
    `r DE–EN` = pair_r(.x, "German","English"),
    `r ES–EN` = pair_r(.x, "Spanish","English"),
    `n items |DIF|>1` = sum(.x$maxgap > 1, na.rm = TRUE))) |>
  ungroup() |> arrange(desc(`r DE–ES`)) |>
  knitr::kable(digits = 2,
               caption = "Cross-language item-difficulty agreement (per-language Rasch)")
```

Mental Rotation is essentially invariant (r ≈ 0.96–0.99 — as expected for a
non-verbal task); Memory and Pattern Matching are high (≈ 0.94–0.95 / mixed);
Math and Hearts & Flowers are a bit lower (≈ 0.81–0.92), reflecting a handful of
items rather than systemic drift. Crucially, none approaches the broken-item
extreme seen in TROG (a 6.5-logit gap).

# Flagged items (largest cross-language gaps)

```{r flagged}
wide |> filter(maxgap > 1) |> arrange(desc(maxgap)) |>
  mutate(across(c(German,Spanish,English,maxgap), \(x) round(x,2))) |>
  select(task_label, item_uid, German, Spanish, English, maxgap) |>
  head(20) |>
  knitr::kable(caption = "Items with > 1-logit max cross-language difficulty gap (centered)")
```

```{r flagged-count}
wide |> group_by(task_label) |>
  summarise(n_items = sum(!is.na(maxgap)),
            n_flagged_gt1 = sum(maxgap > 1, na.rm = TRUE),
            n_flagged_gt1.5 = sum(maxgap > 1.5, na.rm = TRUE),
            max_gap = round(max(maxgap, na.rm = TRUE), 2), .groups = "drop") |>
  knitr::kable(caption = "Flagged-item counts by task")
```

# DIF scatter (German vs Spanish)

```{r scatter, fig.height=6}
wide |> filter(!is.na(German), !is.na(Spanish)) |>
  ggplot(aes(x = Spanish, y = German)) +
  geom_abline(slope = 1, intercept = 0, lty = 3, color = "grey50") +
  geom_point(alpha = 0.5, color = "#268bd2", size = 1) +
  ggrepel::geom_text_repel(data = ~filter(.x, maxgap > 1.3),
    aes(label = str_remove(item_uid, "^(math_|matrix_|mrot_|mg_)")),
    size = 2.2, color = "firebrick", max.overlaps = 12) +
  facet_wrap(~task_label, scales = "free") +
  labs(x = "Item difficulty — Spanish (centered)",
       y = "Item difficulty — German (centered)",
       title = "Cross-language item difficulty: German vs Spanish",
       subtitle = "On-diagonal = invariant; labeled (red) = max pairwise gap > 1.3 logits")
```

# Verdict

- **These five tasks are broadly cross-language invariant.** Mental Rotation is
  near-perfect (non-verbal), Memory and Pattern Matching strong; Math and
  Hearts & Flowers a little lower but driven by specific items, not systemic
  drift.
- **No broken items on the scale of the TROG German embedding item.** The
  largest gaps here (~1.5–2.2 logits) are worth a look but are not catastrophic;
  several are short Memory spans and a few Math items.
- **Hearts & Flowers**: the weakest pair is German–Spanish (r ≈ 0.81), with
  `heart_start` carrying the most DIF — consistent with start trials being
  noisy/low-information (see `tasks/hearts_and_flowers.qmd`).
- **Action:** spot-check the > 1.5-logit items (esp. the top Math and Pattern
  Matching items) for translation/key issues; otherwise these tasks are sound
  for cross-site use. Vocab and the ROAR `by_language` tasks remain to be
  screened with the anchored multigroup method.

```{r save}
write_rds(wide, here::here("data/crosslang_dif_batch.rds"))
```
