---
title: "Math — Bogotá rescoring: a column-order bug, EAP, and ML"
levante:
  summary: "Side-by-side EAP vs ML rescoring of Bogotá Math; surfaced the column-order bug in score_irt (fixed upstream in levantemodels PR #9)."
  status: draft
  task_ids: [egma-math]
  datasets: [pilot_uniandes_co_bogota]
  data:
    - {source: "load_levante_scores", version: "v1_2", table: "scores"}
    - {source: "load_levante_trials", version: "v1_2", table: "trials"}
    - {source: "load_item_parameters", version: "v1_14", table: "item_parameters"}
  depends_on: [00_load_data.qmd]
  findings:
    - "score_irt passed response columns to mirt::fscores() out of model order; fscores matches by position, so θs were wrong."
    - "Fixing column order moves the Bogotá Math Galton slope from 0.19 to 0.81: the apparent compression largely disappears."
    - "EAP vs ML matters little once the bug is fixed (slopes 0.81 vs 0.72)."
  last_run: 2026-07-22
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    fig-width: 11
    fig-height: 5
execute:
  warning: false
  message: false
---

{{< include ../_disclaimer.qmd >}}

# Headline

Mike asked for a side-by-side EAP vs ML rescoring of Bogotá math, expecting
that ML would remove apparent T2 compression due to EAP shrinkage. The
investigation surfaced something more substantive: **`levantemodels::score_irt`
had a column-order bug** that produced wrong θ estimates whenever the
`response.pattern` passed to `mirt::fscores()` had its item columns in a
different order from the model's internal item order. When we fix the
column ordering, the Bogotá math T1 → T2 picture changes from heavy
"regression to the mean" (Galton slope ≈ 0.19) to broad-based growth across
all four T1 quartiles (Galton slope ≈ 0.81). EAP-vs-ML matters much less
than the column fix.

**Update:** this bug was fixed upstream in `levantemodels::score_irt` by
[PR #9](https://github.com/levante-framework/levantemodels/pull/9), so the
installed package no longer reproduces it. To keep the comparison below
intact, we reproduce the pre-fix behavior locally as `score_irt_buggy()`
(defined in `common.R`) — a copy of `score_irt()`'s current logic with the
column-reorder step removed — and compare it against the current, fixed
`levantemodels::score_irt()`. `score_irt()` also now takes a `method`
argument, so EAP and ML scores both come straight from the package rather
than a local reimplementation.

```{r setup}
source(here::here("common.R"))
library(levantemodels)
library(mirt)

scores <- read_rds(here::here("data", "scores_all_sites.rds"))
trials <- load_levante_trials(task_ids = "egma-math")
```

# Demonstrate the bug

`mirt::fscores(..., response.pattern = data)` matches response columns to
model items **by position, not by name**. Same kid, same responses,
reordered columns → wildly different θ.

```{r mirt-bug}
# cached copies of the scoring table + v1_14 math ModelRecord
# (get_model_record's live download path is broken against the current
# redivis package)
spec <- levantemodels:::get_model_spec("math", "pilot_uniandes_co_bogota",
          read_rds(here::here("data/scoring_table_cache.rds")))
mod_rec <- read_rds(here::here("data/math_model_record_v1_14.rds"))

mod_recon <- mirt::multipleGroup(data = mod_rec@data, group = mod_rec@groups,
                                  pars = model_vals(mod_rec), TOL = NaN)
mod_co <- mirt::extract.group(mod_recon, group = "pilot_uniandes_co")

mi <- items(mod_rec)
test_responses <- matrix(NA, 1, length(mi),
                         dimnames = list("TEST", mi))
test_responses[1, 1:30]  <- 1  # correct on first 30 items
test_responses[1, 31:60] <- 0  # incorrect on next 30
test_responses <- as.data.frame(test_responses)

tibble(
  layout = c("Model order", "Reversed", "Shuffled"),
  EAP    = c(
    mirt::fscores(mod_co, method = "EAP",
                  response.pattern = test_responses)[, "F1"],
    mirt::fscores(mod_co, method = "EAP",
                  response.pattern = test_responses[, rev(mi), drop = FALSE])[, "F1"],
    mirt::fscores(mod_co, method = "EAP",
                  response.pattern = test_responses[, sample(mi), drop = FALSE])[, "F1"]
  )
) |> knitr::kable(digits = 3,
                  caption = "Same kid, same response data, three column layouts.")
```

# The bug was in `levantemodels::score_irt`

`score_irt` aligned missing items by appending them to `data_prepped`'s
existing columns, but never reordered to match `items(mod_rec)`:

```{r show-the-bug, eval=FALSE}
# Verbatim from levantemodels::score_irt before PR #9
# (https://github.com/levante-framework/levantemodels/pull/9)
overlap_items <- intersect(colnames(data_prepped), items(mod_rec))
data_aligned  <- select(data_prepped, !!overlap_items)
missing_items <- setdiff(items(mod_rec), colnames(data_prepped))
data_aligned[, missing_items] <- NA
# ↑ at this point, data_aligned columns are
#   [overlap_items_in_data_prepped_order, missing_items_in_setdiff_order]
# which is NOT items(mod_rec) order. The fix, added in PR #9:
data_aligned <- data_aligned[, items(mod_rec)]   # <-- was missing before PR #9
scores <- mirt::fscores(mod, method = "EAP", response.pattern = data_aligned)
```

Our local `score_irt_buggy()` in `common.R` reproduces the pre-fix behavior
(everything above minus that reorder line) so we can still compare against it
now that the installed package is fixed.

# Pipeline vs corrected vs uncorrected: do they agree?

```{r compare-pipelines, cache=TRUE}
co_trials <- trials |> filter(dataset == "pilot_uniandes_co_bogota")
co_recoded <- recode_trials(co_trials)

# 1) Corrected EAP / ML: current (fixed) levantemodels::score_irt
my_eap <- levantemodels::score_irt(co_recoded, as.list(spec), mod_rec, method = "EAP") |>
  transmute(run_id, score = as.numeric(score))
my_ml  <- levantemodels::score_irt(co_recoded, as.list(spec), mod_rec, method = "ML") |>
  transmute(run_id, score = as.numeric(score))

# 2) Buggy: reproduce the pre-PR#9 score_irt() column-order bug locally
# (the installed package no longer has this bug)
official <- score_irt_buggy(co_recoded, as.list(spec), mod_rec) |>
  transmute(run_id, official_score_irt = as.numeric(score))

# 3) Published pipeline scores (from levante_data_latest)
pipeline <- scores |>
  filter(dataset == "pilot_uniandes_co_bogota", task_id == "egma-math") |>
  select(run_id, user_id, age, pipeline_score = score)

all3 <- pipeline |>
  left_join(my_eap |> select(run_id, my_eap = score), by = "run_id") |>
  left_join(my_ml  |> select(run_id, my_ml  = score), by = "run_id") |>
  left_join(official, by = "run_id")
```

```{r pairwise-cor}
all3 |>
  filter(!is.na(my_eap), !is.na(official_score_irt)) |>
  summarise(
    `pipeline ↔ buggy score_irt (reproduced)`    = cor(pipeline_score, official_score_irt),
    `pipeline ↔ corrected EAP`                   = cor(pipeline_score, my_eap),
    `buggy score_irt (reproduced) ↔ corrected EAP` = cor(official_score_irt, my_eap),
    `corrected EAP ↔ corrected ML`                = cor(my_eap, my_ml)
  ) |>
  knitr::kable(digits = 3,
               caption = "Pearson r across scoring approaches")
```

The published pipeline scores match the reproduced buggy `score_irt` (r =
0.95) but disagree with the corrected EAP (r = 0.75) — confirming the
published scores were computed under the same bug (now fixed upstream in
`levantemodels::score_irt`, but not yet reflected in `levante_data_latest`
unless it has been rescored since PR #9).

# Wave-paired rescore data

```{r pairs}
co_meta <- scores |>
  filter(dataset == "pilot_uniandes_co_bogota", task_id == "egma-math") |>
  select(run_id, user_id, age)

paired <- co_meta |>
  left_join(all3 |> select(run_id, pipeline_score, my_eap, my_ml),
            by = "run_id") |>
  group_by(user_id) |>
  arrange(age) |>
  mutate(wave = row_number()) |>
  ungroup() |>
  filter(wave <= 2)

pairs <- paired |>
  group_by(user_id) |>
  filter(n_distinct(wave) == 2) |>
  summarise(
    pipe_t1 = pipeline_score[wave == 1],
    pipe_t2 = pipeline_score[wave == 2],
    eap_t1  = my_eap[wave == 1],
    eap_t2  = my_eap[wave == 2],
    ml_t1   = my_ml[wave == 1],
    ml_t2   = my_ml[wave == 2],
    age_t1  = age[wave == 1],
    age_t2  = age[wave == 2],
    .groups = "drop"
  ) |>
  mutate(t1_quartile = ntile(eap_t1, 4))

cat("Two-wave kids:", nrow(pairs), "\n")
```

# Side-by-side spaghetti

```{r spaghetti, fig.width=12, fig.height=5}
spag_long <- pairs |>
  pivot_longer(c(pipe_t1, pipe_t2, eap_t1, eap_t2, ml_t1, ml_t2),
               names_to = c("estimator", "wave"),
               names_sep = "_",
               values_to = "score") |>
  mutate(estimator = recode(estimator,
                            pipe = "Pipeline (buggy)",
                            eap  = "Corrected EAP",
                            ml   = "Corrected ML"),
         estimator = factor(estimator,
                            levels = c("Pipeline (buggy)",
                                       "Corrected EAP",
                                       "Corrected ML")),
         wave      = recode(wave, t1 = "T1", t2 = "T2"))

spag_clip <- spag_long |> mutate(score = pmin(pmax(score, -6), 6))

ggplot(spag_clip, aes(x = wave, y = score, group = user_id)) +
  geom_line(alpha = 0.25, color = "#268bd2") +
  geom_point(aes(color = wave), alpha = 0.4, size = 1.2) +
  facet_wrap(~estimator) +
  scale_color_manual(values = c(T1 = "#268bd2", T2 = "#cb4b16"),
                     guide = "none") +
  labs(x = NULL, y = "Math θ",
       title = "Bogotá math T1 → T2 spaghetti",
       subtitle = "Same kids, same model, same items. Only the scoring code differs.")
```

# T2 vs T1 scatter

```{r scatter, fig.width=12, fig.height=5}
scatter_data <- bind_rows(
  pairs |> transmute(estimator = "Pipeline (buggy)",
                     t1 = pipe_t1, t2 = pipe_t2, t1_quartile),
  pairs |> transmute(estimator = "Corrected EAP",
                     t1 = eap_t1,  t2 = eap_t2,  t1_quartile),
  pairs |> transmute(estimator = "Corrected ML",
                     t1 = ml_t1,   t2 = ml_t2,   t1_quartile)
) |>
  mutate(estimator = factor(estimator,
                            levels = c("Pipeline (buggy)",
                                       "Corrected EAP",
                                       "Corrected ML")))

scatter_data |>
  ggplot(aes(x = t1, y = t2)) +
  geom_abline(slope = 1, intercept = 0, lty = 3, color = "grey50") +
  geom_point(aes(color = factor(t1_quartile)), alpha = 0.55) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  facet_wrap(~estimator) +
  scale_color_brewer(palette = "RdYlBu", direction = -1,
                     name = "EAP-T1 quartile") +
  coord_equal() +
  labs(x = "T1 score", y = "T2 score",
       title = "T1 vs T2 math scores under three scoring approaches",
       subtitle = "Dotted = identity. Black = OLS fit.") +
  theme(legend.position = "bottom")
```

# Galton slopes & per-quartile changes

```{r galton}
fit_pipe <- lm(pipe_t2 ~ pipe_t1, data = pairs)
fit_eap  <- lm(eap_t2  ~ eap_t1,  data = pairs)
fit_ml   <- lm(ml_t2   ~ ml_t1,   data = pairs)

bind_rows(
  `Pipeline (buggy)` = broom::tidy(fit_pipe, conf.int = TRUE)[2, ],
  `Corrected EAP`    = broom::tidy(fit_eap,  conf.int = TRUE)[2, ],
  `Corrected ML`     = broom::tidy(fit_ml,   conf.int = TRUE)[2, ],
  .id = "estimator"
) |>
  transmute(estimator,
            galton_slope = estimate,
            slope_se     = std.error,
            ci_lo        = conf.low,
            ci_hi        = conf.high) |>
  knitr::kable(digits = 3,
               caption = "Galton slope T2 ~ T1, with 95% CI")
```

```{r quartile-changes}
pairs |>
  group_by(t1_quartile) |>
  summarise(
    n        = n(),
    mean_pipe_t1 = mean(pipe_t1),
    mean_pipe_t2 = mean(pipe_t2),
    delta_pipe   = mean(pipe_t2 - pipe_t1),
    mean_eap_t1  = mean(eap_t1),
    mean_eap_t2  = mean(eap_t2),
    delta_eap    = mean(eap_t2 - eap_t1),
    delta_ml     = mean(ml_t2 - ml_t1),
    .groups = "drop"
  ) |>
  knitr::kable(digits = 2,
               caption = "Per-quartile T1 / T2 / delta under each estimator")
```

# Save outputs

```{r save}
write_rds(pairs,  here::here("data", "math_co_eap_ml_pairs.rds"))
write_rds(all3,   here::here("data", "math_co_all_scoring_approaches.rds"))
```

# Conclusions

1.  **`levantemodels::score_irt` had a column-order bug**, now fixed in
    [PR #9](https://github.com/levante-framework/levantemodels/pull/9). When
    the `response.pattern` data frame's columns were not in `items(mod_rec)`
    order, `mirt::fscores()` matched by position and returned wrong θs; the
    fix reorders `data_aligned` to match `items(mod_rec)` before calling
    `fscores`.

2.  **This bug propagated into the published math scores** for Bogotá
    (and presumably every dataset/task scored by the same code path). The
    published pipeline score correlates with the reproduced buggy `score_irt`
    (r = 0.95) but disagrees with the corrected EAP (r = 0.75).

3.  **Under corrected scoring, the apparent T1 → T2 compression for Bogotá
    math largely disappears.** Galton slope goes from 0.19 (pipeline) to
    0.81 (corrected EAP). All four T1 quartiles show modest positive
    growth; the "decline" for high-T1 kids was an artifact of mismatched
    response patterns.

4.  **EAP vs ML matters little once the bug is fixed.** Corrected EAP
    slope ≈ 0.81, corrected ML slope ≈ 0.72; both show similar growth
    patterns. The original story that "EAP shrinkage is the problem" was
    confounded with the bug.

5.  **Remaining action**: `levantemodels::score_irt` itself is now fixed, but
    unless `levante_data_latest` has been rescored since PR #9, the published
    pipeline scores still reflect the pre-fix behavior. Rescore the math (and
    likely every IRT task) in `levante_data_latest`. We expect non-trivial
    revisions to the cross-sectional and longitudinal math story at Bogotá.

6.  **For the LEVANTE longitudinal project specifically**: 02 and 03
    should use the corrected scores
    (`data/math_co_all_scoring_approaches.rds$my_eap` or `my_ml`) rather
    than the pipeline `score`. Doing so is likely to substantially
    change the math LGC and growth model results.
