---
title: "Memory (Corsi) — item difficulty structure & the '2×2 vs 3×3' question"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 9
    fig-height: 5.5
execute:
  warning: false
  message: false
---

{{< include ../_disclaimer.qmd >}}

# Question

The release notes flagged Memory scoring as not accounting for grid size
(2×2 vs 3×3), and the task was marked DROP. But on corrected (v1.2) data the
Colombia scores look fine, raising the hypothesis that the real culprit was
the **column-order scoring bug** (now fixed) rather than the grid issue.

The Corsi span task has a clean design: sequences of squares light up and the
child reproduces them. Each item is defined by three features encoded in
`item_uid` (`mg_{forward|backward}_{2grid|3grid}_len{N}`):

- **direction** — forward (same order) vs backward (reversed)
- **grid** — 2×2 vs 3×3 (more locations ⇒ harder)
- **length** — span length 2–7

We read the calibrated item difficulties as a function of these features —
which reveals the actual difficulty structure of the task — and ask whether
grid is properly separated and whether the structure replicates across sites.

```{r setup}
source(here::here("common.R"))
ip <- load_item_parameters() |>
  filter(task_id == "memory-game", subset == "all_items") |>
  mutate(direction = str_extract(item_uid, "forward|backward"),
         grid      = str_extract(item_uid, "2grid|3grid"),
         length    = as.integer(str_extract(item_uid, "(?<=len)[0-9]+")))
```

# 1. Difficulty by length × grid × direction

```{r diff-plot, fig.height=5}
ip |>
  ggplot(aes(x = length, y = difficulty, color = grid, linetype = direction)) +
  geom_hline(yintercept = 0, lty = 3, color = "grey70") +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c(`2grid` = "#268bd2", `3grid` = "#cb4b16")) +
  scale_x_continuous(breaks = 2:7) +
  labs(x = "Sequence length", y = "Item difficulty (logits)",
       title = "Memory item difficulty rises orderly with span length",
       subtitle = "Grid is a separate item dimension (3×3 slightly harder); backward ≥ forward",
       color = "Grid", linetype = "Direction")
```

```{r diff-slopes}
ip |> group_by(direction, grid) |>
  summarise(n_items = n(),
            cor_length_difficulty = cor(length, difficulty),
            logits_per_length = coef(lm(difficulty ~ length))[2],
            .groups = "drop") |>
  knitr::kable(digits = 3,
               caption = "Difficulty is near-linear in length within each grid×direction")
```

The span structure is exactly as a working memory task should behave:
difficulty increases ~1 logit per added item, near-perfectly ordered
(r ≈ 0.90–1.00). Backward spans are harder than forward at matched length.

# 2. Is grid size "accounted for"?

The release-notes concern was that scoring ignored 2×2 vs 3×3. But the items
**already separate by grid** — each (direction, length) has distinct 2×2 and
3×3 items with their own calibrated difficulties.

```{r grid-effect}
ip |> select(direction, length, grid, difficulty) |>
  pivot_wider(names_from = grid, values_from = difficulty) |>
  mutate(`3×3 − 2×2` = `3grid` - `2grid`) |>
  filter(!is.na(`3×3 − 2×2`)) |>
  arrange(direction, length) |>
  knitr::kable(digits = 2, caption = "Grid effect at matched direction × length")
```

The 3×3 items are modestly harder than 2×2 at matched length (~+0.2–0.5
logits), as expected from the larger location set — and crucially this is
**captured in the item parameters**, not lost. (The one large reversal is at
length 7, where the 2×2 cell has very few responses.) So the "grid not
accounted for" concern appears **already handled by the item design**; it is
not what was breaking the scores.

# 3. Does the structure replicate across sites?

```{r site-acc, fig.height=5}
tr <- load_levante_trials(task_ids = "memory-game") |>
  filter(dataset %in% c("pilot_mpieva_de_main", "pilot_uniandes_co_bogota"),
         item_uid %in% ip$item_uid) |>
  mutate(correct = as.numeric(correct))

acc <- tr |> group_by(dataset_label, item_uid) |>
  summarise(acc = mean(correct, na.rm = TRUE), n = n(), .groups = "drop") |>
  left_join(ip |> select(item_uid, direction, grid, length), by = "item_uid") |>
  pivot_wider(names_from = dataset_label, values_from = c(acc, n))

acc |>
  ggplot(aes(x = `acc_Colombia — Bogotá`, y = `acc_Germany — Leipzig`,
             color = grid, size = length)) +
  geom_abline(slope = 1, intercept = 0, lty = 3, color = "grey50") +
  geom_point(alpha = 0.8) +
  scale_color_manual(values = c(`2grid` = "#268bd2", `3grid` = "#cb4b16")) +
  coord_equal() +
  labs(x = "Item accuracy — Bogotá", y = "Item accuracy — Leipzig",
       title = "Item accuracy is rank-consistent across sites",
       subtitle = glue::glue("r = {round(cor(acc$`acc_Colombia — Bogotá`, acc$`acc_Germany — Leipzig`, use='complete.obs'),2)}; Leipzig higher overall (above the diagonal)"),
       color = "Grid", size = "Length")
```

Item accuracies correlate **r ≈ 0.90** across the two sites — the difficulty
*ordering* is the same everywhere. Leipzig sits above the diagonal (higher
accuracy), but that is a level difference (sample/age), not disorder or noise
in the measure.

# 4. 2026-08 update: run-level comparability — sumscores vs IRT, and the grid question sharpened

Sections 1–3 established that grid is a calibrated item dimension. But the
sharper comparability question is at the **run level**: in practice, **runs are
pure-grid** — a child gets an all-2×2 or an all-3×3 administration (only 4
mixed runs among ~2,000), and grid assignment differs by site and wave (rural
Colombia is ~99% 3×3; Leipzig and Bogotá are split). So no child ever links
the two grids *within* a run, and cross-grid comparability rests entirely on
the scoring model. Here we test it against model-free sumscores, exploiting
the task's **staircase** design: every run starts at length 2 and stops
upward, so a span level the child never reached can be scored as failed —
"imputing zeros for unreached steps" gives ladder scores that are well-defined
across runs of different lengths.

```{r run-scores}
tr2 <- load_levante_trials(task_ids = "memory-game") |>
  filter(!is.na(item_uid), str_detect(item_uid, "^mg_")) |>
  mutate(direction = str_extract(item_uid, "forward|backward"),
         grid = str_extract(item_uid, "2grid|3grid"),
         len = as.integer(str_extract(item_uid, "(?<=len)[0-9]+")),
         correct = as.numeric(correct)) |>
  filter(len <= 7)

rs <- tr2 |> group_by(dataset, user_id, run_id) |>
  summarise(grid = names(sort(table(grid), decreasing = TRUE))[1],
            ntr = n(), ncorr = sum(correct, na.rm = TRUE), .groups = "drop")
# ladder score: a (direction, length) level counts as passed with >=2 correct;
# levels never reached count 0 automatically (staircase imputation)
lev <- tr2 |> group_by(run_id, direction, len) |>
  summarise(pass = sum(correct, na.rm = TRUE) >= 2, .groups = "drop") |>
  group_by(run_id) |> summarise(levels_passed = sum(pass), .groups = "drop")
span <- tr2 |> filter(correct == 1) |> group_by(run_id, direction) |>
  summarise(mx = max(len), .groups = "drop") |>
  group_by(run_id) |> summarise(maxspan = mean(mx), .groups = "drop")

mem <- read_rds(here::here("data/scores_all_sites.rds")) |>
  filter(task_id == "memory-game", !is.na(score)) |>
  select(run_id, user_id, dataset, dataset_label, age, score, score_se) |>
  inner_join(rs |> select(run_id, grid, ntr, ncorr), by = "run_id") |>
  left_join(lev, by = "run_id") |> left_join(span, by = "run_id")

mem |> group_by(dataset_label, grid) |>
  summarise(runs = n(), mean_age = mean(age), .groups = "drop") |>
  pivot_wider(names_from = grid, values_from = c(runs, mean_age)) |>
  knitr::kable(digits = 1, caption = "Scored memory runs by grid × dataset — runs are pure-grid, and assignment is site- and wave-structured")
```

## 4.1 Sumscores vs IRT θ

```{r sum-vs-theta}
cors <- function(d) tibble(
  n = nrow(d),
  total_correct = cor(d$score, d$ncorr),
  levels_passed = cor(d$score, d$levels_passed),
  max_span = cor(d$score, d$maxspan, use = "pair"))
bind_rows(pooled = cors(mem), `2×2 runs` = cors(mem |> filter(grid == "2grid")),
          `3×3 runs` = cors(mem |> filter(grid == "3grid")), .id = "scope") |>
  knitr::kable(digits = 3, caption = "Correlation of IRT θ with model-free sumscores")
```

Within 3×3 runs, θ is nearly a rescaling of the sumscores (r ≈ 0.96); within
2×2 runs the correlation with total-correct is lower (r ≈ 0.75; max-span does
better at 0.92), reflecting the 2×2 form's compressed range and its
concentration in the younger/earlier administrations. Ordering-wise, the model
and the ladder agree.

## 4.2 The empirical grid test: Leipzig's wave-1→2 grid switch

Leipzig switched most children from 2×2 (wave 1) to 3×3 (wave 2) at a ~5-month
gap, while 42 children stayed 3×3→3×3 — a natural experiment for the model's
cross-grid linkage. If the linkage were wrong, the switchers would show a step
in θ relative to the stayers.

```{r switch}
lw <- mem |> filter(dataset == "pilot_mpieva_de_main") |>
  group_by(user_id) |> arrange(age) |> mutate(wave = row_number(), nw = n()) |>
  filter(nw >= 2, wave <= 2) |> ungroup()
sw <- lw |> select(user_id, wave, grid, score, levels_passed, maxspan, age) |>
  pivot_wider(names_from = wave, values_from = c(grid, score, levels_passed, maxspan, age)) |>
  filter(!is.na(score_2)) |>
  mutate(path = paste0(str_sub(grid_1, 1, 1), "→", str_sub(grid_2, 1, 1)),
         d_theta = score_2 - score_1, d_levels = levels_passed_2 - levels_passed_1,
         d_span = maxspan_2 - maxspan_1, gap = age_2 - age_1)
sw |> group_by(path) |> filter(n() > 5) |>
  summarise(n = n(), gap_yr = mean(gap), `Δ θ` = mean(d_theta),
            `Δ levels` = mean(d_levels), `Δ max-span` = mean(d_span), .groups = "drop") |>
  knitr::kable(digits = 2, caption = "Within-child change across Leipzig waves, by grid path")
```

The sumscores expose their own non-comparability across the switch: **max-span
*drops* (−0.4) while levels-passed *jumps* (+2.2)** for the same children —
raw ladder scores mean different things on different boards. **θ, by contrast,
moves smoothly**: switchers gain ≈ +0.15 over ~5 months while 3×3-stayers gain
≈ +0.01 over ~3 months — both consistent with modest within-year WM growth and
with each other (≈ +0.3–0.4 θ/yr vs ≈ 0), with no step at the grid boundary.
The model's grid linkage passes the one empirical test the data afford.

One caveat stands: at matched levels-passed, pooled θ differs by grid by more
than the ~0.3-logit item offsets alone would imply — but that comparison
confounds grid with site and with the per-dataset EAP priors (2×2 runs live
disproportionately in Leipzig/Bogotá wave 1). The switcher test above is the
cleaner read, and it is reassuring.

# Verdict (updated 2026-08)

- **Memory is well-behaved on corrected data.** Difficulty is cleanly ordered
  by span length, backward > forward, and the structure replicates across
  sites (r ≈ 0.90).
- **The "2×2 vs 3×3" concern is handled at both levels.** Grid is a separate
  calibrated item dimension (§2), and — the sharper run-level question — the
  model's cross-grid linkage passes the Leipzig wave-1→2 switch test (§4.2):
  within-child θ is continuous across the grid change while raw sumscores
  visibly are not. Sumscores and θ agree on ordering (r ≈ 0.88–0.96).
- **Do not substitute raw sumscores across grids** — max-span and
  ladder-total move in opposite directions across a grid switch; if a
  model-free score is ever needed, it must be within-grid only.
- **Recommendation: lift the DROP.** Memory appears suitable for growth/SEM
  on the current (registry v2_3) scores. Residual caveats: the pooled
  grid-offset at matched sumscore partly reflects per-dataset EAP priors (not
  a pure item effect), and a few uncalibrated `len8` trials should be
  excluded.

```{r save}
write_rds(ip, here::here("data", "memory_item_difficulties.rds"))
```
