---
title: "Memory (Corsi) — can we shorten the task?"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 9
    fig-height: 5
execute:
  warning: false
  message: false
---

{{< include ../_disclaimer.qmd >}}

# Scope

The Memory task (`memory-game`, forward + backward Corsi block span) is one
of the longest tasks in the battery, and — because it is an untimed
staircase — its length grows with ability. The DCC question: **can we cut
its length, and how?** The task is not CAT; the levers are the staircase
rules. Candidates: get to backward faster, require fewer correct
repetitions per span length, skip staircase steps, tighten the error
budget. Scores are IRT (multigroup Rasch, backward items simply calibrate
as harder items on the same dimension), so shortened runs can be rescored
on the *same* scale — the question is what we lose in precision, validity,
and test–retest reliability.

Data: `levante_data_latest:e9pf:v1_2` (2026-08 refresh; the four
calibration datasets), model `multigroup_dataset` scalar Rasch from
metadata v2_3. Heavy computation cached by
[`_prep_memory_length.R`](\_prep_memory_length.R).

```{r setup}
source(here::here("common.R"))
library(stringr)

res <- read_rds(here("data/memory_length_results.rds"))

mg <- read_rds(here("data/latest_r2_trials_mg.rds")) |>
  filter(dataset %in% res$priors$group) |>
  mutate(dir = str_extract(item_uid, "(?<=mg_)[a-z]+"),
         grid = str_extract(item_uid, "[23]grid"),
         len = as.integer(str_extract(item_uid, "(?<=len)[0-9]+")),
         correct = as.logical(correct)) |>
  filter(!is.na(correct))

sc <- read_rds(here("data/levante_data_latest__v1_2__scores.rds")) |>
  filter(task_id == "memory-game", !is.na(score))

dataset_pal <- set_names(
  RColorBrewer::brewer.pal(4, "Set1"),
  dataset_labels_named[c("pilot_uniandes_co_bogota", "pilot_uniandes_co_rural",
                         "pilot_mpieva_de_main", "pilot_western_ca_main")])
label_dataset <- \(df) df |>
  mutate(dataset_label = factor(dataset_labels_named[dataset],
                                levels = names(dataset_pal)))
```

# 1. Task anatomy

Reverse-engineered from 1,644 runs of trial data (transition probabilities,
episode structure, block terminations):

- Forward block starts at **length 2**; **+1 length after 3 consecutive
  correct**; an error keeps you at the same length (there is no
  step-down).
- A block ends after **3 cumulative errors** (modern deployments; ~15% of
  runs, mostly early task versions, show larger budgets).
- **Backward restarts at length 2** and runs the same rule. (386 further
  runs stop after forward — attrition across all ability levels, not a
  rule; they are excluded above.)
- Items calibrate as `mg_{direction}_{grid}_len{N}-{attempt}` with
  difficulties **tied across attempts**, so each (direction, grid, length)
  is one difficulty — a clean ~1-logit-per-rung ladder:

```{r ladder}
res$dlook |>
  mutate(difficulty = -d) |>
  select(dir, grid, len, difficulty) |>
  pivot_wider(names_from = len, values_from = difficulty) |>
  arrange(dir, grid) |>
  knitr::kable(digits = 1, caption = "Item difficulty (logits) by direction × grid × span length. Grid (2×2 vs 3×3) is a deployment property of the run, not a within-run phase.")
```

The design consequence: **every rung costs at least 3 correct trials**, no
matter how far below the child's span it is, and strong children pay the
ladder twice (forward, then backward from scratch).

# 2. How long is it, and for whom?

```{r durations}
dur <- mg |>
  group_by(dataset, run_id) |>
  summarise(mins = as.numeric(difftime(max(timestamp), min(timestamp),
                                       units = "mins")),
            n_tr = n(), .groups = "drop") |>
  filter(mins < 20) |>
  inner_join(sc |> select(run_id, score), by = "run_id") |>
  label_dataset()

dur |>
  group_by(dataset_label) |>
  summarise(n = n(), median_min = median(mins), q90_min = quantile(mins, .9),
            r_duration_theta = cor(mins, score), .groups = "drop") |>
  knitr::kable(digits = 2, caption = "Run duration (first-to-last-trial span) and its correlation with ability")
```

```{r duration-quartile}
dur |>
  group_by(q = ntile(score, 4)) |>
  summarise(n = n(), median_min = median(mins), q90_min = quantile(mins, .9),
            median_trials = median(n_tr), .groups = "drop") |>
  knitr::kable(digits = 1, caption = "Pooled duration by ability quartile: the top quartile takes more than twice as long as the bottom")
```

```{r duration-plot, fig.height=4.5}
dur |>
  ggplot(aes(x = score, y = mins, color = dataset_label)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_manual(values = dataset_pal, name = NULL) +
  labs(x = "Memory ability (θ)", y = "Run duration (min)",
       title = "The Memory task is a progressive tax on ability") +
  theme(legend.position = "bottom")
```

The claim checks out: median run ≈ 4.5 min, but duration correlates
r ≈ .5–.8 with θ within dataset; the top ability quartile has a median of
~6 min and a 90th percentile of ~8 min — and trial timestamps stop at the
last trial, so with instructions/practice the experienced length for those
children is comfortably in the reported 8–10 min range.

Where do the minutes go? Per-trial time is mechanical in span length
(sequence presentation is ~1 s per element plus reproduction):

```{r time-per-trial, fig.height=3.8}
res$dt_med |>
  ggplot(aes(x = len, y = med)) +
  geom_line() + geom_point() +
  labs(x = "Span length", y = "Median seconds per trial",
       title = sprintf("Seconds per trial ≈ %.1f + %.1f × length",
                       res$dt_coef[1], res$dt_coef[2]))
```

```{r easy-tax}
easy <- mg |>
  inner_join(res$full |> select(run_id, est), by = "run_id") |>
  left_join(res$dlook, by = c("dir", "grid", "len")) |>
  filter(!is.na(d)) |>
  mutate(p_hat = plogis(est + d),
         secs = res$dt_coef[1] + res$dt_coef[2] * len)

easy |>
  group_by(q = ntile(est, 4)) |>
  summarise(pct_time_easy = sum(secs[p_hat > .9]) / sum(secs),
            pct_time_backward_easy = sum(secs[p_hat > .9 & dir == "backward"]) /
              sum(secs), .groups = "drop") |>
  knitr::kable(digits = 2, caption = "Share of testing time spent on trials the child passes with >90% probability ('easy-trial tax'), by ability quartile")
```

# 3. Rescoring machinery (validation)

All shortened variants are rescored with the calibrated Rasch difficulties
and each dataset's EAP prior via direct quadrature. Full-length rescoring
reproduces the published pipeline scores:

```{r validation}
knitr::kable(res$validation, digits = 3,
             caption = "Quadrature EAP on full runs vs published scores")
```

And the Rasch model is well-calibrated for *simulating* responses —
predicted vs observed accuracy by ability quintile × length:

```{r calib, fig.height=4}
res$calib |>
  ggplot(aes(x = pred_acc, y = obs_acc, color = factor(q))) +
  geom_abline(lty = 3) +
  geom_point(aes(size = n), alpha = 0.7) +
  scale_size_area(max_size = 4, guide = "none") +
  scale_color_viridis_d(name = "θ quintile") +
  labs(x = "Model-predicted accuracy", y = "Observed accuracy",
       title = "Simulation validity: Rasch predictions track observed accuracy")
```

# 4. Exact replay: tightening the error budget

A smaller error budget is a pure truncation of the observed data (the path
up to the Nth error is identical under either rule), so these need no
simulation — we rescore what children actually did.

```{r replay}
retest_pairs <- function(scored, min_gap_mo = 2) {
  scored |>
    inner_join(sc |> select(run_id, age), by = "run_id") |>
    group_by(dataset, user_id) |>
    filter(n() >= 2, !is.na(age)) |>
    arrange(age) |>
    summarise(gap_mo = (age[2] - age[1]) * 12,
              t1 = est[1], t2 = est[2], .groups = "drop") |>
    filter(gap_mo >= min_gap_mo)
}

replay_all <- bind_rows(res$full |> mutate(policy = "budget3 (current)"),
                        res$replay)

replay_summary <- replay_all |>
  group_by(policy) |>
  summarise(median_min = median(mins),
            mean_se = mean(se), .groups = "drop") |>
  left_join(
    replay_all |>
      inner_join(res$full |> select(run_id, full_est = est), by = "run_id") |>
      group_by(policy) |>
      summarise(r_with_full = cor(est, full_est), .groups = "drop"),
    by = "policy") |>
  left_join(
    replay_all |>
      group_by(policy) |>
      group_modify(~{
        p <- retest_pairs(.x)
        tibble(
          retest_de = cor(p$t1[p$dataset == "pilot_mpieva_de_main"],
                          p$t2[p$dataset == "pilot_mpieva_de_main"]),
          n_de = sum(p$dataset == "pilot_mpieva_de_main"),
          stability_bogota = cor(p$t1[p$dataset == "pilot_uniandes_co_bogota"],
                                 p$t2[p$dataset == "pilot_uniandes_co_bogota"]),
          n_bog = sum(p$dataset == "pilot_uniandes_co_bogota"))
      }) |> ungroup(),
    by = "policy")

knitr::kable(replay_summary, digits = 2,
             caption = "Error-budget truncations, rescored from observed data. median_min = predicted time on the retained trials (per-trial time model); DE retest ≈ 3.6 months; Bogotá stability ≈ 16 months.")
```

# 5. Simulation: restructuring the staircase

Path-changing policies are simulated from each run's full-data θ̂ using the
calibrated difficulties (the anchor policy S0 replicates the current rules
— compare it to the observed row to calibrate trust; policy *differences*
within the simulation are the inference target).

Policies:

- **S0** current rules (anchor)
- **S1** promote after **2** consecutive correct (not 3)
- **S2** backward starts at length **3**
- **S3** = S1 + S2
- **S4** = S1 + both blocks start at length 3
- **S5** = S1 + error budget **2** + backward starts at 3 (aggressive)

```{r sims}
sim_summary <- res$sims |>
  group_by(policy) |>
  summarise(median_min = median(mins), q90_min = quantile(mins, .9),
            mean_se = mean(se), .groups = "drop") |>
  left_join(
    res$sims |>
      group_by(policy, rep) |>
      summarise(r = cor(est, theta), .groups = "drop_last") |>
      summarise(r_with_full = mean(r), .groups = "drop"),
    by = "policy")

# simulated retest: per rep, pair each retest user's two waves
sim_pairs <- res$sims |>
  inner_join(sc |> select(run_id, age), by = "run_id") |>
  group_by(policy, rep, dataset, user_id) |>
  filter(n() >= 2, !is.na(age)) |>
  arrange(age) |>
  summarise(gap_mo = (age[2] - age[1]) * 12, t1 = est[1], t2 = est[2],
            .groups = "drop") |>
  filter(gap_mo >= 2)

sim_retest <- sim_pairs |>
  filter(dataset %in% c("pilot_mpieva_de_main", "pilot_uniandes_co_bogota")) |>
  group_by(policy, rep, dataset) |>
  summarise(r = cor(t1, t2), n = n(), .groups = "drop") |>
  group_by(policy, dataset) |>
  summarise(r = mean(r), .groups = "drop") |>
  pivot_wider(names_from = dataset, values_from = r) |>
  rename(sim_retest_de = pilot_mpieva_de_main,
         sim_stability_bog = pilot_uniandes_co_bogota)

sim_table <- sim_summary |> left_join(sim_retest, by = "policy")

obs_row <- tibble(policy = "OBSERVED (current)",
                  median_min = median(res$full$mins),
                  q90_min = quantile(res$full$mins, .9),
                  mean_se = mean(res$full$se),
                  r_with_full = 1,
                  sim_retest_de = replay_summary$retest_de[
                    replay_summary$policy == "budget3 (current)"],
                  sim_stability_bog = replay_summary$stability_bogota[
                    replay_summary$policy == "budget3 (current)"])

knitr::kable(bind_rows(obs_row, sim_table), digits = 2,
             caption = "Simulated staircase policies. Minutes are trial time only (no instructions). The OBSERVED row uses real data; S0 is the simulation anchor — trust differences between S-rows, calibrated by S0 vs OBSERVED.")
```

```{r sim-quartile-minutes}
res$sims |>
  group_by(policy, q = ntile(theta, 4)) |>
  summarise(median_min = median(mins), .groups = "drop") |>
  pivot_wider(names_from = q, values_from = median_min,
              names_prefix = "θ Q") |>
  knitr::kable(digits = 1, caption = "Median simulated minutes by ability quartile: where each policy saves its time")
```

# Verdict

**1. The length complaint is verified and is a tax on ability.** Duration
correlates r ≈ .5–.8 with θ; the top ability quartile has a median of
~6 min and a 90th percentile of ~8 min of pure trial time — with
instructions and practice, that is the reported 8–10 minute experience.
Top-quartile children spend **29% of their testing time on trials they
pass with >90% probability**, mostly re-climbing rungs far below their
span.

**2. Don't touch the error budget — it is where the measurement lives.**
The exact-replay analysis (real data, truncated and rescored) shows that
cutting the per-block error budget from 3 to 2 saves ~27% of time but
costs real reliability: DE 3.6-month retest drops **.58 → .48** and
Bogotá 16-month stability **.46 → .32**. Errors are the informative
trials — they bracket the child's ceiling where Rasch information peaks —
and the three errors per block are most of the task's signal.

**3. The fat is the promotion rule and the restarts.** In the calibrated
simulation (anchor S0 reproduces the current rules; sim retest values are
attenuated relative to observed — compare S-rows to each other, not to
the observed row):

- **Promote after 2 consecutive correct** (S1): −13% median time, −18%
  at the 90th percentile; SE .50 vs .48; retest within simulation noise
  of the anchor.
- **Start backward at length 3** (S2): −4% overall (concentrated in
  able kids); no reliability cost.
- **Both, plus forward also starts at 3** (S4): **−22% median, −25% at
  q90** (top-quartile median 5.4 → 4.0 min of trial time); SE .53 vs
  .48; retest indistinguishable from the anchor (DE .37 vs .41,
  Bogotá .38 vs .37 — differences within rep noise).
- The aggressive add-on (S5: also budget 2) reaches −40% but inherits
  the replay-demonstrated reliability cost of the smaller error budget.

**Recommendation to the DCC:** adopt **S4** — advance after **2**
consecutive correct (not 3), start **both** blocks at length **3**, keep
the 3-error budget. Expected effect: roughly a quarter of task time
removed, concentrated on the longest sessions (the top quartile saves
~1.4 min of trial time; with per-trial overhead the experienced saving is
larger), at a precision cost of ~.05 SD units of SE and no detectable
reliability cost. If a more conservative step is preferred, S3 (promote-2
+ backward-starts-3, forward still starts at 2) yields −16%/−21% with the
same reliability profile. Child-experience risk is low: length 3 is still
very easy for nearly everyone (P(correct) ≈ .8 even at θ = −1), children
below that floor produce interpretable all-error records exactly as they
do now at length 2, and the number of *failures* a child experiences is
governed by the error budget, which is unchanged.

Two caveats. First, the simulation trusts the Rasch model's person
homogeneity (no sequence-specific effects beyond length; the calibration
plot in §3 supports this). Second, changing the staircase changes which
attempt-numbered items exist, so the scoring bank needs the new
(direction × grid × length) items mapped — trivial under the current
tied-difficulty convention, but it should be done deliberately, and a
recalibration pass once ~500 runs of the new form exist would confirm the
difficulty ladder transfers.
