---
title: "ROAR literacy tasks — Sentence (SRE) & Phoneme (PA)"
levante:
  summary: "Are ROAR-Phoneme (by-language Rasch) and ROAR-Sentence (speeded, guessing-adjusted count) valid and well-behaved?"
  status: draft
  task_ids: [pa, sre]
  datasets: all
  data:
    - {source: "load_levante_scores", version: "v1_2", table: "scores"}
    - {source: "load_levante_trials", version: "v1_2", table: "trials"}
    - {source: "load_item_parameters", version: "v1_14", table: "item_parameters"}
  depends_on: [00_load_data.qmd]
  findings:
    - "Both rise with age at every site (PA r ≈ 0.34–0.54; SRE r ≈ 0.27–0.63)."
    - "PA: first-sound easier than last-sound; calibrated by language (no German), so not on one cross-language scale."
    - "SRE is a reading-efficiency count on a task-internal z-scale, not an IRT θ."
  last_run: 2026-07-16
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 9
    fig-height: 5
execute:
  warning: false
  message: false
---

{{< include ../_disclaimer.qmd >}}

# Scope

The two remaining ROAR literacy measures, which use **different scoring** from
the core tasks:

- **ROAR-Phoneme (`pa`)** — phonological awareness, scored with a
  **by-language Rasch** IRT model (separate es-CO and en-US calibrations; no
  German). Items: `pa_{subtype}_{word}` with subtypes **fsm** (first-sound
  matching), **lsm** (last-sound matching), **del** (deletion).
- **ROAR-Sentence (`sre`)** — sentence reading efficiency, a **speeded** task
  scored as a guessing-adjusted, z-scored correct-count (`(correct − incorrect)
  / 180`, standardized) — **not** an IRT θ, so it is not on the same scale as
  the other tasks and individual item difficulties aren't calibrated.

```{r setup}
source(here::here("common.R"))
sc <- read_rds(here("data/scores_all_sites.rds"))
tr <- load_levante_trials(task_ids = c("sre","pa")) |> mutate(correct = as.numeric(correct))
```

# ROAR-Phoneme (PA)

## Item difficulty by subtype

```{r pa-subtype}
pa_ip <- load_item_parameters() |> filter(task_id == "pa") |>
  mutate(subtype = str_extract(item_uid, "(?<=pa_)[a-z]+"),
         subtype = recode(subtype, fsm = "first-sound", lsm = "last-sound",
                          del = "deletion"))
pa_ip |> group_by(subset, subtype) |>
  summarise(n_items = n(), mean_difficulty = mean(difficulty),
            .groups = "drop") |>
  pivot_wider(names_from = subset, values_from = c(n_items, mean_difficulty)) |>
  knitr::kable(digits = 2, caption = "PA mean item difficulty by subtype × language")
```

```{r pa-diff-plot, fig.height=4.5}
pa_ip |>
  ggplot(aes(x = subtype, y = difficulty, color = subset)) +
  geom_hline(yintercept = 0, lty = 3, color = "grey70") +
  geom_jitter(width = 0.15, alpha = 0.6) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.4, linewidth = 0.4) +
  scale_color_brewer(palette = "Set1", name = "Language") +
  labs(x = NULL, y = "Item difficulty (logits)",
       title = "PA item difficulty by phoneme-manipulation subtype")
```

First-sound matching is easiest, last-sound matching harder — the expected
ordering for phonological-awareness manipulations. (Deletion items appear in
the trial data but are sparsely calibrated.)

## Age validity

```{r pa-age, fig.height=4.5}
sc |> filter(task_id == "pa", !is.na(score), !is.na(age)) |>
  ggplot(aes(x = age, y = score, color = dataset_label)) +
  geom_point(alpha = 0.25, size = 0.7) +
  geom_smooth(method = "lm", se = TRUE) +
  scale_color_brewer(palette = "Set1", name = NULL) +
  labs(x = "Age (years)", y = "PA ability (θ)",
       title = "ROAR-Phoneme rises with age") +
  theme(legend.position = "bottom")
```

```{r pa-age-tab}
sc |> filter(task_id == "pa", !is.na(score), !is.na(age)) |>
  group_by(dataset_label) |> filter(n() >= 30) |>
  summarise(n = n(), cor_age = cor(score, age), .groups = "drop") |>
  knitr::kable(digits = 2, caption = "PA age correlation by site (no German PA)")
```

# ROAR-Sentence (SRE)

A speeded efficiency measure: the score is a standardized guessing-adjusted
count, so its scale is task-internal (mean ≈ 0). All four data-rich sites have
it.

```{r sre-dist, fig.height=4}
sc |> filter(task_id == "sre", !is.na(score)) |>
  ggplot(aes(x = score, fill = dataset_label)) +
  geom_density(alpha = 0.4) +
  scale_fill_brewer(palette = "Set1", name = NULL) +
  labs(x = "SRE score (guessing-adjusted, z-scaled)", y = "Density",
       title = "SRE score distribution by site") +
  theme(legend.position = "bottom")
```

```{r sre-age, fig.height=4.5}
sc |> filter(task_id == "sre", !is.na(score), !is.na(age)) |>
  ggplot(aes(x = age, y = score, color = dataset_label)) +
  geom_point(alpha = 0.25, size = 0.7) +
  geom_smooth(method = "lm", se = TRUE) +
  scale_color_brewer(palette = "Set1", name = NULL) +
  labs(x = "Age (years)", y = "SRE score",
       title = "ROAR-Sentence rises with age") +
  theme(legend.position = "bottom")
```

```{r sre-speed}
# the speeded nature: how many sentences reached, and accuracy among reached
tr |> filter(task_id == "sre") |>
  group_by(dataset_label, run_id) |>
  summarise(n_reached = n(), acc = mean(correct, na.rm = TRUE), .groups = "drop") |>
  group_by(dataset_label) |>
  summarise(median_sentences_reached = median(n_reached),
            median_accuracy = round(median(acc), 2), .groups = "drop") |>
  knitr::kable(caption = "SRE is speeded: sentences reached varies; accuracy among reached is high")
```

```{r sre-age-tab}
sc |> filter(task_id == "sre", !is.na(score), !is.na(age)) |>
  group_by(dataset_label) |> filter(n() >= 30) |>
  summarise(n = n(), cor_age = cor(score, age), .groups = "drop") |>
  knitr::kable(digits = 2, caption = "SRE age correlation by site")
```

# Verdict

- **Both ROAR literacy measures are valid** — each rises with age at every
  site (PA r ≈ 0.34–0.54; SRE r ≈ 0.27–0.63).
- **PA** has a sensible internal structure: first-sound < last-sound in
  difficulty. It is calibrated **by language** (es-CO, en-US) with no German
  data, so PA scores are not on a single cross-language scale.
- **SRE** is a speeded, guessing-adjusted **count** measure — interpret it as
  reading *efficiency*, on a task-internal z-scale, not as an IRT θ comparable
  to the other tasks. The speeded design means "sentences reached" is part of
  the signal.
- No scoring red flags for either; both are usable as literacy
  covariates/validators (as they are used in the core-tasks paper).
