---
title: "Same & Different — scoring: IRTrees, chance floors, and a recommendation"
levante:
  summary: "How should Same & Different be scored? Compares IRTree and dichotomous models, guessing floors, and a recode bug, on reliability and validity."
  status: draft
  task_ids: [same-different-selection]
  datasets: [pilot_mpieva_de_main, pilot_uniandes_co_bogota, pilot_uniandes_co_rural, pilot_western_ca_main]
  data:
    - {source: "../levante-pilots/01_fetched_data/task_data_nested.rds", table: "trials"}
    - {source: "../levante-pilots/01_fetched_data/run_data.rds", table: "runs"}
  depends_on: [tasks/_sds_scoring_fits.R]
  findings:
    - "levantemodels::recode_sds() has a repeat-detection bug (compares raw response strings); fix by comparing sorted pairs. This matters more than any modeling choice."
    - "Score with status-aware dichotomous match items plus gate items, no floors on match items, multigroup-scalar Rasch."
    - "Corpus chance metadata for SDS match items matches no guessing model; correct it or document the no-floor choice."
    - "The IRTree is measurement-equivalent (r = .99); keep it for diagnostics."
  last_run: 2026-07-22
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 9
    fig-height: 5.5
execute:
  warning: false
  message: false
---

{{< include ../_disclaimer.qmd >}}

::: {.callout-note}
## Provenance
Transferred from the SDS exploration in
`levante-pilots/03_explore_tasks/blockCAT/sds/` (data prep `_recode_sds.R`;
modeling `model_poly/`, `model_tree_lmer/`, `model_tree_mirt/`), re-based here
on the model calibration dataset. All artifacts are produced by
[`tasks/_sds_scoring_fits.R`](_sds_scoring_fits.R) and cached in
`data/sds_scoring/` — delete that directory to recompute (~15-20 min).
Data: **`../levante-pilots/01_fetched_data/task_data_nested.rds`**
(`item_task == "sds"`), the same source used by `tasks/mrot/`.
:::

```{r setup}
source(here::here("common.R"))
suppressPackageStartupMessages({
  library(mirt); library(boot); library(knitr)
})
set.seed(42)

cache_dir <- here("data/sds_scoring")
if (!file.exists(file.path(cache_dir, "results.rds"))) {
  source(here("tasks/_sds_scoring_fits.R"))
}
coded <- read_rds(file.path(cache_dir, "coded.rds"))
tree <- read_rds(file.path(cache_dir, "tree.rds"))
audit <- read_rds(file.path(cache_dir, "audit.rds"))
mc <- read_rds(file.path(cache_dir, "mc.rds"))
bug <- read_rds(file.path(cache_dir, "bug.rds"))
res <- read_rds(file.path(cache_dir, "results.rds"))
factor_cors <- read_rds(file.path(cache_dir, "factor_cors.rds"))

pilots_data_dir <- here("..", "levante-pilots", "01_fetched_data")
raw <- read_rds(file.path(pilots_data_dir, "task_data_nested.rds")) |>
  filter(item_task == "sds") |>
  unnest(data)
run_data <- read_rds(file.path(pilots_data_dir, "run_data.rds"))
runs_sds <- raw |>
  distinct(run_id, user_id, dataset) |>
  left_join(run_data |> select(run_id, age), by = "run_id")
run_meta <- runs_sds |>
  left_join(raw |> group_by(run_id) |> summarise(t_start = min(timestamp)),
            by = "run_id")
```

# The problem

SDS ("Same & Different Selection") measures cognitive flexibility. In the
*match* blocks (`2match`, `3match`, `4match`), a child sees k+1 cards
containing exactly k matching pairs (each pair matching on a different
dimension: size, color, shape, number, background). The child makes k
successive pair selections: "find two that are the same" then, repeatedly,
"find two that are the same *in a different way*." A selection is credited
correct if it is (a) a genuinely matching pair and (b) not a pair the child
already chose in that trial.

This structure breaks the assumptions of our standard per-trial dichotomous
IRT models in two ways:

1. **Sequential dependence.** What a child faces on selection 2 of a trial
   depends on what they did on selection 1. Both the chance level and the
   psychological demand ("shift to a new dimension") are conditional on the
   response history.
2. **Compound responses.** "Correct" conflates two distinct behaviors: finding
   a *match*, and finding a *new* one (not perseverating on a previous
   choice).

Three approaches were explored in the original levante-pilots work:

- **Polytomous (partial-credit) models** over per-trial summed scores —
  rejected: collapsing the sequence into an ordinal count discards the
  conditional structure and produced disordered category thresholds.
- **IRTree models via `glmer`** (De Boeck & Partchev, 2012, *JSS* 48),
  decomposing each selection into sequential nodes — promising conceptually,
  but `lmer` random-effect scores don't integrate with the mirt scoring
  pipeline.
- **The same tree decomposition in mirt** — evaluated here, against simpler
  item-structure alternatives.

**The IRTree decomposition.** Each selection is decomposed into up to two
Bernoulli nodes:

- **match**: did the child select a matching pair (vs. non-matching)?
- **new | match**: given a match, was it a *new* pair (not previously
  selected)?
- **new | non-match**: given a non-match, was it at least a new error?

Because these nodes partition the outcome space, the likelihood factorizes
cleanly and each node can be treated as a pseudo-item in an ordinary
dichotomous IRT model — which is what makes this framework compatible with
our mirt infrastructure. Nodes that are deterministic given the history (a
"new" node when nothing has been selected yet, chance = 1; an "impossible
new" node when all pairs of that type are exhausted, chance = 0) are dropped.

# Data and recode

Match-block selections from all four pilot datasets, recoded from raw
response/card-set strings by a direct port of the levante-pilots
`_recode_sds.R` logic (trial boundaries reconstructed from `choice1` markers,
cross-validated against the "different" prompt text; trials dropped for
empty/malformed responses, wrong selection counts, inconsistent card sets, or
card sets that don't afford exactly k matching pairs).

```{r data-summary}
sel_n <- nrow(coded); run_n <- n_distinct(coded$run_id)
coded |>
  group_by(dataset) |>
  summarise(runs = n_distinct(run_id), selections = n(),
            `prop match` = mean(match), `prop correct` = mean(correct)) |>
  kable(digits = 2,
        caption = glue("{format(sel_n, big.mark = ',')} selections from {run_n} runs after recode"))
```

Test–retest below uses the Leipzig (`mpieva_de`) children with two SDS
administrations (~3- or ~7-month interval), scored per run from a single
calibration.

## The recode is verified correct {#sec-audit}

Because this recode is slated to be pushed upstream into `levantemodels`, it
was audited with a **fully independent reimplementation** (regex token
extraction instead of JSON parsing, direct attribute-table pair matching
instead of the transpose/table helpers, independent trial sequencing and
status counting; see the `Recode B` block of `_sds_scoring_fits.R`).

**1. Stimulus grammar.** All `r format(audit$n_tokens, big.mark = ",")`
stimulus tokens in every match-block response and card set parse as
`size-color-shape[-number][-texture]` with the expected vocabularies
(`r length(audit$bad_tokens)` exceptions).

**2. Cell-by-cell agreement** between the two implementations:

```{r audit-agreement}
audit$agreement |>
  select(`selections` = n, `match` = match_agree, `new` = new_agree,
         `correct` = correct_agree, `status (prev matches)` = status_m_agree,
         `status (prev non-matches)` = status_n_agree,
         `matching-pair count` = n_matches_agree) |>
  kable(digits = 4, caption = "Proportion agreement, independent reimplementation vs. pipeline recode")
```

(A porting gotcha discovered during the original audit: trials must be
indexed **before** dropping invalid responses — filtering rows first shifts
trial indices whenever a dropped trial contains a `choice1` row.)

**3. Chance formulas.** The conditional guessing formulas
($P(\text{match}) = k/\binom{k+1}{2}$;
$P(\text{new} \mid \text{match}, m\text{ prior matches}) = 1 - m/k$;
$P(\text{new} \mid \text{non-match}, n\text{ prior non-matches}) = 1 - n/(\binom{k+1}{2} - k)$)
were verified by Monte Carlo simulation of the random-pair guessing process
(30,000 simulated trials per block type; all simulated probabilities within
simulation error of the formulas):

```{r mc-table}
mc |>
  select(`k` = match_k, status, `P(match) MC` = p_match_mc,
         `P(match) formula` = p_match_formula,
         `P(new|M) MC` = p_newM_mc, `P(new|M) formula` = p_newM_formula,
         `P(new|N) MC` = p_newN_mc, `P(new|N) formula` = p_newN_formula) |>
  kable(digits = 3)
```

Note these formulas are correct *for the random-guessing model*; whether that
model is appropriate as an IRT floor is a separate question (@sec-floors —
it is not).

## A correctness bug in the production recode {#sec-bug}

`levantemodels::recode_sds()` codes a selection as *new* iff its **raw
response string** has not occurred earlier in the trial. But the response
string records the two cards **in selection order** — so a child who
re-selects the same pair in the opposite order produces a different string
and is credited with a new (hence correct, if matching) response. The
corrected recode sorts the parsed cards before comparison. This is not an
edge case:

```{r bug-quant}
flipped <- bug$cmp |> filter(new_raw != new_sort)
tibble(
  `selections compared` = nrow(bug$cmp),
  `order-variant repeats (mis-credited as new)` = nrow(flipped),
  `% of selections` = round(100 * nrow(flipped) / nrow(bug$cmp), 1),
  `of which matching pairs (correct flips 1 to 0)` = sum(flipped$match),
  `runs affected` = n_distinct(flipped$run_id)
) |> kable()
```

```{r bug-impact, fig.width=6, fig.height=5}
#| fig-cap: "EAP scores under the production (order-sensitive) vs. corrected repeat coding. Same model, same data, only the 'new' coding differs."
bs <- bug$bug_scores |> mutate(across(starts_with("theta"), as.numeric))
ggplot(bs, aes(x = theta_bug, y = theta_fix)) +
  geom_abline(linetype = "dashed", colour = "grey60") +
  geom_point(alpha = 0.4, size = 0.9) +
  labs(x = "Score under production coding (raw-string repeats)",
       y = "Score under corrected coding (sorted-pair repeats)",
       subtitle = glue("r = {round(cor(bs$theta_bug, bs$theta_fix), 3)}, ",
                       "mean |Δθ| = {round(mean(abs(bs$theta_bug - bs$theta_fix)), 2)}, ",
                       "max |Δθ| = {round(max(abs(bs$theta_bug - bs$theta_fix)), 2)}"))
```

**The fix (sorted-pair comparison) must land in `levantemodels` before any
SDS model ships.** Every analysis below uses the corrected coding.

## Children are not random guessers, so theory-derived chance floors are misspecified {#sec-floors}

The LEVANTE convention is Rasch/2PL with a fixed guessing floor
`g = 1/#alts`. The natural extension here sets each node's floor from random
pair-selection (the formulas above). But real children are not random: they
*perseverate*, re-selecting previous pairs far more often than chance
predicts. Several node-items sit **below** their theoretical floor, which a
guessing-floor model cannot fit at all:

```{r below-chance}
props <- tree |>
  summarise(prop = mean(value), n = n(), .by = c(node, item, chance)) |>
  mutate(below = prop < chance)
props |> filter(below) |> arrange(desc(n)) |>
  select(node, item, `chance floor` = chance, `observed prop` = prop, n) |>
  kable(digits = 2)
share_below <- sum(props$n[props$below]) / sum(props$n)
```

`r round(100 * share_below, 1)`% of all node observations fall on below-floor
items — concentrated in exactly the theoretically interesting "new" nodes.
As shown below, the floors measurably *reduce* both internal and test–retest
reliability, because they flatten the likelihood at the low end and discard
stable individual variance in perseveration.

# A framework for choosing the scoring model

The candidate space has four axes:

| Axis | Options |
|---|---|
| **Item structure** | `m0`: block × choice (current production items, e.g. `sds_2match_choice2`) · `m1`: block × choice × response-history status (e.g. `sds_2match_choice2_1m0n`) · `t`: IRTree pseudo-items (match / new nodes × block × status) |
| **Chance floors** | theory-derived random-pair floors · none (g = 0) |
| **Dimensionality** | 1 factor · 2–3 factors (match ability, "new"/shifting ability) |
| **Item model** | Rasch · 2PL |

All models are fit in mirt on identical selections, with
instance-constrained items (repeated encounters share parameters), N(0,3)
difficulty priors, and N(1,0.3) slope priors for 2PL — matching the
production pipeline's conventions. The `m0` variant labeled "corpus floors"
uses the chance values currently in the corpus metadata (0.5 and 0.75 for
2match, 0 elsewhere), which match no principled guessing model.

Evaluation criteria, in the project's priority order: (1) marginal/empirical
reliability; (2) German test–retest r (bootstrap 95% CI); (3) age correlation
as a validity proxy; (4) BIC, compared only among models of the same
responses; (5) simplicity / pipeline fit.

# Model comparison

```{r eval-machinery}
scores_long <- res |>
  select(model, factor, scores) |>
  pmap(\(model, factor, scores) {
    scores |> select(run_id, value = all_of(factor)) |>
      # mirt >= 1.46 fscores columns can arrive as classed 1-col matrices
      mutate(value = as.numeric(value), model = model, factor = factor)
  }) |> list_rbind()
sum_scores <- coded |>
  group_by(run_id) |> summarise(value = mean(correct)) |>
  mutate(model = "sumscore", factor = "F1")
scores_long <- bind_rows(scores_long, sum_scores)
scored <- scores_long |> left_join(run_meta, by = "run_id")

age_cors <- scored |>
  summarise(r_age = cor(value, age, use = "complete.obs"), .by = c(model, factor))

retest_pairs <- scored |>
  filter(dataset == "pilot_mpieva_de_main") |>
  group_by(model, factor, user_id) |>
  filter(n() == 2) |>
  arrange(t_start, .by_group = TRUE) |>
  summarise(t1 = value[1], t2 = value[2], age1 = age[1], .groups = "drop")

boot_ci <- function(t1, t2, R = 2000) {
  b <- boot(tibble(t1, t2), \(d, i) cor(d$t1[i], d$t2[i], use = "complete.obs"), R = R)
  ci <- boot.ci(b, type = "perc")$percent[4:5]
  tibble(r_retest = b$t0, ci_lo = ci[1], ci_hi = ci[2])
}
retest_cors <- retest_pairs |>
  group_by(model, factor) |>
  summarise(boot_ci(t1, t2), n_retest = n(), .groups = "drop")

eval_table <- res |>
  select(model, factor, nfact, invariance, itemtype, converged,
         emp_rxx, marg_rxx, AIC, BIC, npar, logLik) |>
  bind_rows(tibble(model = "sumscore", factor = "F1")) |>
  left_join(age_cors, by = c("model", "factor")) |>
  left_join(retest_cors, by = c("model", "factor"))
n_retest <- retest_cors$n_retest[1]
fmt_retest <- \(r, lo, hi) glue("{sprintf('%.2f', r)} [{sprintf('%.2f', lo)}, {sprintf('%.2f', hi)}]")
gv <- function(m, col, f = "F1") eval_table[[col]][eval_table$model == m & eval_table$factor == f][1]

model_labels <- tribble(
  ~model, ~structure, ~floors, ~itemtype_l,
  "sumscore",  "proportion correct",      "—", "—",
  "m0_rasch",  "block × choice",          "corpus (production)", "Rasch",
  "m0_2pl",    "block × choice",          "corpus (production)", "2PL",
  "m0c_rasch", "block × choice",          "principled (mean)", "Rasch",
  "m0z_rasch", "block × choice",          "none", "Rasch",
  "m1_rasch",  "block × choice × status", "principled", "Rasch",
  "m1_2pl",    "block × choice × status", "principled", "2PL",
  "m1z_rasch", "block × choice × status", "none", "Rasch",
  "m1z_2pl",   "block × choice × status", "none", "2PL",
  "m1z_sig_rasch", "block × choice × status × features", "none", "Rasch",
  "t1_rasch",  "tree nodes, 1 factor",    "principled", "Rasch",
  "t1z_rasch", "tree nodes, 1 factor",    "none", "Rasch",
  "t1z_2pl",   "tree nodes, 1 factor",    "none", "2PL",
  "t2zc_rasch", "tree nodes, 2 corr. factors", "none", "Rasch",
  "t3z_rasch",  "tree nodes, 3 corr. factors", "none", "Rasch")
```

## Pooled single-group models

All models converged. BIC is only comparable between models of the *same
responses*: the **selection universe** (one Bernoulli response per selection:
all `m` models) vs. the **tree universe** (each selection expanded into 1–2
node responses: all `t` models). The gate-augmented model (@sec-gate) adds
responses and is its own universe.

```{r main-table}
eval_table |>
  inner_join(model_labels, by = "model") |>
  mutate(universe = if_else(str_detect(model, "^t"), "tree", "selection"),
         retest = if_else(is.na(r_retest), NA,
                          fmt_retest(r_retest, ci_lo, ci_hi)),
         factor = if_else(factor == "F1", "", factor)) |>
  arrange(universe, BIC) |>
  select(universe, structure, floors, itemtype = itemtype_l, factor, BIC,
         `emp. rxx` = emp_rxx, `marg. rxx` = marg_rxx,
         `r (age)` = r_age, `retest r [95% CI]` = retest) |>
  kable(digits = 2)
```

Reading the table against the axes:

- **Everything beats the sum score on validity** (age r ≈
  `r round(gv("m1z_rasch", "r_age"), 2)` for the IRT variants vs.
  `r round(gv("sumscore", "r_age"), 2)` for proportion correct).
- **Chance floors hurt.** Within every item structure, removing the
  theory-derived floors raises test–retest reliability (status items:
  `r round(gv("m1_rasch", "r_retest"), 2)` with floors →
  `r round(gv("m1z_rasch", "r_retest"), 2)` without; 1-factor tree:
  `r round(gv("t1_rasch", "r_retest"), 2)` →
  `r round(gv("t1z_rasch", "r_retest"), 2)`) and empirical reliability
  (`r round(gv("m1_rasch", "emp_rxx"), 2)` →
  `r round(gv("m1z_rasch", "emp_rxx"), 2)`). This is the empirical footprint
  of @sec-floors. (The apparent retest advantage of the *corpus*-floors model,
  r = `r round(gv("m0_rasch", "r_retest"), 2)`, is dissected in @sec-whym0.)
- **Rasch wins BIC over 2PL and loses nothing** (status items:
  `r format(round(gv("m1z_rasch", "BIC")), big.mark = ",")` vs.
  `r format(round(gv("m1z_2pl", "BIC")), big.mark = ",")`), with
  equal-or-better empirical reliability and equivalent retest. The one
  criterion favoring 2PL — marginal reliability — is the least trustworthy:
  the 2PL's model-implied information overstates its empirical reliability by
  `r round(gv("m1z_2pl", "marg_rxx") - gv("m1z_2pl", "emp_rxx"), 2)`, while
  the Rasch marginal and empirical values agree to
  `r round(abs(gv("m1z_rasch", "marg_rxx") - gv("m1z_rasch", "emp_rxx")), 2)`.
  Simplicity says Rasch.
- **Item structure matters less than expected for the score itself** — with
  floors removed, all structures land at similar retest and age validity;
  status-aware structure buys a little empirical reliability, a clear BIC
  win over block × choice, and models the conditional difficulty honestly.
- **Multi-factor trees split the construct, and only one factor holds up**
  (@sec-tree): the match factor is unreliable and unstable, the newN factor
  is unmeasurable, and multidimensional multigroup models don't fit the
  pipeline.

## Why do the *wrong* production floors look good? {#sec-whym0}

The strangest cell in the table: the model with the production corpus floors
(g = .5 and .75 on the two 2match items, 0 elsewhere) — floors that match no
guessing model — has the best nominal retest
(`r round(gv("m0_rasch", "r_retest"), 2)`), beating both the *correct*
guessing floors (`r round(gv("m0c_rasch", "r_retest"), 2)`) and no floors
(`r round(gv("m0z_rasch", "r_retest"), 2)`). Three things are going on.

**First, the statistics: the advantage is not established.**

```{r whym0-boot}
de_wide <- retest_pairs |>
  filter(model %in% c("m0_rasch", "m1z_rasch"), factor == "F1") |>
  select(model, user_id, t1, t2) |>
  pivot_wider(names_from = model, values_from = c(t1, t2)) |>
  drop_na()
b <- boot(de_wide, \(d, i) {
  d2 <- d[i, ]
  cor(d2$t1_m0_rasch, d2$t2_m0_rasch) - cor(d2$t1_m1z_rasch, d2$t2_m1z_rasch)
}, R = 4000)
ci <- boot.ci(b, type = "perc")$percent[4:5]
```

The bootstrap difference in retest r between the corpus-floors model and the
recommended no-floor model is
`r sprintf("%+.2f [%+.2f, %+.2f]", b$t0, ci[1], ci[2])`
(n = `r nrow(de_wide)` paired), and m0 is the maximum of
`r n_distinct(eval_table$model[!str_detect(eval_table$model, "^mg_|sumscore")])`
fitted models — winner's curse can absorb much of the gap.

**Second, the mechanism: a fixed guessing floor is not really a guessing
parameter — it is an information reshaper.** In a Rasch-with-guess model,
$P(\text{correct}) = g + (1-g)\,\sigma(\theta + d)$, two things happen as g
rises: (i) the item's information shrinks by roughly $(1-g)^2$, and (ii) the
item becomes *one-sided* — a correct response is nearly uninformative, while
an error remains strongly diagnostic of low ability. The corpus floors apply
this surgery to exactly one block:

```{r info-share}
info_share <- function(mod_name) {
  mod <- read_rds(file.path(cache_dir, paste0("fit_", mod_name, ".rds")))
  th <- matrix(seq(-2, 2, length.out = 9))
  items <- extract.mirt(mod, "itemnames")
  ii <- map_dbl(items, \(it) mean(iteminfo(extract.item(mod, it), th)))
  tibble(item = items, info = ii, block = str_extract(items, "[0-9]match")) |>
    group_by(block) |> summarise(info = sum(info), .groups = "drop") |>
    mutate(share = info / sum(info), model = mod_name)
}
bind_rows(info_share("m0_rasch"), info_share("m0z_rasch")) |>
  mutate(model = fct_recode(model, "corpus floors (m0)" = "m0_rasch",
                            "no floors (m0z)" = "m0z_rasch")) |>
  select(model, block, share) |>
  pivot_wider(names_from = block, values_from = share) |>
  kable(digits = 3, caption = "Share of test information by block (θ ∈ [−2, 2])")
```

**Third, 2match is exactly the block you'd want muted for this criterion.**
It is the near-ceiling warm-up block — and its choice2 item sits *below* its
own .75 corpus floor, the same pathology as the tree nodes in @sec-floors —
and its block-level sum score is the least stable of the three:

```{r block-retest}
coded |>
  mutate(item_uid = glue("sds_{item_group}_{item_choice}")) |>
  filter(item_group == "2match") |>
  group_by(item_choice) |>
  summarise(`prop correct` = mean(correct), n = n(), .groups = "drop") |>
  mutate(`corpus floor` = c(0.5, 0.75)) |>
  kable(digits = 2, caption = "2match collapsed items vs. their corpus floors")

block_props <- coded |>
  group_by(run_id, item_group) |>
  summarise(prop = mean(correct), .groups = "drop") |>
  left_join(run_meta, by = "run_id")
block_props |>
  group_by(item_group) |>
  summarise(`mean prop correct` = mean(prop),
            `r (age)` = cor(prop, age, use = "complete.obs"), .groups = "drop") |>
  left_join(
    block_props |> filter(dataset == "pilot_mpieva_de_main") |>
      group_by(item_group, user_id) |> filter(n() == 2) |>
      arrange(t_start, .by_group = TRUE) |>
      summarise(t1 = prop[1], t2 = prop[2], .groups = "drop") |>
      group_by(item_group) |>
      summarise(`retest r` = cor(t1, t2), `n retest` = n(), .groups = "drop"),
    by = "item_group") |>
  kable(digits = 2, caption = "Block-level sum-score properties")
```

2match content is also mostly *match*-node behavior — the unstable component
from the tree analysis (@sec-tree) — because with only 3 cards the "new"
demand is minimal.

So the wrong floors accidentally implement a defensible design decision —
"treat the easy warm-up block as a low-end error detector and score the
mid/high range from 3/4match" — through an indefensible mechanism.
Like-for-like checks show no magic left once you see this: simply *dropping*
2match from the recommended model changes its retest by
`r {no2 <- retest_pairs |> filter(model %in% c("m1z_no2_rasch","m1z_rasch"), factor=="F1") |> select(model, user_id, t1, t2) |> pivot_wider(names_from=model, values_from=c(t1,t2)) |> drop_na(); sprintf("%+.3f", cor(no2$t1_m1z_no2_rasch, no2$t2_m1z_no2_rasch) - cor(no2$t1_m1z_rasch, no2$t2_m1z_rasch))}`
on the common retest subsample, and the residual m0 edge is within the
bootstrap noise above. The **principled versions of the same move are already
in the recommended model**: the gate items supply calibrated low-end error
detection with floors that are *correct* for genuine multiple-choice items
(@sec-gate), and 2PL — if we wanted it — learns a mild version of the block
reweighting from data.

One genuine design implication for block-adaptive testing: for children in
the retest age range, 2match contributes little stable signal, making it a
natural candidate for age-gating or adaptive block skipping — while remaining
essential floor coverage for younger/struggling children, which is precisely
why it should stay in the calibrated item bank rather than be deleted.

## What the tree decomposition teaches us {#sec-tree}

Even though the tree doesn't win as a *scoring* model, it is diagnostic gold:
it shows where the reliable signal in SDS lives. The table pools the
2-factor model (match, new) with correlated factors and the full 3-factor
model that additionally splits new-given-match (newM) from
new-given-non-match (newN).

```{r node-diagnostics}
eval_table |>
  filter(model %in% c("t2zc_rasch", "t3z_rasch")) |>
  mutate(model = fct_recode(model, "2-factor (match, new)" = "t2zc_rasch",
                            "3-factor (match, newM, newN)" = "t3z_rasch")) |>
  select(model, factor, `emp. rxx` = emp_rxx, `r (age)` = r_age,
         `retest r` = r_retest) |>
  kable(digits = 2)
```

Three lessons:

**The "new" component carries the stable signal.** Resisting perseveration
and shifting to an unused pair has retest r ≈
`r round(gv("t2zc_rasch", "r_retest", "new"), 2)` — on par with the best
unidimensional models — while the match component is much noisier (rxx =
`r round(gv("t2zc_rasch", "emp_rxx", "match"), 2)`, retest =
`r round(gv("t2zc_rasch", "r_retest", "match"), 2)`). This makes
psychological sense: with k matching pairs among k+1 cards, finding *some*
match is easy and near ceiling for most children; the executive demand is in
tracking what you already used and shifting dimensions. It also explains why
item structure barely moves the overall score: however you slice the items,
the stable variance comes from the same "new" behavior.

**Match and "new" are close to distinct abilities.** With the factor
covariance freed, they correlate only r =
`r round(factor_cors$cor[[1]]["match", "new"], 2)`. A single SDS θ is
therefore a *blend* of two nearly orthogonal skills, weighted by item
information — worth remembering when interpreting the score as "cognitive
flexibility."

**The 3-node split adds nothing measurable.** The 3-factor model converges
and its BIC essentially ties the 2-factor model
(`r format(round(gv("t3z_rasch", "BIC", "match")), big.mark = ",")` vs.
`r format(round(gv("t2zc_rasch", "BIC", "match")), big.mark = ",")`), but the
newN factor rests on only `r sum(tree$node == "newN")` responses in the whole
dataset (non-matches are rare outside 2match, where the newN node is
deterministic and dropped) — so it is unmeasurable as an individual
difference (rxx = `r round(gv("t3z_rasch", "emp_rxx", "newN"), 2)`). Its
factor correlations are still informative: newN correlates with newM at r =
`r round(factor_cors$cor[[2]]["newM", "newN"], 2)` and with match at r =
`r round(factor_cors$cor[[2]]["match", "newN"], 2)` — not repeating *errors*
and not repeating *successes* share a common perseveration-resistance core
that is unrelated to match-finding. Verdict: use the 2-factor tree for
diagnostics; the 3-node split is theoretically tidy but empirically empty at
these sample sizes.

A practical corollary: **don't discard newN nodes from the score**. Under the
floor model they look "below chance" and useless; without floors they
contribute genuine signal about perseveration.

```{r retest-scatter, fig.width=9, fig.height=4}
#| fig-cap: "German test–retest scores for the two tree factors (correlated 2-factor tree, no floors). The 'new' factor is substantially more stable."
retest_pairs |>
  filter(model == "t2zc_rasch") |>
  mutate(factor = fct_recode(factor, "match ability" = "match",
                             "'new' ability" = "new")) |>
  ggplot(aes(x = t1, y = t2)) +
  facet_wrap(vars(factor)) +
  geom_abline(linetype = "dashed", colour = "grey60") +
  geom_point(aes(colour = age1), alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, colour = "black", linewidth = 0.6) +
  scale_colour_viridis_c(option = "plasma", end = 0.92) +
  labs(x = "Score, time 1", y = "Score, time 2", colour = "Age (yr)")
```

## Multigroup models: invariance ladder and sensitivity {#sec-invariance}

The production pipeline fits multigroup models by site. Two questions: (a)
which invariance level do the data support, and (b) does the multigroup
setting change any pooled conclusion? (Multigroup filtering drops items not
shared by all sites, so these BICs are not comparable with the pooled table.)

```{r ladder-table}
eval_table |>
  filter(str_detect(model, "^mg_m1z_(rasch|2pl)")) |>
  arrange(itemtype, desc(invariance)) |>
  select(itemtype, invariance, converged, npar, AIC, BIC,
         `emp. rxx` = emp_rxx, `r (age)` = r_age, `retest r` = r_retest) |>
  kable(digits = 2, caption = "Invariance ladder (status items, no floors, by site)")
```

**Scalar invariance is the BIC-preferred level for both Rasch and 2PL**, and
Rasch-scalar beats 2PL-scalar overall. A likelihood-ratio test rejects exact
scalar invariance against configural (as it essentially always does at these
sample sizes), so there is *some* real cross-site DIF; but the parsimony
judgment the pipeline applies elsewhere (BIC) lands on scalar, and scores are
essentially insensitive to the level. Refitting the main structural contrast
(m0z vs. m1z vs. t1z, Rasch, scalar-by-site) leaves every pooled conclusion
intact:

```{r mg-table}
eval_table |>
  filter(model %in% c("mg_m0z_rasch_scalar", "mg_m1z_rasch_scalar",
                      "mg_t1z_rasch_scalar")) |>
  mutate(structure = fct_recode(model,
    "block × choice (multigroup scalar)" = "mg_m0z_rasch_scalar",
    "block × choice × status (multigroup scalar)" = "mg_m1z_rasch_scalar",
    "tree nodes, 1 factor (multigroup scalar)" = "mg_t1z_rasch_scalar"),
    retest = fmt_retest(r_retest, ci_lo, ci_hi)) |>
  select(structure, `emp. rxx` = emp_rxx, `r (age)` = r_age,
         `retest r [95% CI]` = retest) |>
  kable(digits = 2)
```

## Binary experiments on the leading model

### Adding feature-knowledge ("gate") items helps {#sec-gate}

The SDS task opens with warm-up blocks excluded from all prior scoring:
`same` (pick the card that is the same as a target; 2AFC) and `dimensions`
(identify a card's value on a named dimension; 3–4AFC). Conceptually these
are a *gate* on the match blocks — a child who can't identify the features
can't find feature matches — and they are near ceiling, so they carry
information mainly about the low end of the scale. Every run in the
match-block data also has gate trials, so adding them as ordinary dichotomous
items costs nothing.

To be explicit about the guessing convention, since it differs by item type
**by design**: gate items keep their **standard multiple-choice floors**
(`same` g = .5; `dimensions` g = .25, or .33 for size), because they are
genuine simultaneous forced choices where random selection is a coherent
floor model and nothing pushes children below it (all gate items sit far
above chance empirically). The match items get **g = 0**, because their
"guessing" process is refuted by the data — children perseverate below the
random-pair floor (@sec-floors). Mixed floors within one model are fully
supported by the pipeline (`chance` is a per-item column).

```{r gate-table}
eval_table |>
  filter(model %in% c("m1z_rasch", "m1z_gate_rasch",
                      "mg_m1z_gate_rasch_scalar")) |>
  mutate(model = fct_recode(model,
    "match blocks only (m1z, pooled)" = "m1z_rasch",
    "+ gate items (pooled)" = "m1z_gate_rasch",
    "+ gate items (multigroup scalar)" = "mg_m1z_gate_rasch_scalar"),
    retest = fmt_retest(r_retest, ci_lo, ci_hi)) |>
  select(model, `emp. rxx` = emp_rxx, `r (age)` = r_age,
         `retest r [95% CI]` = retest) |>
  kable(digits = 2)
```

Adding gate items improves **every** criterion — empirical reliability,
retest, and (largest of any manipulation tested) the age correlation, from
`r round(gv("m1z_rasch", "r_age"), 2)` to
`r round(gv("m1z_gate_rasch", "r_age"), 2)`. (BIC is not comparable — the
model scores additional responses.) The gate-augmented model also converges
as a multigroup scalar model, with scalar again BIC-preferred over
configural
(`r format(round(gv("mg_m1z_gate_rasch_scalar", "BIC")), big.mark = ",")` vs.
`r format(round(gv("mg_m1z_gate_rasch_config", "BIC")), big.mark = ",")`).

```{r gate-scatter, fig.width=6.5, fig.height=5}
#| fig-cap: "Scores with vs. without gate items (pooled Rasch). Gate items mostly re-order the bottom of the scale, where 'did this child know the features at all?' is informative."
gate_cmp <- scores_long |>
  filter(model %in% c("m1z_rasch", "m1z_gate_rasch"), factor == "F1") |>
  select(run_id, model, value) |>
  pivot_wider(names_from = model, values_from = value)
ggplot(gate_cmp, aes(x = m1z_rasch, y = m1z_gate_rasch)) +
  geom_abline(linetype = "dashed", colour = "grey60") +
  geom_point(alpha = 0.4, size = 0.9) +
  labs(x = "Score, match blocks only", y = "Score, match blocks + gate items",
       subtitle = glue("r = {round(cor(gate_cmp$m1z_rasch, gate_cmp$m1z_gate_rasch, use = 'complete.obs'), 3)}"))
```

One caveat: gate items are near ceiling, so their discrimination is
concentrated where few children are — the gains are real but the items mostly
sharpen the *floor* of the scale (which is exactly what a gate should do). If
the task is revised, a couple of *harder* feature-identification items would
strengthen this further.

```{r age-scatter, fig.width=9, fig.height=5}
#| fig-cap: "Recommended model (status-aware match items + gate items, no floors on match items, multigroup scalar Rasch): scores by age and dataset."
scored |>
  filter(model == "mg_m1z_gate_rasch_scalar") |>
  ggplot(aes(x = age, y = value)) +
  geom_point(aes(colour = dataset), size = 0.8, alpha = 0.6) +
  geom_smooth(method = "lm", colour = "black", linewidth = 0.7) +
  scale_colour_ptol() +
  labs(x = "Age (years)", y = "SDS score (θ)", colour = "Dataset")
```

### Disaggregating items by matched features does not help

The last level of item granularity: the specific dimensions in which a card
set contains its matching pairs (e.g., a 2match set whose pairs match on
color and size vs. one whose pairs match on shape and size). The deployed
card sets realize only a few feature signatures per block, so items become
`block × choice × status × signature`:

```{r sig-table}
eval_table |>
  filter(model %in% c("m1z_rasch", "m1z_sig_rasch")) |>
  mutate(model = fct_recode(model, "status items (m1z)" = "m1z_rasch",
                            "status × feature-signature items" = "m1z_sig_rasch"),
         retest = fmt_retest(r_retest, ci_lo, ci_hi)) |>
  select(model, BIC, `emp. rxx` = emp_rxx, `r (age)` = r_age,
         `retest r [95% CI]` = retest) |>
  kable(digits = 2)
```

There is systematic signature-level difficulty variation (same-universe BICs
are comparable here), but reliability, retest, and age validity are
unchanged — at the price of ~2.6× as many item parameters in the registry and
items that depend on deployment-specific card sets. Not worth it for scoring;
worth remembering if specific card sets ever need DIF screening.

# Cross-stack check

This chapter re-derives, on the model calibration dataset
(`../levante-pilots/01_fetched_data`), the analysis originally developed in
`levante-pilots` directly on the four per-site processed datasets (24,799
selections / 811 runs / 168 retest pairs there). Headline numbers from that
original per-site stack, for comparison with the tables above: status items
no floors — rxx .845, age r .30, retest .50; with principled floors — retest
.46; corpus-floors m0 — retest .56; + gate items — rxx .863, age r .35,
retest .53; tree factors — match retest .36, new retest .51, factor r .24.
Every qualitative conclusion — floors hurt, gate items help, Rasch ≥ 2PL,
scalar invariance by BIC, "new" stable / match unstable, signatures useless —
agrees across the two data stacks and the two recode implementations.

# Recommendations

**1. Port the (audited) corrected recode upstream into
`levantemodels::recode_sds()`** (@sec-audit, @sec-bug). The repeat-detection
fix (sorted-pair comparison) matters more than every modeling decision below.
When porting, keep the trial-indexing order (index trials *before* dropping
invalid responses).

**2. Score SDS with status-aware dichotomous match items plus the
`same`/`dimensions` gate items, no floors on the match items, Rasch,
multigroup scalar.** Concretely: `item_uid = sds_{block}_{choice}_{status}`
(status = counts of previously used matching/non-matching pairs, e.g.
`1m0n`), `correct = match & new`, `chance = 0` for match items; `same` and
`dimensions` items enter as ordinary dichotomous items with their
multiple-choice floors. This is the `mg_m1z_gate` row above — the best model
on every criterion, converging cleanly in the production multigroup-scalar
configuration.

**3. Invariance: scalar, by the pipeline's own standard** — BIC-preferred at
every rung tested, with scores insensitive to the level; DIF screening of the
status items across sites remains good hygiene once the recode fix lands.

**4. Correct the corpus chance metadata for SDS match items** (currently
0.5/0.75/0 patterns that match no guessing model) — or make the no-floor
choice explicit in the scoring spec so the metadata is ignored.

**5. Keep the tree decomposition as a research/diagnostic tool.** The
correlated 2-factor tree shows SDS's stable signal is
perseveration-resistance ("new"), not match-finding, and that the two are
nearly orthogonal. Worth (a) reporting as construct evidence, and (b) acting
on if SDS is redesigned — more selections per trial, harder match
discriminations, and a few harder feature-gate items would strengthen the
weak components.

## Limitations and open questions

- **Test–retest is one site, two intervals** (Leipzig, ~3/~7 months); scores
  come from a single calibration including both waves (standard for our
  pipeline, slightly optimistic vs. fixed-parameter scoring of wave 2).
- **Gate items are scored with their released correctness** — they don't pass
  through the match-block recode; their answer keys deserve the same
  below-chance screening as other tasks before the combined model ships
  (nothing looked anomalous here).
- The tiny `3unique` block (Bogotá only) remains excluded.
- Status-aware items condition on the child's own response history, so item
  exposure is endogenous; the tree literature treats this as unproblematic
  for node-level measurement, but item parameters shouldn't be interpreted as
  if statuses were randomly assigned.
- Guessing floors could in principle be *estimated* (3PL) rather than fixed
  or zeroed; with these sample sizes and the perseveration asymmetry, not
  worth the instability risk.
