---
title: "Sentence Understanding (TROG) — grammatical categories & cross-site DIF"
levante:
  summary: "Are TROG grammatical-category difficulty effects consistent across German, Spanish, and English, and do any items carry DIF?"
  status: draft
  task_ids: [trog]
  datasets: all
  data:
    - {source: "load_levante_scores", version: "v1_2", table: "scores"}
    - {source: "load_levante_trials", version: "v1_2", table: "trials"}
    - {source: "load_item_parameters", version: "v1_14", table: "item_parameters"}
  depends_on: [00_load_data.qmd]
  findings:
    - "Category difficulty is consistent across languages (r ≈ 0.88–0.93)."
    - "Item-level DIF is mostly modest (r ≈ 0.82–0.88)."
    - "One broken item: trog_embedding_cat_cow_chase_black (~6.5-logit DIF in German)."
  last_run: 2026-07-16
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 9
    fig-height: 6
execute:
  warning: false
  message: false
---

{{< include ../_disclaimer.qmd >}}

# Question

TROG (Test for Reception of Grammar) probes comprehension of specific
grammatical structures: each item belongs to a **grammatical category**
(`item_group`) such as negation, reversible passive, embedded/relative
clauses, plurals, prepositions, etc. — 31 categories across 99 items, scored
with a **multigroup-scalar Rasch** model (common scale across sites).

Two questions, using cross-sectional first-observation data so we can include
all three languages (German = Leipzig, Spanish = Bogotá + rural, English =
Canada):

1. Are the **grammatical-category difficulty effects consistent** across
   languages?
2. Do any **items carry DIF** (function differently across sites)?

```{r setup}
source(here::here("common.R"))
library(mirt)
sc <- read_rds(here("data/scores_all_sites.rds"))
firstrun <- sc |> filter(task_id == "trog") |>
  mutate(lang = case_when(dataset == "pilot_mpieva_de_main" ~ "German",
                          str_detect(dataset, "uniandes_co") ~ "Spanish",
                          dataset == "pilot_western_ca_main" ~ "English")) |>
  filter(!is.na(lang)) |> group_by(user_id) |> arrange(age) |> slice(1) |>
  ungroup() |> select(run_id, lang)
tr <- load_levante_trials(task_ids = "trog") |>
  inner_join(firstrun, by = "run_id") |>
  mutate(correct = as.numeric(correct), category = item_group)
```

`r firstrun |> count(lang) |> mutate(s=paste0(lang,"=",n)) |> pull(s) |> paste(collapse=", ")`.

# 1. Grammatical-category difficulty structure

Pooled item difficulties (the released multigroup-scalar Rasch), aggregated to
category.

```{r cat-difficulty, fig.height=7}
ip <- load_item_parameters() |> filter(task_id == "trog", subset == "all_items")
catdiff <- ip |> left_join(tr |> distinct(item_uid, category), by = "item_uid") |>
  group_by(category) |>
  summarise(n_items = n(), mean_diff = mean(difficulty), .groups = "drop")
catdiff |>
  ggplot(aes(x = mean_diff, y = reorder(category, mean_diff))) +
  geom_vline(xintercept = 0, lty = 3, color = "grey70") +
  geom_segment(aes(xend = 0, yend = category), color = "grey80") +
  geom_point(aes(size = n_items), color = "#268bd2") +
  labs(x = "Mean item difficulty (logits)", y = NULL, size = "# items",
       title = "TROG difficulty by grammatical category",
       subtitle = "Simple combinations easiest; embedded/conditional/comparative structures hardest")
```

# 2. Are category effects consistent across languages?

Per-language empirical accuracy by category (each child's first session),
correlated across the three languages.

```{r cat-consistency}
catacc <- tr |> group_by(lang, category) |>
  summarise(acc = mean(correct, na.rm = TRUE), n = n(), .groups = "drop") |>
  filter(n >= 30) |> select(lang, category, acc) |>
  pivot_wider(names_from = lang, values_from = acc)
cc <- catacc |> drop_na()
tibble(pair = c("German–Spanish","German–English","Spanish–English"),
       r = c(cor(cc$German, cc$Spanish), cor(cc$German, cc$English),
             cor(cc$Spanish, cc$English))) |>
  knitr::kable(digits = 2,
               caption = glue::glue("Category accuracy correlations across languages (n={nrow(cc)} categories)"))
```

```{r cat-consistency-plot, fig.height=5}
cc |> pivot_longer(c(Spanish, English), names_to = "lang", values_to = "acc") |>
  ggplot(aes(x = German, y = acc, color = lang)) +
  geom_abline(slope = 1, intercept = 0, lty = 3, color = "grey50") +
  geom_point(alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_brewer(palette = "Set1", name = NULL) +
  coord_equal() +
  labs(x = "Category accuracy — German", y = "Category accuracy — other language",
       title = "Grammatical-category difficulty is consistent across languages")
```

Category-level accuracy correlates **r ≈ 0.88–0.93** across the three
languages — the *grammatical* difficulty ordering replicates well (simple
combinations easy everywhere; embedding/conditional/comparative hard
everywhere). The construct travels.

# 3. Item-level DIF

Fit a Rasch model **separately per language** and compare item difficulties.
Differences (after centering each language's scale) are item-level DIF.

```{r dif, cache=TRUE}
fit_site <- function(L) {
  m <- tr |> filter(lang == L) |> distinct(run_id, item_uid, .keep_all = TRUE) |>
    select(run_id, item_uid, correct) |>
    pivot_wider(names_from = item_uid, values_from = correct) |>
    select(-run_id) |> as.matrix()
  m <- m[, colMeans(!is.na(m)) > 0.05]
  m <- m[, apply(m, 2, \(x) length(unique(x[!is.na(x)])) > 1)]
  fit <- mirt(m, 1, itemtype = "Rasch", verbose = FALSE, technical = list(NCYCLES = 2000))
  d <- coef(fit, simplify = TRUE)$items[, "d"]
  tibble(item_uid = names(d), b = -d, lang = L)
}
bdiff <- map_dfr(c("German","Spanish","English"), fit_site) |>
  group_by(lang) |> mutate(b_c = b - mean(b)) |> ungroup()  # center per language
wide_b <- bdiff |> select(item_uid, lang, b_c) |>
  pivot_wider(names_from = lang, values_from = b_c) |>
  left_join(tr |> distinct(item_uid, category), by = "item_uid")

wb <- wide_b |> select(German, Spanish, English) |> drop_na()
tibble(pair = c("German–Spanish","German–English","Spanish–English"),
       r = c(cor(wb$German, wb$Spanish), cor(wb$German, wb$English),
             cor(wb$Spanish, wb$English))) |>
  knitr::kable(digits = 2,
               caption = glue::glue("Item-level difficulty correlations across languages (n={nrow(wb)} items)"))
```

```{r dif-scatter, fig.height=5.5}
wide_b |> filter(!is.na(German), !is.na(Spanish)) |>
  mutate(gap = German - Spanish) |>
  ggplot(aes(x = Spanish, y = German)) +
  geom_abline(slope = 1, intercept = 0, lty = 3, color = "grey50") +
  geom_point(alpha = 0.6, color = "#268bd2") +
  ggrepel::geom_text_repel(data = ~filter(.x, abs(gap) > 1.1),
    aes(label = str_remove(item_uid, "^trog_")), size = 2.5, color = "firebrick",
    max.overlaps = 20) +
  labs(x = "Item difficulty — Spanish (centered)", y = "Item difficulty — German (centered)",
       title = "Item-level DIF: German vs Spanish",
       subtitle = "On-diagonal = invariant. Labeled (red) = |DIF| > 1.1 logits.")
```

```{r dif-top}
wide_b |> filter(!is.na(German), !is.na(Spanish)) |>
  mutate(`DE−ES gap` = German - Spanish) |>
  arrange(desc(abs(`DE−ES gap`))) |>
  select(item_uid, category, German, Spanish, `DE−ES gap`) |>
  head(10) |> knitr::kable(digits = 2, caption = "Largest German-vs-Spanish item DIF")
```

Item difficulties correlate **r ≈ 0.82–0.88** across languages — good, but
lower than the category level, i.e. translation introduces some item-specific
DIF (expected; TROG reached *scalar* invariance overall, not perfect). Most DIF
is modest (≈ 1–1.5 logits), **except one item that is clearly broken:**

```{r broken-item}
broken <- "trog_embedding_cat_cow_chase_black"
tr |> filter(item_uid == broken) |>
  group_by(lang) |>
  summarise(n = n(), accuracy = round(mean(correct, na.rm = TRUE), 2), .groups = "drop") |>
  knitr::kable(caption = glue::glue("{broken}: accuracy by language"))
```

`trog_embedding_cat_cow_chase_black` shows a **~6.5-logit** German-vs-Spanish
difficulty gap — far beyond any other item. The accuracy breakdown shows it
behaves completely differently in one language, the signature of a
mistranslation or wrong answer key. **This item should be reviewed and likely
excluded/fixed.**

# Verdict

- **Grammatical-category effects are consistent across German, Spanish, and
  English** (category accuracy r ≈ 0.88–0.93): TROG measures the same
  difficulty ordering of structures everywhere — good cross-linguistic
  construct validity.
- **Item-level DIF is mostly modest** (r ≈ 0.82–0.88), consistent with the
  task's scalar-invariance status; translation adds some item-specific noise.
- **One item is clearly broken** — `trog_embedding_cat_cow_chase_black`
  (~6.5-logit DIF) — flag for review/exclusion.
- No category systematically carries DIF; the issues are item-specific.

```{r save}
write_rds(wide_b, here::here("data", "trog_item_dif.rds"))
```
