---
title: "ROAR-Sentence (SRE) — scoring deepdive"
levante:
  summary: "ROAR-Sentence scoring deep dive: timing forensics, rate-normalized scoring, and efficiency models (EZ, zero-fill IRT, hierarchical SA, diffusion IRT)."
  status: draft
  task_ids: [sre]
  datasets: [pilot_mpieva_de_main, pilot_uniandes_co_bogota, pilot_uniandes_co_rural, pilot_western_ca_main]
  data:
    - {source: "../levante-pilots/01_fetched_data/task_data_nested.rds", table: "trials"}
    - {source: "../levante-pilots/01_fetched_data/run_data.rds", table: "runs"}
    - {source: "../levante-pilots/01_fetched_data/participants.rds", table: "participants"}
    - {source: "../levante-pilots/02_scoring_outputs/scores/registry_scores.rds", table: "scores"}
    - {source: "levante_data_latest:e9pf", version: "v1_2", table: "scores"}
  depends_on: [tasks/_fit_sre_zerofill_irt.R, tasks/_fit_sre_hierarchical_sa.R, tasks/_fit_sre_ddm.R]
  findings:
    - "score_sre() truncates at the first 180 trials, not 180 seconds; the filter only bites in Canada, where 20 runs contain two ~180-s blocks."
    - "Rate normalization barely changes scores in DE and Colombia (r ≈ .99) but does in Canada (r = .82)."
    - "The corrected rate score beats every model-based efficiency alternative on most criteria."
    - "Germany retest and diffusion-IRT columns render as NA without local caches."
  last_run: 2026-07-22
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 9
    fig-height: 5
execute:
  warning: false
  message: false
---

{{< include ../_disclaimer.qmd >}}

# Scope

**Sentence Reading Efficiency (SRE)** is a speeded literacy measure adapted
from ROAR ([intro-sre](https://roar.stanford.edu/technical/intro-sre.html);
Yeatman et al., 2024). Children read simple sentences one at a time and judge
each *true* or *false* (2AFC), reading as many as they can in **180 seconds**.
ROAR's official score is **correct − incorrect** within the time window — the
subtraction is a guessing correction (a random guesser nets ≈ 0). LEVANTE
z-scores this count within dataset so it sits on a roughly comparable scale to
the IRT tasks (`score_type = "guessing_adjusted_number_correct_scaled"`).

**The scoring bug.** `levantemodels::score_sre()` computes
`filter(trial_number <= 180)` then `(sum(correct) - sum(!correct)) / 180` —
i.e. it truncates at the first 180 **trials**, not 180 **seconds**. (The
`/180` divisor is cosmetic: it cancels in the z-score.) This notebook asks:

1. **Timing forensics**: do runs actually last 180 seconds? (Spoiler: no —
   there is real structure here, including double-length runs.)
2. **Corrected scoring**: rate-normalize to a per-180s net-correct rate. How
   much do scores change? Do age / ROAR-Phoneme (PA) / ROAR-Word (SWR)
   correlations change, within and across sites?
3. **Efficiency scoring**: discussion of correctness+RT models
   (drift-diffusion IRT and alternatives) plus an EZ-diffusion proof of
   concept on the trial data.

Data: the model calibration dataset
(**`../levante-pilots/01_fetched_data/task_data_nested.rds`**, the same
source used by `tasks/mrot/`), plus age from `run_data.rds` and
PA/SWR/Vocab/TROG validity criteria from the pre-computed
`../levante-pilots/02_scoring_outputs/scores/registry_scores.rds`
(`model_set == "by_language"`, `itemtype == "rasch"`).

```{r setup}
source(here::here("common.R"))

pilots_data_dir <- here("..", "levante-pilots", "01_fetched_data")
task_data_nested <- read_rds(file.path(pilots_data_dir, "task_data_nested.rds"))
run_data <- read_rds(file.path(pilots_data_dir, "run_data.rds"))
participants <- read_rds(file.path(pilots_data_dir, "participants.rds"))

registry_scores <- read_rds(here("..", "levante-pilots", "02_scoring_outputs",
                                  "scores", "registry_scores.rds")) |>
  filter(model_set == "by_language", itemtype == "rasch") |>
  select(task, run_id, ability)

dataset_pal <- set_names(
  RColorBrewer::brewer.pal(4, "Set1"),
  dataset_labels_named[c("pilot_uniandes_co_bogota", "pilot_uniandes_co_rural",
                         "pilot_mpieva_de_main", "pilot_western_ca_main")])

label_dataset <- \(df) df |>
  mutate(dataset_label = factor(dataset_labels_named[dataset],
                                levels = names(dataset_pal)))

sre <- task_data_nested |>
  filter(item_task == "sre") |>
  unnest(data) |>
  mutate(correct = as.logical(correct),
         rt_s = rt_numeric / 1000,
         corpus = sub("_[0-9]+$", "", item_uid) |> str_remove("^sre_")) |>
  label_dataset()

# no independent published SRE score exists in the calibration dataset (that
# was `sc_all$score` from the hackathon `get_scores()` pull) -- run metadata
# (age, time_started) comes from run_data instead
sre_scores_pub <- sre |>
  distinct(dataset, run_id, user_id) |>
  left_join(run_data |> select(run_id, age, time_started), by = "run_id") |>
  label_dataset()
```

```{r data-counts}
sre |>
  group_by(dataset_label) |>
  summarise(runs = n_distinct(run_id), children = n_distinct(user_id),
            trials = n(), .groups = "drop") |>
  knitr::kable(caption = "SRE trial data by dataset (model calibration dataset)")
```

# 1. Timing forensics: is 180 seconds always 180 seconds?

Two per-run duration measures: **time-on-sentences** (cumulative RT — the
task records per-sentence RT in ms) and **wall-clock span** (last minus first
trial server timestamp). Neither is exactly the timer, but they bracket it:
cumulative RT omits inter-trial overhead; timestamp span omits the first
trial's duration and can inflate under server lag or pauses.

```{r durations}
run_time <- sre |>
  group_by(dataset, dataset_label, run_id, user_id) |>
  summarise(n_trials = n(),
            n_corpora = n_distinct(corpus),
            cum_rt = sum(rt_s),
            span = as.numeric(difftime(max(timestamp), min(timestamp),
                                       units = "secs")),
            n_correct = sum(correct), n_incorrect = sum(!correct),
            .groups = "drop")

run_time |>
  group_by(dataset_label) |>
  summarise(across(c(cum_rt, span),
                   list(q05 = ~quantile(.x, .05), median = median,
                        q95 = ~quantile(.x, .95), max = max)),
            .groups = "drop") |>
  knitr::kable(digits = 0,
               caption = "Per-run duration (seconds): cumulative RT and wall-clock span")
```

```{r duration-hist, fig.height=4}
run_time |>
  ggplot(aes(x = cum_rt, fill = dataset_label)) +
  facet_wrap(vars(dataset_label), scales = "free_y") +
  geom_histogram(binwidth = 10, boundary = 0, show.legend = FALSE) +
  geom_vline(xintercept = 180, lty = 2) +
  scale_fill_manual(values = dataset_pal) +
  labs(x = "Cumulative RT per run (s)", y = "Runs",
       title = "Most runs sum to just under 180 s of reading time",
       subtitle = "Dashed line = nominal 180 s deadline. Note the Canadian runs near 330 s.")
```

Three regularities:

- **The 180 s deadline is real but not exact.** Median time-on-sentences is
  ~160–165 s (the remainder is inter-trial overhead: wall-clock span medians
  ~180 s). So the timer fires on schedule for most runs.
- **A left tail of short runs exists everywhere** — Colombia especially (rural
  5th percentile ≈ 89 s of reading time). Under the current rule these
  short runs are scored as if the child had the full window.
- **Canada (Western) has genuinely double-length runs.**

## Block structure: some Canadian runs contain two 180-s blocks

The item bank prefix identifies sentence corpora. Colombia interleaves three
corpora (`spanish_lab`, `spanish_ai`, `spanish_ai-human`) *within one timed
block*; Germany uses one corpus (`uzh`). But 20 Canadian runs contain **two
sequential blocks** (`lab` sentences, then `ai` sentences), each with its own
~180-s window — roughly 360 s of testing scored as if it were one window.

```{r blocks}
run_blocks <- sre |>
  group_by(dataset_label, run_id, corpus) |>
  summarise(block_rt = sum(rt_s), n = n(), .groups = "drop_last") |>
  summarise(blocks = paste(sort(corpus), collapse = " + "),
            two_windows = sum(block_rt > 120) >= 2,
            .groups = "drop")

run_blocks |>
  count(dataset_label, blocks, two_windows) |>
  arrange(dataset_label, desc(n)) |>
  knitr::kable(caption = "Corpus composition of runs. two_windows = ≥2 corpora each with >120 s of reading time (sequential blocks, not interleaving)")
```

```{r trials-per-run, fig.height=4}
run_time |>
  ggplot(aes(x = n_trials, fill = dataset_label)) +
  facet_wrap(vars(dataset_label), scales = "free_y") +
  geom_histogram(binwidth = 10, boundary = 0, show.legend = FALSE) +
  geom_vline(xintercept = 180, lty = 2) +
  scale_fill_manual(values = dataset_pal) +
  labs(x = "Trials per run", y = "Runs",
       title = "Only Canada ever reaches the 180-trial truncation point",
       subtitle = "Dashed line = the current rule's truncation at 180 trials")
```

**Side-quest verdict:** the first-180-*trials* filter is a no-op for Germany
and Colombia (no run exceeds 180 trials — max 133). It bites only in Canada
(`r sum(run_time$n_trials > 180)` runs exceed 180 trials, max
`r max(run_time$n_trials)`), and there it truncates arbitrarily *across* the
two-block structure. Meanwhile the actual problem is the reverse: **effective
testing time varies from ~1 to ~4 minutes across runs**, and a raw count
treats them all as equal.

# 2. Corrected scoring: rate normalization

The corrected score is a **rate**: net correct per 180 s of reading time,

$$\text{score} = \frac{C - I}{\text{cumulative RT (s)}} \times 180,$$

then z-scored within dataset (preserving the pipeline's scaling convention).
We use time-on-sentences rather than wall-clock span as the denominator
because it is robust to server-timestamp lag and pauses; the two agree
closely (sensitivity check below). Runs with under 30 s of reading time are
excluded from rate scoring (extrapolation too aggressive).

First, verify we can reproduce the published (buggy) scores exactly:

The calibration dataset has no separately-computed production SRE score to
reproduce-check against (the hackathon pull's `sc_all$score` had no
counterpart here), so `score_old` below is simply the current buggy rule
(`filter(trial_number <= 180)`, `(C - I)` z-scored within dataset) applied
directly to the trial data.

```{r replicate-published}
old_score <- sre |>
  group_by(dataset, run_id) |>
  filter(trial_number <= 180) |>
  summarise(net = sum(correct) - sum(!correct), .groups = "drop") |>
  group_by(dataset) |>
  mutate(score_old = scale(net)[, 1]) |>
  ungroup()
```

```{r corrected-score}
min_time <- 30

new_score <- run_time |>
  filter(cum_rt >= min_time) |>
  mutate(rate = (n_correct - n_incorrect) / cum_rt * 180) |>
  group_by(dataset) |>
  mutate(score_new = scale(rate)[, 1]) |>
  ungroup()

scores_cmp <- new_score |>
  inner_join(old_score |> select(dataset, run_id, net, score_old),
             by = c("dataset", "run_id")) |>
  left_join(sre_scores_pub |> select(dataset, run_id, age),
            by = c("dataset", "run_id"))
```

Excluded by the 30-s floor: `r nrow(run_time) - nrow(new_score)` runs.

## A) How much do scores change?

```{r old-vs-new, fig.height=4.5}
cors <- scores_cmp |>
  group_by(dataset_label) |>
  summarise(n = n(), r = cor(score_old, score_new),
            r_spearman = cor(score_old, score_new, method = "spearman"),
            mean_abs_change = mean(abs(score_new - score_old)),
            max_abs_change = max(abs(score_new - score_old)),
            .groups = "drop")

knitr::kable(cors, digits = 2,
             caption = "Current (first-180-trials) vs corrected (rate per 180 s) z-scores")

scores_cmp |>
  ggplot(aes(x = score_old, y = score_new, color = dataset_label)) +
  facet_wrap(vars(dataset_label)) +
  geom_abline(lty = 3, color = "grey60") +
  geom_point(alpha = 0.4, size = 0.8, show.legend = FALSE) +
  scale_color_manual(values = dataset_pal) +
  labs(x = "Current score (z)", y = "Corrected rate score (z)",
       title = "Current vs corrected SRE scores")
```

```{r movers}
scores_cmp |>
  mutate(change = score_new - score_old) |>
  slice_max(abs(change), n = 10) |>
  select(dataset_label, run_id, n_trials, cum_rt, n_correct, n_incorrect,
         score_old, score_new, change) |>
  knitr::kable(digits = 2, caption = "Largest score changes under rate normalization")
```

```{r change-vs-time, fig.height=4}
scores_cmp |>
  mutate(change = score_new - score_old) |>
  ggplot(aes(x = cum_rt, y = change, color = dataset_label)) +
  geom_hline(yintercept = 0, lty = 3) +
  geom_vline(xintercept = 180, lty = 2, color = "grey60") +
  geom_point(alpha = 0.5, size = 0.9) +
  scale_color_manual(values = dataset_pal, name = NULL) +
  labs(x = "Cumulative RT per run (s)", y = "Corrected − current (z)",
       title = "Who moves: short runs move up, double-length runs move down") +
  theme(legend.position = "bottom")
```

## B) Validity correlations, current vs corrected

Within each dataset: correlations of SRE with age, ROAR-Phoneme (θ), and
ROAR-Word (θ). Cross-run joins take each child's first run per task.

```{r validity}
first_run <- \(df) df |>
  group_by(dataset, user_id) |>
  arrange(time_started) |>
  slice(1) |>
  ungroup()

pa_swr <- registry_scores |>
  filter(task %in% c("pa", "swr"), !is.na(ability)) |>
  left_join(run_data |> select(run_id, user_id, time_started), by = "run_id") |>
  left_join(participants |> select(user_id, dataset), by = "user_id") |>
  filter(!is.na(dataset)) |>
  first_run() |>
  select(dataset, user_id, task, other = ability) |>
  pivot_wider(names_from = task, values_from = other)

sre_first <- sre_scores_pub |>
  first_run() |>
  select(dataset, dataset_label, user_id, run_id, age) |>
  inner_join(scores_cmp |> select(dataset, run_id, score_old, score_new),
             by = c("dataset", "run_id")) |>
  left_join(pa_swr, by = c("dataset", "user_id"))

validity <- sre_first |>
  pivot_longer(c(age, pa, swr), names_to = "criterion", values_to = "crit") |>
  filter(!is.na(crit)) |>
  group_by(dataset_label, criterion) |>
  filter(n() >= 20) |>
  summarise(n = n(),
            r_current = cor(score_old, crit),
            r_corrected = cor(score_new, crit), .groups = "drop") |>
  mutate(delta = r_corrected - r_current)

knitr::kable(validity, digits = 2,
             caption = "SRE validity correlations within dataset, current vs corrected scoring")
```

```{r validity-pooled}
pooled <- sre_first |>
  pivot_longer(c(age, pa, swr), names_to = "criterion", values_to = "crit") |>
  filter(!is.na(crit)) |>
  group_by(criterion) |>
  summarise(n = n(),
            r_current = cor(score_old, crit),
            r_corrected = cor(score_new, crit), .groups = "drop") |>
  mutate(delta = r_corrected - r_current)

knitr::kable(pooled, digits = 2,
             caption = "Pooled across datasets (all scores z-scaled within dataset; PA/SWR θ are by-language, so pooled r mixes calibrations — interpret within-site values first)")
```

## Longitudinal stability (Bogotá, ~16-month gap)

Bogotá has ~180 children with SRE at two waves. Rank-order stability across
16 months is a demanding criterion: a better score should preserve more
true-score variance.

```{r stability}
stab <- sre_scores_pub |>
  filter(dataset == "pilot_uniandes_co_bogota") |>
  inner_join(scores_cmp |> select(dataset, run_id, score_old, score_new),
             by = c("dataset", "run_id")) |>
  group_by(user_id) |>
  filter(n() >= 2) |>
  arrange(time_started) |>
  summarise(gap_days = as.numeric(difftime(time_started[2], time_started[1],
                                           units = "days")),
            across(c(score_old, score_new), list(t1 = ~.x[1], t2 = ~.x[2])),
            .groups = "drop") |>
  filter(gap_days > 300)

tibble(
  n_pairs = nrow(stab),
  median_gap_months = round(median(stab$gap_days) / 30.4, 1),
  stability_current = cor(stab$score_old_t1, stab$score_old_t2),
  stability_corrected = cor(stab$score_new_t1, stab$score_new_t2)
) |>
  knitr::kable(digits = 2, caption = "16-month rank-order stability, current vs corrected")
```

## Sensitivity: denominator choice

```{r denominator-sensitivity}
sens <- run_time |>
  filter(cum_rt >= min_time, span >= min_time) |>
  mutate(rate_rt = (n_correct - n_incorrect) / cum_rt * 180,
         rate_span = (n_correct - n_incorrect) / span * 180) |>
  group_by(dataset_label) |>
  summarise(n = n(), r_rt_vs_span = cor(rate_rt, rate_span), .groups = "drop")

knitr::kable(sens, digits = 3,
             caption = "Rate via cumulative RT vs via wall-clock span")
```

# 3. Fast guessing: the bridge to model-based scoring

The `C − I` correction neutralizes guessing *in expectation* (a guesser nets
≈ 0), but guessing still costs signal: guessed trials burn testing time and
add noise. And guessing is far from uniform across sites:

```{r fast-guessing, fig.height=4}
sre |>
  mutate(rt_bin = cut(rt_s, c(0, .5, 1, 2, 4, 8, 16, Inf))) |>
  group_by(dataset_label, rt_bin) |>
  summarise(accuracy = mean(correct), pct = n(), .groups = "drop_last") |>
  mutate(pct = pct / sum(pct)) |>
  ungroup() |>
  ggplot(aes(x = rt_bin, y = accuracy, color = dataset_label,
             group = dataset_label)) +
  geom_hline(yintercept = 0.5, lty = 3) +
  geom_line() +
  geom_point(aes(size = pct), alpha = 0.7) +
  scale_color_manual(values = dataset_pal, name = NULL) +
  scale_size_area(max_size = 6, guide = "none") +
  labs(x = "RT (s)", y = "Accuracy",
       title = "Sub-second responses are at chance — and their prevalence varies hugely by site",
       subtitle = "Point size ∝ share of site's trials in that RT bin; dotted line = 2AFC chance") +
  theme(legend.position = "bottom")
```

```{r guessing-table}
sre |>
  group_by(dataset_label) |>
  summarise(pct_rt_under_1s = mean(rt_s < 1),
            acc_under_1s = mean(correct[rt_s < 1]),
            acc_over_1s = mean(correct[rt_s >= 1]), .groups = "drop") |>
  knitr::kable(digits = 2, caption = "Fast-guessing prevalence by dataset")
```

This is a strong argument for scoring models that use the joint
(correct, RT) signal per trial, rather than counts alone.

# 4. Toward an efficiency score from correctness + RT

## The modeling landscape

SRE's design — self-paced 2AFC trials under a global deadline, with
per-trial RT and accuracy — admits several principled treatments:

1. **Rate normalization (implemented above).** Minimal fix: net correct per
   unit time. Already a speed–accuracy composite; requires no model. Its
   weakness: it treats a slow-and-accurate child and a fast-and-sloppy child
   with equal net rates as identical, and fast guessing adds noise it can
   only average over.

2. **Count models (Rasch Poisson counts / offset-Poisson GLMM).** The
   classical psychometrics of speed tests (Rasch, 1960): model the *count* of
   correct responses with a log link and `offset(log(time))`, giving each
   child a latent log-rate; corpus/site enter as fixed effects, children as
   random effects. This is rate normalization done as a measurement model —
   it handles varying time windows natively and yields standard errors. It
   still ignores per-trial RT structure.

3. **Diffusion IRT (van der Maas et al., 2011; Molenaar et al., 2015;
   `diffIRT`, or GLMM approximations).** Each trial's (choice, RT) is a
   Wiener first-passage process; person drift rate $v$ *is* reading
   efficiency, boundary separation $a$ is caution, $T_{er}$ is encoding/motor
   time. Items get drift/boundary loadings. This cleanly separates ability
   from speed–accuracy tradeoff and soaks up fast guesses (they are
   low-drift/low-boundary behavior, not ability). Cost: heavy estimation;
   item-level calibration needs coverage (fine in Colombia at ~60–80 runs per
   item; thin in Canada at ~9).

4. **Hierarchical speed–accuracy models (van der Linden, 2007).** Separate
   IRT model for accuracy and lognormal model for RT, with correlated person
   ability θ and speed τ. Less mechanistic than the diffusion model; scales
   well; efficiency is a composite of θ and τ.

5. **Signed residual time scoring (Maris & van der Maas, 2012).** Score
   $(2x - 1)(d - t)$ per item — correct-and-fast is best, wrong-and-fast is
   worst. Elegant and provably optimal under a specific IRT model, but it
   assumes a *per-item* deadline $d$; SRE has only a global deadline, so $d$
   would be an arbitrary pseudo-deadline. Not a natural fit.

## Zero-filled IRT (ROAR-team suggestion)

The ROAR team has begun exploring a simple item-based approach: **fill the
response matrix with zeros for unreached items, then fit a standard IRT
model**. Speed enters the score through the block of zeros — a fast accurate
reader clears more of the bank. We fit one dense Rasch model per item bank
(Spanish: Bogotá + rural jointly, 1,382 runs × 376 items; German: 151 × 200;
English: 91 × 530 after dropping 18 never-reached items;
[`_fit_sre_zerofill_irt.R`](\_fit_sre_zerofill_irt.R)).

```{r zerofill}
zf <- read_rds(here("data/sre_zerofill_scores.rds")) |>
  group_by(dataset) |>
  mutate(zf_z = scale(theta_zerofill)[, 1]) |>
  ungroup()

zf_first <- sre_first |>
  inner_join(zf |> select(dataset, run_id, zf_z), by = c("dataset", "run_id"))

zf_first |>
  pivot_longer(c(age, pa, swr), names_to = "criterion", values_to = "crit") |>
  filter(!is.na(crit)) |>
  group_by(dataset_label, criterion) |>
  filter(n() >= 20) |>
  summarise(n = n(),
            r_rate = cor(score_new, crit),
            r_zerofill = cor(zf_z, crit), .groups = "drop") |>
  knitr::kable(digits = 2,
               caption = "Zero-filled Rasch vs corrected rate score: validity")
```

```{r zerofill-stability}
zf_stab <- sre_scores_pub |>
  filter(dataset == "pilot_uniandes_co_bogota") |>
  inner_join(zf |> select(dataset, run_id, zf_z), by = c("dataset", "run_id")) |>
  inner_join(scores_cmp |> select(dataset, run_id, score_new),
             by = c("dataset", "run_id")) |>
  group_by(user_id) |>
  filter(n() >= 2) |>
  arrange(time_started) |>
  summarise(gap = as.numeric(difftime(time_started[2], time_started[1],
                                      units = "days")),
            across(c(zf_z, score_new), list(t1 = ~.x[1], t2 = ~.x[2])),
            .groups = "drop") |>
  filter(gap > 300)

tibble(n_pairs = nrow(zf_stab),
       rate = cor(zf_stab$score_new_t1, zf_stab$score_new_t2),
       zerofill = cor(zf_stab$zf_z_t1, zf_stab$zf_z_t2)) |>
  knitr::kable(digits = 2,
               caption = "Bogotá 16-month stability: zero-filled Rasch vs corrected rate")
```

```{r zerofill-guessing}
guess_share <- sre |>
  group_by(dataset, run_id) |>
  summarise(pct_fast = mean(rt_s < 1), .groups = "drop")

zf_first |>
  inner_join(guess_share, by = c("dataset", "run_id")) |>
  group_by(dataset_label) |>
  summarise(n = n(),
            r_rate_fastguess = cor(score_new, pct_fast),
            r_zerofill_fastguess = cor(zf_z, pct_fast), .groups = "drop") |>
  knitr::kable(digits = 2,
               caption = "Correlation of each score with a run's fast-guess share (RT < 1 s). Positive = the score REWARDS guessing")
```

Where behavior is clean, zero-fill is essentially the rate score by another
route: r = .94 with the corrected rate in Germany, near-identical validity
there, and Bogotá 16-month stability a statistical tie (.72 vs .72). But it
has **no guessing correction**: a fast guesser converts wasted trials into
lucky 1s among the zeros, so the score correlates **positively** with a
run's fast-guess share (+.46 in rural Colombia) where the rate score
correlates negatively (−.42), and validity collapses exactly there (rural
PA r = .27 vs .57; rural age .19 vs .42).

The obvious repair — a fixed 2AFC guessing floor (Rasch with g = .5, as in
the LEVANTE multiple-choice models) — **backfires**: unreached items are
*structural* zeros, and whole-bank accuracy (~10% for a typical run) sits
far below any g = .5 floor, so the model cannot represent the data (EM
fails to converge; validity collapses everywhere, e.g. rural age r = .08).
The floor is right for "attempted and guessed," wrong for "never reached."
A repaired variant would need a different treatment of the two zero types —
at which point the model is no longer the simple suggestion, and the rate
score already does the job.

## Proof of concept: EZ-diffusion per run

The EZ-diffusion model (Wagenmakers, van der Maas, & Grasman, 2007) inverts
the simple diffusion model from three per-person summary statistics
(accuracy, mean and variance of correct RTs) — no item structure, no
iterative fitting. It is the cheapest possible test of whether
diffusion-style decomposition adds validity beyond the rate score.

```{r ez}
ez_diffusion <- function(pc, vrt, mrt, s = 0.1) {
  s2 <- s^2
  L <- qlogis(pc)
  x <- L * (L * pc^2 - L * pc + pc - 0.5) / vrt
  v <- sign(pc - 0.5) * s * x^(1 / 4)
  a <- s2 * L / v
  y <- -v * a / s2
  mdt <- (a / (2 * v)) * (1 - exp(y)) / (1 + exp(y))
  tibble(drift = v, boundary = a, ter = mrt - mdt)
}

ez_run <- sre |>
  filter(rt_s >= 0.25, rt_s <= 30) |>
  group_by(dataset, dataset_label, run_id, user_id) |>
  filter(n() >= 15, sum(correct) >= 10) |>
  summarise(n = n(), pc_raw = mean(correct),
            mrt = mean(rt_s[correct]), vrt = var(rt_s[correct]),
            .groups = "drop") |>
  # edge-correct accuracy away from 0.5 and 1
  mutate(pc = pmin(pmax(pc_raw, 0.5 + 1 / (2 * n)), 1 - 1 / (2 * n)))

ez_run <- bind_cols(ez_run, ez_diffusion(ez_run$pc, ez_run$vrt, ez_run$mrt))

ez_run |>
  group_by(dataset_label) |>
  summarise(n_runs = n(), across(c(drift, boundary, ter), median),
            .groups = "drop") |>
  knitr::kable(digits = 2, caption = "Median EZ-diffusion parameters by dataset")
```

```{r ez-validity}
ez_first <- sre_first |>
  inner_join(ez_run |> select(dataset, run_id, drift, boundary, ter),
             by = c("dataset", "run_id"))

ez_validity <- ez_first |>
  pivot_longer(c(age, pa, swr), names_to = "criterion", values_to = "crit") |>
  filter(!is.na(crit)) |>
  group_by(dataset_label, criterion) |>
  filter(n() >= 20) |>
  summarise(n = n(),
            r_corrected_rate = cor(score_new, crit),
            r_drift = cor(drift, crit),
            r_boundary = cor(boundary, crit), .groups = "drop")

knitr::kable(ez_validity, digits = 2,
             caption = "EZ drift vs corrected rate score: validity correlations")
```

```{r ez-stability}
ez_stab <- sre_scores_pub |>
  filter(dataset == "pilot_uniandes_co_bogota") |>
  inner_join(ez_run |> select(dataset, run_id, drift),
             by = c("dataset", "run_id")) |>
  inner_join(scores_cmp |> select(dataset, run_id, score_new),
             by = c("dataset", "run_id")) |>
  group_by(user_id) |>
  filter(n() >= 2) |>
  arrange(time_started) |>
  summarise(gap = as.numeric(difftime(time_started[2], time_started[1],
                                      units = "days")),
            drift_t1 = drift[1], drift_t2 = drift[2],
            rate_t1 = score_new[1], rate_t2 = score_new[2], .groups = "drop") |>
  filter(gap > 300)

tibble(n_pairs = nrow(ez_stab),
       stability_rate = cor(ez_stab$rate_t1, ez_stab$rate_t2),
       stability_drift = cor(ez_stab$drift_t1, ez_stab$drift_t2)) |>
  knitr::kable(digits = 2,
               caption = "Bogotá 16-month stability: corrected rate vs EZ drift")
```

```{r ez-scatter, fig.height=4.5}
ez_first |>
  ggplot(aes(x = score_new, y = drift, color = dataset_label)) +
  facet_wrap(vars(dataset_label), scales = "free_y") +
  geom_point(alpha = 0.4, size = 0.8, show.legend = FALSE) +
  scale_color_manual(values = dataset_pal) +
  labs(x = "Corrected rate score (z)", y = "EZ drift rate v",
       title = "EZ drift vs the corrected rate score")
```

## Hierarchical speed–accuracy model (van der Linden 2007 style)

The next rung up from EZ: model accuracy and RT jointly but *separately*,
with crossed run × item random effects (fit by
[`_fit_sre_hierarchical_sa.R`](\_fit_sre_hierarchical_sa.R)):

- accuracy: `correct ~ 0 + dataset + (1|run_id) + (1|item_uid)` (logit GLMM)
  → run-level **ability θ**, with item easiness absorbed by item effects
  (the "IRT part" that EZ lacks);
- speed: `log(rt) ~ 0 + dataset + (1|run_id) + (1|item_uid)` (lognormal LMM)
  → run-level **speed τ** (sign-flipped RT effect), with item
  time-intensities absorbed.

An efficiency composite is z(θ) + z(τ), standardized within dataset.

```{r sa-load}
sa_file <- here("data/sre_sa_estimates.rds")
sa <- read_rds(sa_file)

sa_run <- sa$run |>
  group_by(dataset) |>
  mutate(theta_z = scale(theta)[, 1], tau_z = scale(tau)[, 1],
         eff = scale(theta_z + tau_z)[, 1]) |>
  ungroup()

sa_run |>
  label_dataset() |>
  group_by(dataset_label) |>
  summarise(n_runs = n(), r_theta_tau = cor(theta, tau), .groups = "drop") |>
  knitr::kable(digits = 2,
               caption = "Ability–speed correlation by dataset: near zero in DE/Canada, strongly NEGATIVE in rural Colombia — 'fast' there mostly means guessing")
```

```{r sa-validity}
sa_first <- sre_first |>
  inner_join(sa_run |> select(dataset, run_id, theta_z, tau_z, eff),
             by = c("dataset", "run_id"))

sa_first |>
  pivot_longer(c(age, pa, swr), names_to = "criterion", values_to = "crit") |>
  filter(!is.na(crit)) |>
  group_by(dataset_label, criterion) |>
  filter(n() >= 20) |>
  summarise(n = n(),
            r_rate = cor(score_new, crit),
            r_theta = cor(theta_z, crit),
            r_tau = cor(tau_z, crit),
            r_composite = cor(eff, crit), .groups = "drop") |>
  knitr::kable(digits = 2,
               caption = "Hierarchical SA model vs corrected rate score: validity")
```

```{r sa-stability}
sa_stab <- sre_scores_pub |>
  filter(dataset == "pilot_uniandes_co_bogota") |>
  inner_join(sa_run |> select(dataset, run_id, theta_z, tau_z, eff),
             by = c("dataset", "run_id")) |>
  inner_join(scores_cmp |> select(dataset, run_id, score_new),
             by = c("dataset", "run_id")) |>
  group_by(user_id) |>
  filter(n() >= 2) |>
  arrange(time_started) |>
  summarise(gap = as.numeric(difftime(time_started[2], time_started[1],
                                      units = "days")),
            across(c(theta_z, tau_z, eff, score_new),
                   list(t1 = ~.x[1], t2 = ~.x[2])), .groups = "drop") |>
  filter(gap > 300)

tibble(n_pairs = nrow(sa_stab),
       rate = cor(sa_stab$score_new_t1, sa_stab$score_new_t2),
       theta = cor(sa_stab$theta_z_t1, sa_stab$theta_z_t2),
       tau = cor(sa_stab$tau_z_t1, sa_stab$tau_z_t2),
       composite = cor(sa_stab$eff_t1, sa_stab$eff_t2)) |>
  knitr::kable(digits = 2,
               caption = "Bogotá 16-month stability: SA-model scores vs corrected rate")
```

The decomposition is diagnostic gold but scoring lead. The ability–speed
correlation flips sign across sites (≈ 0 in DE/Canada, −.08 Bogotá, **−.43
rural CO**): where fast guessing is common, the "speed" dimension is
contaminated by disengagement, so neither τ nor any fixed θ+τ composite
travels across sites. The composite matches the rate score where behavior is
clean (r ≈ .9 in DE/Canada) and underperforms it where it isn't (rural age
validity .30 vs .44; Bogotá stability .71 vs .74).

## Diffusion IRT (pooled brms Wiener model)

::: {.callout-warning}
## Stale: fit on the retired hackathon dataset
This section's results (`data/sre_ddm_estimates.rds`) were fit on the
hackathon dataset (`levante_data_pilots_hackathon:fpx0:v1_1`), not on the
model calibration dataset the rest of this notebook now uses, and that cache
isn't present in this environment. A 9.5-hour Sherlock refit wasn't rerun as
part of the switchover, so the numbers below (where they render at all) are
from the old data stack and should be treated as illustrative of the
*modeling approach*, not as current SRE scores. Everything else in this
notebook (rate normalization, zero-fill IRT, EZ-diffusion, the hierarchical
SA model) is fit fresh on the calibration dataset.
:::

The full model: every trial's (choice, RT) as a Wiener process, with
dataset fixed effects plus run *and item* random effects on drift (the IRT
part), run effects on boundary separation and non-decision time, bias fixed
at .5 (~64k trials; item banks are disjoint across sites, so sites are
linked by priors, not anchors). Prototyped locally with brms
([`_fit_sre_ddm.R`](\_fit_sre_ddm.R)), then exported to raw Stan
([`_export_sre_ddm_stan.R`](\_export_sre_ddm_stan.R)) and fit on Sherlock
([`sherlock_sre_ddm/`](sherlock_sre_ddm/); 24 cores, 9.5 h, 4 × 500/500).
Drift-side parameters converged acceptably (R̂ ≤ 1.06); the run-level
non-decision-time hierarchy mixed poorly (R̂ ≈ 1.6) — a known fragility of
ndt under a hard global deadline — which cautions against interpreting the
ndt estimates but leaves the drift comparison intact.

```{r ddm-load}
ddm_file <- here("data/sre_ddm_estimates.rds")
has_ddm <- file.exists(ddm_file)
```

```{r ddm-results, eval=has_ddm}
ddm <- read_rds(ddm_file)

ddm_run <- ddm$run |>
  group_by(dataset) |>
  mutate(drift_z = scale(drift_re)[, 1]) |>
  ungroup()

ddm_first <- sre_first |>
  inner_join(ddm_run |> select(dataset, run_id, drift_z, bs_re, ndt_re),
             by = c("dataset", "run_id"))

ddm_first |>
  pivot_longer(c(age, pa, swr), names_to = "criterion", values_to = "crit") |>
  filter(!is.na(crit)) |>
  group_by(dataset_label, criterion) |>
  filter(n() >= 20) |>
  summarise(n = n(),
            r_rate = cor(score_new, crit),
            r_drift = cor(drift_z, crit),
            r_boundary = cor(bs_re, crit), .groups = "drop") |>
  knitr::kable(digits = 2,
               caption = "Diffusion IRT drift vs corrected rate score: validity")
```

```{r ddm-stability, eval=has_ddm}
ddm_stab <- sre_scores_pub |>
  filter(dataset == "pilot_uniandes_co_bogota") |>
  inner_join(ddm_run |> select(dataset, run_id, drift_z),
             by = c("dataset", "run_id")) |>
  inner_join(scores_cmp |> select(dataset, run_id, score_new),
             by = c("dataset", "run_id")) |>
  group_by(user_id) |>
  filter(n() >= 2) |>
  arrange(time_started) |>
  summarise(gap = as.numeric(difftime(time_started[2], time_started[1],
                                      units = "days")),
            across(c(drift_z, score_new), list(t1 = ~.x[1], t2 = ~.x[2])),
            .groups = "drop") |>
  filter(gap > 300)

tibble(n_pairs = nrow(ddm_stab),
       rate = cor(ddm_stab$score_new_t1, ddm_stab$score_new_t2),
       ddm_drift = cor(ddm_stab$drift_z_t1, ddm_stab$drift_z_t2)) |>
  knitr::kable(digits = 2,
               caption = "Bogotá 16-month stability: diffusion-IRT drift vs corrected rate")
```

```{r ddm-corr, eval=has_ddm, fig.height=4.5}
ddm_first |>
  ggplot(aes(x = score_new, y = drift_z, color = dataset_label)) +
  facet_wrap(vars(dataset_label)) +
  geom_point(alpha = 0.4, size = 0.8, show.legend = FALSE) +
  scale_color_manual(values = dataset_pal) +
  labs(x = "Corrected rate score (z)", y = "DDM-IRT drift (z)",
       title = "Diffusion-IRT drift vs the corrected rate score")
```

The item side of the model behaves exactly as an IRT model should — item
drift effects track empirical item accuracy and (negatively) item RT:

```{r ddm-items, eval=has_ddm, fig.height=4}
item_emp <- sre |>
  group_by(item_uid) |>
  summarise(accuracy = mean(correct), median_rt = median(rt_s), n = n(),
            .groups = "drop") |>
  filter(n >= 20)

ddm$item |>
  inner_join(item_emp, by = "item_uid") |>
  ggplot(aes(x = drift_re, y = accuracy)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(method = "lm", color = "black") +
  labs(x = "Item drift effect (logit-ish)", y = "Empirical item accuracy",
       title = "The IRT part works: item drift effects track item difficulty")
```

So the item-level structure that EZ lacked is genuinely recovered here — and
person-level drift *still* underperforms the rate score where they differ.
That localizes the problem: it is not missing item information, it is that
the diffusion decomposition discards stable, valid between-child variance in
*time allocation* (caution, pacing) that the rate score keeps.

# Verdict

**1. The timer is mostly honest, but the trial-count rule was aimed at the
wrong problem.** The first-180-*trials* filter is a no-op in Germany and
Colombia (max 133 trials/run) and bites only in Canada. The real
heterogeneity is in *time*: 20 Canadian runs contain two sequential ~180-s
blocks (~360 s of testing scored as one window), and there is a left tail of
short runs everywhere (rural Colombia 5th percentile ≈ 89 s of reading time)
that the current rule scores as if they had the full window.

**2. Rate normalization changes little in aggregate, a lot for specific
kids.** Corrected-vs-current r ≈ .99 (DE, both Colombia sites), but **.82 in
Canada** (mean |Δ| = 0.44 z there; the double-block runs drop, short runs
rise). Validity correlations with age/PA/SWR move by |Δr| ≤ .03 everywhere
except Canadian age validity, which improves (.36 → .46); Bogotá 16-month
stability ticks up (.71 → .74). So the correction is about **fairness of
individual scores** (especially any longitudinal use of the Canadian data),
not about rescuing aggregate structure — which was already fine.

**3. Efficiency modeling: a four-rung ladder, and the rate score wins every
rung.** Every model-based alternative is less valid and/or less stable than
the corrected rate score where they differ. The scoreboard below compares
all methods on every available metric.

## Scoreboard: every method × every metric

Validity columns are n-weighted means of within-dataset correlations
(criteria from `registry_scores.rds`: PA/SWR/Vocab/TROG by-language Rasch θ,
first run per child). Reliability uses the best available estimator per
method: odd/even-trial split-half with Spearman–Brown for the
trial-computable scores (rate, EZ), model-based empirical reliability for
the latent-score methods (zero-fill: mirt EAP r~xx~; SA/DDM:
var(θ̂)/(var(θ̂)+mean(SE²))). Germany test–retest uses `levante_data_latest`
v1_2 (the calibration dataset has no DE repeats): n = 23 pairs at ~3.6
months, rescored per method (DDM would need a full refit — left blank; that
DE retest cache isn't present in this environment, so the column below
renders as NA).

```{r scoreboard-methods}
# one frame of run-level scores for every method, z within dataset
methods_run <- scores_cmp |>
  select(dataset, run_id, rate = score_new) |>
  full_join(zf |> select(dataset, run_id, zerofill = zf_z),
            by = c("dataset", "run_id")) |>
  full_join(sa_run |> select(dataset, run_id, sa_composite = eff),
            by = c("dataset", "run_id")) |>
  full_join(ez_run |> group_by(dataset) |>
              mutate(ez_drift = scale(drift)[, 1]) |> ungroup() |>
              select(dataset, run_id, ez_drift),
            by = c("dataset", "run_id"))
if (has_ddm) {
  methods_run <- methods_run |>
    full_join(ddm$run |> group_by(dataset) |>
                mutate(ddm_drift = scale(drift_re)[, 1]) |> ungroup() |>
                select(dataset, run_id, ddm_drift),
              by = c("dataset", "run_id"))
}
method_levels <- intersect(
  c("rate", "zerofill", "sa_composite", "ddm_drift", "ez_drift"),
  names(methods_run))

# criteria: age + other language/reading tasks (first run per child)
crit_wide <- registry_scores |>
  filter(task %in% c("pa", "swr", "vocab", "trog"), !is.na(ability)) |>
  left_join(run_data |> select(run_id, user_id, time_started), by = "run_id") |>
  left_join(participants |> select(user_id, dataset), by = "user_id") |>
  filter(!is.na(dataset)) |>
  first_run() |>
  select(dataset, user_id, task, crit_score = ability) |>
  pivot_wider(names_from = task, values_from = crit_score)

board <- sre_scores_pub |>
  first_run() |>
  select(dataset, user_id, run_id, age) |>
  left_join(crit_wide, by = c("dataset", "user_id")) |>
  inner_join(methods_run, by = c("dataset", "run_id"))

validity <- board |>
  pivot_longer(all_of(method_levels), names_to = "method",
               values_to = "score") |>
  filter(!is.na(score)) |>
  pivot_longer(c(age, pa, swr, vocab, trog), names_to = "criterion",
               values_to = "crit") |>
  filter(!is.na(crit)) |>
  group_by(method, criterion, dataset) |>
  filter(n() >= 20) |>
  summarise(n = n(), r = cor(score, crit), .groups = "drop") |>
  group_by(method, criterion) |>
  summarise(r = weighted.mean(r, n), .groups = "drop") |>
  pivot_wider(names_from = criterion, values_from = r) |>
  select(method, age, pa, swr, vocab, trog)
```

```{r scoreboard-reliability}
# split-half (odd/even trials, Spearman-Brown) for rate and EZ
halves <- sre |>
  mutate(half = ifelse(trial_number %% 2 == 0, "even", "odd")) |>
  group_by(dataset, run_id, half) |>
  summarise(rate_h = (sum(correct) - sum(!correct)) / sum(rt_s) * 180,
            n_h = n(), pc_h = mean(correct),
            mrt_h = mean(rt_s[correct & rt_s >= 0.25 & rt_s <= 30]),
            vrt_h = var(rt_s[correct & rt_s >= 0.25 & rt_s <= 30]),
            nc_h = sum(correct), .groups = "drop")

sb <- \(r) 2 * r / (1 + r)
wmean_by_ds <- \(df, x1, x2) df |>
  group_by(dataset) |>
  filter(n() >= 20) |>
  summarise(n = n(), r = cor({{x1}}, {{x2}}, use = "complete.obs"),
            .groups = "drop") |>
  summarise(rel = sb(weighted.mean(r, n))) |> pull(rel)

rate_halves <- halves |>
  filter(n_h >= 5) |>
  select(dataset, run_id, half, rate_h) |>
  pivot_wider(names_from = half, values_from = rate_h) |>
  filter(!is.na(odd), !is.na(even))
rel_rate <- wmean_by_ds(rate_halves, odd, even)

ez_halves <- halves |>
  filter(n_h >= 8, nc_h >= 5, !is.na(vrt_h), vrt_h > 0) |>
  mutate(pc_adj = pmin(pmax(pc_h, 0.5 + 1 / (2 * n_h)), 1 - 1 / (2 * n_h)),
         drift_h = ez_diffusion(pc_adj, vrt_h, mrt_h)$drift) |>
  select(dataset, run_id, half, drift_h) |>
  pivot_wider(names_from = half, values_from = drift_h) |>
  filter(!is.na(odd), !is.na(even))
rel_ez <- wmean_by_ds(ez_halves, odd, even)

# model-based empirical reliability for the latent-score methods
rel_zf <- zf |> distinct(bank, rxx) |>
  left_join(zf |> count(bank), by = "bank") |>
  summarise(rel = weighted.mean(rxx, n)) |> pull(rel)

rel_sa <- sa_run |>
  group_by(dataset) |>
  summarise(err = mean((theta_se / sd(theta))^2) + mean((tau_se / sd(tau))^2),
            tot = var(scale(theta)[, 1] + scale(tau)[, 1]),
            n = n(), .groups = "drop") |>
  summarise(rel = weighted.mean(pmax(0, 1 - err / tot), n)) |> pull(rel)

rel_ddm <- if (has_ddm) ddm$run |>
  group_by(dataset) |>
  summarise(rel = var(drift_re) / (var(drift_re) + mean(drift_se^2)),
            n = n(), .groups = "drop") |>
  summarise(rel = weighted.mean(rel, n)) |> pull(rel) else NA_real_

reliability <- tibble(
  method = c("rate", "zerofill", "sa_composite", "ddm_drift", "ez_drift"),
  reliability = c(rel_rate, rel_zf, rel_sa, rel_ddm, rel_ez),
  rel_type = c("split-half (SB)", "EAP rxx", "model rxx", "model rxx",
               "split-half (SB)"))
```

```{r de-retest-check}
de_retest_file <- here("data/sre_de_retest_v1_2.rds")
has_de_retest <- file.exists(de_retest_file)
```

```{r scoreboard-retest, eval=has_de_retest}
# Germany test-retest, levante_data_latest v1_2 (n=23 pairs, ~3.6 months)
de <- read_rds(de_retest_file)
de_tr <- de$trials |>
  mutate(correct = as.logical(correct), rt_s = rt_numeric / 1000)

de_scores <- de_tr |>
  group_by(run_id) |>
  summarise(rate = (sum(correct) - sum(!correct)) / sum(rt_s) * 180,
            .groups = "drop")

# zero-fill: score with the calibration-fit DE Rasch model. Matrix starts
# at 0 (unreached AND wrong both 0); set the correct responses to 1.
# Columns follow the model's item order — fscores' response.pattern
# matches by POSITION.
zf_mods <- read_rds(here("data/sre_zerofill_models.rds"))
de_items <- mirt::extract.mirt(zf_mods$de, "itemnames")
de_resp <- matrix(0L, nrow(de_scores), length(de_items),
                  dimnames = list(de_scores$run_id, de_items))
obs <- de_tr |> filter(item_uid %in% de_items, correct)
de_resp[cbind(match(obs$run_id, rownames(de_resp)),
              match(obs$item_uid, de_items))] <- 1L
de_scores$zerofill <- mirt::fscores(zf_mods$de, method = "EAP",
                                    response.pattern = de_resp)[, "F1"]

# EZ per run
de_ez <- de_tr |>
  filter(rt_s >= 0.25, rt_s <= 30) |>
  group_by(run_id) |>
  filter(n() >= 15, sum(correct) >= 10) |>
  summarise(n = n(), pc = mean(correct),
            mrt = mean(rt_s[correct]), vrt = var(rt_s[correct]),
            .groups = "drop") |>
  mutate(pc = pmin(pmax(pc, 0.5 + 1 / (2 * n)), 1 - 1 / (2 * n)),
         ez_drift = ez_diffusion(pc, vrt, mrt)$drift)
de_scores <- de_scores |> left_join(de_ez |> select(run_id, ez_drift),
                                    by = "run_id")

# hierarchical SA refit on the v1_2 DE trials
suppressPackageStartupMessages(library(lme4))
de_lme <- de_tr |> filter(rt_s >= 0.25, rt_s <= 30)
de_acc <- glmer(correct ~ (1 | run_id) + (1 | item_uid), data = de_lme,
                family = binomial, nAGQ = 0)
de_rt <- lmer(log(rt_s) ~ (1 | run_id) + (1 | item_uid), data = de_lme,
              REML = FALSE)
de_sa <- tibble(run_id = rownames(ranef(de_acc)$run_id),
                theta = ranef(de_acc)$run_id[[1]]) |>
  inner_join(tibble(run_id = rownames(ranef(de_rt)$run_id),
                    tau = -ranef(de_rt)$run_id[[1]]), by = "run_id") |>
  mutate(sa_composite = scale(theta)[, 1] + scale(tau)[, 1])
de_scores <- de_scores |> left_join(de_sa |> select(run_id, sa_composite),
                                    by = "run_id")

de_pairs <- de$scores |>
  inner_join(de_scores, by = "run_id") |>
  group_by(user_id) |>
  filter(n() == 2) |>
  arrange(age) |>
  summarise(across(c(rate, zerofill, ez_drift, sa_composite),
                   list(t1 = ~.x[1], t2 = ~.x[2])), .groups = "drop")

retest_de <- tibble(
  method = c("rate", "zerofill", "sa_composite", "ddm_drift", "ez_drift"),
  retest_de = c(cor(de_pairs$rate_t1, de_pairs$rate_t2, use = "complete.obs"),
                cor(de_pairs$zerofill_t1, de_pairs$zerofill_t2,
                    use = "complete.obs"),
                cor(de_pairs$sa_composite_t1, de_pairs$sa_composite_t2,
                    use = "complete.obs"),
                NA_real_,
                cor(de_pairs$ez_drift_t1, de_pairs$ez_drift_t2,
                    use = "complete.obs")))
```

```{r scoreboard-retest-fallback, eval=!has_de_retest}
# data/sre_de_retest_v1_2.rds isn't present in this environment (it's a
# bespoke levante_data_latest v1_2 pull, unrelated to the pilots calibration
# data used elsewhere in this notebook) -- report NA rather than fail
retest_de <- tibble(method = c("rate", "zerofill", "sa_composite", "ddm_drift",
                               "ez_drift"),
                    retest_de = NA_real_)
```

```{r scoreboard-stability}
stab_bog <- sre_scores_pub |>
  filter(dataset == "pilot_uniandes_co_bogota") |>
  inner_join(methods_run, by = c("dataset", "run_id")) |>
  group_by(user_id) |>
  filter(n() >= 2) |>
  arrange(time_started) |>
  summarise(gap = as.numeric(difftime(time_started[2], time_started[1],
                                      units = "days")),
            across(all_of(method_levels), list(t1 = ~.x[1], t2 = ~.x[2])),
            .groups = "drop") |>
  filter(gap > 300)

stability <- tibble(method = method_levels) |>
  rowwise() |>
  mutate(stability_bogota = cor(stab_bog[[paste0(method, "_t1")]],
                                stab_bog[[paste0(method, "_t2")]],
                                use = "complete.obs")) |>
  ungroup()
```

```{r scoreboard-table}
method_labels <- c(
  rate = "Corrected rate (C − I per 180 s)",
  zerofill = "Zero-filled Rasch (ROAR)",
  sa_composite = "Hierarchical SA composite",
  ddm_drift = "Diffusion-IRT drift",
  ez_drift = "EZ-diffusion drift")

scoreboard <- validity |>
  left_join(reliability, by = "method") |>
  left_join(retest_de, by = "method") |>
  left_join(stability, by = "method") |>
  mutate(method = factor(method, levels = names(method_labels),
                         labels = method_labels)) |>
  arrange(method)

knitr::kable(scoreboard, digits = 2,
             col.names = c("Method", "Age", "PA", "SWR", "Vocab", "TROG",
                           "Reliability", "(estimator)",
                           "DE retest (3.6 mo, n=23)",
                           "Bogotá stability (16 mo, n=181)"),
             caption = "Scoring-method scoreboard. Validity = n-weighted mean within-dataset r; all scores z within dataset.")
```

Reading the scoreboard (Diffusion-IRT drift doesn't appear — see the stale
-data callout above; DE retest is NA for every method because that cache
isn't present in this environment):

- **The corrected rate wins most of what matters for a primary score, but
  not uniformly.** Best on age (.52), PA (.49), TROG (.47), and split-half
  reliability (.88); roughly tied with zero-fill and the SA composite on SWR
  (.57 vs .58/.58). It's edged out on Bogotá 16-month stability by zero-fill
  (.74 vs **.76**) and clearly beaten on Vocab by the SA composite (.44 vs
  **.50**).
- **One real nuance: the SA composite's separately-modeled speed dimension
  buys real validity on Vocab** (.50 vs .44 for rate) without a
  comprehension-wide pattern — it doesn't also win TROG (.38 vs rate's .47).
  It underperforms rate on age, PA, reliability, and stability, so this
  reads as a research signal worth having a separate θ/τ decomposition
  available for, not a case for replacing the primary score.
- **Zero-fill's reliability isn't obviously an inflated artifact here** —
  it also posts the best Bogotá stability of any method (.76) — but it's
  still the weakest of the four on TROG (.37 vs .46-.47 for the others),
  consistent with capturing "how much of the bank a fast, accurate reader
  reaches" more than construct-general reading efficiency.
- **EZ-diffusion's Bogotá 16-month stability of .45** (vs .74 for the rate
  score, same 181-pair subsample) is the cleanest available demonstration
  that inverting three summary statistics per run does not produce a usable
  score. (The original DE 3.6-month retest version of this comparison isn't
  available in this environment.)

The diagnosis: SRE's net rate *is* the behaviorally meaningful quantity — it
bundles decision efficiency **and** stable individual differences in
pacing/caution, and every decomposition that strips the latter out (EZ, and
to a lesser extent the SA composite) loses stability. The site-dependent
ability–speed correlation (≈ 0 in DE/Canada, −.08 Bogotá, −.43 in rural CO,
where fast = guessing) additionally means no fixed recombination of
decomposed parts travels across sites. The diffusion-IRT comparison that
previously anchored this section (item drift tracking item accuracy/RT,
drift stability .44→.56, TROG/Vocab wins for drift) was established on the
retired hackathon data stack and hasn't been re-verified on the calibration
dataset — see the callout above.

**Recommendations for the DCC:**

- **Adopt rate normalization now** (patch below). It fixes the Canadian
  double-block and short-run distortions at zero modeling cost and is what
  the ROAR "net correct in 3 minutes" score means.
- **If a measurement model is wanted**, the next step is an
  **offset-Poisson / Rasch-Poisson counts GLMM** (net-correct count with
  `offset(log(time))`, corpus fixed effects, child random effects) — it is
  rate normalization with standard errors and corpus adjustment, not a new
  scoring philosophy.
- **Diffusion IRT: treat the prior "don't use it" conclusion as unconfirmed
  on this data stack, not re-established.** On the retired hackathon data,
  the full hierarchical Wiener fit (9.5 h on Sherlock) yielded drift scores
  that were less stable and mostly less valid than the rate score; that fit
  wasn't rerun here (a 9.5-hour Sherlock job is out of scope for a data-source
  switchover). What *is* re-confirmed on the calibration dataset is the
  narrower claim that decomposing rate into separate ability/speed
  components (EZ-diffusion, the hierarchical SA model) loses test–retest
  stability relative to the simple rate score — consistent with, but not
  proof of, the same conclusion for full diffusion IRT. The decomposition
  remains available in `data/sre_ddm_estimates.rds` (stale) for substantive
  questions about speed–accuracy tradeoff, not as a score.
- **Use RT for data quality regardless**: sub-second responses sit at chance
  and range from 2% of trials (DE) to 13% (rural CO). A fast-guess flag
  (e.g. proportion of RT < 1 s) is a cheap, useful run-level QC covariate.

Proposed patch to `levantemodels::score_sre()`:

```r
score_sre <- \(trial_data_task, dataset) {
  trial_data_task |>
    group_by(.data$run_id) |>
    summarise(elapsed = sum(.data$rt_numeric, na.rm = TRUE) / 1000,
              net = sum(.data$correct) - sum(!.data$correct)) |>
    filter(.data$elapsed >= 30) |>
    mutate(score = scale(.data$net / .data$elapsed * 180)[, 1],
           score_type = "guessing_adjusted_rate_scaled") |>
    select(-"elapsed", -"net")
}
```
