---
title: "Stories (Theory of Mind) — consolidated analysis on the corrected 2026-09 data"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 9
    fig-height: 5
execute:
  warning: false
  message: false
---

{{< include ../_disclaimer.qmd >}}

```{r setup}
source(here::here("common.R"))

dir26 <- here("data/stories_2026-09")
res_items <- read_rds(file.path(dir26, "results_items.rds"))
res_dim   <- read_rds(file.path(dir26, "results_dimensionality.rds"))
res_inv   <- read_rds(file.path(dir26, "results_invariance.rds"))
res_rel   <- read_rds(file.path(dir26, "results_reliability.rds"))
trials    <- read_rds(file.path(dir26, "stories_trials.rds"))
runs      <- read_rds(file.path(dir26, "stories_runs.rds"))
recon     <- read_rds(file.path(dir26, "reconciliation.rds"))

# block-CAT design simulation (tasks/_stories_cat_sim.R)
res_cat <- read_rds(file.path(dir26, "results_cat_sim.rds"))
rule_labels <- c(random = "random", maxinfo = "max information", infomin = "information per minute",
                 rq2 = "top-2 randomesque", disc_a2 = "highest summed discrimination",
                 disc_mean = "highest mean discrimination", oracle = "information at true θ",
                 fixed = "fixed form")
cat_designs <- res_cat$designs |> select(design, design_label = label, n_stories, steps)
# mean realized reliability over the 8 calibration groups; minutes = median over groups
cat_grid <- res_cat$q1 |>
  group_by(design, n_stories, rule) |>
  summarise(rel = mean(rel_realized), med_min = median(median_min), q90_min = median(q90_min),
            max_exp = mean(max_exposure), .groups = "drop")
cg <- function(d, r, col = "rel") {
  x <- cat_grid[[col]][cat_grid$design == d & cat_grid$rule == r]
  stopifnot(length(x) == 1); x
}

# number formatting: n2(0.456) = ".46", n2(1.234) = "1.23"
n2  <- function(x, d = 2) sub("^(-?)0\\.", "\\1.", formatC(x, format = "f", digits = d))
pct <- function(x, d = 0) paste0(formatC(100 * x, format = "f", digits = d), "%")
big <- function(x) format(x, big.mark = ",", trim = TRUE)
rng <- function(x, d = 2) {
  lo <- min(x, na.rm = TRUE); hi <- max(x, na.rm = TRUE)
  paste0(n2(lo, d), if (lo < 0) " to " else "–", n2(hi, d))
}
fp  <- function(p) format.pval(p, digits = 2, eps = 1e-4)          # tables
pe  <- function(p) if (p < 1e-4) "< .0001" else paste("=", n2(p, 3))  # prose: "p `r pe(x)`"
spct <- function(x, d = 0) paste0(if_else(x > 0, "+", ""), pct(x, d))
options(knitr.kable.NA = "")
# pull exactly one value out of a results table (fails loudly otherwise)
pick <- function(df, col, ...) {
  x <- dplyr::filter(df, ...) |> dplyr::pull({{ col }})
  if (length(x) != 1) stop("pick(): expected 1 row, got ", length(x))
  x
}

# item labels: "tom_story4_deception_reality_check_2" -> "S4 rc_2"
short_entry <- function(e) e |>
  str_replace("false_belief", "fb") |> str_replace("reality_check", "rc") |>
  str_replace("emotion_reasoning", "er") |> str_replace("^reference", "ref")
uid_label <- function(uid) {
  m <- str_match(uid, paste0("story(\\d+)_(?:reality_known|moral_reasoning|interpretation|",
                             "deception|reference|second_order)_(.+?)(?:-\\d+)?$"))
  paste0("S", m[, 2], " ", short_entry(m[, 3]))
}

# palettes: languages borrow the colour of their main site in common.R's site_pal
lang_levels <- c("de-DE", "en-US", "en-GB", "es-CO", "es-AR")
lang_pal <- c("de-DE" = site_pal[["pilot_mpieva_de"]],
              "en-US" = site_pal[["pilot_western_ca"]],
              "en-GB" = site_pal[["rfp1_sheffield_gb"]],
              "es-CO" = site_pal[["pilot_uniandes_co"]],
              "es-AR" = site_pal[["rfp1_utdt_ar"]])
qtype_levels <- c("false_belief", "emotion_reasoning", "reference", "reality_check")
qtype_pal <- set_names(ptol_pal()(4), qtype_levels)

# dimensionality samples
dim_labels <- c(de_ib = "Leipzig item bank (de-DE)", de_ib_v2 = "Leipzig, V2 content only",
                de_ib_junelike = "Leipzig, June-style rebuild",
                wca_ib = "Western item bank (en-US)", bog_ib = "Bogotá item bank (es-CO)",
                bog_ib_nofb = "Bogotá, without story-1 FB",
                ar_nhb = "UTDT no-HA retest-B (es-AR)", ar_nhb_nofb = "UTDT, without story-13 FB")
dim_primary <- c("de_ib", "wca_ib", "bog_ib", "ar_nhb")
dim_pal <- c(de_ib = site_pal[["pilot_mpieva_de"]], wca_ib = site_pal[["pilot_western_ca"]],
             bog_ib = site_pal[["pilot_uniandes_co"]], ar_nhb = site_pal[["rfp1_utdt_ar"]])

small_groups <- res_rel$meta$small_groups   # < 25 runs: scored under the DE prior

# what PR #27 adds against the releases: restored trials vs mapped trials in newly included runs
k_add <- list(restored = sum(recon$added_by_kind$n[str_starts(recon$added_by_kind$kind, "restored")]),
              mapped = sum(recon$added_by_kind$n[recon$added_by_kind$kind == "mapped before, run newly included"]),
              flip_runs = sum(recon$run_flips$n[recon$run_flips$before == "few_valid_trials" &
                                                 recon$run_flips$after == "included"]))
stopifnot(k_add$restored + k_add$mapped == sum(recon$per_dataset$added))

# story lists as runs of consecutive numbers: c(2, 7:12) -> "2, 7–12"
story_ranges <- function(s) {
  s <- sort(unique(s)); brk <- c(0, which(diff(s) != 1), length(s))
  paste(map_chr(seq_len(length(brk) - 1), \(i) {
    r <- s[(brk[i] + 1):brk[i + 1]]
    if (length(r) > 1) paste0(r[1], "–", r[length(r)]) else as.character(r)
  }), collapse = ", ")
}

# dataset labels (common.R has no label for rfp1_mpib_de_main)
ds_label <- function(d) coalesce(unname(dataset_labels_named[d]), d)
```

# Purpose and what changed since June

This chapter consolidates the Stories (Theory of Mind, ToM) analyses on the
newest data, with item identities corrected. It supersedes three June
chapters, now retired to `old/` in the
[source repository](https://github.com/levante-framework/levante-analysis):
structure, rescoring and item analysis (`old/stories_tom.qmd`),
dimensionality (`old/tom_dimensionality.qmd`) and multigroup DIF
(`old/tom_dif_multigroup.qmd`). The June
[forensic reconstruction](tom_reality_check_bug.qmd) stays as the appendix
on how the early deployment defects were found. The section on the exclusion
list below revises its account of what went wrong with each item.

The June chapters were frozen on 2026-07-16. They used the June v1_2
release (pilot sites only) and the v1_14 per-language scoring models. Since
then, five things have changed:

1. **Data.** Every raw site dataset with ToM runs, pinned on 2026-09-24:
   `r big(n_distinct(trials$run_id))` included runs from
   `r big(n_distinct(trials$user_id))` children, `r big(nrow(trials))`
   trials, `r n_distinct(trials$dataset)` datasets and
   `r n_distinct(trials$language)` languages. The largest new sample is UTDT
   Argentina (es-AR), which runs only stories 13–18.
2. **Item identity.** levantemodels PR #27 (unmerged) fixes item identities
   in three deployment epochs and sets chance per story question. Against the
   latest releases (for Sheffield, its unreleased next draft of 2026-09-24,
   which matches the pinned raw data) it adds `r big(sum(recon$per_dataset$added))`
   trials: `r big(k_add$restored)` restored by PR #27 and `r k_add$mapped` in
   `r k_add$flip_runs` Bogotá runs that the restores lift above the 10-trial
   minimum. It relabels `r big(sum(recon$per_dataset$relabelled))` and changes
   `correct` for `r sum(recon$per_dataset$correct_differs)` trials.
3. **Item policy.** Items are excluded by story-level flags defined in the
   data build (`tasks/_build_stories_data.R`), not by the generic Airtable
   list. The Spanish reality_known false-belief items are a sensitivity set,
   analysed with and without. The flags are listed below; they are analysis
   choices, and the recommendations say where they should live.
4. **Scoring.** Each dataset with at least 25 runs gets its own group mean
   and variance, as proposed in levantemodels PR #26 (unmerged). Here the
   groups are estimated in one concurrent calibration, on a wider integration
   grid than production uses.
5. **Methods.** The reliability and invariance models fix the guessing floor
   at chance (the LEVANTE convention). The dimensionality models set it to 0
   for items answered at or below chance, and the other two analyses are
   rerun under that rule as a check. Correlations are computed within
   dataset × form rather than pooled over sites. Reliability is reported as
   empirical (EAP) reliability of what children actually answered.
   Dimensionality uses a resampled parallel analysis and bifactor models.

Four analysis scripts produced the results this chapter reads (see
[Reproducing](#reproducing)). The chapter itself fits no models.

```{r key-numbers}
k <- list()
fs <- res_items$flag_summary$by_language
k$cells_ok      <- sum(fs$ok)
k$cells_ceiling <- sum(fs$ceiling)
k$cells_weak    <- sum(fs$weak)
k$cells_poor    <- sum(fs$poor)
k$cells_bcd     <- sum(fs$`below chance, discriminating`)
k$cells_flag    <- k$cells_ok + k$cells_ceiling + k$cells_weak + k$cells_poor + k$cells_bcd
k$acc_13_ar <- pick(res_items$items_lang, accuracy,
                    story_uid == "tom_story13_reality_known_false_belief", language == "es-AR")

vc <- res_items$verdict_counts
vcn <- \(v) sum(vc$n[vc$verdict == v])
k$v_es_disc <- vcn("un-exclude (below chance in Spanish, but discriminates)")
k$v_un   <- vcn("un-exclude") + vcn("un-exclude (weak item)") + k$v_es_disc
k$v_weak <- vcn("un-exclude (weak item)")
k$v_es   <- vcn("un-exclude; flag es-* in analyses, no Airtable row until the translation check")
k$v_scop <- vcn("scoped exclusion")
k$v_keep <- vcn("keep excluded")
k$v_all  <- sum(vc$n)
stopifnot(k$v_un + k$v_es + k$v_scop + k$v_keep == k$v_all)

cr <- res_items$construct_r2
k$r2_qt  <- pick(cr, R2, outcome == "difficulty (IRT, z within language)",
                 unit == "item x language", scheme == "question_type (4)")
k$r2_og  <- pick(cr, R2, outcome == "difficulty (IRT, z within language)",
                 unit == "item x language", scheme == "old_group (6)")
k$r2_nc  <- pick(cr, R2, outcome == "difficulty (IRT, z within language)",
                 unit == "item x language", scheme == "new_construct (6)")
slope_unit <- "item x language (language FE; dR2 over it)"
k$dr2_nc <- pick(cr, dR2, outcome == "age slope (logit per year)", unit == slope_unit,
                 scheme == "new_construct (6)")
k$p_nc   <- pick(cr, p, outcome == "age slope (logit per year)", unit == slope_unit,
                 scheme == "new_construct (6)")

eg <- res_dim$eigen
k$ev_de <- pick(eg, ev1_ev2, sample == "de_ib")
k$ev_other <- eg$ev1_ev2[eg$sample %in% setdiff(dim_primary, "de_ib")]
ld <- res_dim$irt_ld |> filter(family == "guess", sample %in% dim_primary)
k$ecv <- rng(ld$ECV); k$omh <- rng(ld$omega_h); k$cor_uni_g <- rng(ld$cor_theta_uni_general, 3)
k$info_over <- paste0(pct(min(ld$info_overstatement)), "–", pct(max(ld$info_overstatement)))
spec_signs <- res_dim$irt_slopes |>
  filter(family == "guess", has_story_factor, sample %in% dim_primary) |>
  group_by(sample, story) |>
  summarise(mixed = any(lambda_specific < 0) & any(lambda_specific > 0),
            sum_abs = sum(abs(lambda_specific)), abs_sum = abs(sum(lambda_specific)), .groups = "drop")
k$mixed_stories <- sum(spec_signs$mixed); k$n_story_factors <- nrow(spec_signs)
k$de_s2_abs <- pick(spec_signs, sum_abs, sample == "de_ib", story == 2)
k$de_s2_sum <- pick(spec_signs, abs_sum, sample == "de_ib", story == 2)

cc <- res_items$controls_cor
k$r_ct_pooled <- pick(cc, r_control_target, method == "pooled over runs (June method)",
                      item_set == "policy items dropped")
k$r_ct_within <- pick(cc, r_control_target, method == "within dataset x form, Fisher-pooled",
                      item_set == "policy items dropped")

dbt <- res_inv$dif_by_type
k$A_ctrl_x2  <- pick(dbt, share_of_s1_X2, analysis == "A", kind == "control")
k$A_ctrl_it  <- pick(dbt, share_of_items, analysis == "A", kind == "control")
k$B_ctrl_x2  <- pick(dbt, share_of_s1_X2, analysis == "B", kind == "control")
tst <- res_inv$tests
k$A_scal_x2df <- pick(tst, X2_df, analysis == "A", comparison == "scalar vs configural", items == "all")
k$B_scal_x2df <- pick(tst, X2_df, analysis == "B", comparison == "scalar vs configural", items == "all")
k$A_part_p <- pick(tst, p, analysis == "A", comparison == "partial scalar vs configural", items == "all")
k$B_part_p <- pick(tst, p, analysis == "B", comparison == "partial scalar vs configural", items == "all")

# reliability headline: aggregate fixed-form and fCAT rows per dataset
agg_forms <- c("fCAT", "all fixed forms", "all forms")
form_types <- res_rel$rel_by_form |> filter(!form %in% agg_forms) |> distinct(dataset, form_type)
single_type <- form_types |> count(dataset) |> filter(n == 1) |> pull(dataset)
rel_head <- bind_rows(
  res_rel$rel_by_form |> filter(form %in% c("all fixed forms", "fCAT")),
  res_rel$rel_by_form |> filter(form == "all forms", dataset %in% single_type) |>
    select(-form_type) |> left_join(form_types, by = "dataset")
) |>
  filter(!dataset %in% small_groups) |>
  mutate(kind = if_else(str_starts(form_type, "fCAT"), "fCAT (3 random stories)", "fixed forms"))
k$rel_fixed <- rng(rel_head$rel_empirical[rel_head$kind == "fixed forms"])
k$rel_fcat  <- rng(rel_head$rel_empirical[rel_head$kind != "fixed forms"])
# single fixed forms (dataset x form, >= 25 runs)
rel_forms <- res_rel$rel_by_form |>
  filter(form_type == "fixed form", !form %in% agg_forms, n_runs >= 25, !dataset %in% small_groups) |>
  arrange(rel_empirical)
form_lab <- \(r) paste0(ds_label(r$dataset), " ", r$form, ", ", r$n_runs, " runs")
k$low_form  <- form_lab(slice_head(rel_forms, n = 1))
k$high_form <- form_lab(slice_tail(rel_forms, n = 1))
k$rel_forms <- rng(rel_forms$rel_empirical)
dc  <- \(d, col) pick(res_rel$density_check, {{ col }}, dataset == d)
g0s <- \(d, col) pick(res_rel$g0_sensitivity, {{ col }}, dataset == d)
blk <- \(b) res_rel$fcat_blocks$story[res_rel$fcat_blocks$block == b]
cs  <- res_rel$cat_sim_summary
csv <- \(d, col) pick(cs, {{ col }}, dataset == d)
je <- res_rel$june_equivalents
k$de_bank <- pick(je, value, str_starts(quantity, "DE full-bank marginal"))
sb <- res_rel$stability_bins
k$stab_60 <- pick(sb, r, dataset == "pilot_mpieva_de_main", gap_bin == "[60,180)", version == "analysis")
k$stab_60_lo <- pick(sb, r_lo, dataset == "pilot_mpieva_de_main", gap_bin == "[60,180)", version == "analysis")
k$stab_60_hi <- pick(sb, r_hi, dataset == "pilot_mpieva_de_main", gap_bin == "[60,180)", version == "analysis")
jp <- res_dim$june_pool_check
k$fixed_items <- rng(rel_head$median_items[rel_head$kind == "fixed forms"], 0)
k$fcat_min    <- rng(rel_head$median_minutes[rel_head$kind != "fixed forms"], 1)
rc  <- res_rel$released_comparison |> filter(!dataset %in% c(small_groups, "all (pooled)"))
own <- res_rel$released_models |> filter(own_group_in_v2_3) |> pull(dataset)
k$shift_total <- rng(-rc$shift_analysis_minus_released[!rc$dataset %in% own], 1)
k$shift_prior <- rng(-rc$shift_own_minus_ref_prior[!rc$dataset %in% own], 1)
```

## Summary

```{r summary-numbers}
# numbers the summary needs from later sections (computed here, where the summary is)
sum_q3 <- res_dim$q3_grouping_contrast |> filter(sample %in% dim_primary, grouping == "question type")
sum_st <- res_items$slope_tests
sum_fl <- \(col) pick(sum_st, {{ col }}, outcome == "slope_floor", scheme == "question_type")
sum_2pl <- \(a, col) pick(res_inv$tests_2pl, {{ col }}, analysis == a, comparison == "scalar vs metric")
sum_rasch <- \(a, col) pick(res_inv$tests, {{ col }}, analysis == a, comparison == "scalar vs configural", items == "all")
```

- **Item quality.** Of the `r k$cells_flag` item × language cells with at
  least 30 responses, `r k$cells_ok` are fine, `r k$cells_ceiling` are at
  ceiling, `r k$cells_weak` discriminate weakly, `r k$cells_poor` are poor
  and `r k$cells_bcd` are below chance but still discriminate. The poor
  cells are the policy-flagged content plus Spanish reality_known
  false-belief items (story 13 in es-AR: `r n2(k$acc_13_ar, 3)` correct
  with two response options).
- **The exclusion list.** The 12 generic Airtable entries cover
  `r k$v_all` story-level items. `r k$v_un` of them should be un-excluded
  (`r k$v_weak` weak; `r k$v_es_disc` below chance in Spanish but still
  discriminating). `r k$v_scop` need exclusions scoped to one content
  version; for the two on image 4g, a key correction is the alternative, if
  the content team confirms it. `r k$v_keep` should stay excluded
  everywhere: reference_reference in stories 5, 11 and 17, and story 10's
  deception false belief 1, which does not discriminate. Story 13's
  reality_known false belief keeps an analysis flag in Spanish, and any
  Airtable row for it should wait for the translation check. The scoped
  rows cannot be written in today's Airtable schema.
- **What organizes the items.** Question type organizes item difficulty
  (R² `r n2(k$r2_qt, 3)`, against `r n2(k$r2_og, 3)` and `r n2(k$r2_nc, 3)`
  for the two story-construct schemes). Above the guessing floor,
  false-belief targets are more age-sensitive than controls (difference
  `r n2(sum_fl(fb_minus_control), 2)` [`r n2(sum_fl(fb_minus_control_lo), 2)`,
  `r n2(sum_fl(fb_minus_control_hi), 2)`] logits per year), so June's
  "false-belief targets carry the age signal" survives on the
  chance-corrected scale; raw logit slopes do not separate them. The
  clustering of tetrachoric correlations by question type partly reflects
  shared difficulty: on IRT residuals the contrast shrinks to
  `r rng(sum_q3$within_minus_between, 3)`, weak and uneven across
  samples.
- **Dimensionality.** A single score is defensible, most clearly outside
  Leipzig. The Leipzig form's eigenvalue ratio is `r n2(k$ev_de)`, as in
  June (on a partly different item set); the Western, Bogotá and es-AR
  forms are at `r rng(k$ev_other)`. In the bifactor models ECV is `r k$ecv` (below the
  usual .70 in Leipzig and es-AR), and scores barely change
  (r = `r k$cor_uni_g`). Story-level local dependence makes a
  unidimensional model overstate information by `r k$info_over`.
- **Controls.** Keep them. They add reliability and cost almost no age
  validity. Their correlation with targets is `r n2(k$r_ct_within)` within
  dataset × form, lower than the pooled `r n2(k$r_ct_pooled)` that June's
  method gives.
- **Cross-language invariance.** In stories 1–6 (de/en-US/es-CO) the LR
  test rejects scalar invariance (χ²/df `r n2(k$A_scal_x2df, 1)`), but BIC
  does not favour non-invariance under the LEVANTE convention or the 2PL
  (Rasch dBIC `r n2(sum_rasch("A", dBIC), 0)`, 2PL `r n2(sum_2pl("A", dBIC), 0)`);
  it does with a guessing floor of 0 for below-chance items
  (dBIC `r n2(sum_rasch("A_g0", dBIC), 0)`).
  Stories 13–18 (de/es-AR) are non-invariant under both models (dBIC
  `r n2(sum_rasch("B", dBIC), 0)` and `r n2(sum_2pl("B", dBIC), 0)`).
  Whether controls or targets carry more of the DIF depends on the model
  and the design, so June's "controls dominate" is weakened, not simply
  overturned. False-belief items are non-invariant in both analyses, but
  language is confounded with prior exposure (most German children in
  stories 13–18 were on a repeat run), age and site. Two sites that share a
  language differ less (by BIC) and mostly on other items, and German
  children under 9 show the same false-belief DIF, so neither site nor age
  alone explains it.
  Partial scalar models fit, but only as a description.
- **Reliability.** Per administration, reliability is moderate:
  `r k$rel_fixed` per dataset, pooling its fixed forms (single forms
  `r k$rel_forms`), and `r k$rel_fcat` for the deployed fCAT (3 random
  stories, median `r k$fcat_min` minutes). Both are upper bounds. June's
  ".92" was a model-implied value for a `r jp$items_in_model`-item pool that
  no child takes; the same model's reliability for what children answered
  is `r n2(jp$rasch_empirical_rxx_eap, 3)`. In the same children, choosing
  the more informative story per block instead of a random one raises the
  reliability of a 3-story form by a median of
  `r n2(median(cs$rel_adaptive - cs$rel_random))` (up to
  `r n2(max(cs$rel_adaptive - cs$rel_random))`). German alternate-form stability over 2–6 months is
  `r n2(k$stab_60)` [`r n2(k$stab_60_lo)`, `r n2(k$stab_60_hi)`].
- **Block CAT (simulation).** Choosing the most informative story within today's
  three blocks raises mean reliability from `r n2(cg("cur3", "random"))` to
  `r n2(cg("cur3", "maxinfo"))` in the same time; adding a fourth, free story
  reaches `r n2(cg("cur3p1", "maxinfo"))` in about `r n2(cg("cur3p1", "maxinfo", "med_min"), 1)`
  median minutes, more than a fixed six-story form in less time. Select by information at
  the current estimate (not by discrimination), under the group's prior; see
  [Designing the block CAT](#designing-the-block-cat).
- **Scoring.** The means and variances of low-scoring groups are weakly
  identified: they move with the integration grid, the latent density and
  the guessing convention, while scores within a group barely change.
  Datasets without their own group in v2_3 were scored under the German
  prior; with their own prior, their level drops by `r k$shift_prior`
  logits.

# Data and provenance

```{r provenance}
prov <- attr(trials, "provenance")
prov$raw_sources |>
  left_join(prov$release_refs, by = "dataset") |>
  mutate(dataset = ds_label(dataset)) |>
  knitr::kable(caption = "Pinned raw datasets (read) and the latest processed releases (for reconciliation only)")
```

- **Build.** `tasks/_build_stories_data.R` ran `process_trials()` with
  levantemodels from the PR #27 branch (`fix-tom-identity-corrections`,
  commit `r substr(prov$levantemodels[["commit"]], 1, 7)`, which sits
  directly on main `bd24157`). On 2026-09-24, `git ls-remote` gave
  `bd24157` for origin/main (the same as the local main) and `3dc148b`
  for `refs/pull/27/head`. Item identity metadata:
  `r prov$metadata[["ref"]]`.
- **Run and trial rules.** These are the production rules the latest
  releases were built with: `r prov$rules[["run"]]`; trials: `r prov$rules[["trial"]]`;
  then `r prov$rules[["run_after_trials"]]`.
- **Reconciliation.** Without PR #27 the rebuild reproduces the releases
  trial for trial (`r recon$pre_vs_release[["pre_not_in_release"]]` rebuilt
  trials missing from the releases, and
  `r recon$pre_vs_release[["release_not_in_pre"]]` the other way round).
  For Sheffield the comparison is with its unreleased next draft of
  2026-09-24 (`:next` in the table above), because its latest release
  (v3_1) predates the pinned raw data. The table below shows what PR #27
  adds.
- **The fCAT is not adaptive.** The `theory-of-mind-item-bank-cat-test`
  corpus ("fCAT") shows one random story from each of 3 story groups. It
  has no item parameters and makes no θ updates. It is analysed here as a
  random 3-story form (`fcat_random3`).

```{r inventory}
trials |>
  group_by(dataset, form, language) |>
  summarise(runs = n_distinct(run_id), children = n_distinct(user_id),
            trials = n(), median_age = median(age[!duplicated(run_id)], na.rm = TRUE),
            stories = story_ranges(story),
            first = as.Date(min(time_started)), last = as.Date(max(time_started)),
            .groups = "drop") |>
  mutate(dataset = ds_label(dataset)) |>
  knitr::kable(digits = 1, format.args = list(big.mark = ","),
               caption = "Included runs by dataset × form × language")
```

```{r pr27}
recon$per_dataset |>
  select(dataset, release_trials, stories_trials, added, relabelled,
         correct_differs, release_runs, stories_runs) |>
  mutate(dataset = ds_label(dataset)) |>
  knitr::kable(format.args = list(big.mark = ","),
               caption = "What PR #27 changes against the latest releases (trials and runs)")
```

## Item-policy flags

The analyses drop the `policy_exclude` trials by default and analyse the
`spanish_fb_flag` trials both with and without. Neither flag removes rows
from the data file.

```{r policy}
trials |>
  filter(policy_exclude | spanish_fb_flag) |>
  mutate(flag = if_else(policy_exclude, policy_reason,
                        "Spanish reality_known false belief: analyse with and without")) |>
  group_by(flag, item = uid_label(story_uid), item_original) |>
  summarise(trials = n(), datasets = n_distinct(dataset),
            accuracy = mean(correct), chance = first(chance), .groups = "drop") |>
  arrange(flag, item) |>
  knitr::kable(digits = 3, format.args = list(big.mark = ","),
               caption = "Item-policy flags: trials, content code and accuracy")
```

## Western's empty item-bank runs

```{r western}
wca_empty <- runs |>
  filter(dataset == "pilot_western_ca_main", variant_id == "Qx8XB8ZAFwP0RhhiEKw5",
         time_started >= as.POSIXct("2026-04-25", tz = "UTC"), n_trials_raw == 0)
wca_other <- runs |>
  filter(dataset == "pilot_western_ca_main", included,
         user_id %in% wca_empty$user_id)
```

`r nrow(wca_empty)` Western runs on the item-bank variant
(`Qx8XB8ZAFwP0RhhiEKw5`) between `r as.Date(min(wca_empty$time_started))`
and `r as.Date(max(wca_empty$time_started))` logged no trials
(`r sum(wca_empty$stop_reason %in% "taskAbort")` with `stop_reason =
taskAbort`). They are excluded as `no_data`. `r n_distinct(wca_other$user_id)`
of the `r n_distinct(wca_empty$user_id)` children involved have an included
ToM run (at any time) in these data. According to the engineering team's
follow-up, no collected data were lost. The variant still
loaded the item-bank corpus, whose hostile-attribution prompts are missing
from the English item-bank translation, so the task failed its startup
corpus check and aborted before the first trial. The children completed the
other tasks. The failure was written up in an internal incident report on 12 July,
logged by error monitoring on 19 and 28 August, and is related to
core-tasks#437 (open since 8 April). The engineering team believes the
variant has since been updated; the last run on it in these data started on
`r as.Date(max(runs$time_started[runs$variant_id %in% "Qx8XB8ZAFwP0RhhiEKw5"]))`,
which fits an update but does not prove one.
Engineering plans a nightly check of open assignments that
reference stale (unregistered) variants and, in the longer term, moving
existing assignments to new variants when an old one is deprecated. For this
chapter, these are missing administrations, not dropped data.

# Item quality and the exclusion list

## Item screen

Each story item was screened in each language with at least 30 responses:
accuracy against chance (exact binomial tests) and item-rest correlation.
The item-rest r is computed within dataset × form and then Fisher-pooled.
The flag levels are *poor* (below chance without credible discrimination,
or item-rest r reliably negative), *below chance but discriminating*,
*ceiling* (≥ .95), *weak* (not above chance, or item-rest r with a 95% CI
upper bound below .20) and *ok*.[^flagrules]

[^flagrules]: The flag rules were revised once after the first run. The
original weak-item rule (item-rest r point estimate < .10) mostly flagged
small fCAT cells, whose item-rest CIs are wide. The revised rule uses the
CI upper bound, checks ceiling before weak, and gives below-chance items
with a credibly positive r their own level. The original criterion is kept
in `r_ir_below_10`; it would flag
`r sum(res_items$items_lang$r_ir_below_10[res_items$items_lang$flag != "n < 30"], na.rm = TRUE)`
of the `r k$cells_flag` cells.

```{r flag-table}
res_items$flag_summary$by_language |>
  select(language, ok, ceiling, weak, `below chance, discriminating`, poor, `n < 30`) |>
  knitr::kable(caption = "Item × language cells by flag (n ≥ 30 to be flagged)")
```

```{r fig-items, fig.height=13}
il <- res_items$items_lang |>
  filter(flag != "n < 30") |>
  mutate(label = uid_label(story_uid),
         label = paste0(label, case_when(content == "other content" ~ " [other content]",
                                         str_detect(content, "4g") ~ " [4g]",
                                         str_detect(content, "2e") ~ " [2e]",
                                         TRUE ~ "")),
         label = fct_reorder(label, -story),
         question_type = factor(question_type, qtype_levels),
         language = factor(language, lang_levels),
         flag = fct_drop(flag))
ggplot(il, aes(x = accuracy - chance, y = label, colour = language, shape = flag)) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_point(size = 2, alpha = .8) +
  facet_wrap(vars(question_type), scales = "free_y", ncol = 2) +
  scale_colour_manual(values = lang_pal, name = NULL) +
  scale_shape_manual(values = c(ok = 16, ceiling = 17, weak = 1,
                                `below chance, discriminating` = 4, poor = 8), name = NULL) +
  labs(x = "Accuracy minus chance", y = NULL,
       title = "Item accuracy above chance, by language",
       subtitle = "Cells with n ≥ 30; [4g], [2e] and [other content] split the policy-flagged content from the rest") +
  theme(legend.position = "bottom", legend.box = "vertical",
        axis.text.y = element_text(size = 7))
```

```{r poor-not-policy}
res_items$flag_summary$poor_not_policy |>
  transmute(item = uid_label(story_uid), language, n, accuracy, chance,
            p_below = fp(p_below), r_ir, r_ir_lo, flag, datasets) |>
  knitr::kable(digits = 3, caption = "Below-chance cells outside the policy flags")
```

```{r items-numbers}
ia <- function(uid, cont = "all", col = accuracy)
  pick(res_items$items_all, {{ col }}, story_uid == uid, content == cont)
il_acc <- function(uid, lang, col = accuracy)
  pick(res_items$items_lang, {{ col }}, story_uid == uid, language == lang, content == "all")
k$s14_ar    <- il_acc("tom_story14_moral_reasoning_false_belief_1", "es-AR")
k$s14_ar_n  <- il_acc("tom_story14_moral_reasoning_false_belief_1", "es-AR", n)
k$s14_ar_r  <- il_acc("tom_story14_moral_reasoning_false_belief_1", "es-AR", r_ir)
k$s14_ar_lo <- il_acc("tom_story14_moral_reasoning_false_belief_1", "es-AR", r_ir_lo)
k$s14_ar_hi <- il_acc("tom_story14_moral_reasoning_false_belief_1", "es-AR", r_ir_hi)
k$s14_ar_age <- il_acc("tom_story14_moral_reasoning_false_belief_1", "es-AR", median_age)
k$s14_de    <- il_acc("tom_story14_moral_reasoning_false_belief_1", "de-DE")
rk_band <- \(uid, lang, band) pick(res_items$rk_fb_age, accuracy, story_uid == uid, language == lang,
                                   age_band == band)
rk_am <- res_items$rk_fb_age_matched
```

What the screen shows:

- Outside the policy flags, the below-chance cells are Spanish false-belief
  items, above all the reality_known question ("where will X look first?")
  of stories 1, 7 and 13. They are not alike. Story 1 in es-CO is below
  chance but discriminates (item-rest r
  `r n2(il_acc("tom_story1_reality_known_false_belief", "es-CO", r_ir))`
  [`r n2(il_acc("tom_story1_reality_known_false_belief", "es-CO", r_ir_lo))`,
  `r n2(il_acc("tom_story1_reality_known_false_belief", "es-CO", r_ir_hi))`]),
  the pattern of children who answer with the object's real location. Story
  7 in es-CO has a similar r
  (`r n2(il_acc("tom_story7_reality_known_false_belief", "es-CO", r_ir))`
  [`r n2(il_acc("tom_story7_reality_known_false_belief", "es-CO", r_ir_lo))`,
  `r n2(il_acc("tom_story7_reality_known_false_belief", "es-CO", r_ir_hi))`],
  n = `r il_acc("tom_story7_reality_known_false_belief", "es-CO", n)`) and is
  flagged poor only because its interval reaches 0. Story 13 in es-AR carries
  no information: `r n2(k$acc_13_ar, 3)` correct, item-rest r
  `r n2(il_acc("tom_story13_reality_known_false_belief", "es-AR", r_ir), 3)`.
- The failure is not specific to Spanish. Young en-US children are also
  below chance on story 1 (`r n2(rk_band("tom_story1_reality_known_false_belief", "en-US", "<6"))`
  under 6 and `r n2(rk_band("tom_story1_reality_known_false_belief", "en-US", "6-7"))`
  at 6–7 years), and story 13 is at chance in en-GB
  (`r n2(il_acc("tom_story13_reality_known_false_belief", "en-GB"))`,
  n = `r il_acc("tom_story13_reality_known_false_belief", "en-GB", n)`). The
  items "work" only in the older German samples. Compared at the same age,
  though, Spanish-speaking children still answer them less often correctly
  (table below): the Spanish gap is `r rng(-rk_am$es_logit, 1)` logits.
  Age, language, site and sample are confounded, and nobody has checked the
  translation files yet.

```{r fig-rkfb, fig.height=3.8}
res_items$rk_fb_age |>
  filter(n >= 10) |>
  mutate(item = paste0("Story ", story, " reality_known false belief"),
         language = factor(language, lang_levels)) |>
  ggplot(aes(age_band, accuracy, colour = language, group = language)) +
  geom_hline(aes(yintercept = chance), linetype = "dashed") +
  geom_line() + geom_point(aes(size = n)) +
  facet_wrap(vars(item), nrow = 1) +
  scale_colour_manual(values = lang_pal, name = NULL) +
  scale_size_area(max_size = 3, guide = "none") +
  labs(x = "Age band (years)", y = "Accuracy",
       title = "The reality_known false-belief items by age band and language (cells with n ≥ 10; dashed = chance)") +
  theme(legend.position = "bottom")
rk_am |>
  transmute(item = uid_label(story_uid), ages = paste0(n2(age_lo, 1), "–", n2(age_hi, 1)),
            n_es, n_other, other_languages, spanish_logit = es_logit, lo = es_lo, hi = es_hi,
            acc_at_7_other, acc_at_7_es) |>
  knitr::kable(digits = 2, caption = "Age-matched comparison: glm(correct ~ age + Spanish) on the ages both groups cover (the intersection of each group's 5th–95th age percentiles)")
```

- The problem reaches beyond stories 1/7/13. Story 14's
  moral-reasoning false belief ("where will he look first for his soccer
  ball?") is below chance in es-AR (`r n2(k$s14_ar, 3)`,
  n = `r big(k$s14_ar_n)`; de-DE `r n2(k$s14_de, 3)`), yet its item-rest r
  is positive (`r n2(k$s14_ar_r, 3)` [`r n2(k$s14_ar_lo, 3)`,
  `r n2(k$s14_ar_hi, 3)`]). That pattern fits young children (median age
  `r n2(k$s14_ar_age, 1)`) who answer with the real location, not a broken
  key. Age and language are confounded here: the es-AR samples are the
  youngest.
- The weakest items outside the policy flags are yes/no "Is [sibling] mad at
  [sibling]?" questions in the parallel deception stories 4 and 16. They sit at or
  near chance everywhere. Two other targets barely discriminate, and story
  10's deception false belief 1 does not discriminate at all (it is now
  policy-excluded; see the verdicts):

```{r weak-targets}
weak_uids <- c("tom_story4_deception_false_belief_3", "tom_story16_deception_reference_2",
               "tom_story16_deception_reference_5", "tom_story10_deception_false_belief_1",
               "tom_story17_reference_emotion_reasoning_1")
res_items$items_all |>
  filter(story_uid %in% weak_uids, content == "all") |>
  transmute(item = uid_label(story_uid), prompt, n, accuracy, chance,
            r_ir, r_ir_lo, r_ir_hi, lang_weak) |>
  knitr::kable(digits = 3, caption = "Candidates for item revision (not proposed exclusions)")
```

::: {.callout-note collapse="true"}
## All items, pooled over languages (92 rows)

`content` splits the policy-flagged 4g and 2e content from the rest.
Flagged languages are listed by flag level.

```{r items-all}
res_items$items_all |>
  arrange(story, question_type) |>
  transmute(item = uid_label(story_uid), content, question_type, prompt, n,
            accuracy, chance, r_ir, r_ir_lo, r_ir_hi,
            lang_poor, lang_below_disc, lang_weak, lang_ceiling) |>
  knitr::kable(digits = 3)
```
:::

::: {.callout-note collapse="true"}
## Flagged cells by dataset × form

```{r items-cell}
res_items$items_cell |>
  filter(flag %in% c("poor", "below chance, discriminating")) |>
  arrange(story, dataset) |>
  transmute(item = uid_label(story_uid), content, dataset, form, language, n,
            accuracy, chance, median_age, r_ir, flag) |>
  knitr::kable(digits = 3)
```
:::

## The exclusion list, story by story

The Airtable exclusion table holds 12 generic ToM uids, such as
`tom_deception_reality_check_2`. Each names a question type in a story type,
so it applies to all parallel stories. The 2026-09 audit found that none of
them takes effect in the current pipeline: the scoring prep runs its
`anti_join` after `recode_tom()` has rewritten every uid to story level, so
the generic uids match nothing. The table below judges each of the
`r k$v_all` story-level items they cover on the corrected data. `accuracy_kept`
and `r_ir_kept` are the values once the proposed scoped exclusion is
applied.

The verdicts are a check of the data build's policy flags, which were set
from the 2026-09 audit, against the item-screen rules above. For the
"keep excluded" and plain "un-exclude" rows the agreement column is true by
construction. The informative checks are the scoped rows (is the flagged
content below chance while the rest is fine?) and the Spanish rows.

```{r verdicts}
res_items$item_verdicts |>
  transmute(item = uid_label(story_uid), n_all, accuracy_all, chance_all, r_ir_all,
            accuracy_kept, r_ir_kept, verdict, scope) |>
  knitr::kable(digits = 3, caption = "Verdicts for the 28 story-level items behind the 12 generic exclusion uids")
```

```{r verdict-counts}
res_items$verdict_counts |>
  knitr::kable(caption = "Verdict counts (and agreement with the build's policy flags; true by construction except for the scoped and Spanish rows)")
```

```{r b-numbers}
k$rr5  <- ia("tom_story5_reference_reference")
k$rr11 <- ia("tom_story11_reference_reference")
k$rr17 <- ia("tom_story17_reference_reference")
k$rr5_de <- il_acc("tom_story5_reference_reference", "de-DE")
k$rr5_de_n <- il_acc("tom_story5_reference_reference", "de-DE", n)
k$s4rc2_4g <- ia("tom_story4_deception_reality_check_2", "policy-flagged (4g)")
k$s4rc3_4g <- ia("tom_story4_deception_reality_check_3", "policy-flagged (4g)")
k$s4rc2_4d <- ia("tom_story4_deception_reality_check_2", "other content")
k$s4rc3_4d <- ia("tom_story4_deception_reality_check_3", "other content")
k$s2rc1_2e <- ia("tom_story2_moral_reasoning_reality_check_1", "policy-flagged (2e)")
k$s2rc1_2e_n <- ia("tom_story2_moral_reasoning_reality_check_1", "policy-flagged (2e)", n)
k$s2rc1_oth <- ia("tom_story2_moral_reasoning_reality_check_1", "other content")
# story-4 rc_2 on 4g by language, and the 4d level by language
s4 <- \(uid, cont, lang, col = accuracy)
  pick(res_items$items_lang, {{ col }}, story_uid == uid, content == cont, language == lang)
rc2 <- "tom_story4_deception_reality_check_2"
k$rc2_4d_rng <- rng(res_items$items_lang$accuracy[res_items$items_lang$story_uid == rc2 &
                                                   res_items$items_lang$content == "other content"])
# flipped keys (pooled over languages)
fc <- \(uid, part, col) pick(res_items$flip_check, {{ col }}, story_uid == uid, content_part == part,
                             language == "all")
flip_ci <- \(uid, part) paste0(n2(fc(uid, part, r_ir_flipped)), " [", n2(fc(uid, part, r_ir_flipped_lo)),
                               ", ", n2(fc(uid, part, r_ir_flipped_hi)), "]")
s2r <- \(e, cont, resp) pick(res_items$story2_rc_responses, share, entry == e, content == cont,
                             response == resp)
rl <- res_items$reliability
k$ar13_with <- pick(rl, empirical_rxx, sample == "UTDT-AR main, no-HA retest-B form (stories 13-18)",
                    items == "all items", model == "2PL")
k$ar13_without <- pick(rl, empirical_rxx,
                       sample == "UTDT-AR main, no-HA retest-B form, without the flagged es false-belief item",
                       items == "all items", model == "2PL")
```

What the verdicts say:

- **Keep excluded:** reference_reference in stories 5, 11 and 17
  (`r n2(k$rr5, 3)`, `r n2(k$rr11, 3)` and `r n2(k$rr17, 3)` correct;
  chance .33). It is below chance in every language (cells with at least
  30 responses) except German story 5, which is at chance (`r n2(k$rr5_de, 3)`, n = `r k$rr5_de_n`), and its
  item-rest r is negative in every story. The audit found that children pick
  the literal referent, so a key flip would only duplicate the reality check.
  This is a design problem, and the items are still deployed. Story 10's
  deception false belief 1 also stays excluded: its item-rest r is
  `r n2(pick(res_items$item_verdicts, r_ir_all, story_uid == "tom_story10_deception_false_belief_1"))`
  [`r n2(pick(res_items$item_verdicts, r_ir_lo_all, story_uid == "tom_story10_deception_false_belief_1"))`,
  `r n2(pick(res_items$item_verdicts, r_ir_hi_all, story_uid == "tom_story10_deception_false_belief_1"))`]
  (n = `r pick(res_items$item_verdicts, n_all, story_uid == "tom_story10_deception_false_belief_1")`),
  and its interval spans 0 in every language. The same generic uid covers
  stories 4 and 16, which discriminate and are un-excluded.
- **Scoped: story 4 on image 4g.** Reality check 3 on image 4g
  (`r n2(k$s4rc3_4g, 3)`) is below chance in every language. Reality check 2
  (`r n2(k$s4rc2_4g, 3)`) is below chance in de-DE and en-US, weak in en-GB
  (`r n2(s4(rc2, "policy-flagged (4g)", "en-GB"), 3)`, n =
  `r s4(rc2, "policy-flagged (4g)", "en-GB", n)`, one-sided p =
  `r n2(s4(rc2, "policy-flagged (4g)", "en-GB", p_below), 3)`) and at chance in
  es-CO (`r n2(s4(rc2, "policy-flagged (4g)", "es-CO"), 3)`), still well below
  its level on the earlier image 4d (`r k$rc2_4d_rng`). On 4d both questions
  work, although reality check 3 is weak there too (`r n2(k$s4rc3_4d, 3)` on
  two options). Both are keyed "yes" ("Is Mother/Joshua mad at Isabel
  now?"), but the audit found that image 4g shows Mother, Isabel and Joshua
  smiling, and the story never says that Mother learns what Isabel did. With
  the key flipped to "no", the 4g items score
  `r n2(fc(rc2, "flagged: 4g", accuracy_flipped), 2)` and
  `r n2(fc("tom_story4_deception_reality_check_3", "flagged: 4g", accuracy_flipped), 2)`,
  with item-rest r `r flip_ci(rc2, "flagged: 4g")` and
  `r flip_ci("tom_story4_deception_reality_check_3", "flagged: 4g")`:
  positive but weak (reality check 2's interval only just clears 0), a
  third to a half of the 4d values
  (`r n2(fc(rc2, "other content: 4d", r_ir))` and
  `r n2(fc("tom_story4_deception_reality_check_3", "other content: 4d", r_ir))`).
  The data are consistent with a key that conflicts with the scene. A
  scoped correction (re-key 4g to "no") is therefore the preferred option,
  because it keeps the
  `r big(sum(res_items$flip_check$n[res_items$flip_check$content_part == "flagged: 4g" & res_items$flip_check$language == "all"]))`
  trials, but only once the content team confirms that "no" is right for
  the 4g scene; exclusion is the fallback. This chapter keeps the exclusion
  until then.
- **Scoped: story 2 on the V0 Bogotá content.** Reality check 1 on the
  first Bogotá content (2e, May–June 2024) is `r n2(k$s2rc1_2e, 3)`
  (n = `r k$s2rc1_2e_n`) but `r n2(k$s2rc1_oth, 3)` on later content. The
  V0 pattern is the mirror image of the later one:
  `r pct(s2r("reality_check_1", "V0 Bogota (2e)", "coats"))` of V0 children
  chose "coats" (key "shelf"), and
  `r pct(s2r("reality_check_1", "later content", "shelf"))` later chose
  "shelf". A flipped key would give `r n2(fc("tom_story2_moral_reasoning_reality_check_1", "flagged: 2e", accuracy_flipped), 3)`,
  with item-rest r `r flip_ci("tom_story2_moral_reasoning_reality_check_1", "flagged: 2e")`.
  The mechanism is undetermined: an inverted key, a question identity
  swapped in the retroactive V0 map, or the V0 content itself. A clean swap
  of the two reality checks is not supported: the paired reality check 2
  (key "coats") drew "coats" from
  `r pct(s2r("reality_check_2", "V0 Bogota (2e)", "coats"))` of V0 children
  and `r pct(s2r("reality_check_2", "later content", "coats"))` later. No V0
  corpus survives to settle it. The scoped exclusion is the conservative
  choice; a key or identity correction is the alternative if the V0 content
  can be recovered.
- **Spanish reality_known false belief.** Stories 1 and 7 still
  discriminate in es-CO (above), so they are un-excluded without a language
  flag. Story 13 does not discriminate in es-AR or es-CO and stays below the
  other languages at matched age; it keeps the analysis flag, but no Airtable
  row is proposed until the translation check. In es-AR it adds nothing to
  the UTDT form: reliability is `r n2(k$ar13_with, 3)` with it and
  `r n2(k$ar13_without, 3)` without it (2PL + guessing, empirical). The
  build's `spanish_fb_flag`, which the analyses run with and without,
  still covers all three stories.
- **Un-exclude:** the other `r k$v_un - k$v_es_disc` items behave normally;
  of these, story 14's moral reality check 1 is weak.

::: {.callout-note collapse="true"}
## The scoped rows under a flipped key, and story 2's reality checks in V0

```{r flip-check}
res_items$flip_check |>
  transmute(item = uid_label(story_uid), content_part, language, n, accuracy, accuracy_flipped,
            r_ir, r_ir_lo, r_ir_hi, r_ir_flipped, r_ir_flipped_lo, r_ir_flipped_hi) |>
  knitr::kable(digits = 3, caption = "Accuracy and item-rest r (within dataset × form, Fisher-pooled) with the deployed and the flipped key")
res_items$story2_rc_responses |>
  knitr::kable(digits = 3, caption = "Story 2's two reality checks: which option children chose, V0 Bogotá vs later content (DE trials with hostile-attribution answer strings omitted)")
```
:::

::: {.callout-important}
## What this means for the Airtable exclusion table

The DCC's plan is to keep exclusions in the Airtable
table, so they are documented outside code rather than written into
levantemodels. The verdicts above fit that plan, with three changes:

1. Rows need to be **story-level** (for example
   `tom_story17_reference_reference`), not generic. The four keep-excluded
   rows fit today's columns (`item_uid`, `exclude`, `dataset`).
2. Three rows need a **content scope** that those columns cannot express:
   story-4 reality checks 2 and 3 on image 4g, and story-2 reality check 1
   on the V0 content (2e before 2024-07-01), that is, two content scopes
   covering three items. The table would need a content-version (or
   date-window) column. The 4g rows could instead become a key correction,
   which would need a place of its own (see recommendation 2).
3. **No language-scoped rows yet.** Stories 1 and 7 discriminate in Spanish.
   Story 13 should wait for the translation check; if it then needs a
   Spanish row, the table would also need a language column.

The `policy_exclude` and `spanish_fb_flag` flags this chapter uses are
defined in `tasks/_build_stories_data.R`. They stand in for those rows until
the table can hold them.
:::

```{r airtable-rows}
res_items$item_verdicts |>
  filter(!verdict %in% c("un-exclude", "un-exclude (weak item)")) |>
  transmute(item = uid_label(story_uid), verdict, proposed_airtable_row) |>
  knitr::kable(caption = glue("Proposed story-level rows other than plain un-exclusions (the other {k$v_un - k$v_es_disc} items only need their generic rows removed)"))
```

::: {.callout-note collapse="true"}
## Evidence for the scoped rows, by content and language

```{r scoped-split}
res_items$scoped_split |>
  transmute(item = uid_label(story_uid), scope_part, language, n, accuracy, chance,
            p_below = fp(p_below)) |>
  knitr::kable(digits = 3)
```
:::

# What organizes the items

## Difficulty and age sensitivity

Item difficulty comes from a per-language Rasch model with guessing fixed at
chance, z-scored within language. Age slopes come from item × language
logistic regressions of correctness on age with dataset fixed effects. A
slope needs n ≥ 50 and at least 10 correct and 10 incorrect answers.[^minority]
The R² F-tests on item × language cells treat the cells as independent,
although the same item appears in several languages. The item-level rows
(one value per item, averaged over languages) and the random-effects tests
in the next section do not, and the text leads with them.

[^minority]: The minority rule (≥ 10 correct and ≥ 10 incorrect) was added
after the first run, where one near-ceiling item gave a separated slope of
−170. That slope dominated the R² values. Because of the rule, control
slopes come only from samples where controls are not at ceiling. Unfiltered
slopes are kept in `age_slope_raw`.

```{r construct-r2}
res_items$construct_r2 |>
  filter(outcome %in% c("difficulty (IRT, z within language)", "age slope (logit per year)")) |>
  transmute(outcome, unit, scheme, n_items, R2, dR2, p = fp(p)) |>
  knitr::kable(digits = 3, caption = "Variance in item difficulty and item age slope explained by each scheme (for age slopes on cells, dR2 is over language; cell-level p-values treat cells as independent)")
```

```{r fig-schemes, fig.height=6.5}
sm <- res_items$scheme_means |>
  mutate(scheme = recode(scheme, question_type = "Question type", old_group = "Story type (old)",
                         new_construct = "Construct (new)"),
         scheme = factor(scheme, c("Question type", "Story type (old)", "Construct (new)")))
p_slope <- ggplot(sm, aes(x = mean_age_slope, y = fct_reorder(level, mean_age_slope))) +
  geom_linerange(aes(xmin = mean_age_slope - se_age_slope, xmax = mean_age_slope + se_age_slope)) +
  geom_point(size = 2.5) +
  facet_grid(rows = vars(scheme), scales = "free_y", space = "free_y") +
  labs(x = "Raw age slope (logit/yr, ± SE)", y = NULL) +
  theme(strip.text.y = element_blank())
p_diff <- ggplot(sm, aes(x = mean_difficulty_z, y = fct_reorder(level, mean_age_slope))) +
  geom_vline(xintercept = 0, linetype = "dotted") +
  geom_point(size = 2.5) +
  facet_grid(rows = vars(scheme), scales = "free_y", space = "free_y") +
  labs(x = "Difficulty (z within language)", y = NULL) +
  theme(axis.text.y = element_blank())
patchwork::wrap_plots(p_slope, p_diff, nrow = 1) +
  patchwork::plot_annotation(title = "Question type orders difficulty; raw age slopes (no guessing floor) barely differ")
```

```{r qtype-summary}
res_items$qtype_summary |>
  knitr::kable(digits = 3, caption = "By question type (item × language cells; raw logit age slopes)")
```

```{r slope-numbers}
st <- res_items$slope_tests
spq <- res_items$slope_by_qtype |> filter(filter == "symmetric (floor model)")
k$fl_fb   <- pick(spq, mean_slope_floor, question_type == "false_belief")
k$fl_rc   <- pick(spq, mean_slope_floor, question_type == "reality_check")
k$ame_fb  <- pick(spq, mean_ame_floor, question_type == "false_belief")
k$ame_rc  <- pick(spq, mean_ame_floor, question_type == "reality_check")
k$fl_diff <- pick(st, fb_minus_control, outcome == "slope_floor", scheme == "question_type")
k$fl_lo   <- pick(st, fb_minus_control_lo, outcome == "slope_floor", scheme == "question_type")
k$fl_hi   <- pick(st, fb_minus_control_hi, outcome == "slope_floor", scheme == "question_type")
k$fl_p    <- pick(st, p, outcome == "slope_floor", scheme == "question_type")
k$raw_diff <- pick(st, fb_minus_control, outcome == "age_slope", scheme == "question_type")
k$raw_lo   <- pick(st, fb_minus_control_lo, outcome == "age_slope", scheme == "question_type")
k$raw_hi   <- pick(st, fb_minus_control_hi, outcome == "age_slope", scheme == "question_type")
k$raw_p    <- pick(st, p, outcome == "age_slope", scheme == "question_type")
k$nc_raw_p <- pick(st, p, outcome == "age_slope", scheme == "new_construct")
k$nc_fl_p  <- pick(st, p, outcome == "slope_floor", scheme == "new_construct")
ib <- \(sch, col) pick(cr, {{ col }}, outcome == "age slope (logit per year)", unit == "item (mean over languages)",
                       scheme == sch)
isl <- res_rel$item_slopes_by_type
k$a_fb <- pick(isl, median_a, question_type == "false_belief")
k$a_rc <- pick(isl, median_a, question_type == "reality_check")
asf <- res_items$age_slopes_floor
k$n_floor_fail <- sum(is.na(asf$slope_floor) & asf$cc_accuracy >= .1 & asf$cc_accuracy <= .9)
```

- **Difficulty is question type.** Question type explains
  `r n2(pick(cr, R2, outcome == "difficulty (IRT, z within language)", unit == "item (mean over languages)", scheme == "question_type (4)"))`
  of the variance across items (`r n2(k$r2_qt)` across item × language
  cells); control vs target alone explains
  `r n2(pick(cr, R2, outcome == "difficulty (IRT, z within language)", unit == "item x language", scheme == "control vs target (2)"))`
  across cells. Either construct scheme explains about `r n2(k$r2_og, 2)`.
  Reference and false-belief questions are hard and reality checks easy.
  June found the same.
- **Raw age slopes barely differ by question type.** Across items, question
  type explains R² = `r n2(ib("question_type (4)", R2), 3)` of the raw
  slopes (p = `r n2(ib("question_type (4)", p), 2)`), and a random-effects
  comparison puts false belief minus controls at `r n2(k$raw_diff, 3)`
  [`r n2(k$raw_lo, 3)`, `r n2(k$raw_hi, 3)`] logits per year. But raw
  slopes ignore the guessing floor, and their filter drops ceiling cells
  while keeping cells near the floor, which flattens false-belief slopes.
  The next section corrects both.
- The new construct scheme's cell-level test (ΔR² = `r n2(k$dr2_nc, 3)`,
  p = `r n2(k$p_nc, 3)`) does not hold up with item random effects
  (p = `r n2(k$nc_raw_p, 2)` for raw slopes, `r n2(k$nc_fl_p, 2)` above the
  floor).
- Dropping the Spanish reality_known false-belief items changes none of
  this (see the sensitivity table below).

### Age sensitivity above the guessing floor

Each item × language cell is refitted with the floor fixed at chance,
P = c + (1 − c) · logistic(a + b · age), by maximum likelihood. Cells are
kept symmetrically: chance-corrected accuracy between .10 and .90, n ≥ 50,
SE(b) ≤ 1 (`r k$n_floor_fail` cells in that range did not converge). The
probability-scale slope is the average marginal effect over the cell's
children. Question types are compared with a random-effects
meta-regression with crossed item and language intercepts
(`metafor::rma.mv`).

```{r floor-slopes}
res_items$slope_by_qtype |>
  select(filter, question_type, n_cells, n_items, mean_accuracy, mean_slope_logit, mean_slope_floor,
         mean_ame_logit, mean_ame_floor) |>
  knitr::kable(digits = 3, caption = "Age slopes by question type: raw logit, above the floor, and on the probability scale (per year), under the original and the symmetric filter")
res_items$slope_tests |>
  mutate(p = fp(p)) |>
  knitr::kable(digits = 3, caption = "Random-effects tests (items and languages as crossed random intercepts; fb_minus_control from REML, 95% CI)")
```

- **Above the floor, false-belief targets are the most age-sensitive
  items.** Their slope is `r n2(k$fl_fb, 3)` logits per year against
  `r n2(k$fl_rc, 3)` for controls; the difference is `r n2(k$fl_diff, 3)`
  [`r n2(k$fl_lo, 3)`, `r n2(k$fl_hi, 3)`], and question type matters
  (p = `r n2(k$fl_p, 3)`). On the probability scale, false-belief accuracy
  rises by `r n2(k$ame_fb, 3)` per year against `r n2(k$ame_rc, 3)` for
  controls. The 2PL of the reliability model agrees: the median
  false-belief slope is `r n2(k$a_fb)`, against `r n2(k$a_rc)` for controls.
- So June's "false-belief targets carry the age signal" survives, on the
  chance-corrected scale. On raw logits the two cannot be told apart, and
  controls also look flat because many are at ceiling.

::: {.callout-note collapse="true"}
## Sensitivity (Spanish false belief dropped), by-language R², and calibration notes

```{r construct-sens}
res_items$construct_r2_sens |>
  transmute(outcome, unit, scheme, n_items, R2, dR2, p = fp(p)) |>
  knitr::kable(digits = 3)
res_items$construct_r2_by_lang |>
  transmute(language, outcome, scheme, n_items, R2, p = fp(p)) |>
  knitr::kable(digits = 3)
res_items$calib_summary |>
  knitr::kable(caption = "Per-language calibrations used for item difficulty. en-US covers stories 1–6 only; en-GB rests on sparse fCAT data.")
```
:::

## Dimensionality

The factor-analytic samples are complete fixed forms, one run per child,
with policy items dropped. Following June, items need coverage > .4 and
.08 < p < .97. Leipzig is the June sample. Western and Bogotá give the same
stories in other languages, and the UTDT no-HA retest-B form gives
`r big(pick(res_dim$samples, children, sample == "ar_nhb"))` es-AR children
on stories 13–18.

```{r dim-samples}
res_dim$samples |>
  transmute(sample = dim_labels[sample], role, language, stories, datasets, runs, children,
            items, age = paste0(age_min, "–", age_max, " (median ", age_median, ")"),
            n_age, first_run, last_run) |>
  knitr::kable(caption = "Dimensionality samples")
```

::: {.callout-note collapse="true"}
## Items removed from each sample

```{r dim-removed}
res_dim$removed_items |>
  mutate(sample = dim_labels[sample], item = if_else(is.na(item), item, uid_label(item))) |>
  knitr::kable(digits = 3)
```
:::

```{r dim-eigen}
res_dim$eigen |>
  transmute(sample = dim_labels[sample], n_items, n_persons, ev1, ev2, ev1_ev2, ev1_ev2_se,
            ev1_prop, n_ev_gt1, pa_june_nfact, pa_resample_nfact, pa_resample_ncomp, mean_offdiag_r) |>
  knitr::kable(digits = 2, caption = "Eigenvalues of the tetrachoric matrix. ev1_ev2_se is a bootstrap SE (200 person resamples). pa_june_nfact uses June's parallel analysis; pa_resample_nfact (factors) and pa_resample_ncomp (components) compare against column-resampled binary data (20 draws).")
```

```{r fig-scree, fig.height=5}
res_dim$scree |>
  filter(sample %in% dim_primary, component <= 10) |>
  mutate(sample = factor(dim_labels[sample], dim_labels[dim_primary])) |>
  pivot_longer(c(eigenvalue, pa_resampled_pc), names_to = "series", values_to = "value") |>
  mutate(series = recode(series, eigenvalue = "observed",
                         pa_resampled_pc = "resampled-data reference")) |>
  ggplot(aes(component, value, linetype = series)) +
  geom_line() + geom_point(size = 1.5) +
  facet_wrap(vars(sample), nrow = 1) +
  scale_x_continuous(breaks = c(1, 5, 10)) +
  labs(x = "Component", y = "Eigenvalue", linetype = NULL,
       title = "Scree with a resampled parallel-analysis reference") +
  theme(legend.position = "bottom")
```

- The Leipzig item-bank form's ev1/ev2 (`r n2(k$ev_de)`, on
  `r pick(res_dim$eigen, n_items, sample == "de_ib")` items) is close to
  June's 1.90 on 26 items; the June-style rebuild gives
  `r n2(pick(res_dim$eigen, ev1_ev2, sample == "de_ib_junelike"))`. The
  Western, Bogotá and es-AR forms are at
  `r rng(k$ev_other)`. Leipzig is near ceiling: its items average
  p = `r n2(mean(res_dim$item_info$p[res_dim$item_info$sample == "de_ib" & res_dim$item_info$kept]))`,
  and `r sum(res_dim$item_info$p[res_dim$item_info$sample == "de_ib" & res_dim$item_info$kept] > .9)`
  of its `r pick(res_dim$eigen, n_items, sample == "de_ib")` kept items
  exceed .90. Ceiling is a likely cause of the low ratio but not a tested
  one: the samples also differ in stories, language and age.
- June's parallel-analysis method gives
  `r rng(res_dim$eigen$pa_june_nfact, 0)` factors in every sample, as it
  gave June 11. Against resampled binary data, the primary samples need
  `r rng(res_dim$eigen$pa_resample_ncomp[res_dim$eigen$sample %in% dim_primary], 0)`
  components but `r rng(res_dim$eigen$pa_resample_nfact[res_dim$eigen$sample %in% dim_primary], 0)`
  factors (es-AR:
  `r pick(res_dim$eigen, pa_resample_nfact, sample == "ar_nhb")`). Parallel
  analysis on tetrachoric correlations of items with a guessing floor is
  uninformative here: it neither supports June's 11 nor settles on a
  number. The bifactor models below are the better test.

## What the covariance follows: question type, story or construct

For each item pair, the tetrachoric correlation is compared within vs
between groupings (June's method). Intervals are ± 1.96 bootstrap SD over
200 person resamples. In a fixed form, story type ("old group") gives
exactly the same numbers as story, because each form has one story per
story type, so it is omitted. "Question type (different stories)" compares
only pairs from different stories, so it is not confounded with story.

Tetrachoric correlations are not a clean test of grouping here. Question
type explains about a third of item difficulty, and items of similar
difficulty with the same guessing floor correlate more on the tetrachoric
scale even under one factor. The same contrasts are therefore also computed
on Q3, the residual correlations after the unidimensional 2PL + guessing
model of the local-dependence section, which models difficulty and
guessing. Their intervals come from a person bootstrap of the residual
matrix (item parameters and θ held fixed).

```{r grouping}
res_dim$grouping_contrast |>
  filter(sample %in% c(dim_primary, "ar_nhb_nofb", "de_ib_junelike"), grouping != "old group") |>
  transmute(sample = dim_labels[sample], grouping, n_within, n_between, mean_within, mean_between,
            within_minus_between, diff_lo, diff_hi) |>
  knitr::kable(digits = 3, caption = "Within- minus between-group mean tetrachoric r")
res_dim$q3_grouping_contrast |>
  filter(sample %in% dim_primary, grouping != "old group") |>
  transmute(sample = dim_labels[sample], grouping, n_within, n_between, mean_within, mean_between,
            within_minus_between, diff_lo, diff_hi) |>
  knitr::kable(digits = 3, caption = "The same contrasts on Q3 residual correlations (2PL + guessing, unidimensional)")
```

```{r fig-cells, fig.height=5.5}
res_dim$grouping_cells |>
  filter(sample %in% dim_primary) |>
  mutate(sample = factor(dim_labels[sample], dim_labels[dim_primary]),
         panel = if_else(str_detect(cell, "construct"), "Construct", "Story × type"),
         cell = fct_inorder(cell)) |>
  ggplot(aes(x = mean_r, y = fct_rev(cell), colour = sample)) +
  geom_pointrange(aes(xmin = r_lo, xmax = r_hi), position = position_dodge(width = .6)) +
  facet_grid(rows = vars(panel), scales = "free_y", space = "free_y") +
  scale_colour_manual(values = set_names(dim_pal, dim_labels[names(dim_pal)]), name = NULL) +
  labs(x = "Mean tetrachoric r (± 1.96 bootstrap SD)", y = NULL,
       title = "The 'story effect' lives in same-story, same-type pairs") +
  guides(colour = guide_legend(nrow = 2)) +
  theme(legend.position = "bottom")
```

```{r dim-numbers}
gc <- res_dim$grouping_contrast
qt_contrast <- gc$within_minus_between[gc$grouping == "question type" & gc$sample %in% dim_primary]
st_contrast <- gc$within_minus_between[gc$grouping == "story" & gc$sample %in% dim_primary]
nc_contrast <- gc$within_minus_between[gc$grouping == "new construct" & gc$sample %in% dim_primary]
gcell <- res_dim$grouping_cells
q3c <- res_dim$q3_grouping_contrast |> filter(sample %in% dim_primary)
q3_qt <- q3c |> filter(grouping == "question type")
q3_st <- q3c |> filter(grouping == "story")
q3_qt_ds <- q3c |> filter(grouping == "question type (different stories)")
k$q3_qt_le_story <- sum(q3_qt$within_minus_between <= q3_st$within_minus_between[match(q3_qt$sample, q3_st$sample)])
qt_ds <- gc$within_minus_between[gc$grouping == "question type (different stories)" & gc$sample %in% dim_primary]
```

- **Question type** has a reliably positive tetrachoric contrast in every
  sample (`r rng(qt_contrast, 3)`, all pairs; `r rng(qt_ds, 3)` among
  different-story pairs), smaller than June's .101. The June-style rebuild of the Leipzig
  data gives
  `r n2(pick(gc, within_minus_between, sample == "de_ib_junelike", grouping == "question type"), 3)`.
  That rebuild differs from the primary sample in item identities,
  exclusions, runs (all vs first per child) and items all at once, so the
  change cannot be assigned to any one of them. On Q3 residuals the
  question-type contrast shrinks to `r rng(q3_qt$within_minus_between, 3)`
  (`r rng(q3_qt_ds$within_minus_between, 3)` among different-story pairs)
  and is no larger than the story contrast in `r k$q3_qt_le_story` of the
  `r nrow(q3_qt)` samples. It vanishes in Leipzig
  (`r n2(pick(q3_qt, within_minus_between, sample == "de_ib"), 3)`) and
  stays clearly positive in Western
  (`r n2(pick(q3_qt, within_minus_between, sample == "wca_ib"), 3)`). Part of
  the tetrachoric clustering by question type therefore reflects shared
  difficulty and guessing; what remains once they are modelled is weak and
  differs between samples.
- **Story** contrasts are `r rng(st_contrast, 3)` (Q3:
  `r rng(q3_st$within_minus_between, 3)`), and they come entirely
  from pairs of the same story *and* the same question type. In Leipzig,
  same-story pairs of the same type correlate
  `r n2(pick(gcell, mean_r, sample == "de_ib", cell == "same story, same question type"), 3)`;
  same-story pairs of different types correlate
  `r n2(pick(gcell, mean_r, sample == "de_ib", cell == "same story, different question type"), 3)`,
  no more than pairs from different stories.
- **Constructs** organize little (`r rng(nc_contrast, 3)`). The only test
  not confounded with story compares different-story pairs in the same
  construct with pairs across constructs. In the three stories 1–6 forms,
  same-construct pairs correlate less (Leipzig
  `r n2(pick(gcell, mean_r, sample == "de_ib", cell == "different story, same new construct"), 3)`
  vs `r n2(pick(gcell, mean_r, sample == "de_ib", cell == "different story, different new construct"), 3)`).
  es-AR is the exception: its construct contrast is reliably positive
  (`r n2(pick(gc, within_minus_between, sample == "ar_nhb", grouping == "new construct"), 3)`
  [`r n2(pick(gc, diff_lo, sample == "ar_nhb", grouping == "new construct"), 3)`,
  `r n2(pick(gc, diff_hi, sample == "ar_nhb", grouping == "new construct"), 3)`]),
  and it roughly halves without the flagged story-13 item
  (`r n2(pick(gc, within_minus_between, sample == "ar_nhb_nofb", grouping == "new construct"), 3)`).
  June's table showed .045 for constructs, above story's .039, although its
  prose called the construct contrast "near zero". On Leipzig data, that
  contrast reproduces only on the uncorrected, June-style rebuild
  (`r n2(pick(gc, within_minus_between, sample == "de_ib_junelike", grouping == "new construct"), 3)`).
- Cleaning raised the mean inter-item correlation from
  `r n2(pick(res_dim$eigen, mean_offdiag_r, sample == "de_ib_junelike"), 3)`
  (June-style rebuild) to
  `r rng(res_dim$eigen$mean_offdiag_r[res_dim$eigen$sample %in% dim_primary], 3)`.
  The defective items had depressed the between-group baseline.

```{r efa}
res_dim$efa2_summary |>
  filter(sample %in% c(dim_primary, "ar_nhb_nofb", "de_ib_junelike")) |>
  mutate(sample = dim_labels[sample]) |>
  knitr::kable(digits = 3, caption = "Two-factor oblimin EFA of the tetrachoric matrix")
res_dim$efa2_by_qtype |>
  filter(sample %in% dim_primary) |>
  mutate(sample = dim_labels[sample]) |>
  pivot_wider(names_from = sample, values_from = c(n, F1, F2), names_vary = "slowest") |>
  knitr::kable(digits = 2, caption = "Mean loadings on F1/F2 by question type")
```

A two-factor EFA still gives no sub-construct that replicates. What defines
the second factor changes from sample to sample: nothing in Leipzig, an
emotion factor in Western, and a split between controls and targets in
es-AR.

## Local dependence

A 2PL with guessing fixed at chance was fitted as a unidimensional model
and as bifactor models with story testlets or question-type factors. Items
answered at or below chance in the sample get a floor of 0, because a fixed
floor above the observed rate cannot fit; the reliability and invariance
models keep the floor at chance and are rerun under this rule as a check. Every slope has an N(0, 2²) prior (MAP-EM); without
it, fixed-floor fits diverge. AIC and BIC are evaluated at the MAP
estimates. dAIC and dBIC are bifactor minus unidimensional, so negative
values favour the bifactor.

```{r ld}
res_dim$irt_ld |>
  filter(family == "guess", sample %in% c(dim_primary, "ar_nhb_nofb")) |>
  transmute(sample = dim_labels[sample], n_items, n_persons, dAIC_story, dBIC_story,
            dAIC_qtype, dBIC_qtype, ECV, omega_h, omega_total, omega_h_over_t = omega_h / omega_total,
            PUC, rxx_uni,
            cor_theta_uni_general, info_overstatement, q3_within_story, q3_between_story,
            `within-story pairs with Q3 > .2` = paste0(`n_within_q3_gt_.2`, "/", n_within_pairs)) |>
  knitr::kable(digits = 3, caption = "Unidimensional vs bifactor 2PL + guessing")
```

- There is some story-level dependence, but little of consequence. The
  story bifactor improves AIC in every sample. BIC prefers the
  unidimensional model in the three smaller samples and is indifferent in
  es-AR. θ from the two models correlates `r k$cor_uni_g`.
- **How dominant is the general factor?** ECV is `r k$ecv`, below the
  usual .70 guideline in Leipzig
  (`r n2(pick(ld, ECV, sample == "de_ib"))`) and es-AR
  (`r n2(pick(ld, ECV, sample == "ar_nhb"))`). Omega-h (`r k$omh`) comes
  from standardized normal-ogive loadings that ignore the guessing floor, so
  it is not comparable with the observed-score reliabilities elsewhere in
  this chapter. It is also inflated by story factors whose specific
  loadings mix signs (`r k$mixed_stories` of `r k$n_story_factors` story
  factors in the four samples; in Leipzig story 2, Σ|λs| =
  `r n2(k$de_s2_abs)` but |Σλs| = `r n2(k$de_s2_sum)`), and 2-item testlets
  are identified only through the slope prior. The ratio omega-h / omega-t
  (`r rng(ld$omega_h / ld$omega_total)`) is the safer summary. By that
  ratio the general factor dominates in every sample; by ECV it dominates
  clearly only in Western and Bogotá, and in Leipzig and es-AR the case for
  it is weaker.
- The unidimensional model overstates test information by `r k$info_over`,
  in line with the September audit's roughly
  `r pct(pick(res_dim$june_reference, june_value, metric == "information overstatement (uni vs bifactor)"))`
  on the v2_3 calibration data.
- A question-type bifactor fits as well as the story bifactor, and in es-AR
  clearly better. Most of the es-AR advantage comes from the below-chance
  Spanish false-belief item: without story 13's item, dBIC for the
  question-type model goes from
  `r n2(pick(res_dim$irt_ld, dBIC_qtype, family == "guess", sample == "ar_nhb"), 0)` to
  `r n2(pick(res_dim$irt_ld, dBIC_qtype, family == "guess", sample == "ar_nhb_nofb"), 0)`.

The residual dependence sits in a handful of "doublets". Most are yes/no
questions about the same scene with the same key. One pair of location
false-belief questions from different stories behaves the same way:

```{r doublets}
res_dim$ld_pairs_top |>
  filter(sample %in% dim_primary) |>
  transmute(sample = dim_labels[sample], item_i = uid_label(i), item_j = uid_label(j),
            same_story, r, q3, code_i, answer_i, p_i, code_j, answer_j, p_j) |>
  knitr::kable(digits = 2, caption = "Item pairs with Q3 > .2 (2PL + guessing, unidimensional)")
```

::: {.callout-note collapse="true"}
## Robustness: the same models without a guessing floor

```{r ld-noguess}
res_dim$irt_ld |>
  filter(family == "noguess") |>
  transmute(sample = dim_labels[sample], dAIC_story, dBIC_story, dAIC_qtype, dBIC_qtype,
            ECV, omega_h, cor_theta_uni_general, info_overstatement) |>
  knitr::kable(digits = 3)
```
:::

## Sub-scales

```{r subscales}
res_dim$subscales |>
  filter(sample %in% dim_primary) |>
  transmute(sample = dim_labels[sample], scale, n_items, alpha, alpha_sb_to_all,
            twopl_marginal_rxx, rasch_marginal_rxx, age_r, age_r_lo, age_r_hi,
            pct_rand_alpha_below, pct_rand_age_r_below) |>
  knitr::kable(digits = 3, caption = "Sub-scales. alpha_sb_to_all projects alpha to the full scale's length (Spearman–Brown). pct_rand_* is the percentile among 500 random item subsets of the same size.")
```

```{r fig-subscales, fig.height=4.5}
ss <- res_dim$subscales |> filter(sample %in% dim_primary) |>
  mutate(sample = factor(dim_labels[sample], dim_labels[dim_primary]),
         scale = factor(scale, c("all items", "targets only", "false-belief only", "controls only")))
ggplot(ss, aes(x = scale, y = age_r)) +
  geom_linerange(aes(ymin = rand_age_r_q025, ymax = rand_age_r_q975), linewidth = 4,
                 colour = "grey80", data = filter(ss, scale != "all items")) +
  geom_pointrange(aes(ymin = age_r_lo, ymax = age_r_hi)) +
  facet_wrap(vars(sample), nrow = 1) +
  scale_x_discrete(limits = levels(ss$scale)) +
  labs(x = NULL, y = "r(proportion correct, age)",
       title = "Age validity of sub-scales against random item subsets of the same size (grey)") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

June's answer stands: no sub-scale improves on the full set. False-belief
items are the most internally consistent per item (Spearman–Brown-projected
alpha `r rng(res_dim$subscales$alpha_sb_to_all[res_dim$subscales$scale == "false-belief only" & res_dim$subscales$sample %in% dim_primary])`),
but a false-belief-only scale is no more age-valid than random subsets. In
es-AR it is less age-valid than the full scale. The controls are the least
age-valid items (discussed next).

# Controls

This section re-tests June's claims about the reality-check (control)
questions: that they correlate with targets, that they do not gate the
targets' age sensitivity, and that keeping them helps reliability.

```{r controls-cor}
res_items$controls_cor |>
  knitr::kable(digits = 3, caption = "Person-level correlations of control and target accuracy (runs with age)")
```

::: {.callout-note collapse="true"}
## Within each dataset × form

```{r controls-cells}
res_items$controls_cells |>
  mutate(dataset = ds_label(dataset)) |>
  select(-cell) |>
  knitr::kable(digits = 3)
```
:::

Pooled over runs, as June did it, control and target accuracy correlate
`r n2(k$r_ct_pooled)`. Much of that is between-site differences: within
dataset × form the correlation is `r n2(k$r_ct_within)`, and controls
correlate
`r n2(pick(cc, r_control_age, method == "within dataset x form, Fisher-pooled", item_set == "policy items dropped"))`
with age. These are accuracy scores, and fCAT runs have a median of
`r median(res_items$controls_cells$median_ctrl[res_items$controls_cells$form == "fcat_random3"])`
controls each, which attenuates the correlations.

## Gating

```{r gating}
res_items$gating_june |> select(-method) |>
  knitr::kable(digits = 3, caption = "June method: lm(story target accuracy ~ scale(age)) by whether the story's control was passed")
res_items$gating_glmm |>
  mutate(interaction_p = fp(interaction_p)) |>
  knitr::kable(digits = 3, caption = "GLMM on target trials: correct ~ age × ctrl_pass + dataset + (1 | run) + (1 | story_uid). Slopes in logits per year.")
```

```{r fig-gating, fig.height=4.5}
res_items$gating_plot |>
  mutate(language = factor(language, lang_levels),
         ctrl_pass = if_else(ctrl_pass, "control passed", "control failed")) |>
  ggplot(aes(age_bin, target_acc, colour = ctrl_pass)) +
  geom_line() + geom_point(aes(size = n), alpha = .7) +
  facet_wrap(vars(language), nrow = 1) +
  scale_size_area(max_size = 4, guide = "none") +
  scale_colour_manual(values = c("control passed" = "grey30", "control failed" = "#CC6677"),
                      name = NULL) +
  labs(x = "Age (years)", y = "Target accuracy in the story",
       title = "Target accuracy by age, split by whether the story's control was passed") +
  theme(legend.position = "bottom")
```

```{r gating-numbers}
gg <- res_items$gating_glmm
g_all <- filter(gg, sample == "all languages"); g_de <- filter(gg, sample == "de-DE")
g_co  <- filter(gg, sample == "es-CO")
gj <- res_items$gating_june
```

June's method again finds no gating: the target age slope is
`r n2(pick(gj, target_age_beta_perSD, !ctrl_pass), 3)` per SD of age after
a failed control and `r n2(pick(gj, target_age_beta_perSD, ctrl_pass), 3)`
after a passed one. The trial-level GLMM finds a small pooled interaction
(`r n2(g_all$interaction, 3)` logits/yr, p = `r n2(g_all$interaction_p, 3)`).
It comes from de-DE, where targets after a failed control show no age slope
(`r n2(g_de$age_slope_fail, 3)` vs `r n2(g_de$age_slope_pass, 3)`;
`r n2(g_de$pct_instances_fail, 1)`% of story instances failed), and to a
lesser degree from en-US (`r n2(pick(gg, interaction, sample == "en-US"), 3)`,
p = `r n2(pick(gg, interaction_p, sample == "en-US"), 3)`); en-GB and es-AR
are borderline (p = `r n2(pick(gg, interaction_p, sample == "en-GB"), 3)` and
`r n2(pick(gg, interaction_p, sample == "es-AR"), 3)`). es-CO shows none
(`r n2(g_co$pct_instances_fail, 1)`% failed). Passing a control is itself related to ability, so this is a
description, not a causal gate. In a high-scoring sample like Leipzig, a
failed control probably marks a lapse.

## Do controls help the score?

On fixed forms, with guessing fixed at chance and production priors:

```{r controls-rel}
res_items$reliability |>
  mutate(row = row_number()) |>
  filter(row %in% c(1:8, 13:14)) |>
  select(sample, items, model, guessing, n_runs, n_items, n_controls, mean_items_answered,
         marginal_rxx_mirt, marginal_rxx, empirical_rxx, alpha_complete_cases,
         r_theta_age, r_theta_all_vs_targets) |>
  knitr::kable(digits = 3, caption = "Reliability and age validity with and without controls. The last two rows replicate June's method: pure Rasch without priors on June's 58-item Leipzig pool (item bank and retest forms, released identities, policy items kept).")
```

```{r controls-numbers}
rrow <- \(s, it, m, col) pick(res_items$reliability, {{ col }}, sample == s, items == it, model == m)
de_s <- "DE Leipzig, item-bank form (stories 1-6)"
ar_s <- "UTDT-AR main, no-HA retest-B form (stories 13-18)"
jn_s <- "DE Leipzig, June item pool (June method)"
# the items script and the dimensionality script replicate June's .92 on the same pool
stopifnot(abs(rrow(jn_s, "all items", "Rasch", marginal_rxx_mirt) - jp$rasch_marginal_rxx_june_method) < .001,
          abs(rrow(jn_s, "all items", "Rasch", empirical_rxx) - jp$rasch_empirical_rxx_eap) < .001)
k$dage_max <- res_items$reliability |> slice(1:8) |>
  select(sample, model, items, r_theta_age) |>
  pivot_wider(names_from = items, values_from = r_theta_age) |>
  summarise(d = max(abs(`all items` - `targets only`))) |> pull(d)
```

- **Keep the controls.** They raise reliability a little: Leipzig
  `r n2(rrow(de_s, "all items", "Rasch", empirical_rxx), 3)` with them vs
  `r n2(rrow(de_s, "targets only", "Rasch", empirical_rxx), 3)` without; UTDT
  `r n2(rrow(ar_s, "all items", "Rasch", empirical_rxx), 3)` vs
  `r n2(rrow(ar_s, "targets only", "Rasch", empirical_rxx), 3)` (Rasch +
  guessing, empirical). They change r(θ, age) by at most
  `r n2(k$dage_max)`. June reached
  the same conclusion.
- **Per item, controls are the weakest.** The dimensionality analysis finds
  them the least age-valid items. In Leipzig a controls-only score
  correlates
  `r n2(pick(res_dim$subscales, age_r, sample == "de_ib", scale == "controls only"), 3)`
  with age, below
  `r pct(1 - pick(res_dim$subscales, pct_rand_age_r_below, sample == "de_ib", scale == "controls only"))`
  of random `r pick(res_dim$subscales, n_items, sample == "de_ib", scale == "controls only")`-item
  subsets. Once alpha is projected to equal length, targets
  alone match all items in every sample. Controls help the score as extra
  items, not as especially good ones.
- **June's .92 → .89.** That comparison reproduces as mirt's
  `marginal_rxx` on June's sparse
  `r rrow(jn_s, "all items", "Rasch", n_items)`-item Leipzig pool
  (`r n2(rrow(jn_s, "all items", "Rasch", marginal_rxx_mirt), 3)` with
  controls vs `r n2(rrow(jn_s, "targets only", "Rasch", marginal_rxx_mirt), 3)`
  without; June: .915 vs .885). Children answered
  `r n2(rrow(jn_s, "all items", "Rasch", mean_items_answered), 1)` of those
  items on average. The same fit's empirical reliability is
  `r n2(rrow(jn_s, "all items", "Rasch", empirical_rxx), 3)` with controls
  and `r n2(rrow(jn_s, "targets only", "Rasch", empirical_rxx), 3)` without.
  `mirt::marginal_rxx()` integrates TI/(TI + var) over N(0, 1) and assumes
  every item is answered. With a Rasch latent variance of
  `r n2(rrow(jn_s, "all items", "Rasch", latent_var), 2)`, that inflates the
  value. The reliability section returns to this.

# Cross-language invariance

**Methods.** Multigroup Rasch with the guessing floor fixed at chance (the
LEVANTE convention), fitted with mirt. Each child contributes one run: the
earliest that contains the analysed stories. An item needs ≥ 100 responses
per group. The sequence is configural vs scalar, then a DIF screen with all
other items as anchors, then purified and iterated anchors, then partial
scalar. fCAT runs are included; their missing items are missing by design.
Effect sizes `db` are item difficulty for a group minus de-DE, in logits;
positive means harder at matched ability. dBIC = BIC(constrained) −
BIC(less constrained), so positive values favour the less constrained
model.

The ETS classes are adapted, not the ETS rule itself: A (negligible), B
(moderate) and C (large) use the 1- and 1.5-delta cut-offs (0.43 and 0.64
logits) on latent-logit db under the guessing floor, which runs larger than
the observed-score Mantel–Haenszel D-DIF they were defined for.
Significance is the item's BH-adjusted LR test, and class C does not
require |db| to be significantly above 0.43.

A maximum-likelihood 2PL was not usable: its configural models do not
converge, and slopes reach `r n2(max(res_inv$model_choice$max_slope), 0)`.
A 2PL with the production priors (d ~ N(0, 3), a ~ N(1, 0.3), MAP) is
usable and serves as the second model throughout: for the metric and
scalar tests, and for DIF effect sizes.

```{r model-choice}
res_inv$model_choice |>
  select(analysis, model, logLik, npar, BIC, converged, iterations, max_slope,
         prop_slopes_over_4, prop_slopes_negative) |>
  knitr::kable(digits = 2, caption = "Maximum-likelihood 2PL + guessing fits are degenerate")
```

```{r inv-samples}
res_inv$samples |>
  filter(analysis %in% c("A", "B")) |>
  select(analysis, group, n_children, age_median, age_q1, age_q3, age_missing, datasets, forms) |>
  knitr::kable(digits = 1, caption = "Analysis A: stories 1–6, de-DE vs en-US vs es-CO. Analysis B: stories 13–18, de-DE vs es-AR.")
```

```{r inv-exposure}
sB <- res_inv$samples |> filter(analysis == "B")
# how many earlier included ToM runs each analysed child had (the analysed run is the child's
# earliest run containing the analysed stories)
inc_runs <- runs |> filter(included) |> arrange(user_id, time_started) |>
  group_by(user_id) |> mutate(k_run = row_number()) |> ungroup()
first_with <- \(st, langs) trials |>
  filter(story %in% st, language %in% langs, !policy_exclude) |>
  distinct(user_id, run_id, time_started, language) |>
  arrange(time_started, run_id) |> group_by(user_id) |> slice(1) |> ungroup() |>
  left_join(inc_runs |> select(run_id, k_run), by = "run_id")
expB <- first_with(13:18, c("de-DE", "es-AR")) |>
  left_join(runs |> select(run_id, age), by = "run_id")
k$B_de_n <- sum(expB$language == "de-DE")
k$B_de_repeat <- sum(expB$language == "de-DE" & expB$k_run > 1)
k$B_ar_repeat <- sum(expB$language == "es-AR" & expB$k_run > 1)
k$Byoung_n <- sum(expB$language == "de-DE" & !is.na(expB$age) & expB$age < 9)
k$Byoung_repeat <- sum(expB$language == "de-DE" & !is.na(expB$age) & expB$age < 9 & expB$k_run > 1)
stopifnot(k$B_de_n == pick(sB, n_children, group == "de_DE"),
          k$Byoung_n == pick(res_inv$samples, n_children, analysis == "B_young", group == "de_DE"))
```

The groups are not language groups only. In stories 13–18 (analysis B),
`r k$B_de_repeat` of the `r k$B_de_n` German children
(`r pct(k$B_de_repeat / k$B_de_n)`) were on their second or third ToM run,
having taken stories 1–6 or 7–12 earlier;
`r if (k$B_ar_repeat == 0) "none of the es-AR children" else paste(k$B_ar_repeat, "es-AR children")`
were. The German children are older (median
`r n2(pick(sB, age_median, group == "de_DE"), 1)` vs
`r n2(pick(sB, age_median, group == "es_AR"), 1)` years), and es-AR comes
from one site (UTDT). In A, the German children are older too, and es-CO
pools four datasets from 2024–2026, including the V0 content. Group
differences in item difficulty are therefore differences between samples
that differ in language, age, prior exposure and site, and the estimated
group means (numerics note) are not language differences in theory of mind.

```{r inv-tests}
res_inv$tests |>
  filter(analysis %in% c("A", "B")) |>
  transmute(analysis, comparison, items, n_items, X2, df, X2_df, p = fp(p), dBIC) |>
  knitr::kable(digits = 2, caption = "Invariance tests, Rasch + guessing")
res_inv$tests_2pl |>
  transmute(analysis, comparison, df, X2_logLik, X2_logpost, p_logpost = fp(p_logpost), dBIC,
            nested_optimum_better) |>
  knitr::kable(digits = 2, caption = "Invariance tests, 2PL + guessing with production priors (MAP). X2 on the log-likelihood and on the log posterior; dBIC from mirt's BIC. A more constrained model reaching a better optimum (nested_optimum_better) means the less constrained fit found a local optimum, so only metric vs scalar is interpreted.")
```

```{r inv-test-numbers}
t2 <- \(a, cmp, col) pick(res_inv$tests_2pl, {{ col }}, analysis == a, comparison == cmp)
tr_ <- \(a, cmp, col, it = "all") pick(tst, {{ col }}, analysis == a, comparison == cmp, items == it)
isl <- res_rel$item_slopes_by_type
dlab <- res_inv$dif_long |> filter(analysis %in% c("A", "B"))
k$mdd_med <- median(dlab$mdd, na.rm = TRUE)
k$se_med  <- median(dlab$se, na.rm = TRUE)
k$mdd_B   <- median(dlab$mdd[dlab$analysis == "B"], na.rm = TRUE)
```

- **Stories 1–6 (A).** The LR test rejects scalar invariance under Rasch +
  guessing (χ²/df `r n2(k$A_scal_x2df, 1)`), but BIC is indifferent (dBIC
  `r n2(tr_("A", "scalar vs configural", dBIC), 1)`). Under the 2PL, BIC
  prefers the scalar model to the metric one (dBIC
  `r n2(t2("A", "scalar vs metric", dBIC), 0)`; LR on the log posterior
  p = `r n2(t2("A", "scalar vs metric", p_logpost), 2)`). Intercept
  differences are detectable, but neither model's BIC favours
  non-invariance. Without the V0 Bogotá runs, Rasch BIC prefers scalar too
  (dBIC `r n2(tr_("A_noV0", "scalar vs configural", dBIC), 0)`). The
  verdict depends on the guessing convention, though: with the floor set to
  0 for the two items answered at or below chance in a group, Rasch BIC
  rejects scalar invariance (dBIC `r n2(tr_("A_g0", "scalar vs configural", dBIC), 0)`).
  The evidence for non-invariance in stories 1–6 is model-dependent.
- **Stories 13–18 (B)** are non-invariant under both models (Rasch χ²/df
  `r n2(k$B_scal_x2df, 1)`, dBIC `r n2(tr_("B", "scalar vs configural", dBIC), 0)`;
  2PL dBIC `r n2(t2("B", "scalar vs metric", dBIC), 0)`, log-posterior LR
  p `r pe(t2("B", "scalar vs metric", p_logpost))`).
- Rasch DIF can partly be unequal slopes. The Rasch model forces equal
  discrimination, but false-belief items discriminate most (2PL median slope
  `r n2(pick(isl, median_a, question_type == "false_belief"))` vs
  `r n2(pick(isl, median_a, question_type == "reality_check"))` for
  controls), and the groups differ in mean ability.
- **Partial scalar invariance is a description, not a test.** The partial
  scalar models fit (p = `r n2(k$A_part_p)` in A, `r n2(k$B_part_p)` in B),
  but their anchors were chosen by the DIF screen on the same data, so the
  fit is expected. Without V0 Bogotá the partial scalar model is rejected
  (p = `r n2(tr_("A_noV0", "partial scalar vs configural", p), 4)`). Anchor status
  also reflects power: the median SE of db is `r n2(k$se_med)` logits, so
  an item's DIF would have to be about `r n2(k$mdd_med, 1)` logits to be
  detected with 80% power even before the BH adjustment
  (`r n2(k$mdd_B, 1)` in B, where the German group has
  `r k$B_de_n` children). "Class A" means "not detected".

```{r inv-dif}
rasch_db <- res_inv$dif |>
  filter(analysis %in% c("A_rasch", "B_rasch")) |>
  transmute(analysis = str_remove(analysis, "_rasch"), short,
            rasch_db_en_US = db_en_US, rasch_db_es_CO = db_es_CO, rasch_db_es_AR = db_es_AR)
res_inv$dif |>
  filter(analysis %in% c("A", "B"), class != "A") |>
  left_join(rasch_db, by = c("analysis", "short")) |>
  mutate(across(c(db_en_US, db_es_CO, db_es_AR),
                \(x) if_else(boundary & abs(x) > 10, NA_real_, x))) |>
  transmute(analysis, item = uid_label(short), question_type,
            db_en_US, db_es_CO, db_es_AR, rasch_db_en_US, rasch_db_es_CO, rasch_db_es_AR,
            acc_de_DE, acc_en_US, acc_es_CO, acc_es_AR, class,
            note = case_when(boundary ~ "unbounded (below floor); read the Rasch db",
                             near_floor ~ "near floor; Rasch db smaller",
                             TRUE ~ "")) |>
  arrange(analysis, question_type) |>
  knitr::kable(digits = 2, caption = "Non-invariant items (class B or C). db = difficulty vs de-DE (logits, Rasch + guessing); rasch_db = the same from pure Rasch.")
```

```{r fig-dif, fig.height=9}
dl <- res_inv$dif_long |>
  filter(analysis %in% c("A", "B")) |>
  mutate(capped = abs(db) > 5,
         db_plot = pmax(pmin(db, 5), -5),
         item = fct_reorder(uid_label(short), -story),
         question_type = factor(question_type, qtype_levels),
         class = factor(class, c("A", "B", "C")),
         contrast = str_replace_all(contrast, "_", "-"))
ggplot(dl, aes(x = db_plot, y = item, colour = question_type)) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_linerange(aes(xmin = pmax(db - 1.96 * se, -5), xmax = pmin(db + 1.96 * se, 5)),
                 data = filter(dl, !is.na(se), !capped)) +
  geom_point(aes(shape = class), size = 2.3) +
  geom_point(data = filter(dl, capped), shape = 62, size = 5, show.legend = FALSE) +
  facet_wrap(vars(contrast), nrow = 1) +
  scale_y_discrete(limits = levels(dl$item)) +
  scale_colour_manual(values = qtype_pal, name = NULL) +
  scale_shape_manual(values = c(A = 1, B = 17, C = 16), name = "ETS class") +
  labs(x = "db: difficulty vs de-DE (logits; capped at ±5, > = unbounded)", y = NULL,
       title = "Item DIF against de-DE",
       subtitle = "± 1.96 SE; anchors' db and SE come from freeing each anchor in turn") +
  theme(legend.position = "bottom", axis.text.y = element_text(size = 8))
```

```{r inv-numbers}
dq <- res_inv$dif_by_qtype
fb_A <- filter(dq, analysis == "A", question_type == "false_belief")
fb_B <- filter(dq, analysis == "B", question_type == "false_belief")
anc <- res_inv$dif |> filter(analysis %in% c("A", "B")) |>
  group_by(analysis) |> summarise(ctrl_anchor = sum(anchor & is_control),
                                  ctrl = sum(is_control), anchors = sum(anchor))
aba <- res_inv$age_band_acc
s3rc1 <- aba |> filter(short == "story3_interpretation_reality_check_1", n >= 20)
ctrl_flag <- \(a) pick(dbt, n_flagged, analysis == a, kind == "control")
ctrl_large_A <- res_inv$dif |> filter(analysis == "A", is_control, class == "C") |> pull(short)
noV0_class <- \(it) pick(res_inv$dif, class, analysis == "A_noV0", short == it)
dm <- res_inv$dif_by_type_models
dmv <- \(a, kd, col) pick(dm, {{ col }}, analysis == a, kind == kd)
ra <- res_inv$rasch_2pl_agreement
```

## Controls vs targets

June found controls to be the dominant source of cross-language
non-invariance (mean |DIF| ≈ 1.9 for controls vs ≈ 0.5 for targets). The
table below gives that metric (mean over items of the largest pairwise
difficulty gap, boundary items excluded) under both models, next to the
share of the DIF χ².

```{r inv-bytype-models}
dm |>
  transmute(analysis, kind, n_items, rasch_stage1 = rasch_s1, rasch_purified = rasch_s2,
            twopl_stage1 = twopl_s1, twopl_purified, n_large_2pl_purified) |>
  left_join(dbt |> filter(analysis %in% c("A", "B")) |>
              select(analysis, kind, n_flagged, share_of_items, share_of_s1_X2),
            by = c("analysis", "kind")) |>
  knitr::kable(digits = 2, caption = "Mean |db| (logits) for controls and targets: Rasch + guessing (stage 1: all other items as anchors; purified: final anchors) and 2PL with priors (stage 1: all other non-boundary items as anchors; purified: the Rasch anchors minus boundary items). Boundary items excluded.")
ra |>
  knitr::kable(digits = 2, caption = "Agreement of 2PL (stage 1) and Rasch (stage 1) db, by item type, with and without the boundary items")
```

- **Whether controls or targets carry more of the non-invariance depends
  on the model and the design.** In stories 1–6, controls are
  `r pct(k$A_ctrl_it)` of the items but carry `r pct(k$A_ctrl_x2)` of the
  stage-1 DIF χ², and their mean |db| is larger than the targets' under
  both models (Rasch `r n2(dmv("A", "control", rasch_s2))` vs
  `r n2(dmv("A", "target", rasch_s2))`; 2PL
  `r n2(dmv("A", "control", twopl_purified))` vs
  `r n2(dmv("A", "target", twopl_purified))`). Without V0 Bogotá it
  reverses (controls
  `r n2(pick(dbt, mean_abs_db, analysis == "A_noV0", kind == "control"))` vs
  targets `r n2(pick(dbt, mean_abs_db, analysis == "A_noV0", kind == "target"))`,
  Rasch). In stories 13–18 controls carry `r pct(k$B_ctrl_x2)` of the χ²
  and have the smaller Rasch |db|
  (`r n2(dmv("B", "control", rasch_s2))` vs `r n2(dmv("B", "target", rasch_s2))`),
  but the larger 2PL |db| (`r n2(dmv("B", "control", twopl_purified))` vs
  `r n2(dmv("B", "target", twopl_purified))`); for controls in B, the two
  models' db correlate only
  `r n2(pick(ra, r_controls_no_boundary, analysis == "B"))`. In June's own
  design on corrected data (J_policy, below), controls still exceed targets.
  The corrections shrank the controls' DIF, but "controls no longer drive
  the non-invariance" does not hold in general.
- **False-belief targets are non-invariant too.** `r fb_A$n_flagged` of
  `r fb_A$n_items` false-belief items are flagged in A and
  `r fb_B$n_flagged` of `r fb_B$n_items` in B. Most are harder in Spanish
  (story 4 fb_2 goes the other way), and the gaps persist at matched age
  (figure below). The flagged set is wider than the reality_known item:
  moral-reasoning, interpretation and second-order false-belief items are
  also non-invariant. (Story 14 fb_2 and fb_3 are yes/no moral judgments,
  such as "Is Bobby being mean and naughty…?", that the corpus classifies as
  false-belief questions.) Dropping the reality_known item alone changes
  almost nothing (flag agreement
  `r pct(min(res_inv$flag_agreement$agreement[res_inv$flag_agreement$analysis %in% c("A_nofb", "B_nofb")]))`).
- `r length(ctrl_large_A)` controls are large in A:
  `r paste(uid_label(ctrl_large_A), collapse = ", ")`. Story-2 reality
  check 2 and story 5's reference reality check are V0 Bogotá effects: both
  fall to class `r noV0_class("story2_moral_reasoning_reality_check_2")`
  without V0 (next paragraph). **Story-3 reality check 1** ("fountain")
  stays class `r noV0_class("story3_interpretation_reality_check_1")`: it is
  harder for es-CO children at every age (accuracy by age band
  `r rng(s3rc1$acc[s3rc1$language == "es-CO"])` vs
  `r rng(s3rc1$acc[s3rc1$language == "de-DE"])` in de-DE), a candidate for a
  translation or stimulus check.

```{r fig-ageband, fig.height=6.5}
age_items <- c("story1_reality_known_false_belief", "story13_reality_known_false_belief",
               "story2_moral_reasoning_false_belief_1", "story14_moral_reasoning_false_belief_1",
               "story3_interpretation_false_belief", "story15_interpretation_false_belief",
               "story18_second_order_false_belief_1", "story3_interpretation_reality_check_1",
               "story2_moral_reasoning_reality_check_2")
aba |>
  filter(short %in% age_items, n >= 20) |>
  mutate(item = factor(uid_label(short), uid_label(age_items)),
         language = factor(language, lang_levels)) |>
  ggplot(aes(age_band, acc, colour = language, group = language)) +
  geom_hline(aes(yintercept = chance), linetype = "dashed") +
  geom_line() + geom_point(aes(size = n)) +
  facet_wrap(vars(item), ncol = 3) +
  scale_colour_manual(values = lang_pal, name = NULL) +
  scale_size_area(max_size = 3, guide = "none") +
  labs(x = "Age band (years)", y = "Accuracy", title = "Non-invariant items by age band (cells with n ≥ 20; dashed = chance)") +
  theme(legend.position = "bottom")
```

**The V0 Bogotá paragraph.** Story-2 reality check 2 is non-invariant in A
(db for es-CO =
`r n2(pick(res_inv$dif, db_es_CO, analysis == "A", short == "story2_moral_reasoning_reality_check_2"))`).
The effect sits in the first Bogotá content (May–June 2024). Dropping those
runs (A_noV0, n ≥ 90) moves it and story 5's reference reality check to
class `r noV0_class("story5_reference_reality_check")` and cuts the flagged
controls from `r ctrl_flag("A")` to `r ctrl_flag("A_noV0")`. The scalar
χ²/df falls from `r n2(k$A_scal_x2df, 1)` to
`r n2(tr_("A_noV0", "scalar vs configural", X2_df), 1)`, and BIC then
prefers scalar (dBIC `r n2(tr_("A_noV0", "scalar vs configural", dBIC), 0)`),
although the LR test still rejects it
(p `r pe(tr_("A_noV0", "scalar vs configural", p))`). This makes V0
story-2 reality check 2 a candidate for a story- and epoch-scoped row in the
Airtable table, next to V0 reality check 1. It is not applied here.

## Within-language comparisons: is it language?

If the DIF above were about language (for example, translation), two sites
that share a language should show little of it. Two such comparisons are
possible with ≥ 60–100 responses per item: UTDT-AR main vs UTDT-intl
(es-AR, stories 13–18, the same no-HA retest-B form; UTDT-intl children are
younger) and Bogotá vs rural Colombia (es-CO, stories 1–6). A German
subsample matched more closely on age (under 9 years) repeats B.
First-exposure German children in B number only
`r k$B_de_n - k$B_de_repeat`, too few for DIF.

```{r inv-within}
wl <- c("X_esAR", "X_esCO", "B_young")
res_inv$samples |>
  filter(analysis %in% wl) |>
  select(analysis, group, n_children, age_median, age_q1, age_q3, datasets, forms) |>
  knitr::kable(digits = 1, caption = "Within-language and age-restricted samples")
res_inv$tests |>
  filter(analysis %in% c("A", "B", wl), items == "all", comparison == "scalar vs configural") |>
  transmute(analysis, n_items, X2, df, X2_df, p = fp(p), dBIC) |>
  left_join(res_inv$dif_by_type |> filter(analysis %in% c("A", "B", wl)) |>
              group_by(analysis) |>
              summarise(n_flagged = sum(n_flagged), n_large = sum(n_large),
                        mean_abs_db_controls = mean_abs_db[kind == "control"],
                        mean_abs_db_targets = mean_abs_db[kind == "target"], .groups = "drop"),
            by = "analysis") |>
  knitr::kable(digits = 2, caption = "Scalar vs configural and flagged items: between languages (A, B), within a language across sites (X_esAR, X_esCO), and B with German children under 9 (B_young)")
res_inv$dif |>
  filter(analysis %in% c("X_esAR", "X_esCO"), class != "A") |>
  transmute(analysis, item = uid_label(short), question_type, db_utdt_intl, db_rural, class,
            boundary) |>
  knitr::kable(digits = 2, caption = "Flagged items within a language (db = second site minus first)")
```

```{r within-numbers}
wl_t <- \(a, col) pick(tst, {{ col }}, analysis == a, comparison == "scalar vs configural", items == "all")
wl_f <- \(a) sum(res_inv$dif$class[res_inv$dif$analysis == a] != "A")
wl_n <- \(a) sum(res_inv$dif$analysis == a)
wl_qt <- \(a) res_inv$dif |> filter(analysis == a, class != "A") |> count(question_type) |>
  arrange(desc(n)) |> mutate(x = paste(n, str_replace_all(question_type, "_", " "))) |>
  pull(x) |> paste(collapse = ", ")
fa <- res_inv$flag_agreement
```

- **Sites that share a language also differ, but less, and on other
  items.** Within es-AR, the two UTDT samples (same form; UTDT-intl
  children are younger) differ in `r wl_f("X_esAR")` of `r wl_n("X_esAR")`
  items, mostly emotion questions (`r wl_qt("X_esAR")`), and BIC prefers
  scalar invariance (dBIC `r n2(wl_t("X_esAR", dBIC), 0)`). Between de-DE
  and es-AR it is `r wl_f("B")` of `r wl_n("B")` items, mostly false belief
  (`r wl_qt("B")`), and BIC rejects scalar invariance (dBIC
  `r n2(wl_t("B", dBIC), 0)`). Within es-CO, Bogotá and rural Colombia
  differ in `r wl_f("X_esCO")` of `r wl_n("X_esCO")` items, including the two
  V0-driven controls (dBIC `r n2(wl_t("X_esCO", dBIC), 0)`).
- **Age does not explain the false-belief DIF in B.** With German children
  under 9 (median age
  `r n2(pick(res_inv$samples, age_median, analysis == "B_young", group == "de_DE"), 1)`
  vs `r n2(pick(res_inv$samples, age_median, analysis == "B_young", group == "es_AR"), 1)`
  in es-AR), B flags `r wl_f("B_young")` of `r wl_n("B_young")` items,
  largely the same false-belief items (flag agreement with B
  `r pct(pick(fa, agreement, analysis == "B_young"))`). Most of these
  younger German children were on a repeat run too
  (`r k$Byoung_repeat` of `r k$Byoung_n`), so prior exposure is not
  separated.
- The false-belief DIF between German and Argentine children is therefore
  not the generic difference between two sites, and it is not an age
  effect alone. Translation, prior exposure to the task and the samples
  themselves remain possible causes, and these data cannot separate them.

## The June bridge

```{r inv-bytype}
res_inv$dif_by_type |>
  filter(analysis %in% c("J_release", "J_policy", "A", "A_noV0", "A_policy", "B")) |>
  mutate(analysis = factor(analysis, c("J_release", "J_policy", "A", "A_noV0", "A_policy", "B"))) |>
  arrange(analysis, kind) |>
  select(analysis, kind, n_items, n_flagged, n_boundary, share_of_items, share_of_flagged,
         share_of_s1_X2, mean_abs_db) |>
  knitr::kable(digits = 2, caption = "Controls vs targets. J_*: June's design (DE-Leipzig vs Bogotá, pure Rasch, all runs, raw p < .05). J_release = release-equivalent data; J_policy = corrected + policy exclusions. A_policy puts the policy-excluded trials back.")
```

```{r june-bridge}
res_inv$june_bridge |>
  transmute(analysis, items, n_items, X2, df, X2_df, dBIC,
            X2_change_vs_release = pct(X2_change_vs_release),
            dBIC_change_vs_release = pct(dBIC_change_vs_release)) |>
  knitr::kable(digits = 1, caption = "Scalar vs configural in June's design, on today's data")
res_inv$june_reported |>
  # the June chapters moved to old/ after this results file was built
  mutate(source = str_replace(source, "^tasks/tom_dif_multigroup", "old/tom_dif_multigroup")) |>
  knitr::kable(caption = "June's reported values (quoted from the June chapters, not recomputed)")
```

```{r bridge-numbers}
jb <- res_inv$june_bridge
jbv <- \(a, it, col) pick(jb, {{ col }}, analysis == a, items == it)
```

- **June's baseline reproduces on today's release-equivalent data:** χ²/df
  `r n2(jbv("J_release", "all", X2_df), 1)` (June ≈ 29), and mean |DIF|
  `r n2(pick(dbt, mean_abs_db, analysis == "J_release", kind == "control"))`
  for controls vs
  `r n2(pick(dbt, mean_abs_db, analysis == "J_release", kind == "target"))`
  for targets (June ≈ 1.9 vs 0.5). The released trials therefore still carry
  the June defects.
- **Corrections plus policy exclusions remove about half of the misfit in
  June's design:** χ² `r big(round(jbv("J_release", "all", X2)))` →
  `r big(round(jbv("J_policy", "all", X2)))`
  (`r spct(jbv("J_policy", "all", X2_change_vs_release))`), and
  `r pct(-jbv("J_policy", "controls", X2_change_vs_release))` of the
  controls' misfit. June reported about 70% overall. PR #27 alone does not
  reduce misfit (J_corrected
  `r spct(jbv("J_corrected", "all", X2_change_vs_release))`), because it
  restores data to the defective items. The reduction comes from the
  exclusions. June's repair also differed: it flipped keys and dropped the
  hostile-attribution epoch, whose `correct` values the audit found valid.
- In June's DE-vs-Bogotá design on corrected data (J_policy), controls
  still exceed targets (mean |DIF|
  `r n2(pick(dbt, mean_abs_db, analysis == "J_policy", kind == "control"))`
  vs `r n2(pick(dbt, mean_abs_db, analysis == "J_policy", kind == "target"))`),
  consistent with V0 Bogotá content driving part of it.

## Robustness

```{r fig-flags, fig.height=8}
fm <- res_inv$flag_matrix |>
  filter(family %in% c("A", "B")) |>
  pivot_longer(c(A, A_policy, A_nofb, A_fixed, A_noV0, A_rasch, A_g0,
                 B, B_policy, B_nofb, B_rasch, B_g0, B_esco, B_young),
               names_to = "variant", values_to = "class") |>
  filter(str_starts(variant, family)) |>
  mutate(variant = factor(variant, c("A", "A_policy", "A_nofb", "A_fixed", "A_noV0", "A_rasch", "A_g0",
                                     "B", "B_policy", "B_nofb", "B_rasch", "B_g0", "B_esco", "B_young")),
         item = fct_reorder(uid_label(short), -readr::parse_number(str_extract(short, "\\d+"))),
         class = replace_na(class, "not tested"))
ggplot(fm, aes(variant, item, fill = class)) +
  geom_tile(colour = "white") +
  facet_wrap(vars(family), scales = "free", labeller = as_labeller(c(A = "Stories 1–6", B = "Stories 13–18"))) +
  scale_fill_manual(values = c(A = "grey90", B = "#DDCC77", C = "#CC6677", `not tested` = "white"),
                    name = "ETS class") +
  labs(x = NULL, y = NULL, title = "DIF class across analysis variants") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), axis.text.y = element_text(size = 7))
```

```{r flag-agreement}
res_inv$flag_agreement |>
  knitr::kable(digits = 2, caption = "Agreement of flagged / not-flagged status with the primary analysis")
```

The flags hold across the fCAT, guessing-floor (pure Rasch; floor 0 for
items at or below chance, the dimensionality analysis's rule), Spanish
false-belief and policy choices (agreement
`r rng(fa$agreement[fa$analysis %in% c("A_policy", "A_nofb", "A_fixed", "A_rasch", "A_g0", "B_policy", "B_nofb", "B_rasch", "B_g0")])`).
The exceptions are A_noV0
(`r n2(pick(fa, agreement, analysis == "A_noV0"))`), because several
control effects are V0 Bogotá effects, and the underpowered B_esco and
B_young variants.

::: {.callout-note collapse="true"}
## Numerics, anchors, group parameters, and the 2PL cross-check

All `r big(sum(res_inv$convergence$n_fits))` Rasch-family fits converged.
`r sum(res_inv$convergence$n_fits_warned)` fits (in
`r paste(unique(res_inv$convergence$analysis[res_inv$convergence$n_fits_warned > 0]), collapse = ", ")`)
emitted mirt's "log-likelihood decreasing near the ML solution" warning.
Refitting with 121 instead of 61 quadrature points changes identified
difficulties by at most
`r n2(max(res_inv$numerics$max_abs_b_diff_identified), 3)` logits. The two
es-AR items below the guessing floor (story 13 FB, story 14 fb_1) are
unidentified in the Rasch + guessing model; their magnitudes come from
pure Rasch, and they are never 2PL anchors. With the floor set to 0 for
items at or below chance (A_g0, B_g0:
`r paste(uid_label(unique(unlist(res_inv$g0_items[c("A_g0", "B_g0")]))), collapse = ", ")`)
they are identified, and the flags barely change (agreement above). mirt's default θ range ([−6, 6]) was used throughout. The
reliability analysis shows that this range biases the means and variances
of low-scoring groups. The quadrature check varied the number of points,
not the range, so the effect on these fits is untested. It is likely small
for DIF, which the anchors identify, but the group means below should not be
used.

```{r inv-appendix}
res_inv$anchors |>
  filter(analysis %in% c("A", "B")) |>
  transmute(analysis, n_items, n_anchors, anchor_rule, anchors_stable,
            anchors = str_replace_all(anchors, "story", "S")) |>
  knitr::kable()
res_inv$group_pars |> filter(analysis %in% c("A", "B")) |>
  knitr::kable(digits = 2)
res_inv$numerics |> knitr::kable(digits = 4)
res_inv$robust_2pl |>
  transmute(analysis, item = uid_label(str_remove(item, "^tom_")), group, boundary, anchor_2pl, a,
            db_2pl, db_2pl_purified, db_rasch_stage1, db_rasch_stage2) |>
  knitr::kable(digits = 2, caption = "Item DIF under the 2PL with production priors and under Rasch + guessing")
```
:::

# Reliability, stability and age validity

## Scoring model

The analysis model is one 2PL, multigroup by dataset (each dataset with at
least `r res_rel$meta$min_group_runs` runs has its own mean and variance;
DE = N(0, 1)), scalar, with
guessing fixed at chance and production priors (d ~ N(0, 3),
a ~ N(1, 0.3)). It has `r res_rel$meta$n_items_analysis` items after the
policy exclusions and is calibrated concurrently. The θ range is [−12, 8]
with 201 quadrature points.

```{r itemtype}
res_rel$itemtype_bic |>
  select(model, itemtype, grid, runs, items, groups, n_par, logLik, AIC, BIC, converged) |>
  knitr::kable(digits = 1, caption = "Model comparison and variants (all converged)")
res_rel$item_slopes_by_type |>
  knitr::kable(digits = 2, caption = "Item slopes by question type")
```

```{r fig-ab, fig.height=4.5}
res_rel$item_pars |>
  mutate(question_type = factor(question_type, qtype_levels),
         b_plot = pmax(pmin(b, 4), -8)) |>
  ggplot(aes(b_plot, a, colour = question_type)) +
  geom_point(size = 2, alpha = .8) +
  scale_colour_manual(values = qtype_pal, name = NULL) +
  labs(x = "Difficulty b (DE scale; clipped to [−8, 4])", y = "Slope a",
       title = "2PL item parameters: false-belief items discriminate most, emotion items least") +
  theme(legend.position = "bottom")
```

BIC prefers the 2PL over the Rasch version
(dBIC `r n2(pick(res_rel$itemtype_bic, dBIC, model == "analysis", itemtype == "Rasch"), 0)`),
but scores barely depend on the choice (within-dataset r between the two
EAPs `r rng(res_rel$released_comparison$r_analysis_alt_itemtype[!res_rel$released_comparison$dataset %in% c(small_groups, "all (pooled)")], 3)`).
Story 17's emotion question has a slope of
`r n2(pick(res_rel$item_pars, a, str_starts(item, "tom_story17_reference_emotion_reasoning_1")), 3)`.
`r pct(pick(res_rel$item_pars, p_correct, str_starts(item, "tom_story17_reference_emotion_reasoning_1")))`
of children answer it correctly (chance .25), and it carries no
information about ability.

::: {.callout-warning}
## The low tail: group means and variances of low-scoring groups are weakly identified

mirt's default θ range is [−6, 6] with 61 points, and production uses it.
It cuts off the low tail of the lower-scoring groups and pulls their means
and variances toward the German reference. The analysis grid [−12, 8] is
numerically converged: a fit on [−16, 8] changes parameters and EAPs by at
most `r n2(max(res_rel$grid_convergence$max_abs_diff), 4)`. That shows the
integral is accurate, not that the tail is identified. Below θ ≈ −4 a child
who answers at or below guessing has an almost flat likelihood, so there the
normal prior, not the data, sets the lower tail. Two checks show how much
this matters:

- **Latent density.** With the items fixed at the analysis calibration,
  a nonparametric (empirical-histogram) density per group gives variances
  of `r n2(dc("rfp1_utdt_intl_ys", var_eh), 1)` instead of
  `r n2(dc("rfp1_utdt_intl_ys", var_normal), 1)` in UTDT-intl and
  `r n2(dc("pilot_uniandes_co_rural", var_eh), 1)` instead of
  `r n2(dc("pilot_uniandes_co_rural", var_normal), 1)` in rural Colombia,
  while EAPs under the two densities correlate
  `r rng(res_rel$density_check$r_eap_normal_eh)`.
- **Guessing rule.** With the floor set to 0 for the
  `r length(res_rel$meta$g0_items)` items answered at or below chance in some
  group (the dimensionality analysis's rule), the low-scoring groups'
  variances roughly halve (UTDT-intl
  `r n2(g0s("rfp1_utdt_intl_ys", var_analysis), 1)` →
  `r n2(g0s("rfp1_utdt_intl_ys", var_g0), 1)`, Bogotá
  `r n2(g0s("pilot_uniandes_co_bogota", var_analysis), 1)` →
  `r n2(g0s("pilot_uniandes_co_bogota", var_g0), 1)`), and their means rise
  by up to `r n2(max(res_rel$g0_sensitivity$mean_g0 - res_rel$g0_sensitivity$mean_analysis), 1)`.
  BIC prefers this version of the model by
  `r n2(pick(res_rel$itemtype_bic, BIC, model == "analysis", itemtype == "2PL") - pick(res_rel$itemtype_bic, BIC, str_starts(model, "analysis, floor 0")), 0)`,
  so the LEVANTE convention fits these items worse.

Rank order within a dataset barely moves in any of these checks (EAP
correlations `r rng(res_rel$g0_sensitivity$r_analysis_g0)` for the guessing
rule), and empirical reliability changes little. Group levels and
variances of low-scoring groups, however, depend on the grid, the density
and the guessing convention. This applies to the released v2_3 scores and
probably to other multigroup tasks with low-scoring groups; that is
untested. It is a DCC item to investigate before any production change.

```{r grid}
res_rel$grid_sensitivity |>
  transmute(dataset = ds_label(dataset), n_runs,
            mean_default = mean_defaultgrid, var_default = var_defaultgrid,
            mean_wide = mean_analysis, var_wide = var_analysis,
            r_wide_default, max_abs_shift, n_shift_gt_0.25, min_eap_default, min_eap_wide,
            rel_empirical_default, rel_empirical_wide) |>
  knitr::kable(digits = 2, caption = "Default vs wide integration grid")
res_rel$density_check |>
  transmute(dataset = ds_label(dataset), n_runs, mean_normal, var_normal, mean_eh, var_eh,
            post_mass_below_m4, r_eap_normal_eh, rel_empirical_normal, rel_empirical_eh,
            r_age_normal, r_age_eh, em_iterations, em_last_change) |>
  knitr::kable(digits = 3, caption = "Normal vs empirical-histogram group density (items fixed at the analysis calibration). post_mass_below_m4 = mean posterior mass below θ = −4 under the normal prior.")
res_rel$g0_sensitivity |>
  transmute(dataset = ds_label(dataset), n_runs, mean_analysis, mean_g0, var_analysis, var_g0,
            r_analysis_g0, rel_empirical_analysis, rel_empirical_g0, r_age_analysis, r_age_g0) |>
  knitr::kable(digits = 2, caption = "Guessing floor at chance (analysis) vs 0 for items at or below chance in a group (g0)")
```
:::

## Reliability per administration

Empirical reliability is var(EAP) / (var(EAP) + mean SE²) within a
dataset × form. `rel_group` instead uses the model's group variance in the
denominator (1 − mean SE² / σ²_g). All of these come from a
unidimensional model that assumes scalar invariance across datasets.[^upper]

[^upper]: Both assumptions make them upper bounds. The dimensionality
analysis finds that the unidimensional model overstates test information by
`r k$info_over`, which lowers reliability by
`r rng(ld$rxx_uni - ld$rxx_general, 3)` in its samples; and scalar
invariance does not fully hold (see the invariance section).

```{r rel-head}
rel_head |>
  transmute(dataset = ds_label(dataset), kind, n_runs, median_items, median_minutes,
            mean_prop_correct, prop_near_ceiling, prop_at_or_below_guessing,
            rms_se, rel_empirical, rel_group) |>
  arrange(kind, dataset) |>
  knitr::kable(digits = 2, caption = "Reliability per dataset, pooling each dataset's fixed forms, and of the fCAT (datasets with < 25 runs omitted)")
rel_forms |>
  transmute(dataset = ds_label(dataset), form, n_runs, median_items, mean_prop_correct,
            prop_near_ceiling, rel_empirical, rel_group) |>
  knitr::kable(digits = 2, caption = "Reliability of single fixed forms (dataset × form with ≥ 25 runs)")
res_rel$fcat_vs_fixed |>
  filter(form_type == "fixed form") |>
  transmute(dataset = ds_label(dataset), se_ratio_fcat_to_fixed, info_per_minute_ratio) |>
  knitr::kable(digits = 2, caption = "fCAT vs fixed forms within a dataset: RMS SE ratio and information per minute ratio (different children)")
```

```{r fig-se, fig.height=6}
res_rel$scores_plot |>
  filter(!dataset %in% small_groups) |>
  mutate(dataset = ds_label(dataset),
         form_type = if_else(str_starts(form_type, "fCAT"), "fCAT (3 random stories)", "fixed form")) |>
  ggplot(aes(eap, se, colour = form_type)) +
  geom_point(size = .6, alpha = .4) +
  facet_wrap(vars(dataset), nrow = 2, labeller = label_wrap_gen(22)) +
  scale_x_continuous(breaks = c(-6, -3, 0)) +
  scale_colour_manual(values = c("fixed form" = "grey30", "fCAT (3 random stories)" = "#CC6677"),
                      name = NULL) +
  labs(x = "EAP θ (DE scale)", y = "Posterior SD",
       title = "Measurement precision by ability: the story bank is thin above θ = 0") +
  theme(legend.position = "bottom")
```

```{r rel-numbers}
de_fc <- filter(rel_head, dataset == "pilot_mpieva_de_main", kind != "fixed forms")
ffx <- res_rel$fcat_vs_fixed |> filter(form_type == "fixed form")
ffx_de <- pick(ffx, info_per_minute_ratio, dataset == "pilot_mpieva_de_main")
ffx_other <- ffx$info_per_minute_ratio[ffx$dataset != "pilot_mpieva_de_main"]
```

- Pooling each dataset's fixed forms, a fixed form (median
  `r k$fixed_items` items) gives `r k$rel_fixed` per administration. Single
  forms range more widely, from `r n2(min(rel_forms$rel_empirical))`
  (`r k$low_form`) to `r n2(max(rel_forms$rel_empirical))`
  (`r k$high_form`). The deployed fCAT gives `r k$rel_fcat` in a median of
  `r k$fcat_min` minutes.
- Leipzig is the least reliably measured sample because it is near ceiling.
  `r pct(de_fc$prop_near_ceiling)` of its fCAT runs are ≥ 90% correct, and
  its fCAT reliability is `r n2(de_fc$rel_empirical)`.
- Per minute, the fCAT is at least as efficient as the fixed forms outside
  Leipzig (ratio `r rng(ffx_other)`); in Leipzig, where it is at ceiling,
  the ratio is `r n2(ffx_de)`. fCAT and fixed-form runs come from different
  children, who differ in age and ceiling rates, so these ratios are only
  indicative. The within-child simulation below is the cleaner comparison.

### Block CAT: a within-child simulation

The deployed fCAT shows one random story from each of three blocks, always
in the same order (block 1: stories `r paste(blk(1), collapse = ", ")`;
block 2: `r paste(blk(2), collapse = ", ")`; block 3:
`r paste(blk(3), collapse = ", ")`), so block and position are
confounded. To compare story selections in the same children,
every fixed-form run was rescored with the analysis model on (i) its whole
form, (ii) an fCAT-style random draw of one of its stories per block
(averaged over all draws), and (iii) an adaptive draw that takes, block by
block, the story whose items carry the most information at the child's
current EAP. Minutes are the child's own time on the chosen stories.
Reliability here is model-based (1 − mean posterior variance / σ²_g).

```{r cat-sim}
res_rel$cat_sim_summary |>
  transmute(dataset = ds_label(dataset), form, n_runs,
            min_full = median_minutes_full, min_random = median_minutes_random,
            min_adaptive = median_minutes_adaptive,
            rel_full, rel_random, rel_adaptive,
            info_per_min_full, info_per_min_random, info_per_min_adaptive,
            prop_adaptive_better) |>
  knitr::kable(digits = 3, caption = "Fixed-form runs rescored on three stories: random (fCAT-style) vs adaptive selection within blocks, and the whole form")
```

- Three random stories take half the time of the whole form and cost
  `r rng(cs$rel_full - cs$rel_random)` in reliability.
- Choosing the more informative story in each block, in the same time,
  raises reliability by a median of
  `r n2(median(cs$rel_adaptive - cs$rel_random))` (range
  `r rng(cs$rel_adaptive - cs$rel_random)` across forms; UTDT-AR, the
  largest sample: `r n2(csv("rfp1_utdt_ar_main", rel_random))` →
  `r n2(csv("rfp1_utdt_ar_main", rel_adaptive))`, against
  `r n2(csv("rfp1_utdt_ar_main", rel_full))` for the whole form). It gives
  the lower posterior variance for
  `r pct(min(cs$prop_adaptive_better))`–`r pct(max(cs$prop_adaptive_better))`
  of children. The gain is smallest in the Leipzig retest-A form.
- The simulation can only choose between the stories a fixed form contains
  (two per block). The bank has six per block, so real adaptive selection
  could do more, provided the stories are calibrated in each language.

### Reconciling the reliability estimates

Three analyses estimate reliability for the same fixed forms, using
different person sets, item filters and priors:

```{r rel-reconcile}
irel <- \(s, m, col = empirical_rxx)
  pick(res_items$reliability, {{ col }}, sample == s, items == "all items", model == m)
iper <- \(s) paste(irel(s, "Rasch", n_runs), "runs")
iit  <- \(s) as.character(irel(s, "Rasch", n_items))
n_model <- paste(res_rel$meta$n_items_analysis, "in model")
drel <- \(s) pick(res_dim$irt_ld, rxx_uni, family == "guess", sample == s)
dn   <- \(s, col) pick(res_dim$samples, {{ col }}, sample == s)
rrel <- \(d, f) pick(res_rel$rel_by_form, rel_empirical, dataset == d, form == f)
rn   <- \(d, f) pick(res_rel$rel_by_form, n_runs, dataset == d, form == f)
rel_rec <- tribble(
  ~form, ~analysis, ~model, ~persons, ~items, ~reliability,
  "Leipzig item bank", "items & controls", "Rasch + g, single group, production priors", iper(de_s), iit(de_s), irel(de_s, "Rasch"),
  "Leipzig item bank", "items & controls", "2PL + g, single group, production priors", iper(de_s), iit(de_s), irel(de_s, "2PL"),
  "Leipzig item bank", "dimensionality", "2PL + g, single group, slope prior N(0, 2²)", paste(dn("de_ib", children), "children (first run)"), as.character(dn("de_ib", items)), drel("de_ib"),
  "Leipzig item bank", "reliability", "2PL + g, 8-group concurrent calibration", paste(rn("pilot_mpieva_de_main", "item_bank"), "runs"), n_model, rrel("pilot_mpieva_de_main", "item_bank"),
  "Western item bank", "dimensionality", "2PL + g, single group, slope prior N(0, 2²)", paste(dn("wca_ib", children), "children"), as.character(dn("wca_ib", items)), drel("wca_ib"),
  "Western item bank", "reliability", "2PL + g, 8-group concurrent calibration", paste(rn("pilot_western_ca_main", "item_bank"), "runs"), n_model, rrel("pilot_western_ca_main", "item_bank"),
  "Bogotá item bank", "dimensionality", "2PL + g, single group, slope prior N(0, 2²)", paste(dn("bog_ib", children), "children"), as.character(dn("bog_ib", items)), drel("bog_ib"),
  "Bogotá item bank", "reliability", "2PL + g, 8-group concurrent calibration", paste(rn("pilot_uniandes_co_bogota", "item_bank"), "runs"), n_model, rrel("pilot_uniandes_co_bogota", "item_bank"),
  "UTDT no-HA retest-B", "items & controls", "Rasch + g, UTDT-AR main only", iper(ar_s), iit(ar_s), irel(ar_s, "Rasch"),
  "UTDT no-HA retest-B", "items & controls", "2PL + g, UTDT-AR main only", iper(ar_s), iit(ar_s), irel(ar_s, "2PL"),
  "UTDT no-HA retest-B", "dimensionality", "2PL + g, UTDT-AR main + UTDT-intl pooled", paste(dn("ar_nhb", children), "children"), as.character(dn("ar_nhb", items)), drel("ar_nhb"),
  "UTDT no-HA retest-B", "reliability", "2PL + g, 8-group; UTDT-AR main", paste(rn("rfp1_utdt_ar_main", "no_ha_retest_B"), "runs"), n_model, rrel("rfp1_utdt_ar_main", "no_ha_retest_B"),
  "UTDT no-HA retest-B", "reliability", "2PL + g, 8-group; UTDT-intl", paste(rn("rfp1_utdt_intl_ys", "no_ha_retest_B"), "runs"), n_model, rrel("rfp1_utdt_intl_ys", "no_ha_retest_B")
)
rel_rec |>
  knitr::kable(digits = 3, caption = "Empirical (EAP) reliability of the same fixed forms in the three analyses")
```

For these four forms the estimates differ by up to
`r n2(max(tapply(rel_rec$reliability, rel_rec$form, \(x) diff(range(x)))))`
between analyses and range over `r rng(rel_rec$reliability)`. Across all
single fixed forms with at least 25 runs the range is wider
(`r k$rel_forms`, table above); the lowest are Leipzig's retest forms,
taken by a near-ceiling sample. Pooling a dataset's fixed forms gives
`r k$rel_fixed`. On every version of the number, a fixed Stories form
is far from a .9 instrument. The differences between analyses have ordinary
causes, though none was tested separately:

- The dimensionality fits use a weak slope prior within one sample and drop
  near-ceiling items. The reliability model shares item parameters across 8
  groups and shrinks slopes toward 1.
- The dimensionality es-AR sample pools two UTDT datasets whose means
  differ by about
  `r n2(pick(res_rel$group_pars, mean_analysis, dataset == "rfp1_utdt_ar_main") - pick(res_rel$group_pars, mean_analysis, dataset == "rfp1_utdt_intl_ys"), 1)`
  logits in the reliability model
  (`r n2(pick(res_rel$group_pars, mean_analysis, dataset == "rfp1_utdt_ar_main"))`
  vs `r n2(pick(res_rel$group_pars, mean_analysis, dataset == "rfp1_utdt_intl_ys"))`;
  the UTDT-intl mean pools its fCAT and fixed-form runs). The pooling widens
  the ability distribution, which raises reliability.

```{r june-equiv}
res_rel$june_equivalents |>
  knitr::kable(digits = 3, caption = "June-comparable quantities on the corrected model")
res_dim$june_pool_check |>
  knitr::kable(digits = 3, caption = "June's '.92 reliable composite', recomputed with June's method on the Leipzig pool")
```

June's ".92" is the Rasch marginal reliability of the whole pooled Leipzig
item set: `r jp$items_in_model` items across forms, of which children
answered a median of `r jp$items_per_run_median`. It reproduces
(`r n2(jp$rasch_marginal_rxx_june_method, 3)`), but the same model's
reliability for the items children actually answered is
`r n2(jp$rasch_empirical_rxx_eap, 3)`. On the corrected 2PL model, the
full-bank marginal reliability for DE is `r n2(k$de_bank, 3)`. No child
takes the full bank.

## Stability (Leipzig)

Only Leipzig has repeated ToM runs in any number. They are different forms,
months apart, so the numbers below are alternate-form stability, not
test–retest reliability. Only
`r pick(res_rel$stability_forms, n_pairs, dataset == "pilot_mpieva_de_main", gap_bin == "[0,60)")`
pairs are less than 60 days apart.

```{r stability}
res_rel$stability_bins |>
  filter(dataset == "pilot_mpieva_de_main") |>
  transmute(gap_bin, version, n_pairs, median_gap_days, r, r_lo, r_hi, r_partial_age,
            r_disattenuated, diff_vs_released, diff_lo, diff_hi) |>
  knitr::kable(digits = 3, caption = "Leipzig alternate-form stability by gap (bootstrap 95% CIs, 2,000 resamples of pairs)")
res_rel$stability_first2 |>
  filter(dataset == "pilot_mpieva_de_main") |>
  select(-dataset) |>
  knitr::kable(digits = 3, caption = "First two runs per child (Leipzig)")
res_rel$stability_forms |>
  filter(dataset == "pilot_mpieva_de_main", !is.na(r)) |>
  select(-dataset) |>
  knitr::kable(digits = 3, caption = "By form pair")
```

```{r stab-numbers}
sf2 <- res_rel$stability_first2 |> filter(dataset == "pilot_mpieva_de_main")
```

Stability is
`r n2(pick(sb, r, dataset == "pilot_mpieva_de_main", gap_bin == "[60,180)", version == "analysis"))`
at 60–180 days,
`r n2(pick(sb, r, dataset == "pilot_mpieva_de_main", gap_bin == "[180,300)", version == "analysis"))`
at 180–300 days and
`r n2(pick(sb, r, dataset == "pilot_mpieva_de_main", gap_bin == "[300,Inf)", version == "analysis"))`
beyond 300 days. Disattenuated for measurement error it is
`r n2(pick(sb, r_disattenuated, dataset == "pilot_mpieva_de_main", gap_bin == "[60,180)", version == "analysis"))`
at 60–180 days, so most of the instability is measurement error. The
corrections do not raise it: over the first two runs it is
`r n2(pick(sf2, r, version == "analysis"), 3)`, against
`r n2(pick(sf2, r, version == "released v2_3"), 3)` for the released scores
and `r n2(pick(sf2, r, version == "release-equivalent refit"), 3)` for
the same model refitted on release-equivalent data. June reported that its
fix raised stability from about .45 to .53; on these data the corrections
lower it slightly, within noise. The small 60–180-day difference from the
released scores appears already in the "identity fixes only" rows (PR #27's
identity fixes and restored trials), not with the item exclusions; which of
those fixes produces it was not examined.

## Age validity

```{r age-validity}
res_rel$age_validity |>
  filter(form == "all forms", !dataset %in% small_groups) |>
  transmute(dataset = ds_label(dataset), n_with_age, age_sd, age_min, age_max,
            r_age, r_age_lo, r_age_hi, r_age_released, slope_per_year) |>
  arrange(desc(r_age)) |>
  knitr::kable(digits = 2, caption = "r(age, EAP) within dataset, all forms (Fisher 95% CIs)")
```

```{r fig-age, fig.height=6}
res_rel$scores_plot |>
  filter(!dataset %in% small_groups, !is.na(age)) |>
  mutate(dataset = ds_label(dataset),
         form_type = if_else(str_starts(form_type, "fCAT"), "fCAT (3 random stories)", "fixed form")) |>
  ggplot(aes(age, eap, colour = form_type)) +
  geom_point(size = .6, alpha = .3) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(vars(dataset), nrow = 2, labeller = label_wrap_gen(22)) +
  scale_colour_manual(values = c("fixed form" = "grey30", "fCAT (3 random stories)" = "#CC6677"),
                      name = NULL) +
  labs(x = "Age (years)", y = "EAP θ (DE scale)", title = "Stories score by age") +
  theme(legend.position = "bottom")
```

```{r age-numbers}
av <- res_rel$age_validity |> filter(form == "all forms", !dataset %in% small_groups, !is.na(r_age))
de_av_fcat <- pick(res_rel$age_validity, r_age, dataset == "pilot_mpieva_de_main", form == "fCAT")
```

Age validity within a dataset ranges from weak to moderate (r =
`r rng(av$r_age)`; the weakest is `r ds_label(av$dataset[which.min(av$r_age)])`). How
large it is mostly tracks each sample's age range: the SD of age is
`r rng(av$age_sd[str_starts(av$dataset, "pilot")], 1)` years in the pilot
sites and `r rng(av$age_sd[av$dataset %in% c("rfp1_utdt_ar_main", "rfp1_utdt_intl_ys", "rfp1_mpib_intl_ys")], 1)`
in the UTDT and MPIB samples.
Growth per year is
`r rng(av$slope_per_year[av$dataset != "pilot_mpieva_de_main"])` logits
outside Leipzig, and
`r n2(pick(av, slope_per_year, dataset == "pilot_mpieva_de_main"))` in
Leipzig, where the Leipzig fCAT's age correlation is weak
(`r n2(de_av_fcat)`). The corrections barely change age validity (compare
`r_age_released`).

::: {.callout-note collapse="true"}
## Spanish false-belief items: scores with and without

```{r sens-fb}
res_rel$sens_spanish_fb |>
  mutate(dataset = ds_label(dataset)) |>
  knitr::kable(digits = 3, caption = "es-* runs rescored without the flagged reality_known false-belief items (same item parameters)")
```

Removing the flagged items leaves scores, reliability and age validity
essentially unchanged (r with vs without ≥
`r n2(min(res_rel$sens_spanish_fb$r_with_without), 3)`). This rescoring
used the same item parameters, so it says nothing about whether the items
are valid in Spanish.
:::

## What changes from the released scores

```{r released}
res_rel$released_comparison |>
  left_join(res_rel$released_models |> select(dataset, own_group_in_v2_3), by = "dataset") |>
  filter(!dataset %in% small_groups) |>
  transmute(dataset = ds_label(dataset), own_group_in_v2_3, n_runs,
            r_analysis_released, shift_analysis_minus_released,
            shift_refit_minus_released, shift_own_minus_ref_prior, shift_wide_minus_default_grid,
            shift_idfix_minus_refit, shift_analysis_minus_idfix) |>
  knitr::kable(digits = 2, caption = "Analysis EAP vs released v2_3 scores. The shift_* columns split the change: recalibration + own prior + grid (refit), the prior alone and the grid alone (items fixed), PR #27 (idfix), and policy exclusions (analysis minus idfix).")
res_rel$group_pars |>
  filter(!dataset %in% small_groups) |>
  mutate(dataset = ds_label(dataset)) |>
  knitr::kable(digits = 2, caption = "Group means and variances (DE = N(0, 1)) under each model variant, and in v2_3")
```

```{r fig-released, fig.height=6}
res_rel$scores_plot |>
  filter(!dataset %in% small_groups, !is.na(score_released)) |>
  mutate(dataset = ds_label(dataset)) |>
  ggplot(aes(score_released, eap)) +
  geom_abline(linetype = "dashed") +
  geom_point(size = .6, alpha = .3) +
  facet_wrap(vars(dataset), nrow = 2, labeller = label_wrap_gen(22)) +
  labs(x = "Released v2_3 score", y = "Analysis EAP", title = "Analysis scores vs released scores (dashed: y = x)") +
  theme(strip.text = element_text(size = 10))
```

- Rank order within a dataset is largely preserved (r =
  `r rng(rc$r_analysis_released, 3)`), but not unchanged.
- The four datasets that had their own group in v2_3 keep their level
  (shifts `r rng(rc$shift_analysis_minus_released[rc$dataset %in% own])`).
  The datasets v2_3 scored under the German prior drop by
  `r k$shift_total` logits in total (recalibration, own prior, wider grid
  and data corrections together). Scored under their own prior instead of
  the German one, with the items held fixed, they drop by `r k$shift_prior`
  logits, so the prior accounts for most of the change. As the reliability
  section shows, these low-scoring groups' levels are themselves weakly
  identified.
- The data corrections move dataset means little, except in Bogotá. There,
  dropping the policy-flagged items (mainly the V0 2e reality check, but
  also reference_reference) raises the mean by
  `r n2(pick(rc, shift_analysis_minus_idfix, dataset == "pilot_uniandes_co_bogota"))`.
- Group means assume scalar invariance, which does not hold, and they are
  confounded with age. They are not cross-cultural comparisons. In
  Bogotá, UTDT-intl and MPIB-intl,
  `r pct(min(res_rel$rel_by_form$prop_at_or_below_guessing[res_rel$rel_by_form$form == "all forms" & res_rel$rel_by_form$dataset %in% c("rfp1_mpib_intl_ys", "rfp1_utdt_intl_ys", "pilot_uniandes_co_bogota")]))`–`r pct(max(res_rel$rel_by_form$prop_at_or_below_guessing[res_rel$rel_by_form$form == "all forms" & res_rel$rel_by_form$dataset %in% c("rfp1_mpib_intl_ys", "rfp1_utdt_intl_ys", "pilot_uniandes_co_bogota")]))`
  of runs score at or below what guessing would give, so for those children
  the EAP rests largely on the prior.

# Designing the block CAT

```{r cat-setup}
val <- res_cat$validation$fcat3_random
val_time <- res_cat$validation$time
q5m <- res_cat$q5 |> group_by(se_inflation, rule) |>
  summarise(across(c(capitalization, selection_loss, rescoring_gain), mean), .groups = "drop")
q4r <- res_cat$q4_two |> filter(growth == "latent_slope", !history_aware) |>
  group_by(design, rule, gap_years) |> summarise(repeated = mean(share_repeated), .groups = "drop")
q4cost <- res_cat$q4_two |> filter(growth == "latent_slope") |>
  select(dataset, design, rule, gap_years, history_aware, rel_wave2) |>
  pivot_wider(names_from = history_aware, values_from = rel_wave2, names_prefix = "ha_") |>
  group_by(design, rule, gap_years) |> summarise(cost = mean(ha_FALSE - ha_TRUE), .groups = "drop")
```

The deployed fCAT picks one story at random from each of three blocks. To decide how a real block CAT should choose stories, we simulated administrations from the analysis model (`tasks/_stories_cat_sim.R`).

**The simulation.**

- **Items:** the analysis model's items, minus story 4's reality checks 2 and 3 (broken on image 4g). Their time is still counted: `r length(res_cat$meta$bank_items)` items.
- **Children:** each calibration group's latent distribution, with `r big(res_cat$meta$n[["main"]])` simulated children per group and design.
- **Time:** story durations from the deployed fCAT runs, with a per-child speed effect.
- **Procedure:** after each story, ability is updated by EAP under the group's prior, and the next story is chosen within its block.
- **Validation:** the simulated deployed design matches the observed fCAT. Reliability agrees within `r n2(max(abs(val$sim_rel_fcat3_random - val$obs_rel_group_fcat), na.rm = TRUE))` per dataset, and the median time is `r n2(median(val$sim_median_min), 1)` min simulated vs `r n2(val_time$obs_median_min, 1)` observed. The 90th-percentile time is `r n2(val_time$obs_q90_min - median(val$sim_q90_min), 1)` min too short, so read every 90th-percentile figure below as that much optimistic.

This is a model-based simulation. It assumes the 2PL with fixed guessing holds, with scalar invariance except where tested below, and it models no practice effects.

## Precision against time

```{r cat-grid}
cat_grid |>
  filter(rule %in% c("random", "maxinfo", "infomin", "rq2", "fixed")) |>
  mutate(rule = factor(rule_labels[rule], unname(rule_labels))) |>
  select(design, n_stories, rule, rel) |>
  pivot_wider(names_from = rule, values_from = rel) |>
  left_join(cat_grid |> filter(rule %in% c("maxinfo", "fixed")) |>
              transmute(design, minutes = paste0(n2(med_min, 1), " / ", n2(q90_min, 1))),
            by = "design") |>
  left_join(cat_designs |> select(design, design_label, steps), by = "design") |>
  arrange(n_stories, match(design, c("cur3", "canvas3", "best3", "free3", "cur3p1", "canvas4",
                                     "best4", "free4", "fixed_1to6", "fixed_13to18"))) |>
  select(-design) |>
  mutate(across(where(is.double) & !n_stories, \(x) if_else(is.na(x), "", n2(x))),
         n_stories = as.integer(n_stories)) |>
  select(design = design_label, blocks = steps, stories = n_stories,
         any_of(unname(rule_labels[c("random", "maxinfo", "infomin", "rq2", "fixed")])), minutes) |>
  knitr::kable(caption = paste("Mean reliability over the 8 groups by design and selection rule;",
                               "minutes = median / 90th percentile under max information",
                               "(fixed forms: the whole form). RK reality known, IN interpretation,",
                               "MR moral reasoning, SO second order, DC deception, RF reference."))
```

```{r cat-grid-plot, fig.height = 4.5}
cat_grid |>
  filter(rule %in% c("random", "maxinfo", "rq2", "fixed")) |>
  left_join(cat_designs |> select(design, design_label), by = "design") |>
  ggplot(aes(med_min, rel, colour = rule_labels[rule], shape = factor(n_stories))) +
  geom_point(size = 3) +
  ggrepel::geom_text_repel(aes(label = design), size = 3, show.legend = FALSE) +
  labs(x = "Median minutes", y = "Mean reliability (8 groups)", colour = "Selection", shape = "Stories") +
  theme(legend.position = "bottom")
```

- **Adaptive selection within today's blocks is the largest single gain.** Reliability goes from `r n2(cg("cur3", "random"))` to `r n2(cg("cur3", "maxinfo"))` in the same time (`r n2(cg("cur3", "random", "med_min"), 1)` vs `r n2(cg("cur3", "maxinfo", "med_min"), 1)` median minutes).
- **A fourth story adds about as much again.** The current three blocks plus one free story (at most one per story type) reach `r n2(cg("cur3p1", "maxinfo"))` in `r n2(cg("cur3p1", "maxinfo", "med_min"), 1)` median minutes. That beats a fixed six-story form (`r n2(cg("fixed_1to6", "fixed"))`, `r n2(cg("fixed_1to6", "fixed", "med_min"), 1)` min) in about 2.5 fewer minutes.
- **Grouping matters less than selection.**
  - The searched 4-block grouping splits today's moral-reasoning + second-order block into two single-type blocks: `r n2(cg("best4", "maxinfo"))`.
  - No blocks at all, with at most one story per type: `r n2(cg("free4", "maxinfo"))`.
  - The canvas 4-block design: `r n2(cg("canvas4", "maxinfo"))`, lower because every child must take a two-question reality-known story.

```{r cat-by-group}
res_cat$q1 |>
  filter((design == "cur3" & rule %in% c("random", "maxinfo")) |
         (design %in% c("cur3p1", "canvas4", "best4") & rule == "maxinfo") |
         (design == "fixed_1to6" & rule == "fixed")) |>
  mutate(cell = paste(design, rule_labels[rule], sep = ", ")) |>
  select(label, cell, rel_realized) |>
  pivot_wider(names_from = label, values_from = rel_realized) |>
  mutate(across(where(is.double), \(x) n2(x))) |>
  knitr::kable(caption = "Reliability by group for the main designs")
```

```{r cat-blocks}
res_cat$block_balance |>
  filter(design %in% c("cur3", "canvas4", "best4")) |>
  transmute(design, block, stories, types, `expected info, best story` = n2(exp_info_best),
            `expected info, mean story` = n2(exp_info_mean), `median seconds` = median_secs_range) |>
  knitr::kable(caption = "Information and time per block (expected information averaged over the groups)")
```

Today's blocks are unbalanced. The moral-reasoning + second-order block carries about twice the information of the reality-known + interpretation block. The canvas design's reality-known block (stories 1, 7 and 13: two questions each, about 40 seconds) carries the least.

## Choosing stories: information, not discrimination

```{r cat-disc}
res_cat$disc_vs_info_cor |>
  transmute(theta, `rank r, summed discrimination vs information` = n2(spearman_sum_a2),
            `best story` = best_story_info, `highest-discrimination story` = best_story_sum_a2,
            `its share of the best story's information` = pct(info_ratio_top_a2_to_best)) |>
  knitr::kable(caption = "Does a story's discrimination predict its information at a given ability?")
```

Choosing the story with the highest summed discrimination gives `r n2(cg("cur3", "disc_a2"))` against `r n2(cg("cur3", "maxinfo"))` for information-based selection in today's blocks (mean over groups). It does as well as information-based selection for the higher-scoring groups (Leipzig, Sheffield, Western), but for the lowest-scoring samples it is no better than random: `r dq <- res_cat$q1 |> filter(design == "cur3", rule %in% c("random", "disc_a2"), label %in% c("MPIB-intl", "UTDT-intl", "Bogota")) |> select(label, rule, rel_realized) |> pivot_wider(names_from = rule, values_from = rel_realized); paste0(dq$label, " ", n2(dq$disc_a2), " vs ", n2(dq$random), " random", collapse = "; ")`. Discrimination tracks information only near the middle of the scale. Where most young Spanish-speaking children sit, the high-discrimination stories are pitched too hard: story 14 has the highest summed discrimination, while story 18 carries most of the information below θ = −2. What to maximize is each story's Fisher information at the child's current EAP, summed over its questions. Dividing by expected minutes (information per minute) gives about the same reliability in slightly less time.

## Starting value and prior

```{r cat-start}
res_cat$q3 |>
  filter(rule == "maxinfo", design %in% c("cur3", "best4")) |>
  group_by(design, start) |> summarise(rel = mean(rel_realized), .groups = "drop") |>
  mutate(start = recode(start, zero = "θ = 0, group prior", group_mean = "group mean",
                        age = "age-predicted", zero_refprior = "θ = 0 under the N(0, 1) reference prior",
                        pooled_prior = "pooled 8-group prior"),
         rel = n2(rel)) |>
  pivot_wider(names_from = design, values_from = rel) |>
  knitr::kable(caption = "Mean reliability by starting value and the prior used while testing (max information)")
```

Start each child at the group mean, or at an age-predicted value, which gives more variety in the first story. Run the online estimate under the group's prior. Running the whole test under the German N(0, 1) reference prior is the one choice that hurts badly, and it hurts most in the lowest-scoring groups. For a site without its own calibration, the pooled prior is a reasonable fallback, followed by rescoring once the site's own distribution is known.

## Retest: repeated stories

```{r cat-retest}
q4r |>
  filter(gap_years == 1, design %in% c("cur3", "canvas4", "best4", "free4")) |>
  left_join(q4cost |> filter(gap_years == 1), by = c("design", "rule", "gap_years")) |>
  transmute(design, rule = rule_labels[rule], `stories repeated at 12 months` = pct(repeated),
            `reliability cost of forbidding repeats` = n2(abs(round(cost, 3)), 3)) |>
  knitr::kable(caption = "Second wave 12 months later, with growth from the data (latent slopes)")
res_cat$q4_waves_bound |> rename(`waves without any repeat` = bound) |>
  knitr::kable(caption = "How many waves each grouping allows before a story must repeat")
```

Pure information-based selection gives most children the same stories at every wave, because growth of a few tenths of a logit a year barely changes the best story. Choosing at random among the top two about halves the repeats, at little cost. Forbidding repeats costs reliability at every wave. Across six history-aware waves, reliability in the current three blocks falls below random selection by wave 4, because the informative stories are used up. Whether re-seeing a story causes a practice effect is not in these data; only a retest study can answer that.

## Calibration error and rescoring after the fact

```{r cat-calib}
q5m |>
  transmute(`calibration SEs` = if_else(se_inflation == 1, "current", "doubled"),
            rule = rule_labels[rule],
            `overstated reliability (capitalization)` = n2(capitalization, 3),
            `lost to worse story choices` = n2(selection_loss, 3),
            `recovered by rescoring with true parameters` = n2(rescoring_gain, 3)) |>
  knitr::kable(caption = "Selecting with perturbed parameters, scoring with the true ones (mean over groups and designs)")
```

At today's calibration precision, runtime parameter error barely matters. Rescoring after the fact recovers the scoring error, but it cannot change which stories were given; with doubled SEs (a new language on a small calibration sample), that loss is about .01. Recalibrating before deployment on the corrected data is still worthwhile. Rescoring every administration offline, as the score of record, is what makes the runtime parameters low-stakes.

## Cross-language non-invariance

```{r cat-dif}
res_cat$q6 |>
  filter(rule == "maxinfo", design %in% c("cur3", "best4")) |>
  group_by(label, language, variant, design) |>
  summarise(across(c(bias_pooled, bias_rescored, realized_pooled, realized_rescored, realized_oracle), mean),
            .groups = "drop") |>
  mutate(across(where(is.double), \(x) n2(x))) |>
  knitr::kable(caption = paste("Responses generated with group-specific difficulties, stories selected with pooled",
                               "parameters; bias (logits) and realized reliability, scored with pooled vs",
                               "group-specific parameters. Variants bracket the two items at the guessing floor."))
```

The pooled parameters for stories 13–18 are dominated by Argentine data, so under non-invariance it is the German children who are scored off-model on those stories, not the Argentine ones. Rescoring with group-specific parameters removes the bias. This also flags a caveat for this chapter's German scores on stories 13–18.

## Local dependence

Generating responses with story testlet effects, but selecting and scoring with the unidimensional model, overstates reliability by `r n2(min(res_cat$q7 |> filter(rule == "maxinfo", scenario != "independent") |> group_by(scenario) |> summarise(o = mean(predicted - realized)) |> pull(o)), 3)`–`r n2(max(res_cat$q7 |> filter(rule == "maxinfo", scenario != "independent") |> group_by(scenario) |> summarise(o = mean(predicted - realized)) |> pull(o)), 3)`. The ranking of designs is unchanged (rank r ≥ `r n2(min(res_cat$q7_rank$spearman_pred_real))`). A testlet model is not needed for selection. Discount reliability claims slightly.

## Recommendation

1. **Select stories adaptively within blocks,** by Fisher information at the current EAP (or information per minute), not by discrimination.
2. **Give 4 stories:** today's three blocks plus one free story, at most one per story type. That is about `r n2(cg("cur3p1", "maxinfo", "med_min"), 1)` minutes median and about `r n2(cg("cur3p1", "maxinfo", "q90_min") + val_time$obs_q90_min - median(val$sim_q90_min), 1)` at the 90th percentile (adding the simulation's known shortfall). It ties the best 4-block grouping, is the smallest change from today, and allows the most waves before stories must repeat. It is also better than the canvas grouping.
3. **Start at the group mean or an age-predicted value, and run the online estimate under the group's prior.** For new sites, use the pooled prior and rescore afterwards. Never test under N(0, 1).
4. **Use top-2 randomesque selection,** at least until recalibration. It spreads exposure and calibration data across stories and halves repeats at retest. Allow repeats rather than forbidding them until a retest study measures practice effects.
5. **Before deploying:**
   - recalibrate on the corrected data (after levantemodels PR #27);
   - fix or remove the dead items;
   - rescore every administration offline as the score of record, group by group, with group-specific parameters for items known to be non-invariant.
6. **Engineering:**
   - core-tasks needs story-level selection;
   - the runtime ability update must use each item's discrimination (today it is fixed at 1);
   - estimation must be EAP with a prior;
   - θ, SE and the chosen story should be logged per trial.

   The item bank is thin above θ = 0 on the German scale, so better precision for older or higher-scoring children needs harder stories as well as a better selection rule.

# June conclusions → now

Each June conclusion is re-tested on the corrected data and labelled by
these criteria:

- **survives**: same direction and same practical conclusion, with an
  estimate close to June's;
- **survives with changes**: same conclusion, but a materially different
  magnitude or a narrower scope;
- **partly overturned**: the conclusion holds in some samples, models or
  designs and fails in others;
- **different mechanism**: June's number is reproduced, but by a different
  route than June described;
- **not replicated**: June's estimate is not reproduced on these data;
- **overturned**: the conclusion is contradicted;
- **not re-tested**.

```{r june-table}
sm_a <- \(qt) pick(res_items$qtype_summary, mean_age_slope, question_type == qt)
g_en <- filter(gg, sample == "en-US")
q3_rng <- rng(q3_qt$within_minus_between, 3)
jt <- tribble(
  ~source, ~claim, ~june, ~now, ~status,
  "stories_tom §2", "Story-construct schemes explain little item difficulty; question type dominates",
  "R² ≈ .07–.11 (schemes)",
  glue("question type R² {n2(k$r2_qt)}; old group {n2(k$r2_og)}; new construct {n2(k$r2_nc)}"),
  "survives",

  "stories_tom §2", "False-belief targets carry the age signal; controls are flat",
  "boxplot; schemes R² ≈ .06 (n.s.) for age",
  glue("above the guessing floor, FB {n2(k$fl_fb, 3)} vs controls {n2(k$fl_rc, 3)} logit/yr (difference {n2(k$fl_diff, 2)} [{n2(k$fl_lo, 2)}, {n2(k$fl_hi, 2)}]); raw logit slopes {n2(sm_a('false_belief'), 3)} vs {n2(sm_a('reality_check'), 3)} (difference {n2(k$raw_diff, 2)} [{n2(k$raw_lo, 2)}, {n2(k$raw_hi, 2)}]); 2PL slopes {n2(k$a_fb)} vs {n2(k$a_rc)}"),
  "survives on the chance-corrected scale (raw logit slopes do not separate them; controls also flat because at ceiling)",

  "stories_tom §3", "Controls correlate moderately with targets, weakly with age",
  "r = .46 and .27 (pooled)",
  glue("pooled {n2(k$r_ct_pooled)} / {n2(pick(cc, r_control_age, method == 'pooled over runs (June method)', item_set == 'policy items dropped'))}; within dataset × form {n2(k$r_ct_within)} / {n2(pick(cc, r_control_age, method == 'within dataset x form, Fisher-pooled', item_set == 'policy items dropped'))}"),
  "survives with changes (pooled r reproduces; within a site it is lower)",

  "stories_tom §3", "Controls do not gate target age sensitivity",
  "similar target~age slopes after pass and fail",
  glue("June method {n2(pick(gj, target_age_beta_perSD, !ctrl_pass), 3)} vs {n2(pick(gj, target_age_beta_perSD, ctrl_pass), 3)}/SD; GLMM interaction de-DE {n2(g_de$interaction, 2)} (p {pe(g_de$interaction_p)}), en-US {n2(g_en$interaction, 2)} (p = {n2(g_en$interaction_p, 3)}), es-CO {n2(g_co$interaction, 2)} (p = {n2(g_co$interaction_p, 2)})"),
  "partly overturned (gating in de-DE and en-US, not in es-CO; descriptive)",

  "stories_tom §3", "Dropping controls lowers reliability; keep them",
  "marginal .92 → .89",
  glue("June method {n2(rrow(jn_s, 'all items', 'Rasch', marginal_rxx_mirt), 3)} → {n2(rrow(jn_s, 'targets only', 'Rasch', marginal_rxx_mirt), 3)}; empirical Leipzig {n2(rrow(de_s, 'all items', 'Rasch', empirical_rxx), 3)} → {n2(rrow(de_s, 'targets only', 'Rasch', empirical_rxx), 3)}; UTDT {n2(rrow(ar_s, 'all items', 'Rasch', empirical_rxx), 3)} → {n2(rrow(ar_s, 'targets only', 'Rasch', empirical_rxx), 3)}"),
  "survives with changes (direction holds; the magnitudes were model-implied)",

  "stories_tom §5; tom_dimensionality §1", "Not cleanly unidimensional",
  "ev1/ev2 ≈ 1.9 (Leipzig)",
  glue("Leipzig {n2(k$ev_de)}; Western/Bogotá/es-AR {rng(k$ev_other)}; ECV {k$ecv}, omega-h/omega-t {rng(ld$omega_h / ld$omega_total)}"),
  "survives with changes (Leipzig and es-AR; one score defensible elsewhere)",

  "tom_dimensionality §1", "Parallel analysis's 11 factors are over-extraction",
  "11 factors",
  glue("June method {rng(res_dim$eigen$pa_june_nfact, 0)} factors; resampled {rng(res_dim$eigen$pa_resample_ncomp[res_dim$eigen$sample %in% dim_primary], 0)} components, {rng(res_dim$eigen$pa_resample_nfact[res_dim$eigen$sample %in% dim_primary], 0)} factors"),
  "survives (parallel analysis is uninformative here either way)",

  "tom_dimensionality §2", "Question type organizes covariance; story weakly; constructs not at all",
  "within − between: .101 / .039 / .045 (table)",
  glue("tetrachoric: question type {rng(qt_contrast, 3)}; story {rng(st_contrast, 3)}; construct {rng(nc_contrast, 3)}. Q3 residuals: question type {q3_rng}, story {rng(q3_st$within_minus_between, 3)}"),
  "partly overturned (the tetrachoric pattern reproduces but partly reflects shared difficulty; on Q3 residuals question type groups items only in some samples)",

  "tom_dimensionality §2", "Two-factor EFA gives no clean sub-construct",
  "factor r .19, 28% variance",
  glue("factor r {rng(res_dim$efa2_summary$factor_r[res_dim$efa2_summary$sample %in% dim_primary])}; second factor differs by sample"),
  "survives",

  "tom_dimensionality §3", "A false-belief-only scale is less reliable and no more age-valid",
  "Rasch rxx .605 vs .842; age r .401 vs .402",
  glue("Leipzig alpha {n2(pick(res_dim$subscales, alpha, sample == 'de_ib', scale == 'false-belief only'), 3)} vs {n2(pick(res_dim$subscales, alpha, sample == 'de_ib', scale == 'all items'), 3)}; es-AR age r {n2(pick(res_dim$subscales, age_r, sample == 'ar_nhb', scale == 'false-belief only'), 3)} vs {n2(pick(res_dim$subscales, age_r, sample == 'ar_nhb', scale == 'all items'), 3)}"),
  "survives",

  "tom_dimensionality synthesis", "ToM is a reliable composite",
  "reliability .92",
  glue("the .92 is model-implied for a {jp$items_in_model}-item pool ({n2(jp$rasch_marginal_rxx_june_method, 3)}); empirical {n2(jp$rasch_empirical_rxx_eap, 3)}; single fixed forms {k$rel_forms}; fCAT {k$rel_fcat}"),
  "overturned",

  "tom_dimensionality synthesis", "Inter-item correlations are uniformly low",
  "≈ .13",
  glue("June-style rebuild {n2(pick(res_dim$eigen, mean_offdiag_r, sample == 'de_ib_junelike'), 3)}; cleaned {rng(res_dim$eigen$mean_offdiag_r[res_dim$eigen$sample %in% dim_primary], 3)}"),
  "survives with changes",

  "tom_dimensionality caveat", "Story testlet effect is real but weak; a bifactor model is risky",
  "not fitted",
  glue("story bifactor better by AIC in all samples; information overstated {k$info_over}; θ r {k$cor_uni_g}"),
  "new (confirms 'weak')",

  "tom_dif_multigroup", "Scalar invariance is rejected with controls in",
  "χ²/df ≈ 29; BIC prefers configural by ≈ 580",
  glue("release-equivalent June design {n2(jbv('J_release', 'all', X2_df), 1)}, dBIC {round(jbv('J_release', 'all', dBIC))}; corrected June design dBIC {round(jbv('J_policy', 'all', dBIC))}; A χ²/df {n2(k$A_scal_x2df, 1)}, Rasch dBIC {n2(tr_('A', 'scalar vs configural', dBIC), 0)}, 2PL dBIC {n2(t2('A', 'scalar vs metric', dBIC), 0)}, floor-0 rule dBIC {n2(tr_('A_g0', 'scalar vs configural', dBIC), 0)}; B χ²/df {n2(k$B_scal_x2df, 1)}, dBIC {n2(tr_('B', 'scalar vs configural', dBIC), 0)} / {n2(t2('B', 'scalar vs metric', dBIC), 0)}"),
  "partly overturned (holds in June's design and in stories 13–18; in stories 1–6 BIC rejects only under the floor-0 guessing rule)",

  "tom_dif_multigroup; stories_tom §4", "Controls are the dominant source of cross-language non-invariance",
  "mean |DIF| ≈ 1.9 (controls) vs ≈ 0.5 (targets)",
  glue("release-equivalent {n2(pick(dbt, mean_abs_db, analysis == 'J_release', kind == 'control'))} vs {n2(pick(dbt, mean_abs_db, analysis == 'J_release', kind == 'target'))}; corrected June design {n2(pick(dbt, mean_abs_db, analysis == 'J_policy', kind == 'control'))} vs {n2(pick(dbt, mean_abs_db, analysis == 'J_policy', kind == 'target'))}; A Rasch {n2(dmv('A', 'control', rasch_s2))} vs {n2(dmv('A', 'target', rasch_s2))}, 2PL {n2(dmv('A', 'control', twopl_purified))} vs {n2(dmv('A', 'target', twopl_purified))}; B Rasch {n2(dmv('B', 'control', rasch_s2))} vs {n2(dmv('B', 'target', rasch_s2))}, 2PL {n2(dmv('B', 'control', twopl_purified))} vs {n2(dmv('B', 'target', twopl_purified))}"),
  "partly overturned (model- and design-dependent; the gap shrank)",

  "tom_dif_multigroup", "Targets are comparable; false-belief and emotion DIF is small",
  "mean |DIF| ≈ 0.5",
  glue("false belief flagged {fb_A$n_flagged}/{fb_A$n_items} (A), {fb_B$n_flagged}/{fb_B$n_items} (B); story 13 FB es-AR {n2(k$acc_13_ar, 3)}; within es-AR across sites {wl_f('X_esAR')}/{wl_n('X_esAR')} items flagged"),
  "overturned (but language is confounded with age, exposure and site)",

  "tom_dif_multigroup", "Partial scalar invariance is achievable",
  "p ≈ .8 (targets)",
  glue("A p = {n2(k$A_part_p)} ({pick(res_inv$anchors, n_anchors, analysis == 'A')} anchors); B p = {n2(k$B_part_p)} ({pick(res_inv$anchors, n_anchors, analysis == 'B')} anchors); without V0 p = {n2(tr_('A_noV0', 'partial scalar vs configural', p), 4)}"),
  "survives descriptively (anchors chosen on the same data; rejected without V0)",

  "tom_dif_multigroup", "The invariant anchors are mostly false-belief and emotion targets",
  "anchor set of targets",
  glue("anchors include {anc$ctrl_anchor[anc$analysis == 'A']}/{anc$ctrl[anc$analysis == 'A']} controls (A) and {anc$ctrl_anchor[anc$analysis == 'B']}/{anc$ctrl[anc$analysis == 'B']} (B); DIF below ≈ {n2(k$mdd_med, 1)} logits is not detectable"),
  "survives with changes",

  "tom_reality_check_bug", "The 4 provable defects are inverted keys (CO moral rc_1, CO reference, late-DE deception rc_2/3)",
  "below chance, site-specific",
  glue("story 2 rc_1 {n2(k$s2rc1_2e, 3)} on V0 2e only (cause undetermined); story 4 rc_2/3 {n2(k$s4rc2_4g, 3)}/{n2(k$s4rc3_4g, 3)} on image 4g (rc_3 below chance in every language, rc_2 in de-DE and en-US; the key conflicts with the image; flipped r {n2(fc(rc2, 'flagged: 4g', r_ir_flipped))}/{n2(fc('tom_story4_deception_reality_check_3', 'flagged: 4g', r_ir_flipped))}); reference {n2(k$rr5, 3)}/{n2(k$rr11, 3)}/{n2(k$rr17, 3)}, below chance except German story 5 (design)"),
  "partly overturned (the cells are broken, but not site-specific and not all key inversions)",

  "tom_reality_check_bug", "Repairing the defects removes ≈ 70% of scalar non-invariance",
  "χ² 702 → 212",
  glue("{big(round(jbv('J_release', 'all', X2)))} → {big(round(jbv('J_policy', 'all', X2)))} ({spct(jbv('J_policy', 'all', X2_change_vs_release))}); controls {spct(jbv('J_policy', 'controls', X2_change_vs_release))}; PR #27 alone {spct(jbv('J_corrected', 'all', X2_change_vs_release))}"),
  "different mechanism (about half, from exclusions, not key flips)",

  "tom_reality_check_bug; README", "The 12 hand-flagged items are essentially a complete catalog of defects",
  "all 12 map onto defects",
  glue("{k$v_un} of {k$v_all} story-level items un-excluded; {k$v_keep} keep; {k$v_scop} scoped; {k$v_es} held for the translation check"),
  "overturned",

  "stories_tom §1", "Fixing the scoring bug raised Leipzig test–retest",
  "≈ .45 → ≈ .53",
  glue("first two runs {n2(pick(sf2, r, version == 'analysis'), 3)} (released {n2(pick(sf2, r, version == 'released v2_3'), 3)}; release-equivalent refit {n2(pick(sf2, r, version == 'release-equivalent refit'), 3)}); 60–180 d {n2(k$stab_60)}; > 300 d {n2(pick(sb, r, dataset == 'pilot_mpieva_de_main', gap_bin == '[300,Inf)', version == 'analysis'))}"),
  "not replicated (the corrections do not raise it; alternate forms, months apart)",

  "stories_tom §6", "Parallel retest items are broadly similar in difficulty",
  "within-family SD ≈ 0.9 vs 2.3",
  "not re-tested",
  "not re-tested"
)
knitr::kable(jt, col.names = c("June source", "Claim", "June (quoted)", "Now (2026-09 data)", "Status"))
```

# Synthesis and recommendations

```{r rec-numbers}
fb_flag <- res_inv$dif |>
  filter(analysis %in% c("A", "B"), question_type == "false_belief", class != "A") |>
  mutate(story = as.integer(str_extract(short, "(?<=story)\\d+"))) |>
  arrange(story) |> pull(short) |> unique()
k$fb_flag_list <- paste(uid_label(fb_flag), collapse = ", ")
```

**Exclusions and corrections (DCC).**

1. Keep item exclusions in the Airtable table, as the DCC prefers, but
   fix what is there. The 12 generic rows do nothing today (the `anti_join`
   runs after `recode_tom()`), and they would over-exclude if they worked.
   Replace them with story-level rows: un-exclude the `r k$v_un` story-level
   items that behave normally (`r k$v_weak` of them weak); exclude
   reference_reference in stories 5/11/17 and story 10's deception false
   belief 1 everywhere; and add the three content-scoped rows above, which
   need a content-version (or date-window) column. Hold any Spanish row for
   story 13 until the translation check.
2. The same reasoning applies to corrections. PR #27 writes its identity
   remaps (`fix_tom_item_uids()`: a 25-row table keyed on date window, item
   code, answer and occurrence) and three per-story chance values into
   levantemodels code. Those are data properties too. A metadata table, such
   as the item-mapping tables in `levante_metadata_items`, would document
   them next to the exclusions, and could also hold a scoped key correction
   for the story-4 reality checks on image 4g if the content team confirms
   that "no" is right for that scene (that would keep
   `r big(sum(res_items$flip_check$n[res_items$flip_check$content_part == "flagged: 4g" & res_items$flip_check$language == "all"]))`
   trials). This is a design choice for the DCC; the analysis only needs the
   corrections applied somewhere.
3. Consider V0 Bogotá story-2 reality check 2 and story 5's reference
   reality check (non-invariant only because of V0) as further V0-scoped
   rows, next to V0 reality check 1, whose cause stays undetermined.

**Deployed items (core-tasks; item bank).**

4. Re-author story-4 reality checks 2/3 for image 4g (or change their key,
   as above), and redesign reference_reference (stories 5/11/17). Both are
   still in the deployed corpora.
5. Review the near-chance yes/no "Is [sibling] mad at [sibling]?" questions
   (story 4 fb_3, story 16 ref_2), the non-discriminating story-17 emotion
   question and story 10's deception false belief 1, and the same-scene
   doublets (story 6 fb_2/fb_5, both keyed "shoes"; story 4 fb_2/fb_3, both
   "no").
6. Check the Spanish translations (es-CO, es-AR) of the false-belief
   prompts: every false-belief item flagged (class B or C) in the invariance
   analyses (`r k$fb_flag_list`), plus story 7's reality_known item, which
   the item screen flags (stories 7–12 are in no invariance analysis).
   Translation is one hypothesis; prior exposure to the task and the samples
   themselves are others. (Sites that share a language differ less and on
   other items, and the false-belief DIF persists with younger German
   children, so site and age alone do not explain it.) Until the check is done,
   keep the Spanish sensitivity flag and report results with and without
   it; it changes scores very little.

**Scoring.**

7. Score group by group (the PR #26 rule, not yet merged). Before changing
   production's integration grid, investigate how the low-scoring groups'
   means and variances depend on the grid, the latent density and the
   guessing convention: all three move them, while scores within a group
   barely change. For the RfP1 datasets that v2_3 scored under the German
   prior, their own prior alone lowers the level by `r k$shift_prior`
   logits (`r k$shift_total` in total).
8. Recalibrate ToM on the current data once PR #27 is merged. The analysis
   model here uses `r big(pick(res_rel$itemtype_bic, runs, model == "analysis", itemtype == "2PL"))`
   runs in 8 groups. The 2PL is preferred, and local dependence is modest,
   so a testlet model is optional. If it is skipped, discount story
   information by about `r k$info_over` in reliability claims. The block-CAT
   simulation shows it does not change which stories are selected.
9. Treat cross-language comparisons of Stories scores as partial-scalar at
   best, and remember that the language groups also differ in age, prior
   exposure and site. Report group differences only with that caveat and
   adjusted for age.

**Block CAT and the item bank.**

10. The deployed fCAT is a random 3-story form with reliability
    `r k$rel_fcat` in a median of `r k$fcat_min` minutes. In the same
    children, three random stories instead of a six-story form cost
    `r rng(cs$rel_full - cs$rel_random)` in reliability for half the time.
    The 3-vs-4 decision is mostly a trade of minutes for precision; the
    simulation in [Designing the block CAT](#designing-the-block-cat) recommends
    4 stories (today's blocks plus one free story) with adaptive selection.
11. Adaptive selection within blocks helps: choosing the more informative of
    the form's two stories per block raises the reliability of a 3-story
    form by a median of `r n2(median(cs$rel_adaptive - cs$rel_random))`
    (up to `r n2(max(cs$rel_adaptive - cs$rel_random))`) in the same time. With six stories per block it does more (simulation: `r n2(cg("cur3", "random"))` → `r n2(cg("cur3", "maxinfo"))`), but the bank is thin
    above θ = 0 on the DE scale, so real gains also need harder stories and
    items, calibrated in several languages. The fCAT's fixed block order
    confounds block with position; randomize it if blocks are to be
    compared.
12. The block structure need not follow the story constructs, which do not
    organize the item covariance. Question type's tetrachoric clustering
    partly reflects shared difficulty; once difficulty and guessing are
    modelled, what remains is weak and uneven across samples. Blocks can be
    chosen for content coverage and time.
13. Keep scoring the controls. If time binds, they are the least informative
    questions per item. Whether to keep administering them as comprehension
    checks is a design decision; the data show that a failed control makes
    the story's targets less age-informative mainly in Leipzig (and to a
    lesser degree in en-US), not in es-CO.
14. Only Leipzig has repeat data, as alternate forms months apart. A
    short-interval retest (same or parallel forms, about 2 weeks) at a
    high-volume site would give the first real test–retest estimate.

# Methods notes and caveats

- **Post-hoc rules.** Two rules in the items analysis were revised after
  its first run and are disclosed in `res_items$meta$notes`: the weak-item
  flag and the age-slope minority rule. The originals are kept in
  `r_ir_below_10` and `age_slope_raw`. The 2026-09 revision added, after
  review: the floor-corrected age slopes with a symmetric filter, the
  flipped-key check, the per-language discrimination rule for the Spanish
  items, story 10's deception false belief 1 as a policy exclusion, the
  within-language and age-restricted invariance variants, the 2PL tests,
  the latent-density check and the block-CAT simulation.
- **Guessing floor.** The reliability and invariance models fix it at
  chance; the dimensionality models set it to 0 for items answered at or
  below chance in the sample. The invariance and reliability conclusions
  were rechecked under the dimensionality rule (A_g0/B_g0 and the g0
  sensitivity of the reliability model). Neither rule
  can represent a group that answers systematically below chance, as young
  children do on reality_known false-belief questions.
- **`mirt::marginal_rxx()`** integrates TI/(TI + var) over N(0, 1) and
  assumes every item in the matrix is answered. The June chapters' ".92",
  ".89" and ".84" come from it on sparse, multi-form matrices. This chapter
  reports empirical reliability instead.
- **MAP, not ML.** The fixed-floor 2PL and bifactor fits in the
  dimensionality analysis use an N(0, 2²) slope prior, and the invariance
  2PL uses the production priors; their AIC and BIC are evaluated at MAP
  estimates. The bootstrap intervals for ev1/ev2 are biased downward, so the
  chapter reports a bootstrap SE instead of percentile intervals.
- **Upper bounds.** The reliabilities come from unidimensional models that
  assume scalar invariance, so they overstate precision somewhat (footnote
  in the reliability section).
- **Confounds.** Language, site, age, prior exposure, form and deployment
  epoch overlap. The es-AR samples are young
  (`r pick(res_dim$samples, age_min, sample == "ar_nhb")`–`r pick(res_dim$samples, age_max, sample == "ar_nhb")`
  years), the Leipzig samples are older, near ceiling and mostly on repeat
  runs in stories 13–18, and es-CO in the invariance analysis pools
  `r length(str_split_1(pick(res_inv$samples, datasets, analysis == "A", group == "es_CO"), ", "))`
  datasets from 2024–2026. The age-band plots describe; they are not a
  model-based age adjustment.
- **Coverage.** en-GB (Sheffield, fCAT only) has too few responses per item
  for the invariance analyses. en-US items from stories 7–18 have fewer
  than 30 responses. Stories 7–12 are sparse everywhere.
- **Gating** conditions on passing a control, which depends on ability, so
  it is descriptive.
- **Invariance fits** use mirt's default θ range (see the numerics note).
- **Not re-tested:** June's parallel-retest-family analysis (stories_tom §6).

# Reproducing {#reproducing}

The chapter reads only cached files in `data/stories_2026-09/` (gitignored)
and fits no models. To rebuild, run the scripts in the order below with
`Rscript` from a working directory **outside** the repo (the book's renv has
an `mgcv` binary that breaks mirt). Each script takes the book root as its
argument (else `STORIES_BOOK_ROOT`, else the directory above the script's
`tasks/` folder). The build must
come first; the four analysis scripts only read its output and can run in
any order after it.

```{r repro}
tribble(
  ~step, ~script, ~writes, ~built,
  1, "tasks/_build_stories_data.R", "stories_trials.rds, stories_runs.rds, reconciliation.rds, README.md (and, if missing, the pinned raw pulls, release snapshots and the v2_3 model record)", format(prov$built_at),
  2, "tasks/_stories_fits_items.R", "results_items.rds", format(res_items$meta$built_at),
  3, "tasks/_stories_fits_dimensionality.R", "results_dimensionality.rds",
  paste0(res_dim$meta$run_at, " (", round(as.numeric(difftime(as.POSIXct(res_dim$meta$finished_at), as.POSIXct(res_dim$meta$run_at), units = "mins")), 1), " min)"),
  4, "tasks/_stories_fits_invariance.R", "results_invariance.rds",
  paste0(res_inv$meta$created, " (", round(res_inv$meta$runtime_min, 1), " min)"),
  5, "tasks/_stories_fits_reliability.R", "results_reliability.rds, stories_scores_analysis.rds", format(res_rel$meta$built_at)
) |>
  knitr::kable(caption = "Scripts in run order, outputs and build times")
```

```bash
cd /some/dir/outside/the/repo
export LEVANTEMODELS_PATH=/path/to/levantemodels   # checkout of PR #27's branch
Rscript /path/to/levante-analysis/tasks/_build_stories_data.R /path/to/levante-analysis
Rscript /path/to/levante-analysis/tasks/_stories_fits_items.R /path/to/levante-analysis
Rscript /path/to/levante-analysis/tasks/_stories_fits_dimensionality.R /path/to/levante-analysis
STORIES_CORES=6 Rscript /path/to/levante-analysis/tasks/_stories_fits_invariance.R /path/to/levante-analysis
Rscript /path/to/levante-analysis/tasks/_stories_fits_reliability.R /path/to/levante-analysis
```

The data build and the reliability script load levantemodels from the PR #27
source tree with `pkgload::load_all()`. `LEVANTEMODELS_PATH` must point to a
checkout of branch `fix-tom-identity-corrections` (they stop if it is
unset). They record its commit
(`r substr(res_rel$meta$levantemodels[["commit"]], 1, 7)` for the reliability
run). Every results file's `meta` element records the inputs, thresholds,
package versions and `sessionInfo()`. The data build pins every raw dataset
version (table above); the release snapshots are pinned by processed version,
except Sheffield's unreleased next draft, which cannot be pinned.
