---
title: "Test–retest figures (corrected v1.2) — Colombia & Germany"
levante:
  summary: "Test–retest scatter figures (T1 vs T2 per task, coloured by age) for Bogotá and Leipzig, in the style of the core-tasks paper."
  status: draft
  task_ids: [hearts-and-flowers, memory-game, same-different-selection, matrix-reasoning, mental-rotation, egma-math, vocab, trog, theory-of-mind, swr, sre, pa]
  datasets: [pilot_mpieva_de_main, pilot_uniandes_co_bogota]
  data:
    - {source: "load_levante_scores", version: "v1_2", table: "scores"}
  depends_on: [00_load_data.qmd]
  findings:
    - "Figures only; written to figures/test_retest_colombia.png and figures/test_retest_germany.png."
  last_run: 2026-07-16
format:
  html:
    toc: true
    code-fold: true
    fig-width: 12
    fig-height: 10
execute:
  warning: false
  message: false
---

{{< include ../_disclaimer.qmd >}}

# Purpose

Test–retest scatter figures in the style of the core-tasks paper (Figure 4):
each panel is one task, T1 score vs T2 score, points coloured by age, with the
**test–retest correlation r and N** annotated. Two figures:

- **Colombia — Bogotá** (~15-month retest interval)
- **Germany — Leipzig** (~3–7-month retest interval)

On corrected (v1.2) data. Each site shows the tasks for which it has ≥ 20
two-wave children (Colombia has no ToM/Same-and-Different retest; both sites'
thin ROAR cells are dropped where < 20). Scales are per-task (free) because the
tasks are on different score metrics (IRT θ; ROAR-Sentence is a z-scaled
count). PNGs saved to `figures/`.

```{r setup}
source(here::here("common.R"))
fig_dir <- here::here("figures"); fs::dir_create(fig_dir)
sc <- read_rds(here::here("data/scores_all_sites.rds"))

cat_order <- c("Executive function","Reasoning","Spatial cognition",
               "Language","Social cognition","Math","Reading")

build_pairs <- function(ds) {
  sc |> filter(dataset == ds,
               task_id %in% c(core_task_ids, "swr", "sre", "pa"),
               !is.na(score), !is.na(age)) |>
    group_by(task_id, task_label, task_category, user_id) |>
    arrange(age) |> mutate(wave = row_number()) |> ungroup() |>
    filter(wave <= 2) |>
    group_by(task_id, task_label, task_category, user_id) |>
    filter(n_distinct(wave) == 2) |>
    summarise(t1 = score[wave == 1], t2 = score[wave == 2],
              age = age[wave == 1], .groups = "drop") |>
    group_by(task_id, task_label, task_category) |>
    filter(n() >= 20) |>
    mutate(score1 = pmin(pmax(t1, -6), 6), score2 = pmin(pmax(t2, -6), 6)) |>
    ungroup() |>
    mutate(task_category = factor(task_category, levels = cat_order),
           facet = fct_reorder(task_label, as.integer(task_category)))
}

retest_fig <- function(pairs, title) {
  labs_df <- pairs |> group_by(facet) |>
    summarise(r = cor(score1, score2, use = "complete.obs"), n = n(),
              .groups = "drop") |>
    mutate(label = sprintf("r = %.2f\nN = %d", r, n))
  ggplot(pairs, aes(x = score1, y = score2)) +
    geom_abline(slope = 1, intercept = 0, lty = 3, color = "grey55") +
    geom_point(aes(color = age), alpha = 0.55, size = 1.1) +
    geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.7) +
    geom_text(data = labs_df, aes(x = -Inf, y = Inf, label = label),
              hjust = -0.12, vjust = 1.15, size = 3.6, lineheight = 0.95,
              family = .levante_font) +
    facet_wrap(~facet, scales = "free") +
    scale_color_viridis_c(option = "plasma", end = 0.92, name = "Age") +
    labs(x = "Score, time 1", y = "Score, time 2", title = title) +
    theme(legend.position = "bottom",
          strip.text = element_text(face = "bold"))
}
```

# Colombia — Bogotá (~15-month retest)

```{r co-fig, fig.width=12, fig.height=9}
co <- build_pairs("pilot_uniandes_co_bogota")
p_co <- retest_fig(co, "Test–retest — Colombia (Bogotá), ~15 months")
p_co
ggsave(file.path(fig_dir, "test_retest_colombia.png"), p_co,
       width = 12, height = 9, dpi = 300, bg = "white")
```

# Germany — Leipzig (~3–7-month retest)

```{r de-fig, fig.width=12, fig.height=9}
de <- build_pairs("pilot_mpieva_de_main")
p_de <- retest_fig(de, "Test–retest — Germany (Leipzig), ~3–7 months")
p_de
ggsave(file.path(fig_dir, "test_retest_germany.png"), p_de,
       width = 12, height = 9, dpi = 300, bg = "white")
```

# Test–retest correlations

```{r r-table}
bind_rows(
  build_pairs("pilot_uniandes_co_bogota") |> mutate(site = "Colombia (~15mo)"),
  build_pairs("pilot_mpieva_de_main")     |> mutate(site = "Germany (~3–7mo)")
) |>
  group_by(site, task_category, task_label) |>
  summarise(n = n(), r = cor(score1, score2, use = "complete.obs"),
            .groups = "drop") |>
  pivot_wider(names_from = site, values_from = c(n, r)) |>
  arrange(task_category, task_label) |>
  knitr::kable(digits = 2, caption = "Test–retest r by task and site")
```

Figures saved to `figures/test_retest_colombia.png` and
`figures/test_retest_germany.png`.
