---
title: "Stories (ToM) — the reality-check DIF was (mostly) data defects: a forensic reconstruction"
levante:
  summary: "Was the large cross-site DIF on ToM reality-check items a data-flow defect rather than real item functioning? Forensic reconstruction from raw Bogotá and Leipzig data."
  status: draft
  task_ids: [theory-of-mind]
  datasets: [pilot_mpieva_de_main, pilot_uniandes_co_bogota]
  data:
    - {source: "data/tom_raw/", table: "trials"}
  findings:
    - "Largely yes: proven defects include 4 answer-key inversions across both sites, hostile-attribution content mislabeled as ToM (110 DE runs), and trial-map shifts."
    - "Repairing only the provable defects removes ~70% of scalar non-invariance (χ² 702 → 212)."
    - "Every one of the paper's 12 hand-flagged items maps onto an identified defect."
    - "Inputs are caches from a shadow re-processing pipeline that is not versioned in this repo."
  last_run: 2026-07-21
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 10
    fig-height: 6
execute:
  warning: false
  message: false
---

{{< include ../_disclaimer.qmd >}}

# Question & hypothesis

Our ToM analyses (`stories_tom`, `tom_dif_multigroup`) found large cross-site
DIF concentrated on **reality-check (control) items** — and excluding them
bought scalar invariance. But that never made theoretical sense: comprehension
checks should be the *most* translatable items. Hypothesis: the DIF is a
**data-flow bug** (item mismapping in early deployments), not real
item functioning.

To test it we built a **shadow processing pipeline**: raw (unprocessed)
Bogotá + Leipzig data, re-processed with the `levantemodels` internals but keeping
the columns production drops — `item_uid_source`, the raw `item` string
(deployment version codes), `answer`/`distractors` (English stimulus keys —
language-independent ground truth), timestamps, and run-level
`task_version`/`corpus`. Heavy downloads are cached in `data/tom_raw/`;
this notebook re-derives results from those caches.

**Verdict up front: the hypothesis is substantially confirmed — but the
mechanism is richer than one mismap.** We found *provable, site×epoch-specific
defects* (answer-key inversions, hostile-attribution content mislabeled as ToM,
mapping shifts), split across BOTH sites. Repairing just the provable defects
removes **~70% of the scalar non-invariance** (χ² 702 → 212).

```{r setup}
source(here::here("common.R"))
suppressMessages(library(mirt))
loop  <- read_rds(here("data/tom_raw/dif_loop_results.rds"))
prepG <- read_rds(here("data/tom_raw/prepG_results.rds"))
prepH <- read_rds(here("data/tom_raw/prepH_results.rds"))
both  <- loop$both
```

# 1. How ToM item identity is constructed (and where it's fragile)

Tracing `levantemodels`'s `process-trials.R` and `recode_tom`:

- Up to three **competing item-ID sources** per trial: the deployed task's own
  `item_uid`, a curated `trial_items` map (trial_id → uid, assigned
  *retroactively*), and a content-keyed `mapping_items` join. The production
  assertion that these never conflict is **commented out**.
- The story-level uid (`tom_story4_…`) is **constructed at recode time** from
  the raw `item` string's leading digits + corpus-metadata `entry`.
- In the raw data, the `item` string is a **deployment version code** (`1c`,
  `4d`, `4c_new3`) and `answer`/`distractors` are **English stimulus keys**
  (`rug`, `shelf`, `medium_cup`) — both usable as independent ground truth.

```{r identity-sources}
both |> count(site, item_uid_source) |>
  pivot_wider(names_from = site, values_from = n, values_fill = 0L) |>
  knitr::kable(caption = "Which source supplies each trial's item identity")
```

Key asymmetry: **CO has no deployed uids at all** — its ToM identity rests
entirely on the retroactive trial-map (all CO ToM data is from the May–June
2024 beta deployments). DE mixes map-based (early) and deployed (late) ids —
with **no overlap**, so the sources can never cross-validate each other.
Additionally, ~4,200 *valid* CO trials carry no identity information at all
and are silently dropped by production processing.

# 2. Smoking gun #1: hostile-attribution content under ToM labels (DE!)

```{r ha}
ha <- both |> filter(str_detect(answer, "hostile-attribution") |
                     str_detect(distractors, "hostile-attribution"))
ha |> mutate(ym = format(as.Date(server_timestamp), "%Y-%m")) |>
  count(site, ym, name = "n_trials") |>
  knitr::kable(caption = glue("{nrow(ha)} trials in {n_distinct(ha$run_id)} runs carry hostile-attribution answer keys but ToM item ids"))
ha |> count(item_uid, sort = TRUE) |> head(8) |>
  knitr::kable(caption = "ToM uids polluted by ha-keyed trials (top 8)")
```

**110 Leipzig runs (Sept–Oct 2024)** logged hostile-attribution answer keys
under ToM uids — spread across deception, interpretation, and moral-reasoning
items, including the hand-flagged `deception_reality_check_2/_3` and
`moral_reasoning_reality_check_1`. Zero such trials in CO. (Curiously these
trials' *accuracy* patterns match CO's clean values — the labels are corrupt
but the scoring of that epoch behaved like the canonical items; the late-DE
epoch is where the keys go wrong, below.)

# 3. Smoking gun #2: answer-key inversions (below-chance accuracy)

A comprehension question cannot score *below chance* unless its key is wrong.
Binomial screen of every (item × site) cell vs. its chance level (ha-epoch
excluded):

```{r below-chance}
prepG$broken |>
  mutate(across(c(acc, chance), \(x) round(x, 2))) |>
  select(site, item_uid, n, accuracy = acc, chance) |>
  knitr::kable(caption = "Cells significantly BELOW chance (p < .01): provably broken keys")
```

Four provable defects, **split across sites**: CO's
`moral_reasoning_reality_check_1` (8% on a 2AFC! → inverted key: flipping
gives 92% ≈ DE's 98%) and `reference_reference`; DE's (late-epoch)
`deception_reality_check_2/_3` (29%/17% → inverted; the flips land near CO's
values). All four are on the paper's hand-flagged list.

# 4. Content audit: bindings, orders, variants

Using the version-code blocks and English answer keys (modal per
site × block × position):

- **Label↔content bindings mostly agree across sites** (story-2 rc_1↔`shelf`,
  rc_2↔`coats` in both; story-5 reference↔`medium_cup`, rc↔`large_cup` in
  both) — the wholesale "IDs reversed between versions" scenario is ruled out.
- **Question order differs by site** (story 1 plays [rc, fb] in CO,
  [fb, rc] in DE) — harmless for scoring but fatal for any position-based
  retro-mapping.
- **Different option variants under one uid**: story-3's fb/rc_2 use
  `tree`/`fountain` keys in CO but `behind-tree`/`behind-fountain` in DE —
  different stimulus variants pooled under the same item.
- **Within-DE doubling**: in the `2f` block a 47-run sub-epoch maps two
  consecutive positions to the same uid (`false_belief_2` twice).
- **6c/6d mapping shifts**: an 86-run DE subset assigns `6d` trials to
  `false_belief_1/_2` (canonically `6c` items) — with option-count mismatches
  proving the content differs — explaining the hand-flagged
  `second_order_false_belief_2–5`.

# 5. The closed-loop DIF ladder

Fast rest-score logistic DIF (`correct ~ rest_z + site` per item, BH-corrected)
across data preparations:

```{r ladder}
summ <- function(res, lab) res |>
  mutate(qtype = if_else(str_detect(item, "reality_check"), "control", "target")) |>
  group_by(qtype) |>
  summarise(n_items = n(), n_flag = sum(p_adj < .05),
            mean_abs_beta = round(mean(abs(beta_site)), 2), .groups = "drop") |>
  mutate(prep = lab)
bind_rows(
  summ(loop$resA, "A: production identity (baseline)"),
  summ(loop$resC, "C: identity rebuilt from version-code + position"),
  summ(prepG$resG, "G: ha-epoch + broken cells removed"),
  summ(prepH$resH, "H: repaired (keys flipped, unrecoverables dropped)")
) |> select(prep, qtype, n_items, n_flag, mean_abs_beta) |>
  knitr::kable(caption = "DIF by prep: controls vs targets (mean |site beta|, logits)")
```

Reading the ladder honestly: position-rebuilt identity (C) halves control DIF
— but is itself invalid because question *order differs by site*. The
removal/repair preps (G/H) eliminate the proven defects but *concentrate* the
remaining contrast on residual items (the mean over surviving controls rises
even as the proven offenders are fixed). The decisive test is the multigroup
model on the full item set:

# 6. Confirmatory: scalar invariance, controls included

```{r confirmatory}
prepH$mg |> knitr::kable(caption = "Multigroup Rasch, scalar vs configural (controls INCLUDED)")
```

Repairing only the **provable** defects (3 key-flips + 1 cell drop + ha-epoch
exclusion) cuts the scalar-vs-configural misfit by **~70%** (χ² 702 → 212;
ΔBIC 550 → 67). Full scalar invariance is still rejected — see residuals.

# 7. What remains unexplained

```{r residuals}
prepH$resH |> arrange(p_adj) |> head(6) |>
  mutate(beta_site = round(beta_site, 2)) |>
  select(item, beta_site, n, p_adj) |>
  knitr::kable(caption = "Largest residual DIF after repair")
```

The residual profile is suspicious in the same direction: late-DE accuracy of
**98–99%** on several controls (`moral_rc_2`, `interpretation_rc_1`) vs CO at
~65% — near-ceiling one-sided performance on supposedly informative items.
This *smells* like further platform/keying differences (e.g., late-DE
displaying or scoring these as trivially correct), but logs alone can't prove
it. These need task-code / corpus archaeology (compare the deployed
`theory-of-mind` task builds for the CO May-2024 vs DE late-2024 epochs).

# Conclusions & recommendations

1. **The reality-check DIF was substantially manufactured by deployment-era
   data defects** — answer-key inversions (4 proven cells, both sites!),
   hostile-attribution content mislabeled as ToM (110 DE runs), mapping
   shifts (86-run DE 6c/6d subset; 47-run 2f doubling), and variant pooling.
   The published decision to drop controls was right in practice; now we know
   *why* they were pathological.
2. **The "buggy Bogotá" hunch was half-right**: CO's identity is the most
   fragile (fully retro-mapped, 4,200 valid trials dropped, one inverted key,
   one mis-keyed item) — but the largest *provable* corruption (ha-mislabeling,
   inverted deception keys) is in **Leipzig's later epochs**.
3. **Recoverable now**: flip the 3 inverted 2AFC keys (restores ~570 trials of
   signal); exclude the ha-epoch ToM trials; fix the trial-map for the 6c/6d
   and 2f subsets. Then re-test scalar invariance — we project most of the
   remaining gap closes after the residual items are adjudicated.
4. **Pipeline fixes for the DCC**: re-enable the conflict assertion in
   `add_item_ids`; add a below-chance validator rule (any item-site cell
   significantly below chance fails ingest); investigate the 4,200 unmappable
   CO ToM trials; document that ToM item identity is retro-mapped for all
   pre-July-2024 data.
5. The hand-flagged list from the paper is essentially a **complete catalog of
   the defective cells** — every one of the 12 now has a mechanistic
   explanation.
