---
title: "Pattern Matching & Sentence Understanding — training vs development?"
levante:
  summary: "Is the release-notes 'T2 > T1' gain on Pattern Matching and Sentence Understanding a practice effect or development?"
  status: draft
  task_ids: [matrix-reasoning, trog]
  datasets: [pilot_mpieva_de_main, pilot_uniandes_co_bogota]
  data:
    - {source: "load_levante_scores", version: "v1_2", table: "scores"}
    - {source: "load_levante_trials", version: "v1_2", table: "trials"}
  depends_on: [00_load_data.qmd]
  findings:
    - "On corrected data the T2 > T1 effect largely does not replicate; it was mostly the scoring bug."
    - "The one clean cell (Leipzig Pattern Matching, mode-stable, equal length) shows no practice gain."
    - "Not fully resolvable with current data; a fixed-form, fixed-length retest would isolate practice."
  last_run: 2026-07-16
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 9
    fig-height: 5
execute:
  warning: false
  message: false
---

{{< include ../_disclaimer.qmd >}}

# Question

The v1.0 release notes flagged **Pattern Matching** (matrix reasoning) and
**Sentence Understanding** (TROG) as showing systematic **T2 > T1 within both
modes** — possibly a retest/practice (training) effect rather than development.
We revisit this on corrected (v1.2) data using three discriminators:

1. **Observed gain vs. expected development** = within-child T2−T1 vs.
   (cross-sectional age slope × retest interval). Excess beyond development
   ⇒ practice.
2. **Retest interval** (Leipzig has ~3-month and ~7-month groups): a fixed
   practice bump is interval-independent; development scales with interval.
3. **Repeated items**: if accuracy jumps more on items seen at *both* waves
   than on new items, that's item-specific practice/memory.

Two confounds to watch: most longitudinal pairs cross the CAT/non-CAT **mode
boundary**, and **test length** can differ across waves (shorter ⇒ more EAP
shrinkage).

```{r setup}
source(here::here("common.R"))
sc <- read_rds(here("data/scores_all_sites.rds"))
tasks <- c("matrix-reasoning","trog")
w <- sc |> filter(task_id %in% tasks,
                  dataset %in% c("pilot_mpieva_de_main","pilot_uniandes_co_bogota"),
                  !is.na(score), !is.na(age)) |>
  group_by(dataset, task_id, user_id) |> arrange(age) |>
  mutate(wave = row_number()) |> ungroup() |> filter(wave <= 2)
pairs <- w |> group_by(dataset_label, task_label, user_id) |>
  filter(n_distinct(wave) == 2) |>
  summarise(t1 = score[wave==1], t2 = score[wave==2],
            gap = age[wave==2]-age[wave==1],
            a1 = adaptive[wave==1], a2 = adaptive[wave==2],
            na1 = num_attempted[wave==1], na2 = num_attempted[wave==2],
            .groups = "drop") |>
  mutate(gain = t2 - t1, mode = if_else(a1==a2, "mode-stable", "NA→CAT"))
```

# 1. Within-child gain vs expected development

```{r gain-vs-expected}
xsec <- w |> filter(wave==1) |> group_by(dataset_label, task_label) |>
  summarise(xsec_slope = coef(lm(score ~ age))[2], .groups="drop")
pairs |> group_by(dataset_label, task_label) |>
  summarise(n=n(), mean_gain=mean(gain), mean_gap=mean(gap), .groups="drop") |>
  left_join(xsec, by=c("dataset_label","task_label")) |>
  mutate(expected_dev = xsec_slope*mean_gap, excess = mean_gain - expected_dev) |>
  select(dataset_label, task_label, n, mean_gain, xsec_slope, mean_gap,
         expected_dev, excess) |>
  knitr::kable(digits = 2,
               caption = "Observed gain vs cross-sectional expectation (excess > 0 ⇒ practice-like)")
```

The dramatic **T2 > T1 from the release notes does not replicate on corrected
data.** Leipzig now shows small *declines* (Pattern Matching −0.37, Sentence
Understanding −0.11); Bogotá shows gains (PM +0.33, *below* its developmental
expectation +0.55; SU +0.83, *above* its expectation +0.43). So the only
"excess" gain is Bogotá Sentence Understanding (+0.40), and only there.

# 2. The confounds: mode and test length

```{r confounds}
pairs |> group_by(dataset_label, task_label) |>
  summarise(n=n(),
            pct_mode_stable = round(mean(mode=="mode-stable")*100),
            med_items_t1 = median(na1, na.rm=TRUE),
            med_items_t2 = median(na2, na.rm=TRUE),
            .groups="drop") |>
  knitr::kable(caption = "Mode transitions and test length per wave")
```

- **Bogotá** PM & SU are **100% NA→CAT** — every pair crosses the mode
  boundary, so their changes mix development with the mode transition.
- **Leipzig Sentence Understanding** is mostly NA→CAT *and* halves in length at
  T2 (≈ 86 → 41 items) ⇒ heavier EAP shrinkage; its small change is not
  interpretable as practice.
- **Leipzig Pattern Matching is the one clean cell**: mostly mode-stable
  (155 NA→NA + 36 A→A) with **equal test length** (27 vs 28 items) — so its
  decline is a real signal, not an artifact.

# 3. Clean test — Leipzig Pattern Matching (mode-stable, equal length)

```{r leipzig-pm}
lpm <- pairs |> filter(str_detect(dataset_label,"Leipzig"),
                       task_label=="Pattern Matching", mode=="mode-stable") |>
  mutate(interval = if_else(gap < 0.42, "~3 months", "~7 months"))
lpm |> group_by(interval) |>
  summarise(n=n(), mean_gap=round(mean(gap),2), mean_gain=round(mean(gain),2),
            se=round(sd(gain)/sqrt(n()),2), .groups="drop") |>
  knitr::kable(caption = "Leipzig PM mode-stable gain by retest interval")
```

There is **no practice bump** — the change is slightly negative, and (within
noise) does not grow into a positive gain at either interval. A genuine
retest/practice effect would show a positive jump even over a short interval;
we see the opposite. The small decline is most consistent with regression to
the mean plus the modest test-retest reliability of matrix reasoning, not
training.

# 4. Repeated-item check

If practice were memory for specific items, accuracy should rise more on items
seen at *both* waves. Feasible only where items repeat.

```{r repeated-items}
tr <- load_levante_trials(task_ids = tasks) |>
  filter(dataset %in% c("pilot_mpieva_de_main","pilot_uniandes_co_bogota")) |>
  mutate(correct = as.numeric(correct))
rw <- w |> select(run_id, wave)
itw <- tr |> inner_join(rw, by="run_id") |>
  group_by(task_id, user_id, item_uid) |>
  filter(n_distinct(wave) >= 1) |>
  summarise(seen_both = n_distinct(wave)==2,
            acc_t1 = mean(correct[wave==1], na.rm=TRUE),
            acc_t2 = mean(correct[wave==2], na.rm=TRUE), .groups="drop")
# overlap rates
tr |> inner_join(rw, by="run_id") |>
  distinct(task_id, user_id, wave, item_uid) |>
  group_by(task_id, user_id, item_uid) |> summarise(nw=n_distinct(wave), .groups="drop") |>
  group_by(task_id) |> summarise(pct_items_repeated = round(mean(nw==2)*100,1), .groups="drop") |>
  knitr::kable(caption = "Share of items seen at both waves")

# repeated-item accuracy change (trog only — matrix overlap ~4%)
itw |> filter(task_id=="trog", seen_both, !is.nan(acc_t1), !is.nan(acc_t2)) |>
  summarise(n_repeated_item_obs = n(),
            acc_t1 = round(mean(acc_t1),3), acc_t2 = round(mean(acc_t2),3),
            change = round(mean(acc_t2-acc_t1),3)) |>
  knitr::kable(caption = "TROG: accuracy on items seen at BOTH waves")
```

Pattern Matching repeats only ~4% of items (Mars-IB uses parallel retest
forms), so memorization is structurally unlikely. TROG repeats ~19%; on those
repeated items accuracy does rise modestly — but a few-point bump on a fifth of
items can't explain a large overall effect, and the overall TROG change is
confounded by mode/length anyway.

# 5. Real development is present (cross-sectional)

```{r xsec-grad, fig.height=4.5}
sc |> filter(task_id %in% tasks,
             dataset %in% c("pilot_mpieva_de_main","pilot_uniandes_co_bogota"),
             !is.na(score)) |>
  group_by(dataset, task_id, user_id) |> arrange(age) |> slice(1) |> ungroup() |>
  ggplot(aes(x = age, y = score, color = dataset_label)) +
  geom_point(alpha = 0.2, size = 0.7) +
  geom_smooth(method = "lm", se = TRUE) +
  facet_wrap(~task_label) +
  scale_color_brewer(palette = "Set1", name = NULL) +
  labs(x = "Age (years)", y = "Ability (θ)",
       title = "Both tasks show clear cross-sectional development",
       subtitle = "Older children score higher — the age signal is real") +
  theme(legend.position = "bottom")
```

# Verdict

- **The release-notes "T2 > T1 training effect" was largely the scoring bug.**
  On corrected data it does not replicate: Pattern Matching is flat-to-slightly
  declining, and only Bogotá Sentence Understanding shows gain beyond
  cross-sectional expectation.
- **The one clean longitudinal cell — Leipzig Pattern Matching (mode-stable,
  equal length) — shows no practice gain**, arguing against a strong
  retest/training effect for matrix reasoning.
- **It can't be fully resolved** for these tasks with current data: most pairs
  cross the CAT/non-CAT boundary, test length changes across waves (EAP
  shrinkage), and item overlap is low. The honest read is "no robust training
  effect on corrected data; remaining signal is confounded."
- **To answer it cleanly:** a **fixed-form, fixed-length retest** (same items,
  same number, no mode change) would isolate practice from development. Worth
  flagging to the DCC.

```{r save}
write_rds(pairs, here::here("data", "pm_su_pairs.rds"))
```
