---
title: "Sentence Understanding (TROG) — does character skin tone function differently across samples?"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 9
    fig-height: 6
execute:
  warning: false
  message: false
---

{{< include ../_disclaimer.qmd >}}

# Question

The TROG pictures show many children and adults, drawn with a range of skin tones.
If children respond differently to characters who do or don't look like them, an
item's difficulty could depend on the skin tone of the people it depicts, and could
depend on it **differently in different samples** (e.g. rural vs urban Colombia,
Colombia vs Germany). That would be differential item functioning (DIF) tied to
stimulus skin tone, and it would mean the art costs some children points.

This chapter (1) **measures the skin tone of every person drawn in the TROG art** and
(2) asks whether item skin tone predicts **group differences in accuracy** (item DIF)
or **which picture children choose**.

**What "match" can and cannot mean here.** LEVANTE collects no child-level race,
ethnicity or skin-tone data (none in participants or surveys). A match/mismatch
predictor can therefore only be defined at the **sample** level: we ask whether a
sample as a whole is relatively better or worse on items depicting darker- or
lighter-skinned characters than a reference sample. Whether a given direction is a
"match" depends on the skin-tone composition of each sample, which the site teams
know and we do not. The tests below are two-sided for that reason.

```{r setup}
source(here::here("common.R"))
dir <- here("tasks/trog_skin_tone")
res <- read_rds(here("data/trog_skin_tone_fits.rds"))  # tasks/_trog_skin_tone_fits.R
people <- read_csv(file.path(dir, "trog_person_skin.csv"), show_col_types = FALSE)
imgs   <- read_csv(file.path(dir, "trog_image_skin.csv"), show_col_types = FALSE)
ci <- \(est, se) glue("{sprintf('%.3f', est)} [{sprintf('%.3f', est - 1.96 * se)}, {sprintf('%.3f', est + 1.96 * se)}]")
```

# 1. Skin tone in the TROG art

**Method** (details and code in `tasks/trog_skin_tone/README.md`). Images are the
production set (`levante-assets-prod/visual/trog/`, cached 2026-09-04). People were
found with a COCO Mask R-CNN detector, and each person's **face** was segmented with a
human-parsing model (SegFormer, ATR labels). Skin colour is the median CIELAB colour of
the dominant face-pixel cluster, which drops eyes, mouth and blush. It is summarised as
the **Individual Typology Angle** (ITA = atan((L\* − 50)/b\*); higher = lighter).
**Every measurement was checked by eye** on contact sheets: 35 animal/object "faces"
(horses, dogs, monkeys, turtles…) were removed, 6 people seen from behind were measured
on arms/legs, and duplicate readings were merged. ITA bands come from dermatology
photographs, so here they are descriptive labels for stylised art, not claims about
the characters' ethnicity.

```{r people-strip, fig.height=2.6}
bands <- tibble(lo = c(-Inf, -30, 10, 28, 41, 55), hi = c(-30, 10, 28, 41, 55, Inf),
                band = c("dark", "brown", "tan", "intermediate", "light", "very light"))
people <- people |> mutate(hex = rgb(r, g, bl, maxColorValue = 255),
                           band = cut(ITA, c(-Inf, -30, 10, 28, 41, 55, Inf), labels = bands$band)) |>
  arrange(ITA) |> mutate(rank = row_number())
ggplot(people, aes(x = rank, y = 1, fill = hex)) + geom_tile(height = 1) +
  scale_fill_identity() +
  scale_x_continuous(breaks = people |> group_by(band) |> summarise(m = median(rank)) |> pull(m),
                     labels = levels(people$band)) +
  labs(x = NULL, y = NULL,
       title = glue("Measured skin colour of all {nrow(people)} people in the TROG art, sorted by ITA")) +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), axis.line = element_blank())
```

```{r band-table}
people |> count(band) |> mutate(share = scales::percent(n / sum(n), 1)) |>
  knitr::kable(col.names = c("ITA band", "People", "Share"),
               caption = glue("{nrow(people)} people in {n_distinct(people$image)} of {nrow(imgs)} images"))
```

The art is fairly diverse: about one character in six falls in the brown/dark bands and
one in five in the tan band. Items 81–103 (added after Bishop's original set) are, if
anything, slightly darker on average than items 1–80.

```{r item-dist, fig.height=3.5}
res$item_feat |> filter(!is.na(target_ita)) |>
  ggplot(aes(x = target_ita)) + geom_histogram(binwidth = 5, fill = "#268bd2", color = "white") +
  labs(x = "Target-picture skin tone (mean ITA of people in the correct picture)", y = "Items",
       title = glue("{sum(!is.na(res$item_feat$target_ita))} of 99 items have people in the correct picture"))
```

# 2. Data

Each child's **first** TROG run (cross-sectional), 2026-08-24 trials snapshot. Items
are mapped to pictures with the corpus (`trog-item-bank.csv`). The mapping agrees with
children's recorded responses, and corpus-keyed correctness matches the released
`correct` field.

```{r data-table}
res$id_check |> left_join(res$n_runs, by = "grp") |>
  transmute(Sample = grp, Children = n.y, Trials = n.x,
            `Response in item's picture set` = scales::percent(in_set, .1),
            `Corpus key = recorded correct` = scales::percent(agree, .1)) |>
  knitr::kable()
```

Bogotá is the reference sample throughout. **Rural CO vs Bogotá is the cleanest
contrast**: same Spanish translation, same audio, same age range, so the pictures'
content is the main thing that differs in how items are experienced. Leipzig (German)
and Western Canada (English) contrasts also carry translation differences. **CO (MPIB
intl)** is a smaller es-CO sample of much younger children (4.5–8.4 years). Leipzig is
the **only German TROG sample**, so no urban/rural split is possible within Germany yet.

# 3. Does target skin tone predict item DIF?

Explanatory item-response model (Rasch with LEVANTE's fixed 4AFC guessing floor):
`correct ~ group × target skin tone + (1 | child) + (1 + group | item)`, fit to the
items whose correct picture shows people. The **group × skin tone** term is the test.
It is per 10° of ITA, in logits. Because each item has its own random group effect, the
test asks whether skin tone explains item-level DIF beyond ordinary item-to-item DIF.

```{r dif-table}
res$dif_tab |> filter(term %in% c("g", "g:ita10")) |>
  select(contrast, n_runs, n_items, term, est, se, p, sd_item_dif) |>
  pivot_wider(names_from = term, values_from = c(est, se, p)) |>
  transmute(Contrast = paste(contrast, "vs Bogotá"), Children = n_runs, Items = n_items,
            `Overall gap (logits)` = sprintf("%.2f", est_g),
            `Skin tone × group, per 10° ITA [95% CI]` = ci(`est_g:ita10`, `se_g:ita10`),
            p = sprintf("%.2f", `p_g:ita10`),
            `Item DIF SD` = sprintf("%.2f", sd_item_dif)) |>
  knitr::kable(align = "lrrrrrr")
```

```{r dif-plot, fig.height=7}
lv <- paste(c("Rural CO", "Leipzig", "Western CA", "CO (MPIB intl)"), "vs Bogotá")
fx <- res$dif_tab |> select(contrast, term, est) |> pivot_wider(names_from = term, values_from = est) |>
  mutate(contrast = factor(paste(contrast, "vs Bogotá"), levels = lv))
res$item_dif |> mutate(contrast = factor(paste(contrast, "vs Bogotá"), levels = lv)) |>
  ggplot(aes(x = target_ita, y = dif)) +
  geom_hline(data = fx, aes(yintercept = g), lty = 3) +
  geom_point(alpha = .7, color = "#268bd2") +
  geom_abline(data = fx, aes(intercept = g - `g:ita10` * res$ita_ref / 10, slope = `g:ita10` / 10),
              color = "firebrick", linewidth = 1) +
  facet_wrap(~contrast, scales = "free_y") +
  labs(x = "Target-picture skin tone (ITA; left = darker)", y = "Item DIF vs Bogotá (logits)",
       title = "Per-item DIF against the skin tone of the people in the correct picture",
       subtitle = "Points = item estimates (shrunken); red = model slope (CIs in the table); dotted = average gap")
```

# 4. Are children drawn to pictures by skin tone?

The accuracy test uses only the correct picture. Children's **choices** carry more
information: on `r res$n_choice$items` items, at least two of the four pictures show
people, and these usually differ in skin tone. A conditional logit over the pictures on
screen, with a **fixed effect for every picture** (each picture's overall pull, shared
across samples), estimates whether a sample is **more or less drawn to pictures of
darker vs lighter characters than Bogotá children are**. It also includes a
sample-specific pull toward the correct picture, so accuracy differences can't leak into
the skin-tone term. Negative = relatively more drawn to darker-skinned characters.
Standard errors are clustered by item. The **errors-only** model keeps only trials
where the child chose a wrong picture and asks which distractor they picked.

```{r choice-table}
lab_g <- c(Rural.CO = "Rural CO", Leipzig = "Leipzig", Western.CA = "Western CA", CO..MPIB.intl. = "CO (MPIB intl)")
res$choice_tab |> filter(str_detect(term, "^gS_")) |>
  mutate(Sample = lab_g[str_remove(term, "^gS_")]) |>
  transmute(Model = model, Sample, `Skin-tone pull vs Bogotá, per 10° ITA [95% CI]` = ci(coef, `robust se`),
            p = sprintf("%.3f", `Pr(>|z|)`)) |>
  knitr::kable(caption = glue("{res$n_choice$all} trials on {res$n_choice$items} items; errors-only: {res$n_choice$err} trials"))
```

## Is the MPIB-intl signal about the sample or about age?

The MPIB-intl children are much younger than the Bogotá children, so the choice model is
refit on children aged 4.5–8.4 in every sample. The table also shows whether the skin-tone
pull changes with age within Bogotá + rural, and the item-DIF model on the age-matched
subset. That last fit gave an lme4 convergence warning (its overall-gap term is
unreliable), but its skin-tone term agrees with the full-sample fit in §3.

```{r young-table}
res$young_tab |>
  mutate(term = recode(term, gS_CO..MPIB.intl. = "skin × MPIB intl", gS_Rural.CO = "skin × rural",
                       gT_CO..MPIB.intl. = "target × MPIB intl", gT_Rural.CO = "target × rural",
                       skin_age = "skin × age (per year)", skin_rural = "skin × rural",
                       tgt_age = "target × age", tgt_rural = "target × rural",
                       g = "overall gap", `g:ita10` = "skin × MPIB intl (accuracy)")) |>
  filter(str_detect(term, "skin")) |>
  transmute(Model = model, Term = term,
            `Estimate [95% CI]` = ci(coalesce(coef, Estimate), coalesce(`robust se`, `Std. Error`)),
            p = sprintf("%.3f", `Pr(>|z|)`)) |>
  knitr::kable(caption = glue("Age-matched n: {paste(res$young_n$grp, res$young_n$n, sep = ' = ', collapse = ', ')}"))
```

# Verdict

- **The art is diverse.** Of 313 depicted people, about 16% are in the brown/dark ITA
  bands and 20% tan. 52 items show people in the correct picture, 11 of them
  dark-skinned (ITA < 10).
- **Rural vs urban Colombia: no skin-tone DIF.** This is the cleanest contrast, with the
  same translation and audio. Rural children score about 0.47 logits below Bogotá
  children overall, and that gap does not depend on the target's skin tone:
  +0.006 logits per 10° ITA [−0.046, 0.058]. The 11 dark-target items show exactly the
  average gap. Children's choices tell the same story: rural children are drawn to
  darker vs lighter characters about as much as Bogotá children are
  (−0.017 per 10° [−0.039, 0.004]; errors only −0.012). Among children under 8.5 there
  is a borderline lean toward darker characters in rural vs Bogotá children (−0.035,
  p = .05). Controlling for age, it shrinks to −0.011 (p = .32), and it does not show
  up in accuracy. The accuracy data rule out skin-tone effects larger than about
  0.05 logits per 10° of ITA.
- **Leipzig and Western Canada vs Bogotá: no evidence either.** The intervals are wider,
  and these contrasts also carry translation differences. Germany has only one TROG
  sample (Leipzig), so there is no urban/rural German comparison yet.
- **One signal: the CO (MPIB intl) sample.** These young es-CO children (4.5–8.4 years,
  n = 80) are somewhat more drawn to pictures of darker-skinned characters than Bogotá
  children (−0.063 per 10°, p < .001). The effect holds against Bogotá children of the
  same ages (−0.059, p = .003). Within Bogotá + rural, younger children lean slightly
  darker (+0.010 per year, p = .06), but that accounts for only part of the difference.
  The accuracy counterpart has the same sign but is not significant
  (−0.057 [−0.14, 0.03]): MPIB children do relatively better on darker-target items and
  worse on lighter-target ones. Whether this is a "match" effect depends on the sample's
  composition, which the site team would know. It is a small sample, and the signal is
  in choices rather than in accuracy, so it needs replication before it is acted on.
- **Stories** was not analysed: its art is uniformly diverse, and ToM item identities
  are still being repaired (see the Stories chapter).

**Bottom line.** In the samples where the question is best posed, stimulus skin tone
does not produce performance decrements on Sentence Understanding. The one hint of a
skin-tone preference (a young, small Colombian sample) does not yet translate into a
reliable accuracy difference.
