---
title: "Vocabulary — item-level analysis & rescoring (Leipzig + Bogotá)"
levante:
  summary: "Is the apparent within-child Vocabulary decline (non-adaptive → adaptive), strongest at Leipzig, a ceiling effect or a scoring bug?"
  status: draft
  task_ids: [vocab]
  datasets: [pilot_mpieva_de_main, pilot_uniandes_co_bogota]
  data:
    - {source: "load_levante_scores", version: "v1_2", table: "scores"}
    - {source: "load_levante_trials", version: "v1_2", table: "trials"}
    - {source: "load_item_parameters", version: "v1_14", table: "item_parameters"}
  depends_on: [00_load_data.qmd]
  findings:
    - "Mostly the column-order bug: DE mean within-child change is −0.91 logits in the published pipeline vs −0.00 with corrected EAP."
    - "The easy-item ceiling leaves high-ability DE non-adaptive T1 θs with large SEs; propagate score_se rather than treating θ as error-free."
  last_run: 2026-07-22
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    fig-width: 11
    fig-height: 5
execute:
  warning: false
  message: false
---

{{< include ../_disclaimer.qmd >}}

# Purpose

02 showed Vocabulary with the same worrying dynamic as Math: non-adaptive →
adaptive scores appear to *decrease* within child across waves, most starkly
at Leipzig (DE). Two hypotheses were on the table:

1.  **Unrecoverable ceiling.** DE kids in the original (non-adaptive) round
    got *very easy* items, so high-ability kids hit ceiling and their θ is
    poorly estimated / upward-biased — possibly unrecoverable.
2.  **A scoring bug.** CO also shows non-adaptive → adaptive decreases, which
    the ceiling story doesn't obviously explain.

This notebook investigates both at the item level, and applies the
column-order-fixed rescoring (see `tasks/math_eap_vs_ml.qmd` and
`reports/rlevante_handoff.md` for the bug) to see how much of the apparent
decline is real.

The column-order bug was fixed upstream in `levantemodels::score_irt` by
[PR #9](https://github.com/levante-framework/levantemodels/pull/9), so the
installed package no longer has it. "Corrected"/"fixed" below now comes
straight from `levantemodels::score_irt()` (which also takes a `method`
argument for EAP vs ML); "buggy" is reproduced locally as `score_irt_buggy()`
(in `common.R`) to keep this comparison meaningful.

**Bottom line up front:** the apparent DE vocab decline is *mostly the
scoring bug*, not the ceiling. The published pipeline shows a mean within-kid
change of −0.91 logits T1→T2; corrected EAP shows −0.00. The ceiling/easy-item
problem is real and depresses measurement precision at the top, but it is not
what was producing the dramatic spaghetti decline.

```{r setup}
source(here::here("common.R"))
library(levantemodels)
library(mirt)

scores <- read_rds(here::here("data", "scores_all_sites.rds"))
trials <- load_levante_trials(task_ids = "vocab")
ip     <- load_item_parameters() |> filter(task_id == "vocab")

# vocab is scored by_language (single-group Rasch), one model per language
mr_de <- read_rds(here::here("data", "vocab_model_de_v1_14.rds"))
mr_co <- read_rds(here::here("data", "vocab_model_co_v1_14.rds"))

run_adaptive <- scores |> filter(task_id == "vocab") |> select(run_id, adaptive)
```

# 1. The easy-item / ceiling story

DE and CO vocab item banks, and what items kids actually saw under each mode.

```{r item-coverage, fig.height=4.5}
beta_de <- ip |> filter(subset == "de-DE") |> select(item_uid, beta = difficulty)
beta_co <- ip |> filter(subset == "es-CO") |> select(item_uid, beta = difficulty)

de_tr <- trials |> filter(dataset == "pilot_mpieva_de_main") |>
  left_join(beta_de, by = "item_uid") |> left_join(run_adaptive, by = "run_id")
co_tr <- trials |> filter(dataset == "pilot_uniandes_co_bogota") |>
  left_join(beta_co, by = "item_uid") |> left_join(run_adaptive, by = "run_id")

bank <- bind_rows(
  beta_de |> mutate(site = "Leipzig (de-DE)"),
  beta_co |> mutate(site = "Bogotá (es-CO)")
)
seen <- bind_rows(
  de_tr |> filter(!is.na(beta), !is.na(adaptive)) |>
    transmute(site = "Leipzig (de-DE)", beta,
              mode = if_else(adaptive, "adaptive", "non-adaptive")),
  co_tr |> filter(!is.na(beta), !is.na(adaptive)) |>
    transmute(site = "Bogotá (es-CO)", beta,
              mode = if_else(adaptive, "adaptive", "non-adaptive"))
)

ggplot() +
  geom_density(data = bank, aes(x = beta, color = "item bank"),
               linewidth = 1, linetype = "dashed") +
  geom_density(data = seen, aes(x = beta, fill = mode), alpha = 0.4, color = NA) +
  facet_wrap(~site) +
  scale_fill_manual(values = c(`non-adaptive` = "#268bd2", adaptive = "#d33682")) +
  scale_color_manual(values = c(`item bank` = "grey30"), name = NULL) +
  labs(x = "Item difficulty β", y = "Density",
       fill = "Items seen (mode)",
       title = "Vocab: item-bank difficulty vs items children actually saw",
       subtitle = "Non-adaptive forms sampled much easier items than the bank contains")
```

```{r item-coverage-table}
seen |>
  group_by(site, mode) |>
  summarise(median_beta = median(beta), mean_beta = mean(beta),
            p90_beta = quantile(beta, 0.9), .groups = "drop") |>
  left_join(bank |> group_by(site) |>
              summarise(bank_median = median(beta), bank_max = max(beta),
                        .groups = "drop"),
            by = "site") |>
  knitr::kable(digits = 2,
               caption = "Median difficulty seen by mode vs the full bank")
```

The DE non-adaptive form served items at median β ≈ −2.7 while the DE bank
extends to β ≈ +6.7. So high-ability kids saw almost no items near their
ability → little information at the top → θ poorly pinned and prone to
upward ceiling bias. This is real and limits how precisely we can ever
measure the strongest kids from the non-adaptive round. **But** as the next
sections show, it is not the main driver of the apparent T1→T2 decline.

# 2. The scoring bug hits vocab — hard (especially DE)

Same column-order bug as math (`reports/rlevante_handoff.md`): validate the
corrected scoring against each model's calibration-time stored scores
(`@scores`, the ground truth).

```{r validate, cache=TRUE}
validate_vocab <- function(tr_site, dataset, mod_rec) {
  spec    <- levantemodels::get_model_spec("vocab", dataset,
               read_rds(here::here("data/scoring_table_cache.rds")))
  recoded <- recode_trials(tr_site)
  gold <- as.data.frame(mod_rec@scores) |> select(run_id, gold = ability)
  fixed <- levantemodels::score_irt(recoded, as.list(spec), mod_rec, method = "EAP") |>
    select(run_id, fixed = score)
  buggy <- score_irt_buggy(recoded, as.list(spec), mod_rec) |>
    transmute(run_id, buggy = as.numeric(score))
  v <- gold |> inner_join(fixed, by = "run_id") |> inner_join(buggy, by = "run_id")
  tibble(
    dataset = dataset, n = nrow(v),
    cor_fixed_gold = cor(v$gold, v$fixed),
    maxdev_fixed   = max(abs(v$gold - v$fixed)),
    cor_buggy_gold = cor(v$gold, v$buggy),
    maxdev_buggy   = max(abs(v$gold - v$buggy))
  )
}

bind_rows(
  validate_vocab(trials |> filter(dataset == "pilot_mpieva_de_main"),
                 "pilot_mpieva_de_main", mr_de),
  validate_vocab(trials |> filter(dataset == "pilot_uniandes_co_bogota"),
                 "pilot_uniandes_co_bogota", mr_co)
) |>
  knitr::kable(digits = 3,
               caption = "Corrected vs buggy scoring against calibration gold-standard")
```

The corrected scoring reproduces each model's own stored EAP scores exactly
(r = 1.0). The reproduced buggy `score_irt` correlates only **0.59** with
gold for DE vocab (max deviation ~3 logits) — a far worse distortion than
math. The permutation that the bug applies is data-dependent, and DE vocab's
trial ordering happens to scramble the item columns badly.

# 3. Rescore and compare longitudinally

```{r rescore, cache=TRUE}
rescore_site <- function(tr_site, dataset, mod_rec) {
  eap <- score_task_irt(tr_site, "vocab", dataset, mod_rec, method = "EAP") |>
    select(run_id, eap = score)
  ml  <- score_task_irt(tr_site, "vocab", dataset, mod_rec, method = "ML") |>
    select(run_id, ml = score)
  eap |> full_join(ml, by = "run_id")
}

de_rescored <- rescore_site(trials |> filter(dataset == "pilot_mpieva_de_main"),
                            "pilot_mpieva_de_main", mr_de)
co_rescored <- rescore_site(trials |> filter(dataset == "pilot_uniandes_co_bogota"),
                            "pilot_uniandes_co_bogota", mr_co)

build_pairs <- function(dataset, rescored) {
  scores |>
    filter(dataset == !!dataset, task_id == "vocab") |>
    select(run_id, user_id, age, pipeline = score) |>
    left_join(rescored, by = "run_id") |>
    group_by(user_id) |> arrange(age) |> mutate(wave = row_number()) |>
    ungroup() |> filter(wave <= 2) |>
    group_by(user_id) |> filter(n_distinct(wave) == 2) |>
    summarise(pipe_t1 = pipeline[wave == 1], pipe_t2 = pipeline[wave == 2],
              eap_t1 = eap[wave == 1], eap_t2 = eap[wave == 2],
              ml_t1 = ml[wave == 1], ml_t2 = ml[wave == 2],
              .groups = "drop")
}

de_pairs <- build_pairs("pilot_mpieva_de_main", de_rescored) |> mutate(site = "Leipzig")
co_pairs <- build_pairs("pilot_uniandes_co_bogota", co_rescored) |> mutate(site = "Bogotá")
```

```{r change-summary}
bind_rows(de_pairs, co_pairs) |>
  group_by(site) |>
  summarise(
    n = n(),
    mean_change_pipeline = mean(pipe_t2 - pipe_t1, na.rm = TRUE),
    mean_change_EAP      = mean(eap_t2 - eap_t1, na.rm = TRUE),
    mean_change_ML       = mean(ml_t2 - ml_t1, na.rm = TRUE),
    galton_pipeline = coef(lm(pipe_t2 ~ pipe_t1))[2],
    galton_EAP      = coef(lm(eap_t2 ~ eap_t1))[2],
    .groups = "drop"
  ) |>
  knitr::kable(digits = 3,
               caption = "Within-child T1→T2 change and Galton slope, by scoring")
```

# 4. Side-by-side spaghetti (Leipzig)

```{r de-spaghetti, fig.width=12, fig.height=5}
de_spag <- de_pairs |>
  pivot_longer(c(pipe_t1, pipe_t2, eap_t1, eap_t2, ml_t1, ml_t2),
               names_to = c("estimator", "wave"), names_sep = "_",
               values_to = "score") |>
  mutate(estimator = recode(estimator, pipe = "Pipeline (buggy)",
                            eap = "Corrected EAP", ml = "Corrected ML"),
         estimator = factor(estimator, levels = c("Pipeline (buggy)",
                                                   "Corrected EAP",
                                                   "Corrected ML")),
         wave = recode(wave, t1 = "T1", t2 = "T2"),
         score = pmin(pmax(score, -6), 6))

ggplot(de_spag, aes(x = wave, y = score, group = user_id)) +
  geom_line(alpha = 0.25, color = "#268bd2") +
  geom_point(aes(color = wave), alpha = 0.4, size = 1.1) +
  facet_wrap(~estimator) +
  scale_color_manual(values = c(T1 = "#268bd2", T2 = "#cb4b16"), guide = "none") +
  labs(x = NULL, y = "Vocab θ",
       title = "Leipzig vocab T1 → T2 spaghetti by scoring approach",
       subtitle = "The pipeline's downward bend at T2 largely disappears under corrected scoring")
```

```{r de-scatter, fig.width=12, fig.height=5}
de_pairs |>
  pivot_longer(c(pipe_t1, eap_t1, ml_t1), names_to = "est_t1", values_to = "t1") |>
  mutate(est = str_remove(est_t1, "_t1")) |>
  left_join(
    de_pairs |> pivot_longer(c(pipe_t2, eap_t2, ml_t2),
                             names_to = "est_t2", values_to = "t2") |>
      mutate(est = str_remove(est_t2, "_t2")) |>
      select(user_id, est, t2),
    by = c("user_id", "est")
  ) |>
  mutate(est = recode(est, pipe = "Pipeline (buggy)", eap = "Corrected EAP",
                       ml = "Corrected ML"),
         est = factor(est, levels = c("Pipeline (buggy)", "Corrected EAP",
                                      "Corrected ML"))) |>
  ggplot(aes(x = t1, y = t2)) +
  geom_abline(slope = 1, intercept = 0, lty = 3, color = "grey50") +
  geom_point(alpha = 0.5, color = "#268bd2") +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  facet_wrap(~est) +
  coord_equal() +
  labs(x = "T1 vocab θ", y = "T2 vocab θ",
       title = "Leipzig vocab T1 vs T2 by scoring approach",
       subtitle = "Dotted = identity; black = OLS")
```

# 5. How much is recoverable?

The corrected scores remove the spurious decline, but the easy-item ceiling
still limits precision for high-ability DE kids at T1. We can see this in the
score SE: kids whose non-adaptive T1 form contained no items near their
ability have large SE even after the fix.

```{r se-by-ability, fig.height=4.5}
de_eap_full <- score_task_irt(trials |> filter(dataset == "pilot_mpieva_de_main"),
                              "vocab", "pilot_mpieva_de_main",
                              mr_de, method = "EAP") |>
  left_join(run_adaptive, by = "run_id")

de_eap_full |>
  filter(!is.na(adaptive)) |>
  ggplot(aes(x = score, y = score_se, color = adaptive)) +
  geom_point(alpha = 0.5) +
  geom_smooth(se = FALSE) +
  scale_color_manual(values = c(`FALSE` = "#268bd2", `TRUE` = "#d33682"),
                     labels = c("non-adaptive", "adaptive"), name = NULL) +
  labs(x = "Corrected vocab θ", y = "Score SE",
       title = "Leipzig vocab: measurement precision by ability and mode",
       subtitle = "High-ability non-adaptive kids retain large SE — the ceiling limit")
```

```{r ceiling-check}
de_eap_full |>
  filter(!is.na(adaptive)) |>
  mutate(high = score > 1) |>
  group_by(adaptive, high) |>
  summarise(n = n(), median_se = median(score_se, na.rm = TRUE), .groups = "drop") |>
  knitr::kable(digits = 2,
               caption = "Median SE for high vs not-high ability, by mode (DE)")
```

# Findings

1.  **The column-order scoring bug hit vocab, and worse than math** (now
    fixed upstream, [PR #9](https://github.com/levante-framework/levantemodels/pull/9)).
    Corrected scoring reproduces the calibration gold-standard exactly
    (r = 1.0); the reproduced buggy `score_irt` correlates only **0.59** with
    gold for DE vocab (max deviation ~3 logits). CO vocab is much less
    affected (r = 0.99) — the distortion is data-order-dependent.

2.  **The apparent DE vocab T1→T2 decline is mostly the bug, not the
    ceiling.** Mean within-child change:

    | scoring | Leipzig | Bogotá |
    |---|---:|---:|
    | pipeline (buggy) | −0.91 | (small n) |
    | corrected EAP | −0.00 | — |
    | corrected ML | +0.03 | — |

    The Galton slope for DE goes from 0.15 (pipeline) to 0.62 (corrected).
    So the dramatic downward bend in the 02 spaghetti was largely artifactual.

3.  **The easy-item ceiling is real but is a precision problem, not a bias
    that fakes decline.** DE non-adaptive forms sampled items at median
    β ≈ −2.7 (bank max ≈ +6.7), so high-ability kids are measured with
    large SE even after the fix. This caps how confidently we can place the
    strongest kids from the non-adaptive round, but corrected θ are
    unbiased — the scores are *more* recoverable than feared.

# Recommendations

- **Rescore vocab (and all IRT tasks) with the column-order fix** before any
  longitudinal vocab analysis, unless `levante_data_latest` has already been
  rescored since the fix (same fix as math; merged upstream in
  [PR #9](https://github.com/levante-framework/levantemodels/pull/9); see
  also `reports/rlevante_handoff.md`).
- For 02 / 03, use corrected vocab scores. The DE vocab LGC and any
  language-factor growth model should be re-fit; the construct-level decline
  was an artifact.
- The easy-item ceiling means DE non-adaptive T1 vocab θ for high-ability
  kids carry large SE; a measurement-model approach that propagates score SE
  (e.g. plausible values, or using `score_se` as known measurement error in
  the SEM) is more honest than treating θ as observed-without-error.

```{r save}
write_rds(bind_rows(de_pairs, co_pairs),
          here::here("data", "vocab_eap_ml_pairs.rds"))
write_rds(bind_rows(
  de_rescored |> mutate(dataset = "pilot_mpieva_de_main"),
  co_rescored |> mutate(dataset = "pilot_uniandes_co_bogota")
), here::here("data", "vocab_corrected_scores.rds"))
```
