---
title: "Missing `adaptive` flag in `levante_data_latest`"
levante:
  summary: "Report for the DCC: which runs in levante_data_latest are missing the `adaptive` flag, why, and how to impute it."
  status: draft
  task_ids: [egma-math, memory-game]
  datasets: [pilot_mpieva_de_main, pilot_uniandes_co_bogota, pilot_western_ca_main]
  data:
    - {source: "load_levante_scores", version: "v1_0", table: "scores"}
    - {source: "load_levante_trials", version: "v1_0", table: "trials"}
  depends_on: [00_load_data.qmd]
  findings:
    - "Missing flags are concentrated in early-beta task versions of Bogotá Memory and Leipzig Math; all evidence says non-adaptive."
    - "Impute to FALSE until the source is patched; add a validator rule that fails ability rows with no `adaptive` value."
    - "Counts here (196 rows) predate the final tally of 235 runs backfilled in 00."
  last_run: unknown
subtitle: "Short report for the DCC team"
format:
  html:
    toc: false
    code-fold: true
    fig-width: 8
    fig-height: 4
execute:
  warning: false
  message: false
---

```{r setup, echo=FALSE}
source(here::here("common.R"))
scores <- read_rds(here::here("data", "scores_all_sites.rds"))
trials <- load_levante_trials()
```

## Summary

196 of 18,306 scored runs (**1.07%**) have `adaptive` = `NA`. The
missingness is **not random**: it is concentrated in three
(dataset × task) cells and in **early beta `task_version`s**. The trial-level
data strongly suggests every NA run was administered non-adaptively, so the
gap is a reporting/labelling bug rather than missing administration metadata
at the source.

## Where the missingness lives

```{r overall}
scores |>
  count(adaptive, name = "n") |>
  mutate(pct = round(n / sum(n) * 100, 2)) |>
  knitr::kable(caption = "Overall counts.")
```

All 196 NAs sit in just three cells:

```{r by-cell}
scores |>
  filter(is.na(adaptive)) |>
  count(dataset_label, task_label, name = "n_na") |>
  left_join(
    scores |> count(dataset_label, task_label, name = "n_total"),
    by = c("dataset_label", "task_label")
  ) |>
  mutate(pct_na = round(n_na / n_total * 100, 1)) |>
  arrange(desc(n_na)) |>
  knitr::kable(caption = "Cells with any NA `adaptive`.")
```

Three other facts about the NA rows:

- All 196 are scored with `score_type = "ability"` and
  `scoring_model = "multigroup_site_all_items_rasch_f1_scalar"` (i.e. the
  same scoring pipeline as the rest of the data).
- One NA per run — the missingness is a run-level property, not scattered
  across trials.
- All 196 runs are present in the trial table; no run was orphaned between
  the two tables.

## Which `task_version`s are affected?

```{r by-version, fig.height=5}
affected <- scores |>
  filter(task_id %in% c("egma-math", "memory-game"),
         dataset %in% c("pilot_mpieva_de_main",
                        "pilot_uniandes_co_bogota",
                        "pilot_western_ca_main")) |>
  mutate(adapt_lbl = case_when(
    is.na(adaptive)  ~ "NA",
    adaptive == TRUE  ~ "adaptive",
    adaptive == FALSE ~ "non-adapt"
  ))

affected |>
  count(dataset_label, task_label, task_version, adapt_lbl) |>
  pivot_wider(names_from = adapt_lbl, values_from = n, values_fill = 0L) |>
  arrange(dataset_label, task_label, task_version) |>
  knitr::kable()
```

The NAs are confined to specific early-beta task builds:

- **Bogotá / Memory**: `1.0.0-beta.13`, `.14`, `.15` — every run on these
  versions is unflagged. Runs on `1.0.17` and `1.0.18` are properly tagged
  `FALSE`.
- **Leipzig / Math**: mostly `1.0.0-beta.39` (75 NA + 3 FALSE in the same
  version) and `1.0.8` (8 NA + 3 FALSE). Later versions (`1.0.11`, `1.0.12`,
  `1.0.18`, `1.0.19`) are clean.
- **Western / Math**: 3 NA on `1.0.8`.

The mixed-within-version cells (e.g. `1.0.0-beta.39` math with both NA and
`FALSE`) are the strongest evidence that this is a reporting bug rather
than a genuine "we don't know what mode this kid took": within a single
deployed build, identical administrations sometimes emit the flag and
sometimes don't.

## Is the trial table consistent with non-adaptive?

The trial table has no `adaptive` column, but `theta_estimate` is updated
on every trial in CAT administrations.

For the 196 NA runs:

- **0** runs have any populated `theta_estimate` value across their trials.

For comparison (200-run sample of confirmed runs):

- Confirmed `adaptive = TRUE` runs: 78% have `theta_estimate` populated;
  99% of those show evolving (varying) θ across trials.
- Confirmed `adaptive = FALSE` runs: 82% have no `theta_estimate` populated.

`theta_estimate` alone isn't a perfect classifier — about a fifth of
adaptive runs don't write it out — but **zero out of 196** NA runs showing
any θ is well above what you'd expect if these were really CAT runs. The
balance of evidence is that all 196 NA runs were non-adaptive.

## Why this matters for downstream analyses

The two longitudinal sites (Leipzig, Bogotá) are also the ones where the
adaptive-vs-non-adaptive mode shift drives most apparent change in IRT θ
across waves (see `01_data_integrity.html` for the per-task picture; the
Bogotá math "regression to the mean" turns out to be entirely the
non-adaptive → CAT mode shift between waves T1 and T2, with within-item
growth still present in every quartile — see `tasks/math.html`).

Because adaptive flag missingness is concentrated in exactly those two
sites, the imputation choice matters. With the current data:

- **Bogotá Memory**: 110 of 365 runs unflagged (30%). All on `1.0.0-beta.*`
  versions that were *probably* non-adaptive.
- **Leipzig Math**: 83 of 510 runs unflagged (16%). Same pattern.

If these get treated as "adaptive unknown" and dropped from mode-stratified
longitudinal analyses, the effective sample drops noticeably. If they get
imputed to non-adaptive (which the evidence supports), they slot back into
the cleaner pre-CAT-transition cohort.

## Recommended fixes

1. **Backfill at the source**, ideally by re-emitting the `adaptive` flag
   for the affected `task_version`s. The DCC's deployment metadata should
   say definitively whether each beta build shipped a CAT routing
   component; the data team can patch the scoring pipeline to write the
   correct value for retrospective runs.

2. **Until the source is patched**, downstream code can impute with high
   confidence by:

   ```r
   scores |>
     mutate(adaptive = if_else(
       is.na(adaptive),
       FALSE,                # all 196 NA rows are early beta + zero theta
       adaptive
     ))
   ```

   …with a note that 196 rows in three specific cells were imputed.

3. **Validator check** going forward: any new row with
   `score_type == "ability"` and no `adaptive` value should fail validation
   and be flagged for the data engineering team. A schema-level NOT NULL
   constraint on this column for active task builds would have caught the
   beta-39 regression at ingest.

```{r save-summary, echo=FALSE}
write_rds(
  scores |> filter(is.na(adaptive)) |>
    count(dataset_label, task_label, task_version, name = "n_na"),
  here::here("data", "adaptive_missingness_cells.rds")
)
```
