---
title: "Mental rotation: item-set x model-type comparison"
levante:
  summary: "Mental rotation: compare 8 item-collapsing schemes × IRT model types (Rasch/2PL; configural/metric/scalar) on fit, reliability, age correlation, and test–retest."
  status: draft
  task_ids: [mental-rotation]
  datasets: [pilot_mpieva_de_main, pilot_uniandes_co_bogota, pilot_uniandes_co_rural, pilot_western_ca_main]
  data:
    - {source: "../levante-pilots/01_fetched_data/task_data_nested.rds", table: "trials"}
    - {source: "../levante-pilots/01_fetched_data/run_data.rds", table: "runs"}
  depends_on: [tasks/mrot/fit_item_sets.R]
  findings:
    - "Model type matters more than item granularity; 2PL scalar wins BIC in 6 of 8 item sets and has much higher marginal reliability than Rasch (~0.7–0.8 vs ~0.43)."
    - "At three of four sites, marginal reliability declines as items are split more finely; DE stays flat."
    - "Age correlation and test–retest are essentially unaffected by item-set choice."
    - "Not built in the book: needs un-versioned levante-pilots inputs and the fit_item_sets.R registry."
  last_run: 2026-07-20
---

Cross the three item-collapsing dimensions (angle vs. angle_full, stimulus vs.
stimulus_code, chirality in/out) to get 8 item sets ranging from maximally
collapsed (current `item_uid`) to maximally fine-grained (current
`item_original`). For each item set, fit the full `models_multigroup` set
(Rasch/2PL x configural/metric/scalar) and compare BIC, marginal reliability,
empirical reliability, score-age correlation, and test-retest reliability.

Note: polygon items (`polygon1`/`polygon2`) are excluded throughout. Their
`item_original` codes share a `stimulus_code` ("p") and use the `chirality`
character ("1"/"2") to distinguish the two shapes rather than encode true
mirror chirality, so any item set that collapses over chirality without using
`stimulus` would otherwise silently merge the two polygons together.

## Summary of findings

- **Model type matters far more than item granularity, but BIC support for
  2PL is mixed, not decisive.** The 2PL scalar model wins BIC in 6 of 8 item
  sets, by modest margins (roughly 1-150 BIC points over the next-best
  model) -- it actually loses to Rasch scalar in the two item sets that
  collapse `angle_full` with `chirality` retained (by 16.5 and 132 BIC
  points). The stronger case for 2PL is its much higher marginal reliability
  than Rasch (~0.7-0.8 vs. ~0.43), which holds consistently across all 8 item
  sets regardless of which model wins BIC.
- **Aggregate differences across the 8 item-collapsing schemes are modest.**
  For 2PL scalar models, marginal reliability ranges 0.71-0.79, empirical
  reliability is essentially flat (~0.84), and person-level ability scores are
  nearly identical regardless of item set.
- **A per-dataset breakdown reveals a real pattern masked by the aggregate.**
  At three of four sites (CO-Bogota, Western-CA, CO-Rural), marginal
  reliability declines consistently as items are split more finely --
  collapsing over `chirality` costs the most, then `angle` granularity, then
  `stimulus` granularity. DE stays flat (~0.85-0.88) regardless of item set.
  CO-Rural is the lowest-reliability site throughout, but moves in the same
  direction as the others rather than diverging from them.
- **Score-age correlation and test-retest reliability are essentially
  unaffected by item-set choice** within any given site; the variation seen
  in these metrics across sites reflects site differences, not the
  granularity decision (note CO-Rural has no retest data, and the Western-CA
  retest sample is small, n = 30).
- **Bottom line:** the current collapsed `item_uid` scheme loses little
  relative to the maximally fine-grained `item_original` -- if anything, it
  shows the highest reliability among the 8 options tested.

```{r}
#| cache: false

library(tidyverse)
library(here)
library(glue)
library(mirt)
library(levantemodels)

source(here::here("common.R"))
```

## Load data + build item sets

```{r}
pilots_data_dir <- here("..", "levante-pilots", "01_fetched_data")
task_data_nested <- read_rds(file.path(pilots_data_dir, "task_data_nested.rds"))
run_data <- read_rds(file.path(pilots_data_dir, "run_data.rds"))

task_data_mrot <- task_data_nested |> filter(item_task == "mrot")

# at least for this task, NA responses seem to be incorrectly coded as incorrect
task_data <- task_data_mrot |>
  unnest(data) |>
  filter(!is.na(response)) |>
  left_join(run_data |> select(run_id, age))
```

Code item features from `item_original`, excluding polygon stimuli. 3D items
embed an extra exemplar number between the chirality character and the angle
(e.g. `an11240` = code `a` + chirality `n` + exemplar `11` + angle `240`) that
isn't one of the three dimensions we're crossing but has to be preserved for
`stimulus_code` to be well defined -- otherwise items with the same code,
chirality, and angle but different exemplars would be silently merged. We fold
it into `stimulus_code` (2D items have no exemplar digits, so this is a no-op
for them).

```{r}
items_coded <- task_data |>
  select(contains("item")) |> distinct() |>
  separate(item, into = c("stimulus", "angle")) |>
  mutate(angle = as.numeric(angle)) |>
  rename(item = item_original) |>
  mutate(angle_full = item |> str_sub(str_length(item) - 2, str_length(item)) |> as.numeric(),
         stimulus_code = item |> str_sub(1, 1),
         chirality = item |> str_sub(2, 2),
         exemplar = item |> str_sub(3, str_length(item) - 3),
         stimulus_code = paste0(stimulus_code, exemplar)) |>
  filter(!stimulus %in% c("polygon1", "polygon2")) |>
  rename(item_original = item) |>
  select(item_task, item_group, item_original, stimulus, stimulus_code, chirality, angle, angle_full)
```

Cross the three dimensions into 8 item sets and build an `item_original` ->
`item_uid` map for each.

```{r}
make_item_uid_map <- function(angle_var, stimulus_var, use_chirality) {
  items_coded |>
    mutate(angle_str = sprintf("%03d", .data[[angle_var]]),
           stim_val = .data[[stimulus_var]]) |>
    mutate(item_uid_new = if (use_chirality) {
      paste("mrot", item_group, stim_val, chirality, angle_str, sep = "_")
    } else {
      paste("mrot", item_group, stim_val, angle_str, sep = "_")
    }) |>
    select(item_original, item_uid_new)
}

item_set_specs <- expand_grid(
  angle_var = c("angle", "angle_full"),
  stimulus_var = c("stimulus", "stimulus_code"),
  chirality = c(FALSE, TRUE)
) |>
  mutate(item_set = paste(angle_var, stimulus_var,
                           if_else(chirality, "chir", "nochir"), sep = "_"),
         item_uid_map = pmap(list(angle_var, stimulus_var, chirality), make_item_uid_map),
         n_items = map_dbl(item_uid_map, ~ n_distinct(.x$item_uid_new)))

item_set_specs |> select(item_set, angle_var, stimulus_var, chirality, n_items)
```

`angle_stimulus_nochir` is the current `item_uid` (most collapsed);
`angle_full_stimulus_code_chir` is equivalent to the current `item_original`
(most fine-grained, 71 items = distinct `item_original` values after
excluding polygons).

```{r}
build_task_data_x <- function(item_uid_map) {
  task_data |>
    inner_join(item_uid_map, by = "item_original") |>
    select(-item_uid) |>
    rename(item_uid = item_uid_new) |>
    nest(data = -c(dataset, item_task, language))
}

item_set_specs <- item_set_specs |>
  mutate(task_data_x = map(item_uid_map, build_task_data_x))
```

## Fit models

Full multigroup model set (Rasch/2PL x configural/metric/scalar), fit
separately for each of the 8 item sets. This fits 8 x 5 = 40 multigroup
models and will take a while.

```{r}
models_multigroup <- tribble(
  ~nfact, ~itemtype, ~invariance,
  1,   "Rasch", "configural",
  1,   "Rasch", "scalar",
  1,     "2PL", "configural",
  1,     "2PL", "metric",
  1,     "2PL", "scalar",
)

priors <- list(d = c("norm", 0, 3), a1 = c("norm", 1, 0.3))

# scalar 2PL all_items is the last file fit_task_models_multigroup writes per
# item set, so its presence marks that item set as already fit -- lets
# re-renders skip refitting all 40 models from scratch
is_item_set_done <- \(item_set) {
  file.exists(file.path("models/item_sets", item_set, "mrot",
                         "multigroup_dataset", "all_items", "mrot_2pl_f1_scalar.rds"))
}

walk2(item_set_specs$item_set, item_set_specs$task_data_x, \(item_set, task_data_x) {
  if (is_item_set_done(item_set)) return(invisible())
  fit_task_models_multigroup(task_data = task_data_x,
                              models = models_multigroup,
                              priors = priors,
                              task = "mrot",
                              group = dataset,
                              registry_dir = file.path("models/item_sets", item_set))
})
```

## Load fitted models

```{r}
registry_root <- "models/item_sets"

list_item_set_models <- \() {
  mod_files <- list.files(registry_root, recursive = TRUE)
  mod_files |>
    str_split("/", simplify = TRUE) |>
    as_tibble(.name_repair = "minimal") |>
    set_names(c("item_set", "task", "model_set", "subset", "filename")) |>
    bind_cols(path = mod_files)
}

code_item_set_models <- \(mod_df) {
  mod_df |>
    mutate(file = str_remove(filename, "\\.rds$")) |>
    separate_wider_delim(file, names = c("task_dup", "itemtype", "nfact", "invariance"),
                          delim = "_", cols_remove = FALSE, too_few = "align_start") |>
    select(-task_dup) |>
    mutate(invariance = invariance |> replace_na(""))
}

mods <- list_item_set_models() |>
  mutate(mod_rec = map(path, \(p) read_rds(file.path(registry_root, p)))) |>
  code_item_set_models() |>
  mutate(BIC = map_dbl(mod_rec, \(rec) rec@fit$BIC))
```

## Compute comparison metrics

**BIC**: configural models are only fit on `overlap_items` (all-items models
aren't fit for configural invariance, since there's nothing to constrain
across groups), so BIC comparison across all 5 model types uses
`overlap_items` for every model, matching the convention used in
`02_score_data/02_extract_irt/02_compare_models.qmd`.

**Marginal/empirical reliability**: `mirt::extract.group()` only works for
scalar-invariance multigroup fits (metric/configural have cross-group equality
constraints that break it), so these are computed only for the scalar models,
using the `all_items` fit.

**Score-age correlation and test-retest**: computed per dataset (not pooled
across sites, since cross-site mean/scale comparisons are not interpretable)
from each `ModelRecord`'s stored calibration-time EAP scores
(`mod_rec@scores`), which are available for every model type/invariance. For
configural models this uses `overlap_items` scores (the only ones available);
for metric/scalar it uses `all_items` scores.

```{r}
run_meta <- task_data |> distinct(dataset, run_id, user_id, age)

extract.mirt_groups <- \(mod) {
  extract.mirt(mod, "groupNames") |> set_names() |> map(\(gr) extract.group(mod, gr))
}

# marginal + empirical reliability, scalar models only, all_items
mods_scalar <- mods |>
  filter(subset == "all_items", invariance == "scalar")

mods_scalar_rxx <- mods_scalar |>
  mutate(mod = map(mod_rec, model_from_record),
         group_mod = map(mod, extract.mirt_groups)) |>
  select(item_set, itemtype, invariance, group_mod) |>
  unnest_longer(group_mod, indices_to = "group") |>
  mutate(marginal_rxx = map_dbl(group_mod, marginal_rxx),
         scores = map(group_mod, \(m) fscores(m, method = "EAP", full.scores.SE = TRUE)),
         empirical_rxx = map_dbl(scores, empirical_rxx)) |>
  select(item_set, itemtype, invariance, group, marginal_rxx, empirical_rxx)

# score-age correlation + test-retest, all model types, per dataset
mods_scores <- mods |>
  filter(subset %in% c("all_items", "overlap_items")) |>
  # prefer all_items when both exist for a given item_set x itemtype x invariance
  arrange(item_set, itemtype, invariance, desc(subset == "all_items")) |>
  distinct(item_set, itemtype, invariance, .keep_all = TRUE) |>
  mutate(scores = map(mod_rec, \(rec) rec@scores |> left_join(run_meta, by = "run_id")))

age_cor <- mods_scores |>
  select(item_set, itemtype, invariance, subset, scores) |>
  unnest(scores) |>
  group_by(item_set, itemtype, invariance, subset, dataset) |>
  summarise(age_r = cor(ability, age, use = "complete.obs"), n = n(), .groups = "drop")

test_retest <- mods_scores |>
  select(item_set, itemtype, invariance, subset, scores) |>
  unnest(scores) |>
  group_by(item_set, itemtype, invariance, subset, user_id, dataset) |>
  filter(n() > 1) |>
  arrange(age) |>
  mutate(run_number = row_number(), age_gap = max(age) - min(age)) |>
  filter(age_gap > 0.05, run_number <= 2) |>
  ungroup() |>
  select(item_set, itemtype, invariance, subset, dataset, user_id, run_number, ability) |>
  pivot_wider(names_from = run_number, values_from = ability, names_prefix = "run_") |>
  group_by(item_set, itemtype, invariance, subset, dataset) |>
  summarise(retest_r = cor(run_1, run_2, use = "complete.obs"), n_retest = n(), .groups = "drop")
```

## Summary table

```{r}
model_comparison <- mods |>
  filter(subset == "overlap_items") |>
  select(item_set, itemtype, nfact, invariance, BIC) |>
  left_join(mods_scalar_rxx |>
              group_by(item_set, itemtype, invariance) |>
              summarise(marginal_rxx = mean(marginal_rxx),
                        empirical_rxx = mean(empirical_rxx), .groups = "drop")) |>
  left_join(age_cor |>
              group_by(item_set, itemtype, invariance) |>
              summarise(age_r = mean(age_r, na.rm = TRUE), .groups = "drop")) |>
  left_join(test_retest |>
              group_by(item_set, itemtype, invariance) |>
              summarise(retest_r = mean(retest_r, na.rm = TRUE), .groups = "drop")) |>
  left_join(item_set_specs |> select(item_set, angle_var, stimulus_var, chirality, n_items)) |>
  arrange(item_set, itemtype, invariance)

model_comparison
```

```{r}
model_comparison |>
  filter(itemtype == "2pl", invariance == "scalar") |>
  select(item_set, where(is.numeric)) |>
  select(-BIC) |>
  arrange(desc(marginal_rxx)) |>
  mutate(across(where(is.numeric), \(x) round(x, 2)))
```

Reliability, age-correlation, and test-retest values above are averaged across
groups/datasets for a compact overview; per-dataset values are in
`mods_scalar_rxx`, `age_cor`, and `test_retest`.

```{r}
#| fig-width: 10
#| fig-height: 6

ggplot(model_comparison |> group_by(item_set) |> mutate(delta_bic = BIC - min(BIC)),
       aes(x = delta_bic, y = paste(itemtype, invariance))) +
  facet_wrap(vars(item_set), scales = "free_x") +
  geom_vline(xintercept = 0, colour = "grey") +
  geom_crossbar(aes(xmin = delta_bic, xmax = delta_bic, colour = itemtype), width = 0.7,
                linewidth = 0.5) +
  scale_colour_brewer(palette = "Set1", guide = "none") +
  labs(x = "BIC difference from lowest (within item set)", y = "Model")
# ggsave("bic_by_item_set.png", width = 10, height = 6)
```

```{r}
#| fig-width: 8
#| fig-height: 6

ggplot(model_comparison |> filter(invariance == "scalar"),
       aes(x = item_set, y = marginal_rxx, colour = itemtype)) +
  geom_point(size = 2) +
  geom_line(aes(group = itemtype)) +
  scale_colour_brewer(palette = "Set1") +
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  labs(x = "Item set", y = "Marginal reliability (scalar models)", colour = "Item type")
# ggsave("rxx_by_item_set.png", width = 7, height = 4)
```

## Item difficulty comparison across item sets

Compare item difficulties from the finest item set (`item_original`-
equivalent) against the collapsed difficulty for each of the other 7 item
sets, using the scalar 2PL model.

```{r}
finest_set <- "angle_full_stimulus_code_chir"

coefs <- mods |>
  filter(subset == "all_items", itemtype == "2pl", invariance == "scalar") |>
  mutate(coef = map(mod_rec, mod_coefs)) |>
  select(item_set, coef) |>
  unnest(coef) |>
  filter(n_responses > 30) |>
  select(item_set, item, group, a1, difficulty) |>
  distinct()

finest_coef <- coefs |> filter(item_set == finest_set) |>
  select(item_original_uid = item, difficulty_fine = difficulty) |> distinct()

# map each finest item_uid back to its item_original, then to every other
# item set's item_uid, to compare difficulties at different granularities
finest_item_map <- item_set_specs |> filter(item_set == finest_set) |>
  pull(item_uid_map) |> pluck(1) |>
  rename(item_original_uid = item_uid_new)

coef_comp <- coefs |>
  filter(item_set != finest_set) |>
  rename(item_uid_new = item) |>
  left_join(item_set_specs |> select(item_set, item_uid_map) |>
              unnest(item_uid_map), by = c("item_set", "item_uid_new")) |>
  left_join(finest_item_map, by = "item_original") |>
  left_join(finest_coef, by = "item_original_uid") |>
  left_join(items_coded, by = "item_original")
```

```{r}
#| fig-width: 10
#| fig-height: 8

ggplot(coef_comp, aes(x = difficulty_fine, y = difficulty)) +
  facet_wrap(vars(item_set)) +
  coord_fixed() +
  geom_abline(colour = "lightgrey") +
  geom_point(aes(colour = stimulus), size = 1, alpha = 0.6) +
  scale_colour_brewer(palette = "Set2") +
  labs(x = "Item difficulty (finest item set)", y = "Item difficulty (this item set)")
# ggsave("mrot_difficulty_by_item_set.png", width = 9, height = 6)
```

## Score comparison across item sets

Compare person scores (EAP ability) from the finest item set against every
other item set, using the scalar 2PL model.

```{r}
scores_by_set <- mods |>
  filter(subset == "all_items", itemtype == "2pl", invariance == "scalar") |>
  mutate(scores = map(mod_rec, \(rec) rec@scores |> select(run_id, ability))) |>
  select(item_set, scores) |>
  unnest(scores)

scores_wide <- scores_by_set |>
  pivot_wider(names_from = item_set, values_from = ability) |>
  left_join(run_meta, by = "run_id")

score_comp <- scores_wide |>
  select(-run_id, -user_id) |>
  pivot_longer(-c(dataset, age, all_of(finest_set)), names_to = "item_set", values_to = "ability") |>
  rename(ability_fine = all_of(finest_set))
```

```{r}
#| fig-width: 10
#| fig-height: 8

ggplot(score_comp, aes(x = ability_fine, y = ability)) +
  facet_wrap(vars(item_set)) +
  coord_fixed() +
  geom_abline(colour = "lightgrey") +
  geom_point(size = 0.5, alpha = 0.3) +
  labs(x = "Score (finest item set)", y = "Score (this item set)")
# ggsave("mrot_scores_by_item_set.png", width = 9, height = 6)
```
